---
title: "Programming for Applications"
subtitle: "Chapter 12: Preparing Data"
author: "Yu-You Liou (NTU)"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "R in a Nutshell: A Desktop Quick Reference"
logo: "logo.svg"
editor: source
bibliography: bib.bib
suppress-bibliography: true
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch12-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center")
```

# **Where 80% of the Time Goes**

## The Dirty Secret of Data Analysis

The book's author started as a biochemist, fled to computer science to escape endless lab preparation — and landed in data mining, where **~80% of project effort goes to finding, cleaning, and preparing data**, under 5% to actual analysis (the rest to writing it up).

"It came from computers, so it's perfect, right?" In practice data is almost never stored in the right form, and even well-formed data hides surprises. This chapter is the toolbox: combining, transforming, binning, subsetting, summarizing, reshaping, cleaning, and sorting.

# Combining Data Sets

## paste: Concatenating Text

`paste` concatenates character vectors element-wise (coercing other types first):

```{r,echo=TRUE}
x <- c("a", "b", "c", "d", "e")
y <- c("A", "B", "C", "D", "E")
paste(x, y)
paste(x, y, sep="-")
paste(x, y, sep="-", collapse="#")   # collapse: squash into ONE value
```

## cbind: Adding Columns

`cbind` combines objects horizontally. Starting from the NFL salary table of Chapter 11:

```{r,echo=TRUE}
top.5.salaries <- data.frame(
  name.last = c("Manning", "Brady", "Pepper", "Palmer", "Manning"),
  name.first = c("Peyton", "Tom", "Julius", "Carson", "Eli"),
  team = c("Colts", "Patriots", "Panthers", "Bengals", "Giants"),
  position = c("QB", "QB", "DE", "QB", "QB"),
  salary = c(18700000, 14626720, 14137500, 13980000, 12916666))
more.cols <- data.frame(year = rep(2008, 5), rank = 1:5)
cbind(top.5.salaries, more.cols)
```

## rbind: Adding Rows

`rbind` combines vertically — here appending salaries six through eight:

```{r,echo=TRUE}
next.three <- data.frame(
  name.last = c("Favre", "Bailey", "Harrison"),
  name.first = c("Brett", "Champ", "Marvin"),
  team = c("Packers", "Broncos", "Colts"),
  position = c("QB", "CB", "WR"),
  salary = c(12800000, 12690050, 12000000))
rbind(top.5.salaries, next.three)
```

## Extended Example: Fetch and Bind Stock Quotes

The book builds one data frame of quotes for all 30 Dow stocks. `get.quotes` assembles a Yahoo! Finance CSV URL with `paste` (format discovered by trial and error), fetches it with `read.csv`, and `cbind`s on a ticker column:

```{r,eval=FALSE,echo=TRUE}
get.quotes <- function(ticker, from=(Sys.Date()-365), to=(Sys.Date()),
                       interval="d") {
  base <- "http://ichart.finance.yahoo.com/table.csv?"
  symbol <- paste("s=", ticker, sep="")
  from.month <- paste("&a=",                      # months number 00-11!
    formatC(as.integer(format(from,"%m"))-1, width=2, flag="0"), sep="")
  from.day  <- paste("&b=", format(from,"%d"), sep="")
  from.year <- paste("&c=", format(from,"%Y"), sep="")
  to.month  <- paste("&d=",
    formatC(as.integer(format(to,"%m"))-1, width=2, flag="0"), sep="")
  to.day    <- paste("&e=", format(to,"%d"), sep="")
  to.year   <- paste("&f=", format(to,"%Y"), sep="")
  inter     <- paste("&g=", interval, sep="")
  url <- paste(base, symbol, from.month, from.day, from.year,
               to.month, to.day, to.year, inter, "&ignore=.csv", sep="")
  tmp <- read.csv(url)
  cbind(symbol=ticker, tmp)
}
```

## ...and Bind the Tickers Together

`get.multiple.quotes` loops over tickers, `rbind`ing as it goes:

```{r,eval=FALSE,echo=TRUE}
get.multiple.quotes <- function(tkrs, from=(Sys.Date()-365),
                                to=(Sys.Date()), interval="d") {
  tmp <- NULL
  for (tkr in tkrs) {
    if (is.null(tmp)) tmp <- get.quotes(tkr, from, to, interval)
    else tmp <- rbind(tmp, get.quotes(tkr, from, to, interval))
  }
  tmp
}
dow.tickers <- c("MMM","AA","AXP","T","BAC","BA","CAT","CVX","CSCO",
  "KO","DD","XOM","GE","HPQ","HD","INTC","IBM","JNJ","JPM","KFT",
  "MCD","MRK","MSFT","PFE","PG","TRV","UTX","VZ","WMT","DIS")
dow30 <- get.multiple.quotes(dow.tickers)
```

::: {.callout-warning}
**Yahoo! retired this API** — the URL no longer works (use `quantmod` today). The *pattern* — build URLs with `paste`, accumulate with `rbind` — remains everyday practice. To keep the chapter's later examples runnable, we reconstruct the book's small five-stock extract by hand on the next slide.
:::

## Reconstructing my.quotes

The book's five-stock, three-month sample, entered directly:

```{r,echo=TRUE}
my.quotes <- data.frame(
  symbol = rep(c("GE","GOOG","AAPL","AXP","GS"), each=3),
  Date = rep(as.Date(c("2009-03-02","2009-02-02","2009-01-02")), 5),
  Open  = c(8.29,12.03,16.51, 333.33,334.29,308.60, 88.12,89.10,85.88,
            11.68,16.35,18.57, 87.86,78.78,84.02),
  High  = c(11.35,12.90,17.24, 359.16,381.00,352.33, 109.98,103.00,97.17,
            15.24,18.27,21.38, 115.65,98.66,92.20),
  Low   = c(5.87,8.40,11.87, 289.45,329.55,282.75, 82.33,86.51,78.20,
            9.71,11.44,14.72, 72.78,78.57,59.13),
  Close = c(10.11,8.51,12.13, 348.06,337.99,338.53, 105.12,89.31,90.13,
            13.63,12.06,16.73, 106.02,91.08,80.73),
  Volume = c(277426300,194928800,117846700, 5346800,6158100,5727600,
             25963400,27394900,33487900, 31136400,24297100,19110000,
             30196400,28301500,22764300),
  Adj.Close = c(10.11,8.51,11.78, 348.06,337.99,338.53, 105.12,89.31,90.13,
                13.45,11.90,16.51, 106.02,91.08,80.29))
head(my.quotes, 6)
```

## merge: Joining on Common Fields

Baseball player ages live in `Master`; batting statistics in `Batting`; both carry `playerID`. To show batting *with names and ages*, **merge**:

```{r,eval=FALSE,echo=TRUE}
batting <- dbGetQuery(con, "SELECT * FROM Batting")
master  <- dbGetQuery(con, "SELECT * FROM Master")
intersect(names(batting), names(master))
## [1] "playerID"
batting.w.names <- merge(batting, master)   # joins on playerID
```

By default `merge` keys on **all common column names** — here just one, so no extra arguments needed.

## merge: Arguments and Join Semantics

```{r,eval=FALSE,echo=TRUE}
merge(x, y, by =, by.x =, by.y =, all =, all.x =, all.y =,
      sort =, suffixes =, incomparables =, ...)
```

| Argument | Description | Default |
|---|---|---|
| `x`, `y` | the two data frames | |
| `by` / `by.x`, `by.y` | key columns (shared / per side) | common names |
| `all` | keep unmatched rows from both sides (OUTER JOIN) | `FALSE` |
| `all.x` / `all.y` | keep unmatched rows from x / y (LEFT / RIGHT OUTER) | `all` |
| `sort` | sort result by the `by` columns? | `TRUE` |
| `suffixes` | renaming for same-named non-key columns | `c(".x",".y")` |
| `incomparables` | values that never match | `NULL` |

In SQL terms: the default is a NATURAL JOIN; chosen keys give an INNER JOIN; `all=TRUE` a FULL OUTER JOIN — and no matching names at all yields the full **Cartesian product**.

# Transformations

## Reassigning Variables

Fix variables in place with the assignment operator. Imported dates often arrive as text — convert them to real `Date` objects; derived columns join by the same notation:

```{r,echo=TRUE}
my.quotes$Date <- as.Date(my.quotes$Date)   # idempotent here; the book's
class(my.quotes$Date)                        # dow30 needed factor -> Date
my.quotes$mid <- (my.quotes$High + my.quotes$Low) / 2
names(my.quotes)
```

## The transform Function

`transform(data, ...)` applies a set of expressions *inside* the data frame — no `data$` prefixes — and returns the modified frame:

```{r,echo=TRUE}
my.quotes.transformed <- transform(my.quotes, Date = as.Date(Date),
                                   mid = (High + Low) / 2)
names(my.quotes.transformed)
```

## apply: Functions over Array Margins

`apply(X, MARGIN, FUN, ...)` applies `FUN` along chosen dimensions (extra arguments pass through to `FUN`). With `max`, the workings are visible at a glance:

```{r,echo=TRUE}
x <- 1:20
dim(x) <- c(5, 4)
x
apply(X=x, MARGIN=1, FUN=max)   # per row
apply(X=x, MARGIN=2, FUN=max)   # per column
```

## apply in Three Dimensions

`MARGIN` can name several dimensions. Switching to `paste` to *see* which elements group together:

```{r,echo=TRUE}
x <- 1:27
dim(x) <- c(3, 3, 3)
apply(X=x, MARGIN=1, FUN=paste, collapse=",")
apply(X=x, MARGIN=3, FUN=paste, collapse=",")
apply(X=x, MARGIN=c(1,2), FUN=paste, collapse=",")
```

The last call computes, for each (i, j), `FUN` of `x[i, j, 1], x[i, j, 2], x[i, j, 3]`.

## lapply and sapply: Lists, Vectors, Data Frames

`lapply(X, FUN, ...)` applies `FUN` to each element, returning a **list** — and a data frame counts as a list of columns:

```{r,echo=TRUE}
x <- as.list(1:5)
lapply(x, function(x) 2^x)[1:2]
d <- data.frame(x=1:5, y=6:10)
lapply(d, FUN=max)
```

Prefer a vector or matrix? `sapply` simplifies when it can:

```{r,echo=TRUE}
sapply(d, FUN=function(x) 2^x)
```

## mapply: the Multivariate Version

`mapply(FUN, ..., MoreArgs=, SIMPLIFY=, USE.NAMES=)` applies `FUN` to the first elements of all vectors, then the second, and so on:

```{r,echo=TRUE}
mapply(paste, c(1,2,3,4,5), c("a","b","c","d","e"),
       c("A","B","C","D","E"), MoreArgs=list(sep="-"))
```

Arguments: the vectors in `...`; `MoreArgs` — a list of fixed extra arguments for `FUN`; `SIMPLIFY` — simplify the result (default `TRUE`); `USE.NAMES` — name results from the first vector.

## The plyr Library

Confused by the apply zoo — inconsistent names, arguments, returns? The **plyr** package offers 12 logically named functions: input type (array `a` / data frame `d` / list `l`) × output type (`a`/`d`/`l`/`_` to discard):

| Input ↓ Output → | array | data frame | list | discard |
|---|---|---|---|---|
| array | `aaply` | `adply` | `alply` | `a_ply` |
| data frame | `daply` | `ddply` | `dlply` | `d_ply` |
| list | `laply` | `ldply` | `llply` | `l_ply` |

Shared arguments: `.data`, `.fun`, `.progress` (progress bar), `.parallel` (via foreach), `...` to `.fun`; arrays add `.margins`; data frames add `.variables` (split-by) and `.drop`.

```{r,echo=TRUE,message=FALSE}
library(plyr)
llply(.data=d, .fun=function(x) 2^x)$x   # = lapply
aaply(.data=d, .margins=2, .fun=max)     # labeled results
```

(No plyr equivalent of `tapply` exists. Modern note: plyr's ideas grew into `dplyr` and `purrr`.)

# Binning Data

## Shingles and equal.count

Binning groups observations by a variable's value — e.g., daily data summarized monthly. **Shingles** (Chapter 7) represent possibly overlapping intervals, used in lattice for conditioning on numeric variables:

```{r,eval=FALSE,echo=TRUE}
shingle(x, intervals=sort(unique(x)))   # intervals: vector or 2-col matrix
equal.count(x, ...)                     # bins with equal observation counts
```

## cut: Continuous to Discrete

`cut` splits a numeric vector into a **factor**, one level per interval (a Date version exists too):

```{r,eval=FALSE,echo=TRUE}
cut(x, breaks, labels = NULL, include.lowest = FALSE, right = TRUE,
    dig.lab = 3, ordered_result = FALSE, ...)
```

| Argument | Description | Default |
|---|---|---|
| `breaks` | bin count (one integer) or break points (vector) | |
| `labels` | level labels | `NULL` |
| `include.lowest` | does the boundary value join the end bin? | `FALSE` |
| `right` | intervals closed on the right? | `TRUE` |
| `dig.lab` | digits in auto-generated labels | 3 |
| `ordered_result` | return an ordered factor? | `FALSE` |

## cut in Action: Batting Averages

Count players by batting-average range — the book's `batting.2008` data set, fetched per course convention:

```{r,echo=TRUE}
temp <- tempfile(fileext = ".rda")
download.file("https://raw.githubusercontent.com/cran/nutshell/master/data/batting.2008.rda",
              temp, mode = "wb")
load(temp)
batting.2008.AB <- transform(batting.2008, AVG = H/AB)
batting.2008.over100AB <- subset(batting.2008.AB, subset=(AB > 100))
battingavg.2008.bins <- cut(batting.2008.over100AB$AVG, breaks=10)
table(battingavg.2008.bins)
```

## make.groups: Combine with a Source Label

`make.groups` (lattice) stacks similar objects into one data frame, labeling each row's origin:

```{r,echo=TRUE}
library(lattice)
hat.sizes <- seq(from=6.25, to=7.75, by=.25)
pants.sizes <- c(30, 31, 32, 33, 34, 36, 38, 40)
shoe.sizes <- seq(from=7, to=12)
head(make.groups(hat.sizes, pants.sizes, shoe.sizes), 9)
```

# Subsets

## Bracket Notation

Too much data — irrelevant rows, or simply more than memory holds? Slice it. A logical expression indexes the rows; a name vector picks columns:

```{r,echo=TRUE}
batting.2008.short <- batting.2008[batting.2008$yearID==2008,
                       c("nameFirst", "nameLast", "AB", "H", "BB")]
head(batting.2008.short, 4)
```

## The subset Function

`subset(x, subset, select, drop=FALSE)` does nothing brackets can't — but reads better, since column names work *bare* inside it:

```{r,echo=TRUE}
batting.2008.short <- subset(batting.2008, yearID==2008,
                             c("nameFirst", "nameLast", "AB", "H", "BB"))
head(batting.2008.short, 4)
```

Arguments: `subset` — logical expression for rows; `select` — expression for columns; `drop` — passed to `[`.

## Random Sampling

`sample(x, size, replace=FALSE, prob=NULL)` draws a random sample of a *vector's elements*. Nonintuitively, on a data frame it samples **columns** (a data frame is a list of columns!) — so sample *row numbers* instead:

```{r,echo=TRUE}
batting.2008[sample(1:nrow(batting.2008), 5), 9:14]
```

Cleverer subsets follow the same trick — all statistics for three randomly chosen teams:

```{r,echo=TRUE}
batting.2008$teamID <- as.factor(batting.2008$teamID)
batting.2008.3teams <- batting.2008[is.element(batting.2008$teamID,
                         sample(levels(batting.2008$teamID), 3)), ]
summary(batting.2008.3teams$teamID)[1:10]
```

For stratified, cluster, or maximum-entropy sampling, see the `sampling` package.

# Summarizing Functions

## tapply

`tapply(X, INDEX, FUN=, ..., simplify=)` summarizes vector `X` within groups defined by the factor list `INDEX`. Home runs by team:

```{r,echo=TRUE}
tapply(X=batting.2008$HR, INDEX=list(batting.2008$teamID), FUN=sum)
```

Arguments: `X` — the object; `INDEX` — list of factors, each as long as `X`; `FUN`; `...` to `FUN`; `simplify` — array if `FUN` is scalar (default), list otherwise.

## tapply: Richer Summaries, More Dimensions

`FUN` may return several values — `fivenum` (min, lower hinge, median, upper hinge, max) of batting average by league:

```{r,echo=TRUE}
tapply(X=(batting.2008$H/batting.2008$AB),
       INDEX=list(batting.2008$lgID), FUN=fivenum)
```

And `INDEX` may hold several factors — mean home runs by league × batting hand:

```{r,echo=TRUE}
tapply(X=(batting.2008$HR),
       INDEX=list(batting.2008$lgID, batting.2008$bats), FUN=mean)
```

## by: tapply for Data Frames

`by` is `tapply`'s data-frame sibling (`INDICES` replaces `INDEX`) — whole-column means per group:

```{r,echo=TRUE}
by(batting.2008[, c("H", "2B", "3B", "HR")],
   INDICES=list(batting.2008$lgID, batting.2008$bats), FUN=colMeans)
```

## aggregate

`aggregate(x, by, FUN, ...)` summarizes several columns at once (a time-series variant takes `nfrequency`, `ndeltat`, `ts.eps` instead). Team totals:

```{r,echo=TRUE}
aggregate(x=batting.2008[, c("AB", "H", "BB", "2B", "3B", "HR")],
          by=list(batting.2008$teamID), FUN=sum)[1:8, ]
```

Arguments: `x`; `by` — list of grouping elements as long as `x`; `FUN` — scalar summary (time-series default: `sum`); `...` to `FUN`.

## rowsum

Just sums by group? `rowsum(x, group, reorder=TRUE)` is the direct route:

```{r,echo=TRUE}
rowsum(batting.2008[, c("AB", "H", "BB", "2B", "3B", "HR")],
       group=batting.2008$teamID)[1:8, ]
```

## Counting: tabulate and table

`tabulate` counts elements at each integer value (labels not included — add them):

```{r,echo=TRUE}
HR.cnts <- tabulate(batting.2008$HR)
names(HR.cnts) <- 0:(length(HR.cnts) - 1)
HR.cnts[1:12]
```

For **categorical** values, `table` returns a table object of counts (SAS users: this is `PROC FREQ`):

```{r,echo=TRUE}
table(batting.2008$bats)
```

Arguments of `table`: `...` — factors (or coercibles); `exclude` — levels to drop; `useNA` — `"no"`/`"ifany"`/`"always"`; `dnn` — dimension names; `deparse.level` — how to auto-name dimensions.

## Multidimensional Tables and xtabs

More factors, more dimensions — batting hand × throwing hand (add `lgID` for a third):

```{r,echo=TRUE}
table(batting.2008[, c("bats", "throws")])
```

`xtabs` builds the same contingency tables from a **formula** — often less typing:

```{r,echo=TRUE}
xtabs(~bats+lgID, batting.2008)
```

And numeric variables join the party via `cut` first, `table` second — exactly the batting-average binning shown earlier.

# Reshaping Data

## Narrow vs. Wide, and t

Data warehouses often store one *fact* per row ("narrow"); analysis may need one column per concept ("wide") — or vice versa. The simplest reshaper is `t`, transpose (a vector becomes a one-row matrix):

```{r,echo=TRUE}
m <- matrix(1:10, nrow=5)
t(m)
t(1:10)
```

## stack and unstack

Working on the narrow quote table (symbol, date, close):

```{r,echo=TRUE}
my.quotes.narrow <- my.quotes[, c("symbol", "Date", "Close")]
unstacked <- unstack(my.quotes.narrow, form=Close~symbol)
unstacked
stack(unstacked)[1:6, ]
```

In the `form` formula, the right side names the unstacking variable, the left the values. Order survives; the Date column does **not** — `unstack` suits data where only two variables matter.

## reshape: Long to Wide

`reshape` is the powerful (and notoriously two-faced) tool. Rows = dates, columns = stocks:

```{r,echo=TRUE}
my.quotes.wide <- reshape(my.quotes.narrow, idvar="Date",
                          timevar="symbol", direction="wide")
my.quotes.wide
```

Its parameters ride along as attributes (`attributes(my.quotes.wide)$reshapeWide` stores `idvar`, `timevar`, `varying`...) — which is why `reshape(my.quotes.wide)` with **no arguments** reverses the operation.

## reshape: Variations

Swap the roles — rows = stocks, columns = dates; or carry **several** value columns at once:

```{r,echo=TRUE}
reshape(my.quotes.narrow, idvar="symbol", timevar="Date",
        direction="wide")
my.quotes.oc <- my.quotes[, c("symbol", "Date", "Close", "Open")]
reshape(my.quotes.oc, timevar="Date", idvar="symbol",
        direction="wide")[, 1:5]
```

## reshape: Arguments

Two functions in one — `direction="wide"` requires `idvar` + `timevar`; `direction="long"` requires `varying`:

| Argument | Description |
|---|---|
| `data` | data frame to reshape |
| `varying` | wide-format variables that become unique rows in long format |
| `v.names` | long-format variables that become columns in wide format |
| `timevar` / `idvar` | long-format variables identifying observation / group (defaults `"time"` / `"id"`) |
| `ids` / `times` | values for new id / time variables |
| `drop` | variables to exclude |
| `direction` | `"wide"` or `"long"` |
| `new.row.names` | build row names from id × time? |
| `sep` / `split` | how wide names encode variable × time (default sep `"."`) |

## The reshape Package: Melt and Cast

Hadley Wickham's **reshape** package (not the function!) replaces all that with one intuitive model: **melt** a table into transaction-like records, then **cast** the records into any shape:

```{r,echo=TRUE,message=FALSE,warning=FALSE}
library(reshape)
my.molten.quotes <- melt(my.quotes[, 1:8])
head(my.molten.quotes)
```

## cast: Three Shapes from One Molten Frame

```{r,echo=TRUE}
cast(data=my.molten.quotes, variable~Date, subset=(symbol=='GE'))
cast(data=my.molten.quotes, symbol~Date, subset=(variable=='Adj.Close'))
```

A `|` in the formula returns a **list** of tables — `cast(my.molten.quotes, Date~variable|symbol)` yields one quote table per stock.

## melt and cast: Arguments

`melt.data.frame(data, id.vars, measure.vars, variable_name, na.rm, ...)` — `id.vars` identify observations, `measure.vars` are the measurements (unspecified: factors/characters are ids, the rest measures); `variable_name` defaults to `"variable"`. Array and list methods exist too (`melt.array` takes `varnames`; `melt.list` melts recursively).

`cast(data, formula, fun.aggregate=NULL, ..., margins, subset, fill, add.missing, value)`:

- `formula` — `x_1 + x_2 ~ y_1 ~ ... | list_var`, with `...` = "all other variables" and `.` = "none" (default `...~variable`);
- `fun.aggregate` — how to combine multiple values per cell; `margins` — compute margins; `subset` — filter the molten data; `fill` / `add.missing` — handle absent combinations; `value` — name of the value column.

(Modern note: reshape begat `reshape2`, now succeeded by `tidyr`'s `pivot_longer`/`pivot_wider` — same melt/cast ideas.)

# Data Cleaning

## What Cleaning Means

::: {.callout-note}
**Surprises are the rule.** The author's credit-data war story: valid FICO scores span 340–840, yet the data brimmed with 997/998/999 — special codes like "insufficient data," not stellar credit. Hospital records: two patients sharing a name get fused into one record; one patient seeing two doctors becomes two records. Cleaning **does not change data's meaning** — it removes artifacts of collection, processing, and storage so they can't contaminate the analysis.
:::

## Finding and Removing Duplicates

Suppose GE sneaked into the ticker list twice. `duplicated` flags rows that repeat earlier rows; filter with it, or let `unique` do both steps:

```{r,echo=TRUE}
my.quotes.2 <- rbind(my.quotes, my.quotes[my.quotes$symbol == "GE", ])
duplicated(my.quotes.2)
my.quotes.unique <- my.quotes.2[!duplicated(my.quotes.2), ]
my.quotes.unique <- unique(my.quotes.2)        # same result
nrow(my.quotes.unique)
```

# Sorting

## sort

```{r,echo=TRUE}
w <- c(5, 4, 7, 2, 7, 1)
sort(w)
sort(w, decreasing=TRUE)
length(w) <- 7                      # introduce an NA
sort(w)                             # na.last=NA: NAs dropped
sort(w, na.last=TRUE)
sort(w, na.last=FALSE)
```

## order: Sorting Data Frames

`sort` sorts a vector; data frames need a **permutation of row indices**, supplied by `order(..., na.last=, decreasing=)` — it sorts by each argument vector, breaking ties with the next:

```{r,echo=TRUE}
v <- c(11, 12, 13, 15, 14)
order(v)
v[order(v)]
u <- c("pig", "cow", "duck", "horse", "rat")
w <- data.frame(v, u)
w[order(w$v), ]
```

## Sorting by One or More Columns

The quotes by closing price; then by symbol, ties broken by price:

```{r,echo=TRUE}
my.quotes[order(my.quotes$Close), ][1:5, 1:6]
my.quotes[order(my.quotes$symbol, my.quotes$Close), ][1:6, 1:6]
```

## Sorting by All Columns: do.call

Whole-frame sorting holds a trap: `order(my.quotes)` treats the data frame as *one long vector*. `order` wants a list of vectors — deliver the data frame's columns as arguments via `do.call`:

```{r,echo=TRUE}
do.call(order, my.quotes)
my.quotes[do.call(order, my.quotes), ][1:6, 1:6]
```

## Copyright Notice

- **Copyright.** These slides are adapted from *R in a Nutshell: A Desktop Quick Reference* (2nd ed.) by Joseph Adler, O'Reilly Media. All rights reserved by the original author and publisher.

- **Non-commercial use only.** These materials are strictly for educational purposes and may not be used for commercial gain.

- **Attribution.** Any reproduction, distribution, or use of these materials must properly credit the original source.
