---
title: "Programming for Applications"
subtitle: "Chapter 14: Lattice Graphics"
author: "Yu-You Liou (NTU)"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "R in a Nutshell: A Desktop Quick Reference"
logo: "logo.svg"
editor: source
bibliography: bib.bib
suppress-bibliography: true
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch14-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center")
library(lattice)
```

# **Trellis Plots in R**

## History and Idea

The `lattice` package [@sarkar2008lattice] implements **trellis graphics**, Cleveland's framework for visualizing *conditioned* data [@cleveland1993visualizing]: the same panel repeated for every subset of the data, arranged in a grid for instant comparison.

- Formula interface: `y ~ x | condition` reads "y against x, one panel per level of condition."
- Lattice functions **return objects**; the plot appears when the object is printed — remember this inside scripts and functions.

```{r,echo=TRUE,cache=TRUE}
temp <- tempfile(fileext = ".rda")
download.file("https://raw.githubusercontent.com/cran/nutshell/master/data/births2006.smpl.rda", temp, mode = "wb")
load(temp)
dim(births2006.smpl)
```

## A Simple Example

US births in 2006 (a 10% sample): how does birth weight relate to delivery method?

```{r,echo=TRUE,out.width="70%"}
histogram(~ DBWT | DMETH_REC, data = births2006.smpl,
          xlab = "Birth weight (g)")
```

One formula produced three aligned panels — conditioning without loops.

## Using Lattice Functions

All high-level lattice functions share the same anatomy:

```{r,eval=FALSE,echo=TRUE}
xyplot(y ~ x | g,            # formula with optional conditioning
       data = df,            # where the variables live
       groups = h,           # overlay groups within panels
       layout = c(3, 1),     # panel arrangement
       main = "...", xlab = "...", ylab = "...")
```

- `groups=` distinguishes subsets *inside* a panel (with `auto.key=TRUE` for a legend); `|` separates subsets *across* panels.
- The returned `"trellis"` object can be stored, updated with `update()`, and printed later.

## Univariate Plots: Histograms and Densities

```{r,echo=TRUE,out.width="68%"}
densityplot(~ DBWT | DPLURAL, data = births2006.smpl,
            plot.points = FALSE, layout = c(1, 5),
            xlab = "Birth weight (g)")
```

Stacking the panels vertically (one per plurality: singles, twins, triplets...) aligns the x-axes — multiples are born lighter, and the whole story is one column tall.

## Univariate Plots: Strips and Quantiles

- `stripplot()` — one point per observation, ideal for small groups;
- `qqmath()` — quantiles of a sample against a theoretical distribution;
- `bwplot()` — box-and-whisker, the lattice cousin of `boxplot()`:

```{r,echo=TRUE,out.width="68%"}
bwplot(DBWT ~ factor(APGAR5), data = births2006.smpl,
       xlab = "5-minute Apgar score", ylab = "Birth weight (g)")
```

Weight and Apgar score rise together — visible across twenty boxes at once.

## Bivariate Plots

`xyplot()` is the conditioned scatter plot. Average birth weight by day of week, split by delivery method:

```{r,echo=TRUE,out.width="70%"}
wt <- aggregate(DBWT ~ DOB_WK + DMETH_REC, data = births2006.smpl, mean)
xyplot(DBWT ~ DOB_WK | DMETH_REC, data = wt, type = "h",
       layout = c(3, 1), xlab = "Day of week", ylab = "Mean weight (g)")
```

Other bivariate friends: `barchart()`, `dotplot()` (Cleveland's dot plot — often the best replacement for both pies and bars).

## Trivariate Plots

For three numeric dimensions:

```{r,echo=TRUE,out.width="62%"}
x <- seq(-2, 2, length.out = 40)
grid <- expand.grid(x = x, y = x)
grid$z <- with(grid, exp(-(x^2 + y^2)))
levelplot(z ~ x * y, data = grid)
```

- `levelplot()` — colored tiles; `contourplot()` — labeled contour lines;
- `cloud()` — 3-D point clouds; `wireframe()` — 3-D surfaces.

## Custom Panel Functions

A panel function decides what gets drawn inside each rectangle. Wrap the default and add to it:

```{r,echo=TRUE,out.width="65%"}
xyplot(DBWT ~ WTGAIN | SEX,
       data = births2006.smpl[sample(nrow(births2006.smpl), 2000), ],
       alpha = 0.2,
       panel = function(x, y, ...) {
         panel.xyplot(x, y, ...)
         panel.lmline(x, y, col = "red", lwd = 2)
       },
       xlab = "Maternal weight gain (lb)", ylab = "Birth weight (g)")
```

`panel.lmline`, `panel.loess`, `panel.abline`, and `panel.grid` are ready-made building blocks.

## Customizing Lattice

- `trellis.par.get()` / `trellis.par.set()` adjust the theme (colors, fonts, symbols) globally.
- `show.settings()` displays the current theme at a glance.
- Per-plot: `par.settings=` injects theme tweaks into a single call;
- `update(plot_object, ...)` modifies an existing trellis object — handy for polishing a stored plot without recomputing.

```{r,eval=FALSE,echo=TRUE}
my.theme <- trellis.par.get()
my.theme$plot.symbol$pch <- 19
trellis.par.set(my.theme)
```

## Lattice vs. Base Graphics

| | Base | Lattice |
|:---|:---|:---|
| Conditioning | manual loops + `par(mfrow)` | `| g` in the formula |
| Return value | none (ink on paper) | trellis object |
| Legends | manual `legend()` | `auto.key = TRUE` |
| Consistency across panels | your job | automatic |

Lattice shines exactly when the question is "does this pattern hold *within every subgroup*?" — Chapter 15 shows `ggplot2`, a third system built on a full grammar.

## Copyright Notice

- **Copyright.** These slides are adapted from *R in a Nutshell: A Desktop Quick Reference* (2nd ed.) by Joseph Adler, O'Reilly Media. All rights reserved by the original author and publisher.

- **Non-commercial use only.** These materials are strictly for educational purposes and may not be used for commercial gain.

- **Attribution.** Any reproduction, distribution, or use of these materials must properly credit the original source.
