---
title: "Programming for Applications"
subtitle: "Chapter 14: Lattice Graphics"
author: "Yu-You Liou (NTU)"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "R in a Nutshell: A Desktop Quick Reference"
logo: "logo.svg"
editor: source
bibliography: bib.bib
suppress-bibliography: true
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch14-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center", fig.height = 4.2)
library(lattice)
```

# **Trellis Graphics**

## History and Strengths

In the early 1990s, Richard Becker and William Cleveland at Bell Labs built **Trellis graphics**, grounded in Cleveland's research on how people actually perceive information. The **lattice** package is R's implementation (some function names still say "trellis").

Lattice's real strength: **splitting a chart into panels** (a grid) or **groups** (colors/symbols) using conditioning or grouping variables — plus clean, readable output by default. Familiar charts (scatter, bar, histogram) and new friends (dot plots, strip plots, Q-Q plots) all included.

## How Lattice Works

The vocabulary: rectangular drawing areas are **panels**; the data sent to a panel is a **packet**; **panel functions** actually draw packets. The lifecycle:

1. you call a high-level lattice function;
2. it assembles and *returns* a lattice object (class `"trellis"` — hence `print.trellis`, `plot.trellis`);
3. printing the object (automatic at the console) is what draws it;
4. `plot.trellis` lays out the panel matrix, assigns packets, and calls the **panel function** for each.

::: {.callout-warning}
**Inside functions and scripts, lattice plots nothing unless you `print` the object** — the classic lattice gotcha (first met in Chapter 3).
:::

## A Simple Example: Formula, Conditioning, Groups

Lattice arguments are consistent: a **formula** first (positional — its name has changed before!), then `data`. The same toy data three ways:

```{r,echo=TRUE}
d <- data.frame(x=c(0:9), y=c(1:10), z=c(rep(c("a", "b"), times=5)))
xyplot(y~x, data=d)
xyplot(y~x|z, data=d)            # conditioning: panels split by z
xyplot(y~x, groups=z, data=d)    # grouping: colors/symbols by z
```

`y~x|z` reads "y against x, conditioned on z."

## Graphics-to-Lattice Dictionary

| graphics function | lattice function | Chart |
|---|---|---|
| `barplot` | `barchart` | bar/column charts |
| `dotchart` | `dotplot` | Cleveland dot plots |
| `hist` | `histogram` | histograms |
| `density`+`plot` | `densityplot` | kernel density plots |
| `stripchart` | `stripplot` | strip charts |
| `qqnorm` (stats) | `qqmath` | theoretical Q-Q plots |
| `plot` | `xyplot` | scatter plots |
| `qqplot` (stats) | `qq` | two-sample Q-Q plots |
| `pairs` | `splom` | scatter plot matrices |
| `image` | `levelplot` | image plots |
| `contour` | `contourplot` | contour plots |
| `persp` | `cloud`, `wireframe` | 3-D perspective |

Most functions accept a formula + data frame; many also take arrays, matrices, tables, or vectors directly (e.g., `barchart` happily eats a `table`).

## Custom Panel Functions

To superimpose extra elements, replace the **panel function** — typically a wrapper that adds something, then delegates:

```{r,echo=TRUE}
xyplot(y~x|z, data=d,
  panel=function(...){
    panel.abline(a=1, b=1)
    panel.xyplot(...)
  })
```

# Univariate Trellis Plots

## The Births Data

Running example: every US birth of 2006 (CDC natality file — 3.1 GB raw; the book preprocessed it with a Perl `unpack` script, recoded factors in R, and sampled 10% into `births2006.smpl`). Variables include `DOB_MM`/`DOB_WK` (month/weekday of birth), `MAGER` (mother's age), `WTGAIN`, `SEX`, `APGAR5`, `DMEDUC`, `UPREVIS`, `ESTGEST`, `DMETH_REC` (delivery method), `DPLURAL` (single/twin/...), `DBWT` (birth weight, grams):

```{r,echo=TRUE}
temp <- tempfile(fileext = ".rda")
download.file("https://raw.githubusercontent.com/cran/nutshell/master/data/births2006.smpl.rda",
              temp, mode = "wb")
load(temp)
dim(births2006.smpl)
```

## barchart

Births by day of week (1 = Sunday):

```{r,echo=TRUE}
births.dow <- table(births2006.smpl$DOB_WK)
barchart(births.dow)
```

Far more babies arrive on weekdays — surprising! Is it the *delivery method*? Tabulate by weekday × method (dropping unknowns):

```{r,echo=TRUE}
births2006.dm <- transform(
  births2006.smpl[births2006.smpl$DMETH_REC != "Unknown",],
  DMETH_REC=as.factor(as.character(DMETH_REC)))
dob.dm.tbl <- table(WK=births2006.dm$DOB_WK, MM=births2006.dm$DMETH_REC)
barchart(dob.dm.tbl)            # default: stacked groups, no key
```

## Iterating to the Right Display

Unstack and add a key; still hard to compare — so drop the color groups and switch to columns in **two panels**:

```{r,echo=TRUE}
barchart(dob.dm.tbl, stack=FALSE, auto.key=TRUE)
barchart(dob.dm.tbl, horizontal=FALSE, groups=FALSE)
```

Now it's plain: vaginal births dip ~25–30% on weekends, **C-sections drop 50–60%**. Lattice's quick iteration — stack/unstack, group/panel, flip — is exactly how you zero in on the story.

## dotplot

Cleveland **dot plots** show one point per category — calmer than bars for bigger tables. Is the weekday pattern seasonal? Count by week × month × method and plot:

```{r,echo=TRUE}
dob.dm.tbl.alt <- table(WEEK=births2006.dm$DOB_WK,
                        MONTH=births2006.dm$DOB_MM,
                        METHOD=births2006.dm$DMETH_REC)
dotplot(dob.dm.tbl.alt, stack=FALSE, auto.key=TRUE, groups=TRUE)
```

Slight seasonal wiggles; the weekly pattern persists year-round.

## dotplot: the Tire Failure Data

NHTSA's 2003 "Stepped-Up Speed to Failure" tests (post-Firestone recall, run in hot Phoenix): tires run at increasing speeds until failure. Six tire types (B–L, various sizes/brands). Variables: `Time_To_Failure`, `Speed_At_Failure_km_h` (a stepped value — treat as a factor!), `Tire_Type`:

```{r,echo=TRUE}
temp <- tempfile(fileext = ".rda")
download.file("https://raw.githubusercontent.com/cran/nutshell/master/data/tires.sus.rda",
              temp, mode = "wb")
load(temp)
dotplot(as.factor(Speed_At_Failure_km_h)~Time_To_Failure|Tire_Type,
        data=tires.sus)
```

All type D tires died fast at 180 km/h; some type H tires endured. (Revisited in Chapter 18.)

## histogram

Birth weight by plurality — note alphabetical panel order (hence the numbered level names) and the space-filling square panels:

```{r,echo=TRUE}
histogram(~DBWT|DPLURAL, data=births2006.smpl)
```

Stack the panels with `layout` for comparability:

```{r,echo=TRUE,fig.height=5}
histogram(~DBWT|DPLURAL, data=births2006.smpl, layout=c(1, 5))
```

Roughly normal within groups; mean weight falls as plurality rises. (Other arguments: `type` = `"percent"`/`"count"`/`"density"`, `nint`/`breaks`/`endpoints`/`equal.widths` for binning.)

## densityplot

One smooth line instead of bins — and stackable. With 427k observations we suppress the per-point strip (`plot.points=FALSE`):

```{r,echo=TRUE,fig.height=5}
densityplot(~DBWT|DPLURAL, data=births2006.smpl,
            layout=c(1,5), plot.points=FALSE)
```

Density plots *superimpose* gracefully — turn the conditioning variable into a grouping one:

```{r,echo=TRUE}
densityplot(~DBWT, groups=DPLURAL, data=births2006.smpl,
            plot.points=FALSE, auto.key=TRUE)
```

## stripplot

**Strip plots** are one-dimensional scatter plots — many values per category (vs. dot plots' one), ideal for *small* data. Weights of quadruplets-and-up (44 observations), jittered:

```{r,echo=TRUE}
stripplot(~DBWT, data=births2006.smpl,
          subset=(DPLURAL=="5 Quintuplet or higher" |
                  DPLURAL=="4 Quadruplet"),
          jitter.data=TRUE)
```

## qqmath

`qqmath` compares sample quantiles to a theoretical distribution (default: normal via `qnorm`; also `f.value`, `distribution`, `qtype`). Sanity check with actual normals, then the births:

```{r,echo=TRUE}
qqmath(rnorm(100000))
qqmath(~DBWT|DPLURAL,
       data=births2006.smpl[sample(1:nrow(births2006.smpl), 50000), ],
       pch=19, cex=0.25,
       subset=(DPLURAL != "5 Quintuplet or higher"))
```

Birth weights: not quite normal.

## The San Francisco Home Sales Data

Second running example: 3,281 San Francisco home sales, 2008-02-13 to 2009-07-14, compiled by the author from newspaper listings, geocoded, and joined to Zillow neighborhood shapefiles (with `sp`/`maptools` and `point.in.polygon`). Variables: `street`, `city`, `zip`, `saledate`, `price`, `bedrooms`, `squarefeet`, `lotsize`, `yearbuilt`, `condolike`, `latitude`/`longitude`, `neighborhood`. Real data — imperfect, partially missing:

```{r,echo=TRUE}
temp <- tempfile(fileext = ".rda")
download.file("https://raw.githubusercontent.com/cran/nutshell/master/data/sanfrancisco.home.sales.rda",
              temp, mode = "wb")
load(temp)
```

## qqmath on Prices: Finding the Log-Normal

Prices range $100,000–$9,500,000 — intuitively not normal:

```{r,echo=TRUE}
qqmath(~price, data=sanfrancisco.home.sales)
qqmath(~log(price), data=sanfrancisco.home.sales)
```

The log transform straightens the line nicely. Now split by bedrooms — groups, smooth lines instead of points (`type="smooth"`, an argument that flows to the panel function), explicit subset to keep the key clean:

```{r,echo=TRUE}
qqmath(~log(price), groups=bedrooms,
       data=subset(sanfrancisco.home.sales,
                   !is.na(bedrooms) & bedrooms > 0 & bedrooms < 7),
       auto.key=TRUE, drop.unused.levels=TRUE, type="smooth")
```

Separated lines, higher for more bedrooms — the same trick works for square-footage sextiles (`Hmisc::cut2(squarefeet, g=6)` as `groups`) and neighborhoods.

# Bivariate Trellis Plots

## xyplot

Price vs. size — first naively, then trimmed (price < $4M, < 6,000 sq ft) and conditioned by zip code with custom strips:

```{r,echo=TRUE}
xyplot(price~squarefeet, data=sanfrancisco.home.sales)
trellis.par.set(fontsize=list(text=7))
xyplot(price~squarefeet|zip, data=sanfrancisco.home.sales,
       subset=(zip!=94100 & zip!=94104 & zip!=94108 &
               zip!=94111 & zip!=94133 & zip!=94158 &
               price < 4000000 &
               ifelse(is.na(squarefeet), FALSE, squarefeet < 6000)),
       strip=strip.custom(strip.levels=TRUE))
trellis.par.set(fontsize=list(text=12))
```

The linear relationship sharpens within each zip — neighborhood conditioning (`price~squarefeet|neighborhood`, with `pch=19, cex=.2`) tells the same story with better labels.

## bwplot: Box Plots by Month

Distribution of prices over time — round dates to months with `cut`. The raw attempt drowns in outliers; log-transform and rotate the labels:

```{r,echo=TRUE}
bwplot(log(price)~cut(saledate, "month"),
       data=sanfrancisco.home.sales,
       scales=list(x=list(rot=90)))
```

Medians wandered; the interquartile range moved less; the high end visibly thinned — yet the basic shape stayed stable month to month. (Volume note: the data starts mid-February and ends mid-July — don't over-read those months.)

## splom and qq

- **`splom(x, data, ...)`** — scatter plot matrices (lattice's `pairs`); panels drawn by `panel.splom`, the matrix orchestrated by `superpanel=panel.pairs`, axis detail via `pscales`, names via `varnames`.
- **`qq(x, data, ...)`** — *two-sample* quantile-quantile plots, comparing two observed distributions (formula syntax `factor ~ value`).

# Trivariate Plots

## levelplot

Flat grids colored by a third value (drawn by `panel.levelplot`; key arguments `at`, `cuts=15`, `region`, `colorkey`, `col.regions`, `contour`). Coordinates are too precise to plot raw — bin first with `cut` + `table`:

```{r,echo=TRUE}
attach(sanfrancisco.home.sales)
levelplot(table(cut(longitude, breaks=40), cut(latitude, breaks=40)),
          scales=list(y=list(cex=.5), x=list(rot=90, cex=.5)))
```

Mean *price* by area? Swap `table` for `tapply`:

```{r,echo=TRUE}
levelplot(tapply(price,
                 INDEX=list(cut(longitude, breaks=40),
                            cut(latitude, breaks=40)),
                 FUN=mean),
          scales=list(draw=FALSE))
```

## levelplot with Conditioning; contourplot

A third factor makes panels — sales location by (capped) bedroom count:

```{r,echo=TRUE}
bedrooms.capped <- ifelse(bedrooms < 5, bedrooms, 5)
levelplot(table(cut(longitude, breaks=25),
                cut(latitude, breaks=25),
                bedrooms.capped),
          scales=list(draw=FALSE))
detach(sanfrancisco.home.sales)
```

**`contourplot`** is the topographic-map sibling (`cuts=7`, `labels`, `contour=TRUE`, `region=FALSE`).

## cloud and wireframe

- **`cloud`** — 3-D point clouds (projections of 3-D points onto 2-D); arguments include `aspect=c(1,1)`, `panel.aspect`, drawing handled by `panel.cloud` → `panel.3dscatter`.
- **`wireframe`** — 3-D surfaces, lattice's `persp` (`panel.3dwire` underneath).
- Other lattice plots worth knowing: **`rfs`** (residual and fit spread plots for a model, optionally with `distribution=qnorm`) and **`parallel`** (parallel-coordinate plots).

# Customizing Lattice Graphics

## Common Arguments to Lattice Functions

Shared by the high-level functions:

| Argument | Description |
|---|---|
| `x` (formula), `data` | what to plot, where the variables live |
| `allow.multiple`, `outer` | handling of `y1+y2~x` formulas: superpose or separate panels |
| `groups` / `subset` / `subscripts` | grouping variable; row filter; expose indices to panels |
| `panel` | the panel function — *the* customization hook |
| `aspect` | panel aspect ratio (`"fill"`, `"xy"` banking to 45°, numeric) |
| `auto.key` | quick legend (logical or list) |
| `prepanel` | function determining per-panel space/limits |
| `strip` | strip-drawing function (e.g., `strip.custom`); `strip.left` for left strips |
| `xlab`, `ylab`, `main`, `sub`, `xlim`, `ylim` | annotation and limits |
| `scales` | axis control (next slide) |
| `layout`, `between`, `skip` | panel arrangement, spacing |
| `drop.unused.levels`, `default.scales`, `lattice.options` | the long tail (passed via `trellis.skeleton`) |

## Controlling Axes: scales

`scales` is a list (optionally with `x` and `y` sublists):

- `relation` — `"same"` (shared limits), `"free"` (per-panel), `"sliced"` (equal-length slices);
- `draw`, `alternating`, `tick.number`, `at`, `labels`, `cex`, `rot`, `font`, `col`, `log`, `abbreviate`...

Seen in action earlier: `scales=list(x=list(rot=90))` for rotated month labels, `scales=list(draw=FALSE)` to silence map axes.

## Parameters: trellis.par.get / set

Lattice ignores `par`; its own settings live in a deep parameter list:

```{r,echo=TRUE}
trellis.par.get("axis.text")
```

```{r,eval=FALSE,echo=TRUE}
trellis.par.set(list(axis.text = list(cex = 0.5)))
show.settings()                    # visual overview of current settings
names(trellis.par.get())           # 30+ groups, ~380 parameters
```

Groups include `superpose.symbol`/`superpose.line` (groups), `plot.polygon` (bars/histograms), `box.rectangle`/`box.umbrella` (bwplot), `plot.line`/`plot.symbol` (xyplot et al.), `reference.line`, `regions`+`shade.colors` (levelplot/wireframe), `strip.background`/`strip.shingle`, `fontsize`, `par.main.text` and friends.

## plot.trellis, strip.default, simpleKey

- **`plot.trellis`** — arguments controlling drawing: `position`/`split`/`more` (plot several trellis objects on one page), `panel.height`/`panel.width`, `packet.panel` (which packet lands in which panel; default `packet.panel.default`), `panel.error`.
- **`strip.default(which.given, which.panel, var.name, factor.levels, shingle.intervals, strip.names, strip.levels, style, bg, fg, par.strip.text)`** — how strips display conditioning levels; customize via `strip.custom`.
- **`simpleKey(text, points=TRUE, rectangles=FALSE, lines=FALSE, col, cex, ...)`** — builds the legend that `auto.key` requests.

## Low-Level and Panel Functions

Grid-based primitives mirror base graphics: `llines`/`panel.lines`, `lpoints`/`panel.points`, `ltext`/`panel.text`, `lsegments`/`panel.segments`, `lpolygon`/`panel.polygon`, `larrows`/`panel.arrows`, `lrect`/`panel.rect`, plus `panel.axis` and `panel.superpose`.

Ready-made panel helpers to compose into custom panels:

- `panel.abline`, `panel.curve`, `panel.rug`, `panel.grid`, `panel.fill`;
- `panel.mathdensity` (overlay a theoretical density), `panel.average`;
- `panel.loess` (smoother), `panel.lmline` (regression line);
- and each high-level function's `panel.*` (e.g., `panel.xyplot`, `panel.barchart`, `panel.histogram`) for delegation — the pattern from our first custom panel.

## Copyright Notice

- **Copyright.** These slides are adapted from *R in a Nutshell: A Desktop Quick Reference* (2nd ed.) by Joseph Adler, O'Reilly Media. All rights reserved by the original author and publisher.

- **Non-commercial use only.** These materials are strictly for educational purposes and may not be used for commercial gain.

- **Attribution.** Any reproduction, distribution, or use of these materials must properly credit the original source.
