---
title: "Programming for Applications"
subtitle: "Chapter 21: Classification Models"
author: "Yu-You Liou (NTU)"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "R in a Nutshell: A Desktop Quick Reference"
logo: "logo.svg"
editor: source
bibliography: bib.bib
suppress-bibliography: true
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch21-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center", fig.height = 4.2)
```

# **Predicting Categories**

## Regression's Categorical Cousin

A **classification model** predicts a *categorical* response — spam or not, disease or healthy, which of several classes. The toolkit parallels Chapter 20, but the response is a factor. The chapter's running example is **Spambase**: 4,601 emails with 57 numeric features, each labeled spam or not.

We move from linear methods (logistic regression, discriminant analysis) to trees and ensembles (bagging, boosting, random forests) to neural networks and SVMs.

# Linear Classification Models

## Logistic Regression

The workhorse for a **binary** response: a `glm` with `family=binomial`, modeling the log-odds of the outcome:

```{r,eval=FALSE,echo=TRUE}
spam.model <- glm(is_spam ~ ., data=spambase, family=binomial(link="logit"))
summary(spam.model)
predict(spam.model, newdata, type="response")   # predicted probabilities
```

- `summary` reports coefficients, z-values, p-values, plus **null and residual deviance** and **AIC** (not R²).
- `type="response"` returns probabilities; threshold them (e.g., at 0.5) to get class labels.
- "Number of Fisher Scoring iterations" reports how quickly the IRLS fit converged.
- Multi-class extension: `nnet::multinom` (multinomial log-linear model), or ordered outcomes via `MASS::polr`.

## Linear and Flexible Discriminant Analysis

Discriminant analysis finds the (linear) combinations of predictors that best *separate* the classes — `MASS::lda` (and `qda` for quadratic boundaries):

```{r,eval=FALSE,echo=TRUE}
library(MASS)
iris.lda <- lda(Species ~ ., data=iris)
predict(iris.lda, iris)$class          # predicted classes
```

Richer variants in package `mda`:

- **`fda`** — *flexible* discriminant analysis (nonlinear boundaries via regression);
- **`mda`** — *mixture* discriminant analysis (each class a mixture of Gaussians).

LDA assumes equal class covariances and roughly normal predictors; QDA relaxes the equal-covariance assumption.

## Log-Linear Models

For counts in a contingency table, **log-linear models** test associations among categorical variables — `stats::loglin(table, margin, ...)` (and the friendlier formula interface `MASS::loglm`):

```{r,eval=FALSE,echo=TRUE}
library(MASS)
loglm(~ Sex + Class + Survived, data=Titanic)   # mutual independence
loglm(~ Sex*Class + Survived, data=Titanic)     # Sex,Class associated
```

Arguments to `loglin`: the table, a list of margin vectors to fit, an optional starting estimate, and `maxit` (iteration cap for the IPF algorithm).

# Trees and Ensembles

## Classification Trees: rpart

`rpart` builds a tree by **recursive partitioning** — repeatedly splitting on the feature that best separates classes:

```{r,eval=FALSE,echo=TRUE}
library(rpart)
spam.tree <- rpart(is_spam ~ ., data=spambase)
spam.tree                                  # text description of the splits
printcp(spam.tree)                         # CP table: complexity vs. error
plot(spam.tree); text(spam.tree)           # draw it
spam.tree.pruned <- prune(spam.tree, cp=0.01)   # prune to avoid overfitting
```

The `printcp`/`plotcp` output (CP, nsplit, rel error, xerror) guides **pruning**: pick the complexity parameter that minimizes cross-validated error.

## Bagging and Boosting

Single trees are unstable; **ensembles** of trees are far more accurate:

- **Bagging** (bootstrap aggregating) — `ipred::bagging` or `adabag::bagging`: fit many trees to bootstrap samples, average/vote. Reduces variance.
- **Boosting** — `ada::ada` or `adabag::boosting` (AdaBoost), `gbm::gbm` (gradient boosting): fit trees *sequentially*, each focusing on the previous one's errors. Reduces bias and variance.

```{r,eval=FALSE,echo=TRUE}
library(adabag)
spam.bag  <- bagging(is_spam ~ ., data=spambase)
spam.boost <- boosting(is_spam ~ ., data=spambase)
predict(spam.boost, newdata)$class
```

These are embarrassingly parallel — each tree fits independently (Chapter 26).

## Random Forests

`randomForest` bags trees *and* randomizes the features considered at each split, decorrelating them — usually the strongest off-the-shelf classifier:

```{r,eval=FALSE,echo=TRUE}
library(randomForest)
spam.rf <- randomForest(is_spam ~ ., data=spambase, ntree=500)
spam.rf                                # OOB error estimate + confusion matrix
importance(spam.rf)                    # variable importance
predict(spam.rf, newdata)
```

Key knobs: `ntree` (number of trees), `mtry` (features tried per split). The **out-of-bag (OOB)** error is a free, honest accuracy estimate — no separate test set needed.

# Neural Networks and SVMs

## Neural Networks

`nnet::nnet` fits a single-hidden-layer feed-forward network:

```{r,eval=FALSE,echo=TRUE}
library(nnet)
spam.nn <- nnet(is_spam ~ ., data=spambase, size=10, decay=0.01, maxit=200)
predict(spam.nn, newdata, type="class")
```

`size` sets the hidden units, `decay` is weight regularization (curbing overfitting), `maxit` caps iterations. (Modern deep learning lives in `keras`/`torch`, but the idea is the same, scaled up.)

## Support Vector Machines

`e1071::svm` finds the maximum-margin boundary, kernelized for nonlinearity:

```{r,eval=FALSE,echo=TRUE}
library(e1071)
spam.svm <- svm(is_spam ~ ., data=spambase,
                type="C-classification", kernel="radial")
predict(spam.svm, newdata)
naiveBayes(is_spam ~ ., data=spambase)   # also in e1071: naive Bayes
```

`type="C-classification"` is the standard classifier; `kernel` may be `"linear"`, `"polynomial"`, `"radial"` (RBF), or `"sigmoid"`. Tune cost and kernel parameters with `tune`. The same package's `naiveBayes` gives a fast probabilistic baseline.

## k-Nearest Neighbors, and Choosing a Model

The simplest classifier of all — `class::knn(train, test, cl, k)` — labels each point by majority vote of its `k` nearest neighbors; no model is "fit" at all.

::: {.callout-tip}
**No universal best.** Logistic regression is interpretable; LDA needs few assumptions checked; trees are readable; **random forests and boosting usually win on raw accuracy**; SVMs shine in high dimensions; kNN is a quick baseline. Compare honestly with a held-out test set or cross-validation, and judge with a **confusion matrix**, not just overall accuracy.
:::

## Copyright Notice

- **Copyright.** These slides are adapted from *R in a Nutshell: A Desktop Quick Reference* (2nd ed.) by Joseph Adler, O'Reilly Media. All rights reserved by the original author and publisher.

- **Non-commercial use only.** These materials are strictly for educational purposes and may not be used for commercial gain.

- **Attribution.** Any reproduction, distribution, or use of these materials must properly credit the original source.
