---
title: "Programming for Applications"
subtitle: "Chapter 21: Classification Models"
author: "Yu-You Liou (NTU)"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "R in a Nutshell: A Desktop Quick Reference"
logo: "logo.svg"
editor: source
bibliography: bib.bib
suppress-bibliography: true
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch21-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center")
```

# **Predicting Categories**

## The Spam Problem

Classification predicts a **category** rather than a number. Our running example: 4,601 emails, 57 numeric features, one label — spam or not:

```{r,echo=TRUE,cache=TRUE}
temp <- tempfile(fileext = ".rda")
download.file("https://raw.githubusercontent.com/cran/nutshell/master/data/spambase.rda", temp, mode = "wb")
load(temp)
dim(spambase)
table(spambase$is_spam)
```

Split into training and test sets — the only honest way to measure a classifier:

```{r,echo=TRUE}
set.seed(11)
idx   <- sample(nrow(spambase), 3000)
train <- spambase[idx, ]
test  <- spambase[-idx, ]
```

## Logistic Regression

Model the *probability* of spam through the logit link:

$$\log \frac{p}{1-p} = c_0 + c_1 x_1 + \cdots + c_n x_n$$

```{r,echo=TRUE,warning=FALSE}
spam.glm <- glm(is_spam ~ word_freq_free + word_freq_money +
                  capital_run_length_total + char_freq_exclamation,
                family = binomial, data = train)
coef(summary(spam.glm))
```

Positive coefficients raise the spam odds; `exp(coef)` gives odds ratios.

## From Probabilities to Decisions

```{r,echo=TRUE}
p.hat <- predict(spam.glm, newdata = test, type = "response")
pred  <- ifelse(p.hat > 0.5, 1, 0)
tab   <- table(actual = test$is_spam, predicted = pred)
tab
sum(diag(tab)) / sum(tab)        # accuracy
```

The **confusion matrix** separates false positives (good mail flagged) from false negatives (spam passed) — in most applications the two errors cost differently, so choose the cutoff accordingly.

## Linear Discriminant Analysis

LDA models each class as a Gaussian and draws the optimal linear boundary:

```{r,echo=TRUE,message=FALSE}
library(MASS)
spam.lda <- lda(is_spam ~ word_freq_free + word_freq_money +
                  capital_run_length_total + char_freq_exclamation,
                data = train)
mean(predict(spam.lda, test)$class == test$is_spam)
```

- Stable when classes are well-separated and roughly Gaussian; `qda()` allows curved boundaries.
- Log-linear models (`loglin()`/`glm` with `poisson`) play the analogous role for pure contingency-table data.

## k Nearest Neighbors

No model at all: classify each point by a vote of its $k$ closest training points:

```{r,echo=TRUE,message=FALSE}
library(class)
vars <- c("word_freq_free", "word_freq_money",
          "capital_run_length_total", "char_freq_exclamation")
pred.knn <- knn(scale(train[, vars]), scale(test[, vars]),
                cl = train$is_spam, k = 5)
mean(pred.knn == test$is_spam)
```

::: {.callout-tip}
**Scale your features for kNN.** Distance-based methods are dominated by whichever variable has the biggest units; standardize first.
:::

## Classification Trees

Trees split the feature space by a sequence of yes/no questions — fully interpretable:

```{r,echo=TRUE,out.width="62%"}
library(rpart)
spam.tree <- rpart(factor(is_spam) ~ ., data = train)
plot(spam.tree, uniform = TRUE, margin = 0.05)
text(spam.tree, cex = 0.7)
mean(predict(spam.tree, test, type = "class") == test$is_spam)
```

`printcp()` and `prune()` control overfitting by cost-complexity pruning.

## Neural Networks and SVMs

```{r,echo=TRUE,message=FALSE,results='hide'}
library(nnet)
set.seed(11)
spam.nn <- nnet(factor(is_spam) ~ ., data = train,
                size = 4, decay = 0.1, maxit = 200, MaxNWts = 2000)
```

```{r,echo=TRUE}
mean(predict(spam.nn, test, type = "class") == test$is_spam)
```

- `nnet()` fits a single-hidden-layer network — the classical workhorse.
- Support vector machines (`svm()` in the `e1071` package) find maximum-margin boundaries and handle nonlinearity through kernels.

## Random Forests

Hundreds of trees on bootstrap samples, votes averaged:

```{r,echo=TRUE,message=FALSE}
library(randomForest)
set.seed(11)
spam.rf <- randomForest(factor(is_spam) ~ ., data = train, ntree = 200)
mean(predict(spam.rf, test) == test$is_spam)
```

```{r,echo=TRUE,out.width="60%"}
varImpPlot(spam.rf, n.var = 10, main = "Top 10 features")
```

Usually the best out-of-the-box accuracy, plus a free feature-importance ranking — at the price of interpretability.

## Choosing a Classifier

| Need | Reach for |
|:---|:---|
| interpretable coefficients / odds | logistic regression |
| interpretable rules | classification tree |
| best plug-and-play accuracy | random forest |
| small data, smooth boundary | LDA / kNN |
| high-dimensional, margins matter | SVM |

Whatever the model: train/test split (or cross-validation), confusion matrix, and error costs come first [@hastie2009elements]. Chapter 22 widens the view to the rest of machine learning.

## Copyright Notice

- **Copyright.** These slides are adapted from *R in a Nutshell: A Desktop Quick Reference* (2nd ed.) by Joseph Adler, O'Reilly Media. All rights reserved by the original author and publisher.

- **Non-commercial use only.** These materials are strictly for educational purposes and may not be used for commercial gain.

- **Attribution.** Any reproduction, distribution, or use of these materials must properly credit the original source.
