---
title: "Programming for Applications"
subtitle: "Chapter 20: Regression Models"
author: "Yu-You Liou (NTU)"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "R in a Nutshell: A Desktop Quick Reference"
logo: "logo.svg"
editor: source
bibliography: bib.bib
suppress-bibliography: true
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch20-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center", fig.height = 4.2)
```

# **Predicting Continuous Values**

## What Regression Is For

A **regression model** relates a continuous **response** (dependent variable) to a set of **predictors** (independent/stimulus variables). Three uses:

- **prediction** — a warfarin dose from a patient's weight; a credit customer's balance;
- **explanation** — how weight relates to food intake, *adjusting for* age, exercise, genetics;
- **inference** — visualizing trends, ANOVA, testing variable significance.

This chapter starts with the simplest model — **linear regression by ordinary least squares (OLS)** — then builds, evaluates, refines it, and surveys richer model types. Classification waits for Chapter 21.

# A Simple Linear Model

## The Data and the Question

Which batting metrics predict a team's runs? Team batting 2000–2008 (extracted by SQL in Chapter 16, packaged as `team.batting.00to08`):

```{r,echo=TRUE}
temp <- tempfile(fileext = ".rda")
download.file("https://raw.githubusercontent.com/cran/nutshell/master/data/team.batting.00to08.rda",
              temp, mode = "wb")
load(temp)
```

A lattice scatter of runs against each variable hints at importance — home runs and *walks* both track runs strongly:

```{r,echo=TRUE,fig.height=3.6}
library(lattice)
attach(team.batting.00to08)
forplot <- make.groups(
  singles=data.frame(value=singles, runs), doubles=data.frame(value=doubles, runs),
  triples=data.frame(value=triples, runs), homeruns=data.frame(value=homeruns, runs),
  walks=data.frame(value=walks, runs))
detach(team.batting.00to08)
xyplot(runs~value|which, data=forplot, scales=list(relation="free"),
       pch=19, cex=.2, layout=c(5,1))
```

## lm: Fitting

`lm` fits a linear model by OLS. Note: in formulas `+ * - ^` are *special*; wrap literal arithmetic in `I()`, and generate polynomial terms with `poly(x, degree=)`.

```{r,echo=TRUE}
runs.mdl <- lm(
  formula=runs~singles+doubles+triples+homeruns+
            walks+hitbypitch+sacrificeflies+
            stolenbases+caughtstealing,
  data=team.batting.00to08)
runs.mdl
```

R prints almost nothing on fitting — **information comes from helper functions** (a deliberate contrast with SAS/SPSS).

## Inspecting a Model: summary

```{r,echo=TRUE}
summary(runs.mdl)
```

The report: residual distribution; a **coefficient table** (estimate, std. error, t-value, p-value, significance stars); and fit statistics — **R² = 0.914**, adjusted R² = 0.911, an overwhelmingly significant F-statistic. Singles, doubles, triples, home runs, walks, HBP, sac flies all matter; stolen bases and caught-stealing do not.

## Model Accessors

Beyond `summary`, a family of extractors:

```{r,echo=TRUE}
formula(runs.mdl)          # the formula used
coef(runs.mdl)[1:4]        # coefficients (alias: coefficients)
head(residuals(runs.mdl))  # residuals; fitted() gives fitted values
confint(runs.mdl)[1:4, ]   # confidence intervals for coefficients
```

Others: `effects` (orthogonal effects), `vcov` (variance-covariance matrix), `deviance`, `influence` / `influence.measures`, `anova` (ANOVA table).

## Predicting New Values

`predict(object, newdata, interval=c("none","confidence","prediction"), level=0.95, ...)` applies the model to fresh data — optionally with intervals:

```{r,echo=TRUE}
newdata <- team.batting.00to08[1, ]
predict(runs.mdl, newdata, interval="confidence")
```

`anova(model1, model2, ...)` compares nested models; `update(model, formula)` refits after a formula change — far cheaper than building from scratch on large data.

## Diagnostic Plots

`plot(lm.object)` draws the standard diagnostics — residuals vs. fitted, a normal Q-Q of residuals, scale-location, and residuals vs. leverage:

```{r,echo=TRUE,fig.height=4.5}
par(mfrow=c(2,2))
plot(runs.mdl)
par(mfrow=c(1,1))
```

Look for: structure in residuals (nonlinearity), departures from the Q-Q line (non-normality), a fanning scale-location (heteroscedasticity), and high-leverage outliers.

## lm in Detail

```{r,eval=FALSE,echo=TRUE}
lm(formula, data, subset, weights, na.action,
   method = "qr", model = TRUE, x = FALSE, y = FALSE,
   qr = TRUE, singular.ok = TRUE, contrasts = NULL, offset, ...)
```

The model is $y = c_0 + c_1 x_1 + \cdots + c_n x_n + \varepsilon$; OLS chooses the $c_i$ minimizing the sum of squared **residuals**. Key arguments: `formula` (required); `data`, `subset`, `weights`, `na.action`; `contrasts` controls factor coding (treatment, Helmert, ...); `weights` gives weighted least squares. Companions `glm`, `nls`, `loess`, `rlm` share many of these argument names.

# When OLS Isn't Enough

## Assumptions and Their Tests

OLS is technically valid only under assumptions — yet often predicts well even when they're bent:

- **linearity** of the response in the predictors;
- **constant error variance** (homoscedasticity) — test with `car::ncvTest` (non-constant variance score test);
- predictors not collinear (singular fits);
- normally distributed, independent errors.

When they fail badly, reach for the alternatives below.

## Robust and Resistant Regression

Outliers wreck OLS. Two defenses (package `MASS`):

- **Resistant regression** — `lqs` (least trimmed squares, least median of squares): fits using a *subset* of points, ignoring extreme ones;
- **Robust regression** — `rlm`: M-estimation, down-weighting (rather than discarding) large residuals iteratively.

```{r,eval=FALSE,echo=TRUE}
library(MASS)
lqs(runs~homeruns+walks, data=team.batting.00to08)   # resistant
rlm(runs~homeruns+walks, data=team.batting.00to08)   # robust
```

## Subset Selection and Shrinkage

Too many predictors? Three remedies:

- **Stepwise selection** — `step` (and `MASS::stepAIC`, wider model range) add/drop terms by AIC.
- **Ridge regression** — `MASS::lm.ridge(formula, data, lambda=)` shrinks coefficients toward zero (L2 penalty), taming collinearity.
- **Lasso / least angle regression** — `lars::lars(x, y, type=c("lasso","lar","forward.stagewise","stepwise"))` (L1 penalty) shrinks *and* selects, zeroing some coefficients; it computes the entire path at once.

Ridge and lasso are both **penalized regression**; the modern unified tool is `glmnet`.

```{r,eval=FALSE,echo=TRUE}
step(runs.mdl)                          # stepwise by AIC
MASS::lm.ridge(runs~., team.batting.00to08, lambda=seq(0,1,0.1))
```

## Principal Components and PLS Regression

When predictors are many and correlated, regress on *components* instead — package `pls`, both via `mvr` (aliases `pcr`, `plsr`):

- **Principal components regression (`pcr`)** — regress on the leading principal components of the predictors;
- **Partial least squares (`plsr`)** — components chosen to also correlate with the response.

# Generalized and Nonlinear Models

## glm: Generalized Linear Models

`glm` extends linear models to non-normal responses via a **family** and a **link function** — predicting the link of the expected response:

```{r,eval=FALSE,echo=TRUE}
glm(formula, family = gaussian, data, weights, subset,
    na.action, start = NULL, ...)
```

| Family | Typical link | Use |
|---|---|---|
| `gaussian` | identity | ordinary linear regression |
| `binomial` | logit | binary / proportion outcomes (Chapter 21) |
| `poisson` | log | counts |
| `Gamma` | inverse | positive continuous, skewed |
| `inverse.gaussian` | 1/μ² | |

(`glmnet` also fits *penalized* GLMs — `family`, `lambda`, `alpha` choosing ridge↔lasso.)

## Nonlinear Models

When the relationship isn't linear in the parameters:

- **`nls(formula, data, start, ...)`** — nonlinear least squares; you supply the functional form and starting values.
- **Survival models** — `survival::survreg` (parametric) and `coxph` (Cox proportional hazards), reporting Wald tests on coefficients.

```{r,eval=FALSE,echo=TRUE}
nls(y ~ a * exp(b * x), data=d, start=list(a=1, b=0.1))
library(survival)
coxph(Surv(time, status) ~ age + sex, data=lung)
```

## Smoothing: Splines, Surfaces, Kernels

For flexible, nonparametric trends:

- **Splines** — `smooth.spline`, and `splines::ns`/`bs` for use inside formulas;
- **Polynomial surfaces** — `loess` (local regression) fits a smooth surface from local polynomial fits;
- **Kernel smoothing** — `ksmooth`, `density`.

```{r,echo=TRUE,fig.height=3.4}
plot(cars$speed, cars$dist)
lines(smooth.spline(cars$speed, cars$dist), col="blue")
lines(loess.smooth(cars$speed, cars$dist), col="red")
```

## Machine-Learning Regression

The chapter closes by previewing tree-based regression (full treatment: Chapters 21–22):

- **`rpart`** — recursive partitioning (regression trees);
- **bagging**, **boosting** (`gbm`), **random forests** (`randomForest`) — ensembles of trees, each highly parallelizable (Chapter 26);
- **PRIM** (patient rule induction method) — bump hunting.

::: {.callout-tip}
**A workflow, not a function.** Fit with `lm`, *read* `summary`, *look* at `plot`, then iterate with `update` — adding, dropping, or transforming terms. Reach for robust/penalized/nonlinear tools only when the diagnostics demand it.
:::

## Copyright Notice

- **Copyright.** These slides are adapted from *R in a Nutshell: A Desktop Quick Reference* (2nd ed.) by Joseph Adler, O'Reilly Media. All rights reserved by the original author and publisher.

- **Non-commercial use only.** These materials are strictly for educational purposes and may not be used for commercial gain.

- **Attribution.** Any reproduction, distribution, or use of these materials must properly credit the original source.
