---
title: "Programming for Applications"
subtitle: "Chapter 20: Regression Models"
author: "Yu-You Liou (NTU)"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "R in a Nutshell: A Desktop Quick Reference"
logo: "logo.svg"
editor: source
bibliography: bib.bib
suppress-bibliography: true
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch20-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center")
```

# **Modeling Relationships**

## A Question to Model

How many runs does a baseball team score, given its hitting statistics? Team batting data for 2000–2008, loaded the course way:

```{r,echo=TRUE,cache=TRUE}
temp <- tempfile(fileext = ".rda")
download.file("https://raw.githubusercontent.com/cran/nutshell/master/data/team.batting.00to08.rda", temp, mode = "wb")
load(temp)
names(team.batting.00to08)
```

The linear model assumes

$$\text{runs} = c_0 + c_1 x_1 + \cdots + c_n x_n + \varepsilon,$$

and `lm()` estimates the coefficients by least squares.

## Fitting a Model

```{r,echo=TRUE}
runs.mdl <- lm(runs ~ singles + doubles + triples + homeruns +
                 walks + hitbypitch + sacrificeflies +
                 stolenbases + caughtstealing,
               data = team.batting.00to08)
runs.mdl
```

Each coefficient reads "extra runs per extra unit, holding the rest fixed" — a home run is worth roughly three times a single.

## Evaluating the Fit

```{r,echo=TRUE}
summary(runs.mdl)
```

Read in this order: residual spread, coefficient significance (stars), then $R^2$ — the share of variance explained.

## Diagnostic Plots

`plot(model)` produces the standard panel:

```{r,echo=TRUE,out.width="72%"}
par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
plot(runs.mdl)
par(mfrow = c(1, 1))
```

Look for: curvature in *Residuals vs. Fitted* (nonlinearity), departures in the *Q–Q plot* (non-normal errors), funnels in *Scale–Location* (heteroskedasticity), and influential points in *Leverage*.

## Refining the Model

Drop weak terms and compare nested models:

```{r,echo=TRUE}
runs.mdl2 <- update(runs.mdl, . ~ . - caughtstealing - stolenbases)
anova(runs.mdl2, runs.mdl)
AIC(runs.mdl, runs.mdl2)
```

- `update()` edits a formula without retyping it.
- `anova()` tests whether the extra terms earn their keep; `AIC`/`BIC` balance fit against complexity.
- `step(runs.mdl)` automates the search by AIC.

## Details of lm

Useful arguments and extractors:

```{r,eval=FALSE,echo=TRUE}
lm(formula, data, subset = year > 2003,   # fit on a subset
   weights = w, na.action = na.exclude)
```

```{r,echo=TRUE}
coef(runs.mdl2)[1:4]
head(residuals(runs.mdl2), 3)
predict(runs.mdl2, newdata = team.batting.00to08[1:2, ],
        interval = "prediction")
```

`confint()`, `fitted()`, `model.matrix()` complete the extraction toolkit.

## Robust and Resistant Regression

Outliers can drag least squares anywhere. Robust alternatives down-weight them:

```{r,echo=TRUE,message=FALSE}
library(MASS)
runs.rlm <- rlm(runs ~ singles + doubles + triples + homeruns + walks,
                data = team.batting.00to08)
round(coef(runs.rlm), 3)
```

- `rlm()` — M-estimation (`MASS`); `lqs()` — resistant fits like least trimmed squares.
- If robust and ordinary fits disagree sharply, investigate the influential observations before trusting either.

## Subset Selection and Shrinkage

With many correlated predictors, three strategies tame variance:

- **Subset selection**: `step()`, or exhaustive search via the `leaps` package.
- **Ridge regression**: keep all terms, shrink coefficients ($L_2$ penalty) — `lm.ridge()` in `MASS`.
- **Lasso**: shrink *and* select ($L_1$ penalty) — the `glmnet` package fits whole regularization paths efficiently.

```{r,eval=FALSE,echo=TRUE}
library(glmnet)
fit <- glmnet(x = as.matrix(predictors), y = runs, alpha = 1)  # lasso
cv  <- cv.glmnet(as.matrix(predictors), runs)   # pick lambda by CV
```

Principal-components and partial-least-squares regression (`pls` package) are the projection-based cousins.

## Generalized Linear Models

`glm()` extends the linear model through a **link function** and an error **family** — one interface for counts, rates, and probabilities:

```{r,echo=TRUE}
counts <- data.frame(events = c(2, 1, 4, 7, 9, 14),
                     exposure = 1:6)
fit.pois <- glm(events ~ exposure, family = poisson, data = counts)
coef(summary(fit.pois))
```

- `gaussian` reproduces `lm()`; `poisson` models counts; `binomial` gives logistic regression (Chapter 21).
- Coefficients live on the link scale — exponentiate Poisson/logit coefficients to interpret.

## Nonlinear and Smooth Models

When the relationship is genuinely curved:

```{r,echo=TRUE,out.width="60%",message=FALSE}
fit.lo <- loess(dist ~ speed, data = cars)
plot(cars, pch = 19)
lines(seq(4, 25), predict(fit.lo, data.frame(speed = seq(4, 25))),
      col = "red", lwd = 2)
```

- `nls()` fits parametric nonlinear curves (you supply the formula and starting values).
- `loess()` fits local polynomials; `smooth.spline()` penalized splines; `mgcv::gam()` full generalized additive models.

## Survival Models

Time-to-event data with censoring use the `survival` package:

```{r,echo=TRUE,message=FALSE,out.width="60%"}
library(survival)
fit.km <- survfit(Surv(time, status) ~ x, data = aml)
plot(fit.km, lty = 1:2, xlab = "Weeks", ylab = "Survival")
legend("topright", c("Maintained", "Nonmaintained"), lty = 1:2)
```

- `survfit()` — Kaplan–Meier curves; `survdiff()` — log-rank test; `coxph()` — Cox proportional hazards regression.
- The `Surv(time, status)` object encodes censoring; ignoring censoring biases everything.

## A Modeling Recipe

1. Plot first (Chapters 13–15); guess the functional form.
2. Fit the simplest plausible model with `lm()`/`glm()`.
3. Diagnose: `plot(model)`, residuals, influence.
4. Refine: transform, add/drop terms, or regularize.
5. Validate on held-out data (Chapter 22's discipline).
6. Report effect sizes with `confint()`, not just stars [@hastie2009elements].

## Copyright Notice

- **Copyright.** These slides are adapted from *R in a Nutshell: A Desktop Quick Reference* (2nd ed.) by Joseph Adler, O'Reilly Media. All rights reserved by the original author and publisher.

- **Non-commercial use only.** These materials are strictly for educational purposes and may not be used for commercial gain.

- **Attribution.** Any reproduction, distribution, or use of these materials must properly credit the original source.
