---
title: "Programming for Applications"
subtitle: "Chapter 25: Bioconductor"
author: "Yu-You Liou (NTU)"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "R in a Nutshell: A Desktop Quick Reference"
logo: "logo.svg"
editor: source
bibliography: bib.bib
suppress-bibliography: true
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch25-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center")
```

# **R for Genomics**

## What Is Bioconductor?

**Bioconductor** is a separate ecosystem of R packages for the analysis of high-throughput biological data — gene expression, sequencing, flow cytometry, and more. It has its own repository, release cycle, and installation route.

::: {.callout-note}
**A parallel CRAN.** Bioconductor packages are not installed with `install.packages()` from CRAN. They share specialized data structures so tools interoperate, and they are versioned together for reproducibility.
:::

This chapter is a conceptual tour; the code is shown for reference and does not run without the (large) Bioconductor packages installed.

## Installation

The modern entry point is the `BiocManager` package:

```{r,eval=FALSE,echo=TRUE}
install.packages("BiocManager")        # once, from CRAN
BiocManager::install()                 # core infrastructure
BiocManager::install(c("Biobase",      # base classes
                       "affy",         # microarray preprocessing
                       "limma",        # differential expression
                       "GEOquery"))    # download public data
```

`BiocManager::version()` reports the Bioconductor release; `BiocManager::valid()` checks that your installed packages are mutually compatible.

## Core Data Structures

Bioconductor's power comes from shared S4 classes (Chapter 10) that bundle related data together:

- **`ExpressionSet` / `eSet`** — an assay matrix (genes × samples) plus **phenotype data** (sample annotations) and **feature data** (gene annotations) in one object;
- **`SummarizedExperiment`** — the modern successor, the standard container for sequencing assays;
- **`GRanges`** — genomic intervals (chromosome, start, end, strand) with metadata.

Keeping measurements and their annotations in a single object is what prevents the classic bioinformatics error: sample labels drifting out of sync with sample data.

## A Typical Workflow

```{r,eval=FALSE,echo=TRUE}
library(Biobase); library(GEOquery)
# 1. Download a public expression study from GEO
gse <- getGEO("GSE2034")[[1]]

# 2. Inspect the bundled object
exprs(gse)[1:5, 1:3]      # the expression matrix
pData(gse)[1:3, 1:4]      # phenotype (sample) data
fData(gse)[1:3, 1:2]      # feature (gene) data
```

`exprs()`, `pData()`, and `fData()` are the three accessor verbs you use constantly — assay values, sample annotations, and gene annotations, respectively.

## Differential Expression with limma

The signature analysis: which genes differ between two conditions?

```{r,eval=FALSE,echo=TRUE}
library(limma)
group  <- factor(pData(gse)$status)         # e.g. tumor vs. normal
design <- model.matrix(~ group)
fit    <- lmFit(exprs(gse), design)         # a linear model PER GENE
fit    <- eBayes(fit)                       # borrow strength across genes
topTable(fit, coef = 2, number = 10)        # top differential genes
```

`limma` fits the regression of Chapter 20 to *every gene at once*, then uses an empirical-Bayes step to stabilize variance estimates across thousands of genes — essential when samples are few.

## The Course Connection

The companion `nutshell` package includes a microarray data set used in the book's Bioconductor example; we load it the course way, no Bioconductor required:

```{r,echo=TRUE,eval=FALSE}
temp <- tempfile(fileext = ".rda")
download.file("https://raw.githubusercontent.com/cran/nutshell/master/data/GSE2034.rda", temp, mode = "wb")
load(temp)
# GSE2034: breast-cancer expression measurements used to illustrate
# the analysis pipeline described above.
```

This lets us discuss the *shape* of genomic data — a wide matrix of genes against samples — without installing the full Bioconductor stack.

## Why It Matters Beyond Biology

Even outside genomics, Bioconductor teaches transferable lessons:

- **Bundle data with its metadata** in one object — applicable to any analysis where labels can desync.
- **Borrow strength across many parallel models** (the empirical-Bayes idea) — useful wherever you fit thousands of small models.
- **Reproducibility through coordinated versioning** — a discipline the wider R world increasingly adopts.

For genuine bioinformatics work, the Bioconductor website's workflows and vignettes are the definitive, continually updated reference.

## Copyright Notice

- **Copyright.** These slides are adapted from *R in a Nutshell: A Desktop Quick Reference* (2nd ed.) by Joseph Adler, O'Reilly Media. All rights reserved by the original author and publisher.

- **Non-commercial use only.** These materials are strictly for educational purposes and may not be used for commercial gain.

- **Attribution.** Any reproduction, distribution, or use of these materials must properly credit the original source.
