---
title: "Programming for Applications"
subtitle: "Chapter 25: Bioconductor"
author: "Yu-You Liou (NTU)"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "R in a Nutshell: A Desktop Quick Reference"
logo: "logo.svg"
editor: source
bibliography: bib.bib
suppress-bibliography: true
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch25-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center", fig.height = 4.2)
```

# **R for Bioinformatics**

## What Bioconductor Is

Most of this book is field-agnostic; this chapter is the exception, focused on **bioinformatics**. **Bioconductor** is an open-source project of R packages for analyzing high-throughput genomic data — microarrays, next-generation sequencing, and more. It has its own release cycle, its own repository (separate from CRAN), and shared data structures that let its hundreds of packages interoperate.

::: {.callout-note}
**Installation differs from CRAN.** Bioconductor packages install through its own installer — historically `biocLite`, today `BiocManager`:

```{r,eval=FALSE,echo=TRUE}
install.packages("BiocManager")
BiocManager::install(c("affy", "GEOquery", "limma"))
```
:::

# A Worked Example

## Loading Raw Expression Data

The example analyzes a breast-cancer microarray study (GEO accession GSE2034, Affymetrix HG-U133A platform). Raw Affymetrix data lives in **`.CEL`** files, read in a batch with `ReadAffy` from the `affy` package:

```{r,eval=FALSE,echo=TRUE}
library(affy)
GSE2034 <- ReadAffy()                          # reads all .CEL in the working dir
# or name them explicitly:
GSE2034.smpl <- ReadAffy(filenames=paste0("GSM", 36777:36800, ".CEL"))
GSE2034                                         # an AffyBatch object
```

`ReadAffy` returns an **`AffyBatch`** — unprocessed probe-level intensities plus the chip-definition (`cdf`) that maps probes to the 22,283 probe sets on HG-U133A.

## From Raw Data to Expression

Raw intensities must be **quality-checked, background-corrected, normalized, summarized by probe set, and log-transformed** before analysis:

```{r,eval=FALSE,echo=TRUE}
library(simpleaffy)
qc(GSE2034)                          # quality-control report

# normalize + summarize in one step (several methods available):
eset <- rma(GSE2034)                 # robust multi-array average -> ExpressionSet
# or: vsnrma(GSE2034), expresso(GSE2034, ...)
```

`rma` (and relatives `vsnrma`, `expresso`) collapse the AffyBatch into an **`ExpressionSet`** — the central Bioconductor container.

## Loading Preprocessed Data from GEO

Often the processed expression set is already public — `GEOquery::getGEO` downloads it directly:

```{r,eval=FALSE,echo=TRUE}
library(GEOquery)
GSE2034.geo <- getGEO("GSE2034")     # a list of ExpressionSet objects
class(GSE2034.geo[[1]])
## [1] "ExpressionSet"
```

No `.CEL` files needed — the analysis can start from a single download.

# The ExpressionSet

## Anatomy of the Central Data Structure

An **`ExpressionSet`** bundles everything one experiment needs, with accessor functions:

| Component | Accessor | Contents |
|---|---|---|
| Expression matrix | `exprs()` | genes (rows) × samples (columns) |
| Phenotype data | `pData()` / `phenoData()` | per-*sample* metadata (treatment, outcome...) |
| Feature data | `fData()` / `featureData()` | per-*gene* annotation |
| Experiment info | `experimentData()` | study-level description |
| Annotation | `annotation()` | the platform, e.g. `"GPL96"` |

```{r,eval=FALSE,echo=TRUE}
dim(exprs(eset))            # genes x samples
head(pData(eset))           # sample metadata
```

## Matching Phenotype Data

Real studies need clinical/phenotype data joined to the expression matrix. Replace or merge the `pData`:

```{r,eval=FALSE,echo=TRUE}
# build phenotype data and attach it
pData(eset) <- new("AnnotatedDataFrame", data=clinical.df)
# or merge GEO phenotype with your own:
merged <- merge(pData(GSE2034.geo[[1]]), my.clinical, by="sample")
pData(GSE2034.geo[[1]]) <- merged
```

Aligning samples between the two sources is the fiddly part — the book spends real effort getting the row order to match.

## Downstream: Differential Expression

Once you have an `ExpressionSet` with matched phenotypes, the standard next step is finding **differentially expressed genes** — package `limma` (linear models for microarrays):

```{r,eval=FALSE,echo=TRUE}
library(limma)
design <- model.matrix(~ pData(eset)$outcome)
fit <- lmFit(eset, design)           # fit a linear model per gene
fit <- eBayes(fit)                   # empirical-Bayes moderation
topTable(fit, number=20)             # top differentially expressed genes
```

::: {.callout-tip}
**Why a separate ecosystem?** Genomic data is huge, structured (genes × samples × annotation), and method-rich. Bioconductor's shared `ExpressionSet` (and successors like `SummarizedExperiment`) let dozens of packages — normalization, testing, annotation, visualization — plug together. The R you learned transfers directly; only the data structures and the installer are special.
:::

## Copyright Notice

- **Copyright.** These slides are adapted from *R in a Nutshell: A Desktop Quick Reference* (2nd ed.) by Joseph Adler, O'Reilly Media. All rights reserved by the original author and publisher.

- **Non-commercial use only.** These materials are strictly for educational purposes and may not be used for commercial gain.

- **Attribution.** Any reproduction, distribution, or use of these materials must properly credit the original source.
