---
title: "Programming for Applications"
subtitle: "Chapter 22: Machine Learning"
author: "Yu-You Liou (NTU)"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "R in a Nutshell: A Desktop Quick Reference"
logo: "logo.svg"
editor: source
bibliography: bib.bib
suppress-bibliography: true
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch22-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center", fig.height = 4.2)
```

# **Finding Structure**

## Beyond Supervised Learning

Chapters 20–21 were **supervised** — predict a known response. This chapter covers two **unsupervised** tasks, where there is no response to predict, only structure to discover:

- **market basket analysis** — which items occur together (association rules);
- **clustering** — which observations are similar enough to group.

# Market Basket Analysis

## Association Rules and apriori

Given a set of **transactions** (a shopping basket, a listening history), association-rule mining finds **items frequently associated** with each other — "customers who bought X also bought Y." The classic algorithm is **a priori**, in package `arules`:

```{r,eval=FALSE,echo=TRUE}
library(arules)
apriori(data, parameter = NULL, appearance = NULL, control = NULL)
```

- `data` — a `transactions` object (or a matrix/data frame coercible to one);
- `parameter` — thresholds, chiefly **support** (how often items appear together) and **confidence** (how reliable the rule is);
- `appearance` constrains which items may appear on each side; `control` tunes the algorithm.

`arules` is built on S4 classes (a `transactions` class for the data) and the well-engineered apriori C implementation by Christian Borgelt.

## Example: Audioscrobbler

The book mines Audioscrobbler listening data (now part of Last.fm). Load with `read.transactions`, mine, inspect:

```{r,eval=FALSE,echo=TRUE}
library(arules)
playlists <- read.transactions("audioscrobbler.csv", format="basket", sep=",")
rules <- apriori(playlists,
                 parameter=list(support=0.01, confidence=0.5))
inspect(head(sort(rules, by="lift"), 10))   # strongest rules by lift
```

Resulting rules read like "listeners of Green Day and Nirvana also listen to Red Hot Chili Peppers," each with support, confidence, and **lift** (how much more often than chance). A related, faster algorithm for frequent itemsets is `eclat`.

# Clustering

## Distance Measures

Clustering groups observations so that members of a cluster are more similar to each other than to outsiders — and similarity starts with **distance**. `stats::dist` builds a distance matrix:

```{r,eval=FALSE,echo=TRUE}
dist(x, method = "euclidean", diag = FALSE, upper = FALSE, p = 2)
```

`method` may be `"euclidean"`, `"manhattan"`, `"maximum"`, `"canberra"`, `"binary"`, or `"minkowski"` (with parameter `p`). The right metric matters: scale and units shape every cluster that follows, so standardize variables first when they differ in range.

```{r,echo=TRUE}
d <- dist(scale(iris[, 1:4]))     # standardize, then Euclidean distance
as.matrix(d)[1:4, 1:4]
```

## Partitioning: kmeans, pam, clara

**Partitioning** algorithms split data into a chosen number `k` of clusters:

- **`kmeans(x, centers, ...)`** — fast, minimizes within-cluster sum of squares; sensitive to outliers and starting centers.
- **`cluster::pam(x, k)`** — partitioning around medoids; uses actual data points as centers, more robust than k-means.
- **`cluster::clara(x, k)`** — pam scaled to large data via sampling.

```{r,echo=TRUE}
set.seed(1)
km <- kmeans(scale(iris[, 1:4]), centers=3)
table(km$cluster, iris$Species)   # clusters vs. true species
km$centers[, 1:2]
```

## Cluster Quality and Hierarchical Clustering

- **`cluster::silhouette`** measures how well each point fits its cluster (−1 to 1); the average silhouette width helps **choose `k`**.
- **Hierarchical clustering** needs no `k` up front — `hclust(dist_object, method=)` builds a tree (dendrogram), merging closest clusters step by step; `cluster::agnes` (agglomerative) and `diana` (divisive) are richer alternatives.

```{r,echo=TRUE,fig.height=3.6}
hc <- hclust(dist(scale(iris[, 1:4])), method="ward.D2")
plot(hc, labels=FALSE, main="Hierarchical clustering of iris")
rect.hclust(hc, k=3)              # cut into 3 clusters
```

`method` ranges over `"complete"`, `"average"`, `"single"`, `"ward.D2"`, ... — each defining "distance between clusters" differently, and each producing differently shaped clusters.

::: {.callout-tip}
**Unsupervised work has no answer key.** There is no accuracy to optimize — only structure to judge. Vary the distance metric, the algorithm, and `k`; validate with silhouette widths, domain knowledge, and whether the clusters are *useful*, not merely whether they exist.
:::

## Copyright Notice

- **Copyright.** These slides are adapted from *R in a Nutshell: A Desktop Quick Reference* (2nd ed.) by Joseph Adler, O'Reilly Media. All rights reserved by the original author and publisher.

- **Non-commercial use only.** These materials are strictly for educational purposes and may not be used for commercial gain.

- **Attribution.** Any reproduction, distribution, or use of these materials must properly credit the original source.
