---
title: "Programming for Applications"
subtitle: "Chapter 22: Machine Learning"
author: "Yu-You Liou (NTU)"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "R in a Nutshell: A Desktop Quick Reference"
logo: "logo.svg"
editor: source
bibliography: bib.bib
suppress-bibliography: true
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch22-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center")
```

# **Learning Without Labels**

## Supervised vs. Unsupervised

Chapters 20–21 *supervised* learning: a known outcome guided the fit. This chapter is the **unsupervised** side — no labels, only structure to discover:

- **Market basket analysis**: which items co-occur in transactions?
- **Clustering**: which observations belong together?

Both are exploration tools: they generate hypotheses rather than test them [@hastie2009elements].

## Market Basket Analysis

The `arules` package mines **association rules** of the form $\{A, B\} \Rightarrow \{C\}$ from transaction data:

```{r,echo=TRUE,message=FALSE}
library(arules)
data(Groceries)             # built into arules
summary(Groceries)@lengths[1:5]
```

```{r,echo=TRUE,results='hide',message=FALSE,warning=FALSE}
rules <- apriori(Groceries,
                 parameter = list(supp = 0.005, conf = 0.5))
```

```{r,echo=TRUE}
inspect(head(sort(rules, by = "lift"), 3))
```

## Reading a Rule

Three numbers qualify every rule:

::: {.callout-note}
**Support** — share of transactions containing all the rule's items. **Confidence** — $P(\text{rhs} \mid \text{lhs})$: how often the right side appears when the left side does. **Lift** — confidence divided by the right side's base rate; lift $> 1$ means genuine association, not popularity.
:::

- High-confidence, low-lift rules ("⇒ whole milk") merely restate that milk is popular — lift is the filter that catches this.
- Thresholds in `apriori()` trade rule quantity against quality.

## k-Means Clustering

Partition $n$ observations into $k$ compact groups:

```{r,echo=TRUE}
set.seed(5)
km <- kmeans(scale(USArrests), centers = 4, nstart = 25)
table(km$cluster)
km$centers
```

```{r,echo=TRUE,out.width="60%"}
plot(prcomp(scale(USArrests))$x[, 1:2], col = km$cluster, pch = 19)
```

- Scale the variables first; `nstart = 25` restarts guard against bad local optima.
- Choose $k$ by the elbow of `km$tot.withinss` across candidate values, or by silhouette width (`cluster` package).

## Hierarchical Clustering

No $k$ needed in advance — build a tree of merges and cut it later:

```{r,echo=TRUE,out.width="68%"}
d  <- dist(scale(USArrests))
hc <- hclust(d, method = "ward.D2")
plot(hc, cex = 0.55, main = "US states by crime profile")
rect.hclust(hc, k = 4, border = "red")
```

- `dist()` offers euclidean, manhattan, and more; linkage (`single`, `complete`, `average`, `ward.D2`) changes the tree's character.
- `cutree(hc, k = 4)` extracts cluster labels for downstream use.

## Other Unsupervised Tools

- **Model-based clustering**: the `mclust` package fits Gaussian mixtures and selects $k$ by BIC.
- **Density-based**: `dbscan` finds arbitrarily shaped clusters and flags noise points.
- **Dimension reduction as preprocessing**: PCA (Chapter 16) before clustering often improves both speed and quality.
- **Self-organizing maps** (`kohonen`) project high-dimensional data onto a 2-D grid.

## Honest Evaluation

Unsupervised results need skepticism most:

1. Stability — do clusters survive resampling or a different seed?
2. Silhouette / within-cluster variance — are groups actually compact?
3. Interpretation — can a domain expert name each cluster?
4. For rules: validate top rules on a held-out period before acting on them.

A clustering that cannot be named, or a rule that disappears next quarter, is noise wearing a costume.

## Copyright Notice

- **Copyright.** These slides are adapted from *R in a Nutshell: A Desktop Quick Reference* (2nd ed.) by Joseph Adler, O'Reilly Media. All rights reserved by the original author and publisher.

- **Non-commercial use only.** These materials are strictly for educational purposes and may not be used for commercial gain.

- **Attribution.** Any reproduction, distribution, or use of these materials must properly credit the original source.
