---
title: "Programming for Applications"
subtitle: "Chapter 7: R Objects"
author: "Yu-You Liou (NTU)"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "R in a Nutshell: A Desktop Quick Reference"
logo: "logo.svg"
editor: source
bibliography: bib.bib
suppress-bibliography: true
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch7-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center")
```

# **The Building Blocks**

## Types and Classes

All objects in R rest on a basic set of built-in objects:

- an object's **type** defines how it is *stored*;
- its **class** defines what information it *contains* and how it may be *used*.

This chapter covers the built-in objects themselves — the object-oriented machinery (class definitions, inheritance, methods) waits until Chapter 10. Note that "object-oriented programming" means more than "programming with objects"!

## A Map of the Primitive Types

The built-in object types sort into a few categories:

- **Basic vectors** — single-type collections: integers, doubles, complex, character, logical, raw.
- **Compound objects** — containers of named objects: lists, pairlists, S4 objects, environments.
- **Special objects** — `any`, `NULL`, `...`: each means something important in context, but you never create one yourself.
- **R language objects** — objects representing R code, evaluable to other objects.
- **Functions** — the workhorses: take arguments, return outputs, sometimes cause side effects (plots, files, network).
- **Internal / bytecode** — types defined by R but normally invisible to user-level code.

## Primitive Types (1): Basic Vectors

| Type | Description | Example |
|---|---|---|
| `integer` | Whole numbers; produced naturally by sequences | `5:5`, `integer(5)` |
| `double` | Floating-point (8 bytes on modern platforms); the default for numeric values | `1`, `-1`, `2^50` |
| `complex` | Complex numbers; both parts must be written, even a zero real part | `2+3i`, `exp(0+1i*pi)` |
| `character` | Text strings (a "string" in other languages) | `"Hello world."` |
| `logical` | Boolean values | `TRUE`, `FALSE` |
| `raw` | Raw bytes; for encoding objects from outside R | `raw(8)`, `charToRaw("Hello")` |

## Primitive Types (2): Compound Objects

| Type | Description | Example |
|---|---|---|
| `list` | A possibly heterogeneous, optionally named collection; data frames are built on lists | `list(1, 2, "hat")` |
| `pairlist` | Name–value pairs, mainly internal; deprecated in user code (plain lists are as efficient and more flexible) | `.Options` |
| `S4` | Objects supporting modern OO: inheritance, methods (Chapter 10) | — |
| `environment` | The set of symbol–value pairs in a context, plus a pointer to an enclosing environment | `.GlobalEnv`, `new.env()` |

## Primitive Types (3): Special and Language

| Type | Description | Example |
|---|---|---|
| `any` | "Any type is OK" — prevents coercion; used in S4 slots and generic signatures | `representation(data="ANY")` |
| `NULL` | "There is no object"; can have no attributes | `NULL` |
| `...` | Variable-length argument lists, especially arguments passed through to other functions | — |
| `symbol` | A language object referring to another object | `as.name("x")`, `quote(x)` |
| `promise` | Evaluated on *first use*, not creation; implements delayed loading in packages | `delayedAssign("v", c(x,y,z))` |
| `language` | Objects representing R code itself | `quote(function(x) {x+1})` |
| `expression` | An unevaluated expression; create with `expression()`, run with `eval()` | `expression(1 + 2)` |

## Primitive Types (4): Functions and Internal

| Type | Description | Example |
|---|---|---|
| `closure` | Functions written in R: user-defined, most of base R, most packages | `function(x) {x+1}`, `print` |
| `special` | Internal functions whose arguments are *not necessarily evaluated* | `if`, `[` |
| `builtin` | Internal functions that evaluate their arguments | `+`, `^` |
| `bytecode` | Compiled R functions from the `compiler` package | `cmpfun(function(x) {x^2})` |
| `char` | Scalar "string"; character vectors are made of these (users never touch them) | — |
| `externalptr` | External pointer, used in C code | — |
| `weakref` | Weak reference (internal only) | — |

# Vectors

## Building Vectors: c and Coercion

The six basic vector types appear constantly. The simplest constructor is `c`, which **combines** its arguments — and coerces them all to a single type:

```{r,echo=TRUE}
v <- c(.295, .300, .250, .287, .215)
v
v <- c(.295, .300, .250, .287, "zilch")   # one character value...
v                                          # ...and everything is character
```

## Recursive Combination

`c` can assemble a vector from nested structures with `recursive=TRUE` — but beware: a *list* argument without it gives you back a list:

```{r,echo=TRUE}
v <- c(.295, .300, .250, .287, list(.102, .200, .303), recursive=TRUE)
v
typeof(v)
v <- c(.295, .300, .250, .287, list(1, 2, 3))
typeof(v)
class(v)
```

## Sequences and the length Attribute

Two more vector builders — the `:` operator and the more flexible `seq`:

```{r,echo=TRUE}
1:10
seq(from=5, to=25, by=5)
```

A vector's length can be manipulated *directly* — shrinking discards, expanding fills with `NA`:

```{r,echo=TRUE}
w <- 1:10
length(w) <- 5
w
length(w) <- 10
w
```

# Lists

## Ordered, Optionally Named, Heterogeneous

A list is an ordered collection of objects, indexable by position — recall the `[` vs. `[[` distinction:

```{r,echo=TRUE}
l <- list(1, 2, 3, 4, 5)
l[1]      # a sublist
l[[1]]    # the element
```

Named elements model real things naturally. A physical parcel — destination New York, 2×6×9 inches, $12.95 postage — mixes three data types in one object:

```{r,echo=TRUE}
parcel <- list(destination="New York", dimensions=c(2, 6, 9), price=12.95)
parcel$price
```

Lists are *the* building block for heterogeneous structures — data frames are built on them.

# Other Objects

## Matrices

A **matrix** extends a vector to two dimensions, holding two-dimensional data of one type. The clean constructor is `matrix`, here with named rows and columns:

```{r,echo=TRUE}
m <- matrix(data=1:12, nrow=4, ncol=3,
            dimnames=list(c("r1", "r2", "r3", "r4"),
                          c("c1", "c2", "c3")))
m
```

- Other structures convert via `as.matrix`.
- **Implementation note:** a matrix is stored as a *vector*, not as a vector of vectors — subscripts are an access convention, not the storage layout. And unlike most classes, matrices carry no explicit `class` attribute (attributes: later this chapter).

## Arrays

An **array** extends a vector to *any* number of dimensions:

```{r,echo=TRUE}
a <- array(data=1:24, dim=c(3, 4, 2))
a
```

Like matrices, arrays are stored as plain vectors underneath, and have no explicit class attribute.

## Factors: Categorical Values

Categorical data could live in a character vector — but repeating long strings is wasteful. A **factor** is an ordered collection of items whose possible values are called **levels**:

```{r,echo=TRUE}
eye.colors <- factor(c("brown", "blue", "blue", "green",
                       "brown", "brown", "brown"))
levels(eye.colors)
eye.colors
```

Note the print format: no quotes, and the levels listed explicitly — visibly *not* a character vector.

## Ordered Factors

Sometimes level order matters. A survey asks for reactions to "melon is delicious with an omelet": Strongly Disagree ... Strongly Agree. Coding these 1–5 *implies* equal spacing and meaningful averages — is Disagree + Agree really = Neutral? An **ordered factor** captures the ranking without the arithmetic claims:

```{r,echo=TRUE}
survey.results <- factor(
  c("Disagree", "Neutral", "Strongly Disagree",
    "Neutral", "Agree", "Strongly Agree",
    "Disagree", "Strongly Agree", "Neutral",
    "Strongly Disagree", "Neutral", "Agree"),
  levels=c("Strongly Disagree", "Disagree", "Neutral",
           "Agree", "Strongly Agree"),
  ordered=TRUE)
survey.results
```

## Factors Under the Hood

Internally a factor is an **integer vector** plus a `levels` attribute mapping integers to labels (integers are small and fixed-size — efficient). Strip the class and the implementation shows:

```{r,echo=TRUE}
eye.colors
class(eye.colors)
eye.colors.integer.vector <- unclass(eye.colors)
eye.colors.integer.vector
class(eye.colors.integer.vector)
```

Restore the class attribute, and it's a factor again:

```{r,echo=TRUE}
class(eye.colors.integer.vector) <- "factor"
eye.colors.integer.vector
```

## Data Frames

A **data frame** represents a table: scientific observations with several measurements each, or a database table of rows and typed columns. Each column ("variable") may have a different type, but all must have the same length:

```{r,error=TRUE,echo=TRUE}
data.frame(a=c(1, 2, 3, 4, 5), b=c(1, 2, 3, 4))
```

The book's example — where users search most for the word "bacon" (Google Insights, 2004–2009):

```{r,echo=TRUE}
top.bacon.searching.cities <- data.frame(
  city = c("Seattle", "Washington", "Chicago", "New York", "Portland",
           "St Louis", "Denver", "Boston", "Minneapolis", "Austin",
           "Philadelphia", "San Francisco", "Atlanta", "Los Angeles",
           "Richardson"),
  rank = c(100, 96, 94, 93, 93, 92, 90, 90, 89, 87, 85, 84, 82, 80, 80)
)
head(top.bacon.searching.cities, 8)
```

## Data Frames Are Lists

```{r,echo=TRUE}
typeof(top.bacon.searching.cities)
class(top.bacon.searching.cities)
```

A data frame is implemented as a **list with class `data.frame`** — so list methods work unchanged: `top.bacon.searching.cities$rank` extracts the rank column.

## Formulas

To express a relationship between variables — for a plot or a model — R provides the **formula** class:

```{r,echo=TRUE}
sample.formula <- as.formula(y ~ x1 + x2 + x3)
class(sample.formula)
typeof(sample.formula)
```

This reads "y is a function of x1, x2, and x3." Chapter 3's lattice example, `Amount ~ Year | Food`, reads "Amount as a function of Year, conditioned on Food."

## Formula Vocabulary

| Item | Meaning |
|---|---|
| variable names | the variables themselves |
| `~` | separates response (left) from stimulus variables (right) |
| `+` | a linear relationship between variables |
| `0` | added to a formula: omit the intercept, e.g. `y~u+w+v+0` |
| `|` | conditioning variables (lattice formulas, Chapter 14) |
| `I()` | interpret the enclosed expression *arithmetically*: `a+b` means both variables; `I(a+b)` means their sum |
| `*` | interactions: `y~(u+v)*w` ≡ `y~u+v+w+I(u*w)+I(v*w)` |
| `^` | crossing to a degree: `y~(u+w)^2` ≡ `y~(u+w)*(u+w)` |
| `f(x)` | a function of variables as a term, e.g. `y~log(u)+sin(v)+w` |

Some functions add their own vocabulary (e.g., `s()` for smoothing splines in `gam`). Formulas return in Chapters 14 and 20.

## Time Series

Many problems track a variable **over time**; class `ts` represents this, feeding regression functions (`ar`, `arima`) and specialized plot methods.

```{r,eval=FALSE,echo=TRUE}
ts(data = NA, start = 1, end = numeric(0), frequency = 1,
   deltat = 1, ts.eps = getOption("ts.eps"), class = , names = )
```

| Argument | Description |
|---|---|
| `data` | the observations (vector or matrix) |
| `start` / `end` | time of first/last observation: one number, or (unit, offset) |
| `frequency` | observations per unit of time |
| `deltat` | sampling fraction between observations (`frequency = 1/deltat`) |
| `ts.eps` | comparison tolerance for frequencies |
| `class` | `"ts"` for one series; `c("mts","ts")` for several |
| `names` | names of each series in a multiseries object |

Eight quarters starting Q2 2008 — the print method knows about quarters and months:

```{r,echo=TRUE}
ts(1:8, start=c(2008, 2), frequency=4)
```

## A Real Time Series: Turkey Prices

The USDA tracks retail meat prices (supermarkets covering ~20% of the US market, averaged by month). The book packages monthly turkey prices as `turkey.price.ts`:

```{r,echo=TRUE}
temp <- tempfile(fileext = ".rda")
download.file("https://raw.githubusercontent.com/cran/nutshell/master/data/turkey.price.ts.rda",
              temp, mode = "wb")
load(temp)
turkey.price.ts
```

Utility functions read off the structure:

```{r,echo=TRUE}
start(turkey.price.ts)
end(turkey.price.ts)
frequency(turkey.price.ts)
deltat(turkey.price.ts)
```

## Shingles

A **shingle** generalizes a factor to a *continuous* variable: a numeric vector plus a set of intervals which — like roof shingles — may **overlap**. They let a continuous variable serve as a conditioning or grouping variable, and are used extensively in the `lattice` package (Chapter 14).

## Dates and Times

Three classes represent moments in time:

- **`Date`** — dates without times;
- **`POSIXct`** — date-times as seconds since 1970-01-01 00:00;
- **`POSIXlt`** — date-times as separate vectors: `sec` (0–61, allowing leap seconds!), `min`, `hour`, `mday`, `mon` (0–11), `year` (since 1900), `wday`, `yday`, `isdst`.

Store dates as date objects, not strings or numbers — the classes support arithmetic, and many plot functions require them:

```{r,echo=TRUE}
date.I.started.writing <- as.Date("2/13/2009", "%m/%d/%Y")
date.I.started.writing
today <- Sys.Date()
today - date.I.started.writing
```

## Connections

**Connections** move data between R and the outside world — like file pointers in C or filehandles in Perl. Targets include files, URLs, zip/gzip/bzip-compressed files, pipes, network sockets, FIFOs, even the system clipboard.

The lifecycle: create → open → use → close. Loading a saved (gzip-compressed) data file explicitly:

```{r,eval=FALSE,echo=TRUE}
consumption.connection <- gzfile(description="consumption.RData", open="r")
load(consumption.connection)
close(consumption.connection)
```

Usually you skip all this: `save`, `load`, `read.table` open connections **implicitly** when handed a filename or URL. Explicit connections earn their keep with nonstandard sources. See `?connection`.

# Attributes

## Properties Attached to Objects

Objects carry named properties — **attributes** — explaining what the object represents and how to interpret it. Often, two similar objects differ *only* in their attributes. (Why attributes rather than lists or S4? History: attributes predate R's modern object systems.)

| Attribute | Description |
|---|---|
| `class` | the class of the object |
| `comment` | a comment, often a description |
| `dim` | dimensions |
| `dimnames` | names along each dimension |
| `names` | names of elements (columns of a data frame, etc.) |
| `row.names` | names of rows (related to `dimnames`) |
| `tsp` | start time (time series) |
| `levels` | levels of a factor |

For attribute `a` of object `x`: query with `a(x)`, set with `a(x) <- value`. (Changes affect the current environment only.)

## Inspecting Attributes

All attributes at once via `attributes`; individually via accessor functions:

```{r,echo=TRUE}
m <- matrix(data=1:12, nrow=4, ncol=3,
            dimnames=list(c("r1", "r2", "r3", "r4"),
                          c("c1", "c2", "c3")))
attributes(m)
dim(m)
colnames(m)
rownames(m)
```

## Attributes Define the Object

Change the attributes, change the class — remove `dim` and a matrix collapses to a vector:

```{r,echo=TRUE}
dim(m) <- NULL
m
class(m)
typeof(m)
```

## Same Data, Different Attributes

An array and a vector with identical contents:

```{r,error=TRUE,echo=TRUE}
a <- array(1:12, dim=c(3, 4))
b <- 1:12
a[2, 2]
b[2, 2]      # error: b has no dimensions
```

Are they "the same"? `==` compares cell-wise (and answers in the shape of `a`):

```{r,echo=TRUE}
a == b
```

## all.equal and identical

`all.equal` compares data *and* attributes, explaining any differences; `identical` answers strictly yes or no:

```{r,echo=TRUE}
all.equal(a, b)
identical(a, b)
```

Give `b` the missing attribute and the difference evaporates:

```{r,echo=TRUE}
dim(b) <- c(3, 4)
b[2, 2]
all.equal(a, b)
identical(a, b)
```

## Class Is an Attribute

For simple objects, class and type align; for compound objects they diverge. Query with `class` and `typeof` (for matrices/arrays the class is implicit):

```{r,echo=TRUE}
x <- c(1, 2, 3)
typeof(x)
class(x)
```

Because class is just an attribute, you can build a factor *by hand* from an integer vector:

```{r,echo=TRUE}
v <- as.integer(c(1, 1, 1, 2, 1, 2, 2, 3, 1))
levels(v) <- c("what", "who", "why")
class(v) <- "factor"
v
```

(No guarantee the internal implementation of factors never changes — treat this trick as a demonstration, not a technique.)

## Quoting and the Limits of Inspection

To ask about a *symbol itself* rather than what it points to, quote it:

```{r,echo=TRUE}
class(quote(x))
typeof(quote(x))
```

Some types resist inspection entirely: there is no way to isolate an `any`, `...`, `char`, or `promise` object — checking a promise's type would force its evaluation, converting it into an ordinary object.

## Copyright Notice

- **Copyright.** These slides are adapted from *R in a Nutshell: A Desktop Quick Reference* (2nd ed.) by Joseph Adler, O'Reilly Media. All rights reserved by the original author and publisher.

- **Non-commercial use only.** These materials are strictly for educational purposes and may not be used for commercial gain.

- **Attribution.** Any reproduction, distribution, or use of these materials must properly credit the original source.
