---
title: "Programming for Applications"
subtitle: "Chapter 7: R Objects"
author: "Yu-You Liou (NTU)"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "R in a Nutshell: A Desktop Quick Reference"
logo: "logo.svg"
editor: source
bibliography: bib.bib
suppress-bibliography: true
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch7-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center")
```

# **The Object Catalog**

## Primitive Object Types

Every R object has an underlying **type**, visible with `typeof()`. The catalog falls into natural families:

- **Basic vectors**: `logical`, `integer`, `double`, `complex`, `character`, `raw`;
- **Compound objects**: `list` and everything built on lists;
- **Language objects**: `symbol`, `call`, `expression`;
- **Functions**: `closure` (R-level), `builtin` and `special` (implemented in C);
- **Special objects**: `NULL`, environments, and a few internals.

```{r,echo=TRUE}
typeof(1.5); typeof(1L); typeof("a"); typeof(list()); typeof(mean)
```

## Vectors

Vectors are R's atoms: ordered, homogeneous collections. `c()` flattens its arguments recursively:

```{r,echo=TRUE}
v <- c(1, 2, c(3, 4, c(5)))
v
length(v)
```

Names can travel with the elements:

```{r,echo=TRUE}
prices <- c(apple = 30, banana = 12, cherry = 95)
prices["banana"]
```

## Lists

Lists are heterogeneous containers, the backbone of every complex structure in R:

```{r,echo=TRUE}
rec <- list(name = "Kim", scores = c(88, 92), pass = TRUE)
rec$scores
rec[["name"]]
rec[1]          # single brackets return a (sub)list
```

::: {.callout-note}
**`[ ]` vs. `[[ ]]`.** Single brackets *slice* a list and return a list; double brackets *extract* one component and return the component itself. Confusing the two is the most common list error.
:::

## Matrices and Arrays

A matrix is a vector with a `dim` attribute of length 2; arrays generalize to any number of dimensions:

```{r,echo=TRUE}
m <- matrix(1:6, nrow = 2)
m
dim(m)
t(m)            # transpose
m %*% t(m)      # matrix product
```

Row and column names make matrices self-documenting: `dimnames(m) <- list(c("r1","r2"), c("a","b","c"))`.

## Factors

A **factor** stores categorical data: integer codes plus a `levels` attribute mapping codes to labels.

```{r,echo=TRUE}
sizes <- factor(c("S", "L", "M", "L", "S"), levels = c("S", "M", "L"))
sizes
table(sizes)
as.integer(sizes)     # the codes underneath
```

- `ordered = TRUE` creates ordinal factors (`S < M < L`), which comparison operators respect.
- Factors are how regression functions know a column is categorical — they generate dummy variables automatically.

## Data Frames

The data frame — one row per observation, one column per variable — is technically a list of equal-length vectors with class `data.frame`:

```{r,echo=TRUE}
df <- data.frame(player = c("A", "B", "C"),
                 yards  = c(43, 28, 51),
                 good   = c(TRUE, FALSE, TRUE))
str(df)
df[df$good, "yards"]
```

`str()` is the single most useful inspection tool in R: structure, types, and a preview in one view.

## Formulas

A formula captures a model description *without evaluating it*:

```{r,echo=TRUE}
f <- yards ~ player + good
class(f)
```

Formula syntax: `+` adds terms, `-` removes them, `*` crosses (`a*b` = `a + b + a:b`), `:` is an interaction, `.` means "all other columns", and `I()` protects arithmetic (`I(x^2)`). Models, lattice plots, and aggregation functions all accept formulas.

## Time Series

`ts` objects attach regular time metadata to a vector. The course convention loads the textbook's turkey-price series from GitHub:

```{r,echo=TRUE}
temp <- tempfile(fileext = ".rda")
download.file("https://raw.githubusercontent.com/cran/nutshell/master/data/turkey.price.ts.rda", temp, mode = "wb")
load(temp)
class(turkey.price.ts)
start(turkey.price.ts); frequency(turkey.price.ts)
```

```{r,echo=TRUE,out.width="60%"}
plot(turkey.price.ts)
```

## Dates and Times

```{r,echo=TRUE}
d <- as.Date("2026-09-22")
d + 7                          # date arithmetic in days
weekdays(d)
now <- Sys.time()
class(now)                     # POSIXct: seconds since 1970
format(now, "%Y-%m-%d %H:%M")
```

- `Date` for daily data; `POSIXct` for timestamps.
- `difftime` objects represent durations; `seq(d, by = "month", length.out = 3)` builds calendars.

## Connections

Connections abstract over data sources — files, URLs, compressed archives, pipes — so reading functions need not care where bytes come from:

```{r,eval=FALSE,echo=TRUE}
temp <- tempfile(fileext = ".rda")
download.file("https://raw.githubusercontent.com/cran/nutshell/master/data/dow30.rda",
              temp, mode = "wb")
load(temp)           # our GitHub convention: download, then load

gz <- gzfile("big.csv.gz")    # transparent decompression
read.csv(gz)
```

The course convention — `download.file()` to a temporary file, then `load()` — uses connections under the hood; `mode = "wb"` keeps the binary file intact on every platform.

## Attributes

Attributes are the metadata that turn plain vectors into richer structures:

```{r,echo=TRUE}
x <- 1:6
attr(x, "dim") <- c(2, 3)     # now it's a matrix!
x
class(x)
attributes(x)
```

`names`, `dim`, `dimnames`, `levels`, and `class` are all just attributes — the entire object system of Chapter 10 rests on this one mechanism.

## Copyright Notice

- **Copyright.** These slides are adapted from *R in a Nutshell: A Desktop Quick Reference* (2nd ed.) by Joseph Adler, O'Reilly Media. All rights reserved by the original author and publisher.

- **Non-commercial use only.** These materials are strictly for educational purposes and may not be used for commercial gain.

- **Attribution.** Any reproduction, distribution, or use of these materials must properly credit the original source.
