---
title: "Prompt"
author: "Yu-You Liou"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r,echo=FALSE}
print(uuid::UUIDgenerate())
```

## Role

You are an expert academic assistant and an experienced instructor of R programming.

## Task

Generate a comprehensive, plagiarism-free lecture slide for *Chapter N* based on the official textbook **R in a Nutshell: A Desktop Quick Reference** (2nd ed.), located at the repository root.

## Instructions

1.	**Structural Alignment**: Replicate the exact formatting, structural hierarchy, and stylistic layout of an existing chapter directory (e.g., `ch3/`) to ensure visual and pedagogical consistency across chapters. Produce both `chN.qmd` (English) and `chN-zh.qmd` (Traditional Chinese), structurally identical and cross-linked from the title slide.

(*Note*: Do not include section numbers like x.x or x.x.x in slide titles, except for the first page.)

2.	**Exhaustive Coverage**: The handout must be meticulously detailed. Do not summarize or omit topics. Every code chunk, function, argument, and conceptual point introduced in the target chapter must be included and explained.

3.	**Paraphrasing & Academic Integrity**: To prevent plagiarism, completely rewrite and paraphrase all textual descriptions, explanations, and commentary from the textbook. The narrative flow and wording must be entirely original, academic, and tailored for a student audience, while strictly preserving the technical accuracy of the original content.

4.	**Code Execution**: Ensure all R code syntax is updated, correct, and properly formatted in markdown code blocks. Chunks that install software or require unavailable services must use `eval=FALSE`.

5.	**Dataset Loading Convention (course-specific)**: The textbook loads example data with `library(nutshell)` followed by `data(xxx)`. The `nutshell` package is no longer on CRAN, so **never** use it. Do **not** use `load(url(...))` either — it fails on some platforms. Instead, download each dataset from the CRAN GitHub mirror to a temporary file, then load it:

    ```r
    temp <- tempfile(fileext = ".rda")
    download.file("https://raw.githubusercontent.com/cran/nutshell/master/data/xxx.rda",
                  temp, mode = "wb")
    load(temp)
    ```

    When a chunk loads several datasets, declare `temp` once and repeat the `download.file()` + `load()` pair. Datasets from the companion packages use the same pattern with `nutshell.bbdb` or `nutshell.audioscrobbler` in place of `nutshell`.

6.	**Directory and Storage Architecture**: Replicate the established file structure utilized in the previous module. Dynamically generate and store the output within a newly designated directory (e.g., `chN/`), mirroring the organization and naming conventions of the existing chapter repositories (shared `bib.bib` and `logo.svg`).
