---
title: "Programming for Applications"
subtitle: "Chapter 4: R Packages"
author: "Yu-You Liou (NTU)"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "R in a Nutshell: A Desktop Quick Reference"
logo: "logo.svg"
editor: source
bibliography: bib.bib
suppress-bibliography: true
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch4-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center")
```

# **Extending R**

## What Is a Package?

A **package** bundles related functions, documentation, and data sets into a single installable unit — the R counterpart of a Perl module, a C library, or a Java class collection.

- Some packages ship with R itself (`stats`, `graphics`, `utils`, ...).
- Thousands more live in public repositories, covering everything from finance to genomics to the latest machine-learning methods.
- You can also build your own.

::: {.callout-note}
**Two separate steps.** A package must first be *installed* into a local library on disk (once), and then *loaded* into each session that uses it. Keeping the two apart saves memory, keeps the help system fast, and avoids name collisions between packages.
:::

## Libraries

A **library** is a directory on disk where installed packages live.

- R always has a system-level library; you can add personal libraries on top (useful when you lack admin rights).
- `.libPaths()` shows the libraries R is currently searching.

```{r,echo=TRUE}
.libPaths()
```

Note the vocabulary: you install a *package* into a *library* — despite the name of the `library()` function.

## Listing Packages

```{r,echo=TRUE,eval=FALSE}
library()                    # what is installed?
installed.packages()[1:5, c("Package", "Version")]
```

```{r,echo=TRUE}
# what is loaded right now?
(.packages())
```

- `installed.packages()` returns a matrix with one row per installed package — name, version, dependencies, license.
- A fresh session automatically loads a small set of base packages.

## Loading Packages

Load an installed package with `library()` (or `require()`, which returns `FALSE` instead of an error when the package is missing):

```{r,echo=TRUE}
library(lattice)         # now lattice functions are visible
"package:lattice" %in% search()
```

- Loading attaches the package to the **search path**, the list of places R looks when resolving a name (Chapter 8).
- `detach("package:lattice")` removes it again.
- On all platforms the command is the same; the GUIs merely offer menu shortcuts for it.

## CRAN

The **Comprehensive R Archive Network** is the central public repository:

- Over 20,000 contributed packages, each with documentation and platform-specific builds.
- Every package page lists its dependencies, manuals, and vignettes.
- **Task Views** (<https://cran.r-project.org/web/views/>) group packages by field — Econometrics, MachineLearning, TimeSeries, Graphics — and are the best way to discover what exists.

Other repositories you will meet: **Bioconductor** (genomics; Chapter 25) and **GitHub** (development versions).

## Installing Packages

One function does the work, downloading from CRAN and resolving dependencies automatically:

```{r,eval=FALSE,echo=TRUE}
install.packages("ggplot2")                     # one package
install.packages(c("dplyr", "lattice"))         # several at once
update.packages()                               # refresh everything
```

- In RStudio: *Tools → Install Packages...* is the same call behind a dialog box.
- Install once per machine; `library()` once per session.
- To install from a non-CRAN repository, pass `repos=`; for GitHub development versions, the `remotes` package provides `install_github()`.

## Package Data and the Course Convention

Many packages carry example data sets, loaded with `data()`:

```{r,echo=TRUE}
data(cars)        # from the datasets package
head(cars, 3)
```

The textbook's companion packages (`nutshell`, `nutshell.bbdb`, `nutshell.audioscrobbler`) were archived from CRAN, so this course streams their data sets directly from the CRAN GitHub mirror instead:

```{r,echo=TRUE}
temp <- tempfile(fileext = ".rda")
download.file("https://raw.githubusercontent.com/cran/nutshell/master/data/doctorates.rda", temp, mode = "wb")
load(temp)
doctorates[, 1:6]
```

## Custom Packages

Building a package is the natural endpoint once you accumulate functions you reuse:

- A package is a directory with a `DESCRIPTION` file (metadata), an `R/` folder (code), `man/` (help pages), and optionally `data/`.
- `package.skeleton()` generates the layout from objects in your workspace; modern workflows use the `devtools` and `roxygen2` packages.
- `R CMD build` creates the distributable archive; `R CMD check` runs an extensive quality audit.

```{bash,eval=FALSE,echo=TRUE}
R CMD build mypackage/
R CMD check mypackage_1.0.0.tar.gz
```

Even if you never publish to CRAN, packaging is the cleanest way to share code with teammates — or with your future self.

## Copyright Notice

- **Copyright.** These slides are adapted from *R in a Nutshell: A Desktop Quick Reference* (2nd ed.) by Joseph Adler, O'Reilly Media. All rights reserved by the original author and publisher.

- **Non-commercial use only.** These materials are strictly for educational purposes and may not be used for commercial gain.

- **Attribution.** Any reproduction, distribution, or use of these materials must properly credit the original source.
