---
title: "Programming for Applications"
subtitle: "Chapter 4: R Packages"
author: "Yu-You Liou (NTU)"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "R in a Nutshell: A Desktop Quick Reference"
logo: "logo.svg"
editor: source
bibliography: bib.bib
suppress-bibliography: true
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch4-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center")
```

# **The Package System**

## What Is a Package?

A **package** is a bundled set of related functions, help files, and data files — the unit in which R functionality is shared.

- Rough analogues elsewhere: modules in Perl, libraries in C/C++, classes in Java. (Careful, C programmers: in R, "library" means the *place packages live*, not the package itself.)
- Typically the contents are thematically related: the `stats` package collects statistical analysis functions.
- The ecosystem is enormous: graphics packages, statistical tests, cutting-edge machine learning — and industry-specific tools for microarray data, credit risk modeling, the social sciences...
- Some packages ship with R; thousands more live in public repositories; and you can build your own. This chapter covers all three.

## Libraries, and Why Packages Must Be Loaded

Two steps precede using any package:

1. it must be **installed** into a local **library** (by default one system-level library; you can add more);
2. it must be **loaded** into the current session.

Why insist on explicit loading instead of loading everything?

- **Help-system speed:** every loaded package slows help searches (the book's author learned this writing with dozens loaded).
- **Name conflicts:** two packages may define objects with the same name — you could silently call the wrong `fit`, with strange and unexpected results. Loading only what you need minimizes the risk.

# Listing Packages

## What Is Loaded, What Is Available

Three quick diagnostics:

```{r,echo=TRUE}
getOption("defaultPackages")   # loaded at startup (omits base, always loaded)
(.packages())                  # currently loaded (note the outer parentheses)
```

```{r,eval=FALSE,echo=TRUE}
(.packages(all.available=TRUE))  # everything installed in your libraries
library()                        # same idea, shown in a window
```

- `base` is missing from the default list because it implements core language features and is *always* loaded.

## Included Packages (1): Infrastructure

R ships with a curated set of packages. The first group implements the language and environment themselves:

| Package | Default | Purpose |
|---|---|---|
| `base` | ✓ | Core language: arithmetic, I/O, programming support |
| `utils` | ✓ | Utilities: package management, file I/O, editing |
| `methods` | ✓ | Formal (S4) classes and methods |
| `stats` | ✓ | Statistical tests, distributions, modeling tools |
| `graphics` | ✓ | Base graphics functions |
| `grDevices` | ✓ | Graphics devices for base and grid graphics |
| `datasets` | ✓ | Famous example data sets |
| `grid` |  | Lower-level engine for sophisticated graphics |
| `compiler` |  | Byte-code compiler for R |
| `parallel` |  | Parallel computation, parallel RNG |
| `tools` |  | Package-development tools |
| `tcltk` |  | Tcl/Tk interface for platform-independent UIs |
| `codetools` |  | Code-analysis tools |

## Included Packages (2): Statistics and Modeling

The second group provides widely used statistical methods — for many analyses you never need anything else:

| Package | Purpose |
|---|---|
| `MASS` | Functions and data from Venables & Ripley's *Modern Applied Statistics with S* — a trove of useful statistics |
| `cluster` | Clustering algorithms |
| `class` | Classification: nearest neighbors, SOMs, LVQ |
| `nnet` | Feed-forward neural networks, multinomial log-linear models |
| `rpart` | Recursive partitioning and regression trees |
| `survival` | Survival analysis |
| `nlme` | Linear and nonlinear mixed-effects models |
| `mgcv` | Generalized additive (mixed) models |
| `splines` | Regression splines |
| `boot` | Bootstrap resampling |
| `KernSmooth` | Kernel smoothing |
| `spatial` | Kriging, point-pattern analysis |
| `foreign` | Reading Stata, SAS, SPSS files |
| `lattice` | Trellis graphics — prettier than the defaults |
| `stats4` | Statistics as S4 classes/methods |

# Loading Packages

## The `library()` Command

Use a function from an unloaded package and R refuses:

```{r,error=TRUE,echo=TRUE}
fit <- rpart(Kyphosis ~ Age + Number + Start, data = kyphosis)
```

Load the package first, and all is well:

```{r,echo=TRUE}
library(rpart)
```

- A sibling command, `require()`, takes slightly different arguments (and returns `FALSE` instead of an error when the package is missing) — see the help files.
- GUIs offer menu-driven loading too — but **always put the `library()` call in your script anyway**, or the script will fail when run elsewhere.

## Loading via the GUI, per Platform

- **Windows / Linux GUI:** choose *Load package* from the *Packages* menu; a window lists the packages available to load.
- **macOS:** fancier — choose *Package Manager* from the *Packages & Data* menu. The Package Manager shows which packages are loaded, loads them with a click, and even browses each package's help file.
- RStudio users: the *Packages* pane does the same job with checkboxes — and writes the `library()` call to the console so you can copy it into your script.

# Exploring Package Repositories

## CRAN, Bioconductor, R-Forge

Thousands of packages live online; the two biggest sources:

- **CRAN** (Comprehensive R Archive Network) — hosted by the R Foundation, mirrored worldwide (pick a nearby mirror: faster for you, kinder to the servers). The book counted 1,698 packages in February 2009; today the count exceeds 20,000. If you know Perl: CRAN is R's CPAN.
- **Bioconductor** — an open-source project for genomic-data analysis tools, built on R, distributed as R packages through its own channel (mostly *not* on CRAN).
- **R-Forge** — collaborative development site for works in progress; interesting but read the disclaimers — many projects there are unfinished.

R can install from arbitrary repositories, but in practice nearly all projects publish on CRAN (some books even distribute sample code and data there). The modern exception: development versions on **GitHub** — covered shortly.

## Finding Packages on the Web

R's own GUI installs packages well but *searches* them poorly. Use the browser:

| Repository | URL |
|---|---|
| CRAN | <https://cran.r-project.org/web/packages/> (authoritative list; use your local mirror) |
| Bioconductor | <https://www.bioconductor.org/packages/release/BiocViews.html> |
| R-Forge | <https://r-forge.r-project.org/> |

A search engine works too: the book's tip is to search for **"R package" + the application name**. Its example: searching *"R package multivariate additive regression splines"* surfaces the `mda` package with its `mars` function — though the author later found the `earth` package a better choice for that algorithm. (Today, sites like <https://r-universe.dev> and CRAN Task Views make this search far easier.)

# Installing Packages

## Windows and Linux GUIs

Installing through the Windows GUI:

1. *(Optional)* By default R fetches from the "CRAN" and "CRAN (extra)" categories; pick more via *Select repositories* on the *Packages* menu.
2. From the *Packages* menu, choose *Install package(s)*.
3. First install of the session? R asks you to pick a **mirror** — choose one nearby.
4. Click the package you want and press OK; R downloads and installs it.

Permissions matter here exactly as in Chapter 1: an Administrator account installs anywhere; a standard account may need R installed in its own directory, or R run as Administrator, to add packages.

## macOS GUI

The macOS installer interface shows more information but is a bit more confusing:

1. From the *Packages & Data* menu, select *Package Installer*.
2. *(Optional)* A menu at the top-left selects the package category — initially "CRAN (binaries)".
3. Click **Get List** to display available packages.
4. The search box filters the list — but only *after* you've clicked Get List.
5. Select the packages you want and press **Install Selected**.

Two cautions from the book: R installs **system-wide by default** (choose an alternative location if you lack permissions), and this interface does **not** install dependencies automatically — loading a package whose dependencies are missing yields an error.

## Installing from the R Console

The console route works everywhere — and is the one to put in scripts. The book's example, installing `tree` and `maptree` in one call:

```{r,eval=FALSE,echo=TRUE}
install.packages(c("tree", "maptree"))
```

R reports the download of each package and stores them in the default library (the location named by `.Library`). To uninstall, name the packages *and* the library they went into:

```{r,eval=FALSE,echo=TRUE}
remove.packages(c("tree", "maptree"), .Library)
```

::: {.callout-note}
Modern `install.packages()` resolves and installs **dependencies automatically** — one of the quiet improvements since the book's edition.
:::

## The Package-Management Vocabulary

The full family of console commands:

| Command | Description |
|---|---|
| `installed.packages` | Matrix describing all installed packages |
| `available.packages` | All packages available on the repository |
| `old.packages` | Installed packages with newer versions available |
| `new.packages` | Available packages not yet installed |
| `download.packages` | Download packages to a local directory |
| `install.packages` | Install packages from the repository |
| `remove.packages` | Remove installed packages |
| `update.packages` | Update installed packages to latest versions |
| `setRepositories` | Set the list of package repositories |

`update.packages()` at the start of term keeps your toolbox current.

## Installing from the System Command Line

Downloaded a package file yourself? Install it without launching the R shell, via `R CMD INSTALL`. The book's example with `aplpack` ("Another PLotting PACKage"):

```{bash,eval=FALSE,echo=TRUE}
$ R CMD INSTALL aplpack_1.1.1.tgz
```

A successful run reports:

```
* Installing to library '/Library/Frameworks/R.framework/Resources/library'
* Installing *binary* package 'aplpack' ...
* DONE (aplpack)
```

## Installing from Other Repositories: devtools

Not everything lives on CRAN. The **devtools** package installs from git repositories and other URLs — the book's example is the development version of ggplot2, which Hadley Wickham hosts on GitHub:

```{r,eval=FALSE,echo=TRUE}
# first time only:
install.packages("devtools")

library(devtools)
install_github("tidyverse/ggplot2")   # modern repo path: owner/repo
```

(The book's call was `install_github("ggplot2")`; the function now requires the `owner/repo` form. The lighter-weight `remotes` package provides the same `install_github()` today.)

# Custom Packages

## Why Build Your Own?

Building a package is the right move when you want to **share code or data** with others — or simply to pack your own work into a reusable, documented form. The book practices what it preaches: its data sets ship as the `nutshell` package, created exactly this way.

The easy path has three stops: create the directory skeleton, edit the metadata, build and check.

## Creating a Package Directory: `package.skeleton`

`package.skeleton` lays out the required directory structure and can copy a set of R objects into it:

```{r,eval=FALSE,echo=TRUE}
package.skeleton(name = "anRpackage", list, environment = .GlobalEnv,
                 path = ".", force = FALSE, namespace = FALSE,
                 code_files = character())
```

| Argument | Description | Default |
|---|---|---|
| `name` | Name for the new package | `"anRpackage"` (the book: possibly the least useful default in all of R) |
| `list` | Names of R objects to include | |
| `environment` | Where to evaluate `list` | `.GlobalEnv` |
| `path` | Filesystem location | `"."` |
| `force` | Overwrite an existing directory? | `FALSE` |
| `namespace` | Add a namespace? | `FALSE` |
| `code_files` | Paths of files containing R code | `character()` |

## What the Skeleton Contains

The book's own invocation:

```{r,eval=FALSE,echo=TRUE}
package.skeleton(name = "nutshell", path = "~/Documents/book/current/")
```

R reports creating directories, a DESCRIPTION file, a Read-and-delete-me, saved functions and data, and help files. The key directories:

- **`man/`** — help files;
- **`R/`** — R source files;
- **`data/`** — data files, in any of three forms: R data files (`save()` output, suffix `.rda`/`.Rdata`), comma-separated `.csv` files, or R source `.R` files;
- **`DESCRIPTION`** — the package's metadata, at the root.

## The DESCRIPTION File

The generated template — every line meant to be edited:

```
Package: nutshell
Type: Package
Title: What the package does (short line)
Version: 1.0
Date: 2012-03-13
Author: Who wrote it
Maintainer: Whom to complain to <yourfault@somewhere.net>
Description: More about what it does (maybe more than one line)
License: What license is it under?
```

Useful optional fields:

- **`LazyLoad`** — if `yes` (default), data objects load as lightweight *promises* rather than into memory; full access, (almost) no space.
- **`LazyData`** — the same idea, specifically for data files.
- **`Depends`** — requirements on R or other packages, e.g. `Depends: R(>= 2.8), nnet`.

## Documenting the Package

R automates much of the help-file boilerplate with the `prompt` family:

- `prompt` — generic documentation for an object;
- `promptData` — for data files;
- `promptMethods` — for methods of a generic function;
- `promptClass` — for a class.

Each writes a skeleton `.Rd` file into `man/` for you to complete. See their help pages for details. (Most modern developers generate `.Rd` files from inline comments with the `roxygen2` package instead.)

## Building and Checking

With all materials in place, finish from the system command line (not the R shell). First verify the package complies with the rules and builds correctly:

```{bash,eval=FALSE,echo=TRUE}
$ R CMD check nutshell
```

Then build it:

```{bash,eval=FALSE,echo=TRUE}
$ R CMD build nutshell
```

Both accept `--help` for options. The definitive reference for serious package authors is the CRAN manual *Writing R Extensions*: <https://cran.r-project.org/doc/manuals/R-exts.pdf>.

## Copyright Notice

- **Copyright.** These slides are adapted from *R in a Nutshell: A Desktop Quick Reference* (2nd ed.) by Joseph Adler, O'Reilly Media. All rights reserved by the original author and publisher.

- **Non-commercial use only.** These materials are strictly for educational purposes and may not be used for commercial gain.

- **Attribution.** Any reproduction, distribution, or use of these materials must properly credit the original source.
