---
title: "Programming for Applications"
subtitle: "Chapter 13: Graphics"
author: "Yu-You Liou (NTU)"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "R in a Nutshell: A Desktop Quick Reference"
logo: "logo.svg"
editor: source
bibliography: bib.bib
suppress-bibliography: true
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch13-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center")
```

# **Base Graphics**

## The Painter's Model

Base graphics follow a "pen on paper" model: high-level functions draw a complete chart, low-level functions add ink on top, and nothing can be erased — only redrawn [@tukey1977exploratory; @cleveland1993visualizing].

- **High-level**: `plot()`, `hist()`, `barplot()`, `pie()`, `boxplot()`, ...
- **Low-level**: `points()`, `lines()`, `text()`, `legend()`, `abline()`, ...
- Output goes to the active **graphics device** (screen, PDF, PNG).

Our example data, loaded the course way:

```{r,echo=TRUE,cache=TRUE}
temp <- tempfile(fileext = ".rda")
download.file("https://raw.githubusercontent.com/cran/nutshell/master/data/toxins.and.cancer.rda", temp, mode = "wb")
load(temp)
download.file("https://raw.githubusercontent.com/cran/nutshell/master/data/batting.2008.rda", temp, mode = "wb")
load(temp)
download.file("https://raw.githubusercontent.com/cran/nutshell/master/data/turkey.price.ts.rda", temp, mode = "wb")
load(temp)
download.file("https://raw.githubusercontent.com/cran/nutshell/master/data/yosemite.rda", temp, mode = "wb")
load(temp)
```

## Scatter Plots

Does environmental toxin release relate to cancer mortality? One `plot()` call per question:

```{r,echo=TRUE,out.width="65%"}
plot(toxins.and.cancer$total_toxic_chemicals / toxins.and.cancer$Surface_Area,
     toxins.and.cancer$deaths_total / toxins.and.cancer$Population,
     xlab = "Toxins released per sq mile",
     ylab = "Deaths per capita")
```

`plot(x, y)` is the universal entry point; formula notation `plot(y ~ x, data = df)` works too.

## Identifying Points

Label interesting observations with low-level additions:

```{r,echo=TRUE,out.width="62%"}
attach(toxins.and.cancer)
plot(air_on_site / Surface_Area, deaths_lung / Population,
     xlab = "Air toxins", ylab = "Lung-cancer deaths per capita")
text(air_on_site / Surface_Area, deaths_lung / Population,
     labels = State_Abbrev, cex = 0.6, pos = 4)
detach(toxins.and.cancer)
```

Interactively, `identify()` and `locator()` let you click on the device to find points and coordinates.

## Plotting Time Series

`plot()` knows what a `ts` object is and draws lines against the calendar:

```{r,echo=TRUE,out.width="65%"}
plot(turkey.price.ts, ylab = "US cents per pound",
     main = "Average turkey price by month")
```

Holiday demand is visible at a glance — prices dip every November.

## Bar Charts

```{r,echo=TRUE,out.width="65%"}
doubles.by.team <- tapply(batting.2008$`2B`, batting.2008$teamID, sum)
barplot(sort(doubles.by.team, decreasing = TRUE)[1:8],
        las = 2, main = "Doubles by team, 2008 (top 8)")
```

- `horiz=TRUE` flips the bars; matrix input draws grouped or stacked bars (`beside=`).
- `las=2` rotates axis labels — small flag, big readability gain.

## Pie Charts and Mosaic Plots

```{r,echo=TRUE,out.width="55%"}
pie(c(Good = 787, Missed = 163, Blocked = 24),
    main = "Field goal outcomes, 2005")
```

- Perception research is brutal to pie charts: humans compare lengths better than angles. Prefer bars when values are close.
- For two categorical variables, `mosaicplot(~ cut + color, data = ...)` tiles the plane proportionally to the cross-tabulation.

## Three-Dimensional Data

Surfaces and grids have three friends: `image()`, `contour()`, and `persp()`. Yosemite Valley's elevation grid:

```{r,echo=TRUE,out.width="60%"}
image(yosemite, col = terrain.colors(64), asp = 1,
      main = "Half Dome (elevation raster)")
```

```{r,echo=TRUE,out.width="55%"}
persp(yosemite[seq(1, nrow(yosemite), 6), seq(1, ncol(yosemite), 6)],
      expand = 0.2, theta = 225, phi = 20, shade = 0.6, border = NA)
```

## Plotting Distributions

```{r,echo=TRUE,out.width="60%"}
hist(batting.2008$AB[batting.2008$AB > 100], breaks = 30,
     xlab = "Batting average", main = "Hitters with 100+ at-bats")
```

```{r,echo=TRUE,out.width="60%"}
plot(density(batting.2008$AB[batting.2008$AB > 100]),
     main = "Kernel density estimate")
rug(batting.2008$AB[batting.2008$AB > 100])
```

`qqnorm()` compares a sample against the normal distribution — points on the line mean "plausibly normal."

## Box Plots

The five-number summary as a picture, one box per group:

```{r,echo=TRUE,out.width="68%"}
boxplot(AB ~ teamID, data = batting.2008[batting.2008$AB > 100, ],
        las = 2, cex.axis = 0.7, ylab = "Batting average")
```

Whiskers extend to 1.5 × IQR; dots beyond are flagged as outliers — Tukey's recipe, unchanged since 1977.

## Graphics Devices

Charts render on a **device**. The screen is the default; files are devices you open and close:

```{r,eval=FALSE,echo=TRUE}
pdf("figure.pdf", width = 7, height = 5)    # vector: crisp at any size
plot(turkey.price.ts)
dev.off()                                    # ALWAYS close the device

png("figure.png", width = 800, height = 600) # raster: for web pages
plot(turkey.price.ts)
dev.off()
```

Forgetting `dev.off()` leaves a half-written file — the most common graphics bug in scripts.

## Customizing Charts

Most chart functions share arguments; `par()` sets session-wide graphical parameters:

```{r,echo=TRUE,out.width="60%"}
oldpar <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))   # 1x2 layout
plot(cars, pch = 19, col = "steelblue", main = "pch & col")
plot(cars, type = "l", lty = 2, lwd = 2, main = "lines")
par(oldpar)
```

- Per-call: `pch` (symbol), `col`, `lty`/`lwd` (line type/width), `cex` (size), `xlim`/`ylim`, `main`/`xlab`/`ylab`.
- Session: `par(mfrow=)` for panel grids, `par(mar=)` for margins. Save and restore `par()` like we did above.

## Annotation Functions

Build up a figure layer by layer:

```{r,echo=TRUE,out.width="60%"}
plot(cars, pch = 19)
abline(lm(dist ~ speed, data = cars), col = "red", lwd = 2)
text(8, 100, "fitted line", col = "red")
legend("topleft", legend = c("data", "fit"),
       pch = c(19, NA), lty = c(NA, 1), col = c("black", "red"))
title(main = "Stopping distance vs. speed")
```

`points()`, `lines()`, `segments()`, `arrows()`, `rect()`, and `polygon()` complete the ink-adding toolkit.

## Copyright Notice

- **Copyright.** These slides are adapted from *R in a Nutshell: A Desktop Quick Reference* (2nd ed.) by Joseph Adler, O'Reilly Media. All rights reserved by the original author and publisher.

- **Non-commercial use only.** These materials are strictly for educational purposes and may not be used for commercial gain.

- **Attribution.** Any reproduction, distribution, or use of these materials must properly credit the original source.
