---
title: "Programming for Applications"
subtitle: "Chapter 13: Graphics"
author: "Yu-You Liou (NTU)"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "R in a Nutshell: A Desktop Quick Reference"
logo: "logo.svg"
editor: source
bibliography: bib.bib
suppress-bibliography: true
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch13-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center", fig.height = 4.2)
```

# **Base Graphics**

## Three Plotting Systems, One Chapter Each

R plots graphics many ways; the book focuses on the three most popular packages:

- **graphics** (this chapter) — the classic system: easy to customize, easy to modify, interactive on screen;
- **lattice** (Chapter 14) — built for *splitting data by a conditioning variable*;
- **ggplot2** (Chapter 15) — a different metaphor that makes complex, pretty charts easy.

This chapter: a tour of the common chart types, then graphics devices, then the full machinery of customization (`par`, low-level functions).

# Scatter Plots

## plot: the Generic Entry Point

Our example data: 2008 cancer cases (American Cancer Society) and 2006 toxic releases by state (EPA), packaged as `toxins.and.cancer`:

```{r,echo=TRUE}
temp <- tempfile(fileext = ".rda")
download.file("https://raw.githubusercontent.com/cran/nutshell/master/data/toxins.and.cancer.rda",
              temp, mode = "wb")
load(temp)
```

`plot` is generic — vectors, tables, time series all plot. For two vectors, `plot.default` runs. Cancer death rate vs. toxin release rate:

```{r,echo=TRUE}
attach(toxins.and.cancer)
plot(total_toxic_chemicals/Surface_Area, deaths_total/Population)
```

## plot.default: Arguments

```{r,eval=FALSE,echo=TRUE}
plot(x, y = NULL, type = "p", xlim = NULL, ylim = NULL, log = "",
     main = NULL, sub = NULL, xlab = NULL, ylab = NULL,
     ann = par("ann"), axes = TRUE, frame.plot = axes,
     panel.first = NULL, panel.last = NULL, asp = NA, ...)
```

| Argument | Description |
|---|---|
| `x`, `y` | data: two vectors, or a time series / formula / list / two-column matrix (see `?xy.coords`) |
| `type` | `"p"` points, `"l"` lines, `"o"` overplotted, `"b"` both, `"s"` steps, `"h"` vertical lines, `"n"` nothing |
| `xlim` / `ylim` | axis limits |
| `log` | logarithmic axes: `""`, `"x"`, `"y"`, `"xy"` |
| `main` / `sub` / `xlab` / `ylab` | titles and axis labels |
| `ann` / `axes` / `frame.plot` | draw annotation / axes / surrounding box? |
| `panel.first` / `panel.last` | expressions evaluated before / after points are drawn |
| `asp` | aspect ratio y/x |
| `...` | graphical parameters (see below) |

## Interactive Identification, and Labels

A stronger correlation appears between *airborne* toxins and *lung* cancer:

```{r,echo=TRUE}
plot(air_on_site/Surface_Area, deaths_lung/Population,
     xlab="Air Release Rate of Toxic Chemicals",
     ylab="Lung Cancer Death Rate")
text(air_on_site/Surface_Area, deaths_lung/Population,
     labels=State_Abbrev, cex=0.5, adj=c(0,-1))
```

- Interactively: `locator(1)` prints the coordinates of a clicked point; `identify(x, y, labels)` labels points as you click them.
- For all labels at once, `text` does it (here resized with `cex`, repositioned with `adj`). The relationship *is* statistically significant (Chapter 18) — though that alone argues no causal story.

## matplot, smoothScatter, pairs

Three relatives for special situations:

- **`matplot(x, y, type=, lty=, lwd=, pch=, col=, cex=, ..., add=)`** plots all columns of one matrix against all columns of another; the usual `par`-style arguments accept *vectors*, one entry per series.
- **`smoothScatter`** shades regions by point *density* — the cure for overplotting with very many points (see Chapter 16).
- **`pairs`** draws a scatter plot matrix for every pair of variables:

```{r,echo=TRUE}
temp <- tempfile(fileext = ".rda")
download.file("https://raw.githubusercontent.com/cran/nutshell/master/data/batting.2008.rda",
              temp, mode = "wb")
load(temp)
pairs(batting.2008[batting.2008$AB>100, c("H", "R", "SO", "BB", "HR")])
```

# Plotting Time Series

## plot.ts and acf

`plot` has a time-series method (`plot.type` = `"multiple"`/`"single"`, `panel`, `nc`, `yax.flip`, margin controls...). The turkey prices of Chapter 7:

```{r,echo=TRUE}
temp <- tempfile(fileext = ".rda")
download.file("https://raw.githubusercontent.com/cran/nutshell/master/data/turkey.price.ts.rda",
              temp, mode = "wb")
load(temp)
plot(turkey.price.ts)
```

Strongly seasonal: November/December sales (Thanksgiving, Christmas), minor spring dips (Easter).

## The Correlogram

The **autocorrelation function** shows how correlated points are with their past, by lag — `acf` computes it and plots by default:

```{r,echo=TRUE}
acf(turkey.price.ts)
```

Correlated at 12-month cycles, inversely at 6-month cycles. (Time series analysis proper: Chapter 23.)

# Bar Charts

## barplot Basics

US doctoral degrees, 2001–2006 (Statistical Abstract), entered directly:

```{r,echo=TRUE}
doctorates <- data.frame(
  year=c(2001, 2002, 2003, 2004, 2005, 2006),
  engineering=c(5323, 5511, 5079, 5280, 5777, 6425),
  science=c(20643, 20017, 19529, 20001, 20498, 21564),
  education=c(6436, 6349, 6503, 6643, 6635, 6226),
  health=c(1591, 1541, 1654, 1633, 1720, 1785),
  humanities=c(5213, 5178, 5051, 5020, 5013, 4949),
  other=c(2159, 2141, 2209, 2180, 2480, 2436))
doctorates.m <- as.matrix(doctorates[2:7])   # barplot wants a matrix,
rownames(doctorates.m) <- doctorates[, 1]    # not a data frame
barplot(doctorates.m[1, ])
```

Default behavior: y-axis drawn, no x-axis line, bars named from column names.

## Grouped and Stacked Bars

Side-by-side horizontal bars with a legend; then stacked columns (transpose so columns = years, widen `ylim` to fit the legend):

```{r,echo=TRUE}
barplot(doctorates.m, beside=TRUE, horiz=TRUE, legend=TRUE, cex.names=.75)
barplot(t(doctorates.m), legend=TRUE, ylim=c(0, 66000))
```

## barplot: Arguments

Key arguments (of many):

| Argument | Description |
|---|---|
| `height` | vector, or matrix (stacked if `beside=FALSE`, grouped if `TRUE`) |
| `width` / `space` | bar widths / gaps (with `beside=TRUE`: `c(within, between)` groups) |
| `names.arg` / `legend.text` | bar labels / legend (logical = use rownames) |
| `beside` / `horiz` | grouped vs. stacked / horizontal vs. vertical |
| `density` / `angle` | shading lines per inch / their slope |
| `col` / `border` | bar colors / border color |
| `main`, `sub`, `xlab`, `ylab`, `xlim`, `ylim`, `log`, `axes` | the usual annotations |
| `xpd` | may bars overflow the plot region? |
| `cex.axis` / `cex.names` / `axisnames` / `axis.lty` | axis text sizing and styling |
| `inside` / `offset` / `add` / `args.legend` | separators, baseline shift, overplot, legend options |

# Pie Charts

## pie — Use Sparingly

What happened to US-caught fish in 2006 (millions of pounds):

```{r,echo=TRUE}
domestic.catch.2006 <- c(7752, 1166, 463, 108)
names(domestic.catch.2006) <- c("Fresh and frozen",
  "Reduced to meal, oil, etc.", "Canned", "Cured")
pie(domestic.catch.2006, init.angle=100, cex=.6)
```

Arguments: `x` (nonnegative values), `labels`, `edges` (polygon smoothness), `radius`, `clockwise` + `init.angle`, `density`/`angle` shading, `col` (default: six pastels), `border`, `lty`, `main`.

::: {.callout-warning}
**The help file itself says it:** "Pie charts are a very bad way of displaying information. The eye is good at judging linear measures and bad at judging relative areas. A bar chart or dot chart is a preferable way of displaying this type of data."
:::

# Plotting Categorical Data

## Conditional Density: cdplot

How does a *categorical* outcome vary along a *numeric* axis? `cdplot` smooths with kernel density estimates (`density` underneath; arguments `bw`, `n`, `from`, `to` pass through; `ylevels` reorders; formula+`data`+`subset` supported). Batting hand by batting average:

```{r,echo=TRUE}
batting.w.names.2008 <- transform(batting.2008,
  AVG=H/AB, bats=as.factor(bats), throws=as.factor(throws))
cdplot(bats~AVG, data=batting.w.names.2008,
       subset=(batting.w.names.2008$AB>100))
```

The proportion of switch hitters (`B`) rises with batting average.

## Mosaic, Spine, and Association Plots

For two factors, three displays of the contingency structure:

```{r,echo=TRUE}
mosaicplot(formula=bats~throws, data=batting.w.names.2008, color=TRUE)
```

- `mosaicplot` boxes are proportional to counts; further arguments: `sort`, `off` (spacing), `dir` (orientations), `shade` (extended plots coloring log-linear residuals), `margin`, `type` (`"pearson"`/`"deviance"`/`"FT"` residuals).
- `spineplot(bats~throws, data=...)` — the close sibling.
- `assocplot(table(bats, throws))` — bars showing each cell's *deviation from independence* (Cohen–Friendly).
- See also `stars` and `fourfoldplot`.

# Three-Dimensional Data

## The Yosemite Elevation Data

All 3-D functions plot a **matrix**: row index = x, column index = y, cell = z. The book's example: USGS elevation data for Yosemite Valley, downloaded as GridFloat (a 562×253 stream of 4-byte floats read with `readBin(..., size=4, endian="little")`, then given `dim`); packaged as `yosemite`:

```{r,echo=TRUE}
temp <- tempfile(fileext = ".rda")
download.file("https://raw.githubusercontent.com/cran/nutshell/master/data/yosemite.rda",
              temp, mode = "wb")
load(temp)
dim(yosemite)
```

## persp: Surfaces in Perspective

Key arguments: `x`/`y` (coordinates of rows/columns), `z` (the matrix), `theta`/`phi` (viewing azimuth/colatitude), `r`/`d` (viewpoint distance, perspective strength), `expand` (z exaggeration), `col`/`border`, `ltheta`/`lphi` (light source), `shade`, `box`/`axes`/`nticks`/`ticktype`.

Looking toward Half Dome — flip the columns (persp draws y bottom-up), take a square 253-column subset (mind the fencepost `+1`!), rotate and light:

```{r,echo=TRUE}
halfdome <- yosemite[(nrow(yosemite) - ncol(yosemite) + 1):562,
                     seq(from=253, to=1)]
persp(halfdome, col=grey(.25), border=NA, expand=.15,
      theta=225, phi=20, ltheta=45, lphi=20, shade=.75)
```

## image and heatmap

`image` color-codes the matrix as a grid (arguments: `zlim`, `col` — typically `heat.colors`, `topo.colors`, `terrain.colors`, `breaks`, `oldstyle`, `add`):

```{r,echo=TRUE}
image(yosemite, asp=253/562, ylim=c(1,0), col=sapply((0:32)/32, gray))
```

(Tweaks: aspect ratio matching the data, reversed y, 32 grays.) The biology-flavored relative **`heatmap`** adds optional **dendrograms** along the margins — `Rowv`/`Colv`, `distfun`/`hclustfun`, `scale`, `margins`, `cexRow`/`cexCol` and friends.

## contour

`contour` joins equal values with lines — arguments `nlevels`/`levels`, `labels`/`labcex`/`drawlabels`/`method` (label placement: `"simple"`, `"edge"`, `"flattest"`), `vfont`, plus the usual `col`/`lty`/`lwd`/`add`:

```{r,echo=TRUE}
contour(yosemite, asp=253/562, ylim=c(1, 0))
```

Contours are commonly *added to* an image plot (`add=TRUE`).

# Plotting Distributions

## Histograms

Plate appearances (PA = AB+BB+HBP+SF+SH) per batter, 2008:

```{r,echo=TRUE}
batting.2008 <- transform(batting.2008, PA=AB+BB+HBP+SF+SH)
hist(batting.2008$PA)
```

A horde of sub-50-PA players — worth excluding from rate statistics. Zoom in and refine the bins:

```{r,echo=TRUE}
hist(batting.2008[batting.2008$PA>25, "PA"], breaks=50, cex.main=.8)
```

## Density Plots and Rugs

Many statisticians prefer **density plots** — more robust, easier to read. Compute with `density`, draw with `plot`, garnish with a `rug` (a strip plot along the axis):

```{r,echo=TRUE}
plot(density(batting.2008[batting.2008$PA>25, "PA"]))
rug(batting.2008[batting.2008$PA>25, "PA"])
```

## Quantile-Quantile Plots

A **Q-Q plot** charts sample quantiles against theoretical ones — points on the 45° diagonal mean the distributions match. `qqnorm` compares against the normal; `qqplot` compares two arbitrary distributions:

```{r,echo=TRUE}
qqnorm(batting.2008$AB)
```

At-bats are decidedly *not* normal — the bowed shape says it instantly.

# Box Plots

## Anatomy and Function

A **box plot** compacts a distribution: box = interquartile range (25th–75th percentile), inner line = median, whiskers = *adjacent values* (largest observation ≤ Q3 + 1.5·IQR; smallest ≥ Q1 − 1.5·IQR), loose points = outside values.

Key `boxplot` arguments: formula `y~grp` + `data` + `subset` (or vectors directly); `range` (whisker reach), `width`/`varwidth` (∝ √n), `notch` (±1.58·IQR/√n intervals), `outline`, `names`, `plot` (FALSE returns statistics), `border`/`col`, `log`, `horizontal`, `add`, `at`, `pars`.

## OBP by Team

On-base percentage by American League team, regulars only:

```{r,echo=TRUE}
batting.2008 <- transform(batting.2008, OBP=(H+BB+HBP)/(AB+BB+HBP+SF))
boxplot(OBP~teamID,
        data=batting.2008[batting.2008$PA>100 & batting.2008$lgID=="AL",],
        cex.axis=.7)
```

# Graphics Devices

## Screens and Files

Graphics land on a **device**. Defaults by platform: `windows` (Windows), `X11` (Unix), `quartz` (macOS). File devices: `bmp`, `jpeg`, `png`, `tiff`, plus `postscript`, `pdf`, `pictex` (LaTeX), `xfig`, `bitmap`.

- Most accept `width`, `height`, `pointsize`; file devices take a `file` argument. (Pedantically `filename` for bmp/jpeg/png/tiff — but partial matching lets `file` work everywhere.)
- **Close the device with `dev.off()`** to actually write the file. The book's own figures were produced exactly this way:

```{r,eval=FALSE,echo=TRUE}
png("scatter.1.png", width=4.3, height=4.3, units="in", res=72)
plot(total_toxic_chemicals/Surface_Area, deaths_total/Population)
dev.off()
```

# Customizing Charts

## Four Routes, and Common Arguments

Customize via (1) arguments to charting functions, (2) session parameters (`par`), (3) functions that modify an existing chart, or (4) writing your own functions. Arguments most chart functions share: `add`, `axes`, `log`, `type`, `xlab`/`ylab`, `main`, `sub`.

**`par`** queries and sets parameters on the active device — set once, affect every subsequent plot:

```{r,echo=TRUE}
par("bg")
```

```{r,eval=FALSE,echo=TRUE}
par(bg="white")           # set
par()                     # list everything
```

All parameters are read-write except the read-only `cin`, `cra`, `csi`, `cxy`, `din`. Caution: some functions reuse parameter names with different meanings (e.g., `bg` in `points` = point background).

## Layout: Margins and Multiple Plots

- **Margins:** `mai` (inches) / `mar` (text lines, default `c(5.1, 4.1, 4.1, 2.1)` for bottom/left/top/right) / `mex` (line-size factor); title-label spacing via `mgp`; device size via read-only `din`; square plot region via `pty="s"`.
- **Multiple figures:** `mfcol=c(3,2)` (fill by column) or `mfrow` (by row) split the device; `mfg` jumps to a chosen cell. Outer margins: `omi`/`oma`/`omd`. Per-figure geometry: `pin`, `plt`, `fig`. Clipping: `xpd`.

```{r,echo=TRUE,fig.height=5}
par(mfcol=c(3, 2))
pie(c(5, 4, 3)); plot(x=c(1,2,3,4,5), y=c(1.1,1.9,3,3.9,6))
barplot(c(1,2,3,4,5)); barplot(c(1,2,3,4,5), horiz=TRUE)
pie(c(5,4,3,2,1)); plot(c(1,2,3,4,5,6), c(4,3,6,2,1,1))
par(mfcol=c(1, 1))
```

(Easier alternatives: `layout`, `split.screen` — or the grid/lattice packages.)

## Text, Lines, Colors, Axes, Points

- **Text size:** `ps` (point size) × `cex` (global scale) × `cex.axis`/`cex.lab`/`cex.main`/`cex.sub`. **Typeface:** `family`; style via `font` (+ `.axis`, `.lab`, `.main`, `.sub`). Alignment `adj`, line spacing `lheight`, rotation `crt` (characters) / `srt` (strings).
- **Lines:** type `lty` (0=blank, 1=solid, 2=dashed, 3=dotted, 4=dotdash, 5=longdash, 6=twodash), width `lwd`, ends `lend`, joins `ljoin`/`lmitre`; plot box style `bty` (`"o"`, `"l"`, `"7"`, `"]"`, `"c"`, `"u"`, `"n"`).
- **Colors:** `bg`, `fg`, `col` (+ `.axis`, `.lab`, `.main`, `.sub`). Specify by name (`colors()`), `"#RRGGBB"`, or palette index (`palette()`); generators: `rgb`, `hsv`, `hcl`, `gray`, `rainbow`.
- **Axes:** `lab` (tick counts), `las` (label orientation), `mgp`, tick sizes `tcl`/`tck`, tick positions `xaxp`/`yaxp`, interval style `xaxs`/`yaxs` (`"r"` extends 4%, `"i"` internal), suppress with `xaxt="n"`/`yaxt="n"`; log flags `xlog`/`ylog`.
- **Points:** symbol via `pch` (19 solid circle, 20 bullet, 21 filled circle, 22 square, 23 diamond, 24/25 triangles; see `?points`); symbol size historically `mkh`.
- Others: `ann` (annotation on/off), `ask` (prompt before new page), `new` (overplot pretender), `usr` (user-coordinate extremes), `smo`, `err`.

# Basic Graphics Functions

## The Low-Level Layer

High-level functions are *assembled from* low-level ones — `plot` calls `title`, `plot.new`, `plot.window`, `points`, `lines`, `axis`, `box`; `pie` calls `polygon`, `lines`, `text`, `title`; `boxplot` delegates to `bxp`; and so on. Knowing the low level means you can pass customization *through* the high level, or build charts from scratch.

- **`points(x, y, type="p", ...)`** — add points (useful: `col`, `bg`, `pch`, `cex`, `lwd`); matrix version `matpoints`.
- **`lines(x, y, ...)`** — add connected segments (useful: `lty`, `lwd`, `col`, `lend`, `ljoin`, `lmitre`); matrix version `matlines`.
- **`curve(expr, from, to, n=101, add=, ...)`** — plot an expression in `x` or a function name.

## Straight Lines, Shapes, Segments

**`abline`** draws one line across the plot — by intercept+slope (`a`, `b`), horizontal (`h`), vertical (`v`), regression object (`reg`), or `coef`; `untf` un-transforms on log axes:

```{r,echo=TRUE}
plot(x=c(0, 10), y=c(0, 10))
abline(h=4); abline(v=3)
abline(a=1, b=1); abline(coef=c(10, -1))
polygon(x=c(2, 2, 4, 4), y=c(2, 4, 4, 2))
segments(x0=6, y0=6, x1=8, y1=9)
```

Vector arguments draw many lines at once (`abline(h=1:10, v=1:10)`) — though for a grid, use `grid(nx, ny, col, lty, lwd)`. Shapes: `polygon(x, y, density, angle, border, col, lty)`; rectangles via `rect(xleft, ybottom, xright, ytop, ...)`; point-pair segments via `segments(x0, y0, x1, y1, ...)`; arrows via `arrows`.

## text, mtext, title, legend

- **`text(x, y, labels, adj, pos, offset, vfont, cex, col, font)`** — text at coordinates (`pos`: 1 below, 2 left, 3 above, 4 right).
- **`mtext(text, side=3, line=0, outer=, at=, ...)`** — text in a *margin* (sides 1–4 = bottom/left/top/right).
- **`title(main, sub, xlab, ylab, line=, outer=)`** — annotate after the fact.
- **`legend(x, y, legend, ...)`** — the kitchen sink: `fill` (color boxes), `col`/`lty`/`lwd`/`pch` (sample lines/points), `bty`/`bg`/`box.*` (the legend box), `pt.bg`/`pt.cex`/`pt.lwd`, `xjust`/`yjust`, `x.intersp`/`y.intersp`/`adj`/`text.width`/`text.col`, `merge`, `ncol`/`horiz`, `title`, `inset`, `xpd`.

## axis, box, trans3d

- **`axis(side, at, labels, tick, line, pos, outer, font, lty, lwd, lwd.ticks, col, col.ticks, hadj, padj)`** — draw an axis where and how you want (sides 1–4 as usual).
- **`box(which="plot", lty="solid")`** — frame the `"plot"`, `"figure"`, `"inner"`, or `"outer"` region.
- **`trans3d(x, y, z, pmat)`** — project 3-D points onto a `persp` plot, using the perspective matrix that `persp` returns. The bridge between 3-D surfaces and 2-D annotation.

## Copyright Notice

- **Copyright.** These slides are adapted from *R in a Nutshell: A Desktop Quick Reference* (2nd ed.) by Joseph Adler, O'Reilly Media. All rights reserved by the original author and publisher.

- **Non-commercial use only.** These materials are strictly for educational purposes and may not be used for commercial gain.

- **Attribution.** Any reproduction, distribution, or use of these materials must properly credit the original source.
