---
title: "Programming for Applications"
subtitle: "Chapter 2: The R User Interface"
author: "Yu-You Liou (NTU)"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "R in a Nutshell: A Desktop Quick Reference"
logo: "logo.svg"
editor: source
bibliography: bib.bib
suppress-bibliography: true
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch2-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center")
```

# **Meeting the Interface**

## Why You Are Here

You presumably have a problem you hope R will solve. Typical examples:

- checking the statistical significance of experimental results;
- plotting data to understand it better;
- analyzing genome data — or sales data, survey data, sensor data...

R is a *software environment* for statistical computing and graphics, made of many components: the language, its interpreter, a graphics system, and the desktop application that bundles them. This chapter tours the parts you physically interact with — the **graphical user interface (GUI)** on each platform, and then the most important part of all: the **R console**.

If you have never used an interactive language before, this chapter teaches the survival skills; if you have, it will still tell you where everything lives.

# The R Graphical User Interface

## Launching R for the First Time

- On **Windows** and **macOS**, opening the R application brings up a command window plus menu bars.
- On most **Linux** systems, R simply starts on the command line — no window at all.
- The look differs by platform, but the working parts are the same everywhere: a menu bar, possibly a toolbar with common functions, and — front and center — the R console.

## Windows

- R installs under `%ProgramFiles%\R` (usually `C:\Program Files\R`) with a Start-menu entry in the group *R*.
- Launching it opens the **R GUI window**, which contains a menu bar, a toolbar, and the R console.
- The screenshots in the textbook date from Windows XP — kept in the second edition precisely because R itself had changed so little. The layout today remains essentially identical.

## macOS

- The installer adds an application named **R** to the Applications folder; run it like any other Mac program.
- You get the same trio: menu bar, toolbar with common functions, and an R console window.
- macOS users have a second route: run R **from the terminal** without the GUI. Open Terminal (in *Applications → Utilities*) and type `R` at the prompt.

## Linux and Unix

Start R from the command line by typing:

```{bash,eval=FALSE,echo=TRUE}
$ R
```

- Capital "R" — Linux filenames are case sensitive. (And the `$` is the Unix prompt; don't type it.)
- Unlike Windows/macOS, this starts an interactive session *in the terminal itself*.
- The book's era offered a windowed alternative, `R -g Tk &`, which launched R in its own Tk window (menu bar but no toolbar) running in the background. Modern Linux users wanting a windowed experience generally use RStudio instead.

## How Little the Standard GUI Does

::: {.callout-note}
**A surprise for desktop users:** the standard R GUI implements only rudimentary menu functionality — reading help, managing graphics windows, editing some source and data files. There are no menus, buttons, or palettes for loading data, transforming it, plotting it, or building models. Commercial packages like SAS and SPSS ship far richer interfaces.
:::

All real work in plain R happens by *typing commands*. That is not a defect — it is a design choice with real benefits, as we will see. But it explains why several projects built friendlier front ends...

## Additional R GUIs

Several projects layer an easier interface on top of R:

- **Rcmdr** (R Commander): an R package providing an alternative GUI — buttons for loading data, menu items for many common functions. Install it like any package.
- **RKWard**: a polished front end with palettes, menu-driven analysis, data-editing tools, and an IDE for R code; historically strongest on Linux.
- **RStudio**: the open-source IDE that became the de facto standard — more shortly.
- The book also mentions Revolution Computing's *R Productivity Environment*, a commercial Windows IDE of that era (script editor, object browser, visual debugger).

These slides don't cover the front ends in detail — but everything in this course works in all of them, because they all sit on the same R functions underneath.

# The R Console

## The Most Important Tool

The **console** is where you type commands and watch R respond.

- What you type are **expressions**; the part of R called the **interpreter** reads each expression and answers with a result or an error message.
- If you have used a command line (`cmd.exe`, a Unix shell) or an interactive interpreter such as LISP, this will feel familiar. (R and LISP are genuine relatives: both can compute on the language itself, both use similar internal structures for data, and both love parentheses.)

::: {.callout-tip}
**Why command lines win for data analysis.** After finishing a problem you want a record of *every* step — how the data was loaded, whether a random sample was taken and how, which variables were created, which model parameters were used. A typed transcript gives you exactly that, and lets you re-create the analysis later.
:::

## The Startup Message

Launching R prints a banner like this (modern version shown):

```
R version 4.4.2 (2024-10-31) -- "Pile of Leaves"
Copyright (C) 2024 The R Foundation for Statistical Computing
Platform: x86_64-apple-darwin20

R is free software and comes with ABSOLUTELY NO WARRANTY.
You are welcome to redistribute it under certain conditions.
Type 'license()' or 'licence()' for distribution details.

Type 'demo()' for some demos, 'help()' for on-line help, or
'help.start()' for an HTML browser interface to help.
Type 'q()' to quit R.
```

It tells you: the version you are running, license information, quick reminders for getting help — and then hands you a **command prompt**.

## The Prompt and Your First Expression

When R is waiting for input, it shows a greater-than sign at the start of an empty line: the **prompt** `>`. Type `17 + 3` and press Enter:

```{r,echo=TRUE}
17 + 3
```

Reading the exchange:

- you entered `17 + 3` at the prompt;
- R answered `[1] 20` (the meaning of `[1]` is explained in Chapter 3).

In the textbook's convention, **boldface** marks what the user typed; everything else is R's output. Your own terminal won't bold your input — but it may color it.

## Continuation: the `+` Prompt

When an expression doesn't fit on one line, R notices it is incomplete and switches the prompt to a plus sign (`+`), inviting you to finish:

```{r,echo=TRUE}
1 * 2 * 3 * 4 * 5 *
  6 * 7 * 8 * 9 * 10
```

- In the console the second line appears after a `+` prompt — easy to confuse with addition in long sums or inequalities, so watch for it.
- To help you tell prompt, input, and output apart, most platforms color them differently:

| Platform | Command prompt | User input | R output |
|---|---|---|---|
| macOS | Purple | Blue | Black |
| Windows | Red | Red | Blue |
| Linux | Black | Black | Black |

## Command-Line Editing

R saves you typing with line-editing tools:

- **Up/down arrows** scroll through previous commands — the single most useful trick. Repeat a command with a small change (different parameter, fixed parenthesis) without retyping it.
- `history()` lists previously typed commands.
- The **Tab** key auto-completes function names and filenames, showing a list of possible completions.
- Under the hood, Linux and macOS use the GNU `readline` library, which supports a large set of editing commands; Windows supports a smaller set.

```{r,eval=FALSE,echo=TRUE}
history()    # show recent commands
```

# Batch Mode

## Beyond the Interactive Session

Interactive mode suits ad hoc analysis, but suppose you must repeat the same pipeline — load experiment data, transform it, emit three PDF plots, quit — many times. **Batch mode** runs a whole file of commands in sequence and saves the results.

From the *system* command line (not the R console):

```{bash,eval=FALSE,echo=TRUE}
$ R CMD BATCH generate_graphs.R
```

- R executes everything in `generate_graphs.R` and writes the output to `generate_graphs.Rout`.
- You can name the output file yourself — for instance, stamping it with today's date:

```{bash,eval=FALSE,echo=TRUE}
$ R CMD BATCH generate_graphs.R generate_graphs_`date "+%y%m%d"`.log
```

- Generating graphics in batch mode? Remember to specify the output device and filenames. For all command-line options: `R --help`.

## Rscript and Executable Scripts

`R CMD BATCH` has one notable limitation: the script cannot read the system's standard input. The second batch command, **Rscript**, lifts that restriction:

```{bash,eval=FALSE,echo=TRUE}
$ Rscript generate_graphs.R
```

You can even make R scripts directly executable (Linux/macOS). Create `hello_world.R`:

```{bash,eval=FALSE,echo=TRUE}
#! /usr/bin/env Rscript
print("Hello world!")
```

then mark it executable and run it like any program:

```{bash,eval=FALSE,echo=TRUE}
$ chmod +x hello_world.R
$ ./hello_world.R
[1] "Hello world!"
```

Finally, batch execution is available *from inside R* too: the `source()` function runs the commands in a file within your current session — see `?source`.

# Other Ways to Run R

## R Inside Microsoft Excel

For heavy Excel users on Windows, the **RExcel** software (from the rcom project) let you run R directly inside Excel. The book's installation sequence:

```{r,eval=FALSE,echo=TRUE}
install.packages("RExcelInstaller", "rcom", "rsproxy")
library(rcom)
comRegisterRegistry()       # configure the RCOM server
library(RExcelInstaller)
installstatconnDCOM()       # install RDCOM
installRExcel()             # launch the RExcel installer
```

The workflow: enter values in a column → *Put R Var → Array* names them as an R object (say `v`) → select a blank cell → *Get R Value → Array* and enter an expression such as `(v - mean(v)) / sd(v)` → the rescaled values land back in Excel. You can also call R from Excel formulas and macros, and plot R graphics inside Excel.

::: {.callout-warning}
**Dated technology.** RExcel was Windows-only and is no longer actively maintained. The modern route runs the other way: read and write Excel files *from R* with packages like `readxl` and `openxlsx` (Chapter 11).
:::

## RStudio

**RStudio** is a free, open-source integrated development environment (IDE) for R — and today the most popular way to run it.

- Unlike the standard GUI, RStudio tiles the screen into panes and tabs: console, source editor, environment/history, and files/plots/help.
- It can also be installed **on a Linux server**, letting you use R from a web browser.
- Download: <https://posit.co> (the company formerly named RStudio is now Posit).

We will use RStudio throughout this course: it wraps everything from this chapter — console, history, tab completion, script execution — in one coherent window.

## R as a Server, on the Web, in Emacs

Open-source projects connect R to other applications:

- **As a server — Rserve.** Other programs (e.g., Java) send computations to a running R instance over the network. One beefy Rserve machine, with plenty of CPU and memory, can serve calculations users couldn't run on their own desktops.
- **As a web application — rApache.** Embeds R analyses in a web server, e.g., to publish sophisticated reports rendered with R graphics. (The modern descendant of this idea is **Shiny**.)
- **Inside Emacs — ESS** (*Emacs Speaks Statistics*). An add-on that runs R within the Emacs editor, beloved by long-time statisticians.

Whatever the wrapper, it is the same interpreter underneath — which is why one reference book covers them all.

## Looking Ahead

- You now know where to type (the console), how to read the prompt (`>` vs. `+`), and how to save keystrokes (history, Tab).
- Chapter 3 puts the console to work: a hands-on tutorial of expressions, vectors, functions, data structures, models, and charts.

::: {.callout-tip}
**Try it now.** Open R (or RStudio), type a few arithmetic expressions, press the up arrow, edit, re-run. Five minutes of play makes the next chapter twice as easy.
:::

## Copyright Notice

- **Copyright.** These slides are adapted from *R in a Nutshell: A Desktop Quick Reference* (2nd ed.) by Joseph Adler, O'Reilly Media. All rights reserved by the original author and publisher.

- **Non-commercial use only.** These materials are strictly for educational purposes and may not be used for commercial gain.

- **Attribution.** Any reproduction, distribution, or use of these materials must properly credit the original source.
