---
title: "Programming for Applications"
subtitle: "Chapter 2: The R User Interface"
author: "Yu-You Liou (NTU)"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "R in a Nutshell: A Desktop Quick Reference"
logo: "logo.svg"
editor: source
bibliography: bib.bib
suppress-bibliography: true
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch2-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center")
```

# **Meeting R Face to Face**

## Many Faces of One System

When people say they "use R," they may be interacting with it in quite different ways. This chapter tours the main options:

- the **graphical user interface (GUI)** that ships with R on Windows and macOS;
- the **R console**, the interactive prompt at the heart of every interface;
- **batch mode**, for running scripts without any interaction;
- **RStudio**, the integrated development environment (IDE) we will use in this course;
- a few other front ends you may encounter in the wild.

Whichever wrapper you choose, the engine underneath — the interpreter — is identical.

# The R Graphical User Interface

## Windows

- The Windows build of R installs a program called **RGui**. It opens a window containing a menu bar, a toolbar, and — most importantly — the R console.
- The menus offer conveniences such as loading and saving workspaces, installing packages, and basic script editing.
- The GUI is serviceable but minimal; almost everyone moves to RStudio quickly.

## macOS

- The macOS installer provides **R.app**, a native application with a console window, a simple editor with syntax highlighting, and menus for package management and plotting devices.
- Plots appear in a separate Quartz graphics window.
- As on Windows, the bundled GUI covers the basics, and serious work tends to happen in an IDE.

## Linux and Unix

- On most Linux systems there is no bundled GUI at all: typing `R` in a terminal starts the console directly on the command line.
- Plots open in an X11 window; everything else happens at the prompt.
- This is a perfectly complete way to use R — the GUI elements are conveniences, not requirements.

# The R Console

## The Interactive Prompt

The console is where R actually happens. You type an expression after the `>` prompt; R evaluates it and prints the result:

```{r,echo=TRUE}
17 %% 5
"hello, world"
```

- If a line is syntactically incomplete, the prompt changes to `+` and R waits for the rest.
- Results that are not assigned to a variable are simply printed and discarded.

::: {.callout-note}
**The read–eval–print loop (REPL).** The console endlessly repeats three steps: *read* what you typed, *evaluate* it, and *print* the value. Interactive languages like R make exploration cheap — you can poke at your data one line at a time.
:::

## Command-Line Editing

The console includes line-editing conveniences that save enormous amounts of typing:

- **Up/Down arrows** scroll through your command history; recall a previous command, edit it, and resubmit.
- **Tab completion** finishes function and variable names — type `summ` then *Tab*.
- `history()` displays recent commands.
- On all platforms, *Ctrl+C* (or *Esc*) interrupts a running computation and returns you to the prompt.

## Batch Mode

For repetitive jobs — nightly reports, simulations, grading scripts — you do not want an interactive session at all. R can execute a script from the operating-system shell:

```{bash,eval=FALSE,echo=TRUE}
# classic interface
R CMD BATCH myscript.R output.txt

# modern, more flexible interface
Rscript myscript.R arg1 arg2
```

- `Rscript` prints to standard output and accepts command-line arguments, making it the natural choice for automation and pipelines.
- Everything you can do interactively can be done in a script.

## R Inside Other Applications

- R can be embedded in **Microsoft Excel** through third-party add-ins, letting spreadsheet users call R functions on worksheet ranges.
- Database systems, web servers, and other languages (Python, Java) all have bridges to R.
- The common pattern: another program prepares the data, R does the statistics, and results flow back.

# RStudio

## The IDE We Will Use

**RStudio Desktop** is a free, open-source IDE that has become the de facto standard front end for R. One window organizes four panes:

- **Source editor** — scripts and documents, with completion and highlighting;
- **Console** — the same R prompt as always;
- **Environment / History** — the objects in your workspace;
- **Files / Plots / Packages / Help** — everything else.

```{r,eval=FALSE,echo=TRUE}
# RStudio is an interface to R, not a replacement for it:
# install R first (Chapter 1), then RStudio from
# https://posit.co/download/rstudio-desktop
```

## Why an IDE Matters

- One-keystroke execution of the current line or selection (*Ctrl/Cmd+Enter*) keeps you in flow between editor and console.
- Integrated help, plots, and package management remove constant window-switching.
- Projects keep each analysis in its own folder with its own workspace — a habit that will keep your coursework organized.
- RStudio also builds the course's document formats (R Markdown, Quarto) with a single click.

## Other Ways to Run R

You may also meet:

- **ESS (Emacs Speaks Statistics)** — R inside the Emacs editor; beloved by long-time users.
- **Jupyter notebooks** — R kernels run inside the notebook interface popular in data science.
- **Visual Studio Code** — extensions provide R editing and an integrated terminal.

All of them drive the same interpreter; choose whatever keeps you productive. In this course, demonstrations use RStudio.

## Copyright Notice

- **Copyright.** These slides are adapted from *R in a Nutshell: A Desktop Quick Reference* (2nd ed.) by Joseph Adler, O'Reilly Media. All rights reserved by the original author and publisher.

- **Non-commercial use only.** These materials are strictly for educational purposes and may not be used for commercial gain.

- **Attribution.** Any reproduction, distribution, or use of these materials must properly credit the original source.
