---
title: "Programming for Applications"
subtitle: "Chapter 10: Object-Oriented Programming"
author: "Yu-You Liou (NTU)"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "R in a Nutshell: A Desktop Quick Reference"
logo: "logo.svg"
editor: source
bibliography: bib.bib
suppress-bibliography: true
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch10-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center")
```

# **Objects with Manners**

## Two Systems, One Idea

R is functional at heart, but ships with object-oriented machinery — twice over:

- **S3**: informal, lightweight, ubiquitous. A class is just a `class` attribute; dispatch is by naming convention.
- **S4**: formal classes with declared slots, validity checking, and registered methods [@chambers2008software].

::: {.callout-note}
**Key ideas.** A **class** describes a kind of object; an object is an **instance** of its class. A **generic function** (like `print` or `summary`) decides what to do by **dispatching** to a class-specific **method**. **Inheritance** lets one class extend another.
:::

S3 powers most of base R; S4 powers Bioconductor and code that needs rigor. We look at S4 first, then S3.

## A Running Example

Suppose we track people's body-weight history over time. A natural design:

- a `TimeSeries` class holding values plus start/end timestamps;
- a `WeightHistory` class that **extends** `TimeSeries` with a person's name;
- generic operations — `period()`, `series()` — implemented per class.

This example exercises every S4 feature in the chapter.

## Defining an S4 Class

`setClass()` declares the class name and its **slots** (typed fields):

```{r,echo=TRUE}
library(methods)
setClass("TimeSeries",
  representation(data = "numeric", start = "POSIXct", end = "POSIXct"))

my.ts <- new("TimeSeries",
  data  = c(1, 2, 1, 4, 5),
  start = as.POSIXct("2026-09-22 09:00:00"),
  end   = as.POSIXct("2026-09-22 09:04:00"))
my.ts@data            # slots are accessed with @
```

`new()` constructs instances; printing an S4 object lists its slots.

## Validity Checking

A validity function rejects malformed objects at construction time:

```{r,echo=TRUE,error=TRUE}
setValidity("TimeSeries", function(object) {
  if (object@start <= object@end) TRUE
  else "start must precede end"
})
new("TimeSeries", data = 1:3 + 0,
    start = as.POSIXct("2026-01-02"), end = as.POSIXct("2026-01-01"))
```

This is the major practical advantage of S4 over S3: the system enforces your invariants for you.

## Generic Functions and Methods (S4)

`setGeneric()` declares a generic; `setMethod()` registers an implementation for a class:

```{r,echo=TRUE}
setGeneric("period", function(object) standardGeneric("period"))
setMethod("period", "TimeSeries", function(object) {
  (object@end - object@start) / (length(object@data) - 1)
})
period(my.ts)
```

You can also add methods for *existing* generics — e.g., a `summary` method for `TimeSeries` — and R will dispatch automatically.

## Inheritance (S4)

`contains=` makes one class extend another; the child inherits slots, validity, and methods:

```{r,echo=TRUE}
setClass("WeightHistory",
  contains = "TimeSeries",
  representation(name = "character"))

kim <- new("WeightHistory",
  data  = c(63, 62.5, 62),
  start = as.POSIXct("2026-01-01"), end = as.POSIXct("2026-03-01"),
  name  = "Kim")
period(kim)        # inherited method just works
kim@name
```

A class can inherit from several parents; method dispatch picks the most specific match.

## S3: Classes by Convention

In S3, the "class" is nothing but an attribute, and a method is nothing but a function named `generic.class`:

```{r,echo=TRUE}
cat_record <- list(name = "Mochi", weight = 4.2)
class(cat_record) <- "cat"

print.cat <- function(x, ...) {
  cat("Cat:", x$name, "—", x$weight, "kg\n")
}
print(cat_record)        # print() dispatches to print.cat
```

No declarations, no validation — convention all the way down. `UseMethod()` inside a function makes it a generic.

## How S3 Dispatch Works

```{r,echo=TRUE}
describe <- function(x) UseMethod("describe")
describe.default <- function(x) "some object"
describe.numeric <- function(x) paste("numbers, mean =", mean(x))
describe(letters)
describe(c(1, 2, 3))
```

- The `class()` vector is searched left to right for a matching `generic.class` function; `.default` is the fallback.
- `methods("describe")` lists available methods; `inherits(x, "cat")` tests class membership.

## S3 and S4 Compared

| Aspect | S3 | S4 |
|:---|:---|:---|
| Class definition | `class(x) <- "name"` | `setClass()` with typed slots |
| Field access | `x$field` | `x@slot` |
| Method | function named `generic.class` | `setMethod()` |
| Validation | none | `setValidity()` |
| Cost | minimal | ceremony |

- Use **S3** for everyday work and quick polymorphism; it is what `lm`, `data.frame`, and most of CRAN use.
- Use **S4** when correctness matters more than brevity — large systems, complex hierarchies, Bioconductor.
- Newer systems exist (R5 reference classes, R6, S7), all built on the same attribute machinery.

## Copyright Notice

- **Copyright.** These slides are adapted from *R in a Nutshell: A Desktop Quick Reference* (2nd ed.) by Joseph Adler, O'Reilly Media. All rights reserved by the original author and publisher.

- **Non-commercial use only.** These materials are strictly for educational purposes and may not be used for commercial gain.

- **Attribution.** Any reproduction, distribution, or use of these materials must properly credit the original source.
