---
title: "Programming for Applications"
subtitle: "Chapter 10: Object-Oriented Programming"
author: "Yu-You Liou (NTU)"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "R in a Nutshell: A Desktop Quick Reference"
logo: "logo.svg"
editor: source
bibliography: bib.bib
suppress-bibliography: true
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch10-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center")
```

# **Two Object Systems**

## A Functional Language with OOP Support

At heart R is a **functional** language — and writing functional programs is fine practice. But R also supports **object-oriented programming (OOP)**, the dominant paradigm of Java, C#, Ruby, C++... Many R packages are built with R objects, including the core `stats` package, `lattice`, and `ggplot2`. OOP shines when representing complicated things: statistical models, charts.

Two mechanisms coexist, both inherited from S:

- **S3** (~1990): class attributes enabling single-argument methods. The classic modeling software uses S3, so it is everywhere.
- **S4**: formal classes and methods — multiple dispatch, abstract types, sophisticated inheritance. Used by many newer packages, extensively in Bioconductor.

The book's advice: build *new* abstractions with **S4**; learn **S3** because you must read and extend the vast code that uses it.

## Key Ideas, via Time Series

A time series — equally spaced measurements over time, with a start, end, frequency — makes every OOP concept concrete:

- a **class** is the formal definition; each actual time series is an **instance**; a function operating on the class is a **method**;
- hiding the storage details (data frame? vector? text?) behind methods is **encapsulation**;
- a weight-history class that *reuses* the time-series machinery, adding only what differs, exhibits **inheritance** — time series is the **superclass**, weight history the **subclass**;
- one name (`period`) doing the right thing for different classes is **polymorphism**;
- building a class out of several component classes is **composition** — and inheriting from more than one class is **multiple inheritance**, which R permits.

(Yes, R already has a time-series class — `ts`, an S3 class in `stats`. That is deliberate: we will rebuild it the S4 way, then dissect the S3 original.)

# An S4 Worked Example

## setClass: Defining TimeSeries

Slots are where an object stores information. We represent a time series by three slots — the data, a start time, an end time (units, frequency, and period are derivable):

```{r,echo=TRUE}
setClass("TimeSeries",
  representation(
    data="numeric",
    start="POSIXct",
    end="POSIXct"
  )
)
```

The `representation` names the class of object each slot holds.

## new: Creating an Instance

`new` is the generic constructor: first argument the class name, the rest fill the slots. The default S4 print method shows every slot:

```{r,echo=TRUE}
my.TimeSeries <- new("TimeSeries",
  data=c(1, 2, 3, 4, 5, 6),
  start=as.POSIXct("07/01/2009 0:00:00", tz="GMT",
                   format="%m/%d/%Y %H:%M:%S"),
  end=as.POSIXct("07/01/2009 0:05:00", tz="GMT",
                 format="%m/%d/%Y %H:%M:%S"))
my.TimeSeries
```

## setValidity: Rejecting Bad Objects

Not all slot values make sense: `end` must not precede `start`, and both must have length 1. Register a validity function with `setValidity` and check with `validObject`:

```{r,echo=TRUE}
setValidity("TimeSeries",
  function(object) {
    object@start <= object@end &&
    length(object@start) == 1 &&
    length(object@end) == 1
  }
)
validObject(my.TimeSeries)
```

(You can also supply the validity function when calling `setClass` — see its full definition.)

## Validation in Action

From now on, `new` checks each candidate and rejects the invalid:

```{r,error=TRUE,echo=TRUE}
good.TimeSeries <- new("TimeSeries",
  data=c(7, 8, 9, 10, 11, 12),
  start=as.POSIXct("07/01/2009 0:06:00", tz="GMT",
                   format="%m/%d/%Y %H:%M:%S"),
  end=as.POSIXct("07/01/2009 0:11:00", tz="GMT",
                 format="%m/%d/%Y %H:%M:%S"))
bad.TimeSeries <- new("TimeSeries",
  data=c(7, 8, 9, 10, 11, 12),
  start=as.POSIXct("07/01/2009 0:06:00", tz="GMT",
                   format="%m/%d/%Y %H:%M:%S"),
  end=as.POSIXct("07/01/1999 0:11:00", tz="GMT",
                 format="%m/%d/%Y %H:%M:%S"))   # ends before it starts!
```

## A First Method: period

A plain function that computes the period from the slots:

```{r,echo=TRUE}
period.TimeSeries <- function(object) {
  if (length(object@data) > 1) {
    (object@end - object@start) / (length(object@data) - 1)
  } else {
    Inf
  }
}
```

The `@` operator reaches into slots — the implementer's view. Users should get nicer verbs; enter generic functions.

## setGeneric: series

To extract the data series from *any* suitable object — polymorphism — define a function and promote it to a **generic**; the old body becomes the default method:

```{r,echo=TRUE}
series <- function(object) { object@data }
setGeneric("series")
series(my.TimeSeries)
showMethods("series")
```

(In OOP terms this is *overloading* the function name.)

## setMethod: period for TimeSeries

Create a generic `period`, then register `period.TimeSeries` as its method for the `TimeSeries` signature:

```{r,echo=TRUE}
period <- function(object) { object@period }
setGeneric("period")
setMethod(period, signature=c("TimeSeries"), definition=period.TimeSeries)
period(my.TimeSeries)
```

Calling the generic dispatches to the right definition by class.

## Methods for Existing Generics — Even Operators

You can attach methods to generics that already exist, like `summary` — or to **operators**:

```{r,echo=TRUE}
setMethod("summary", signature="TimeSeries",
  definition=function(object, ...) {
    print(paste(object@start, " to ", object@end, sep="", collapse=""))
    print(paste(object@data, sep="", collapse=","))
  }
)
summary(my.TimeSeries)
setMethod("[", signature=c("TimeSeries"),
  definition=function(x, i, j, ..., drop) { x@data[i] }
)
my.TimeSeries[3]
```

::: {.callout-note}
This works for only *some* built-ins — `print`, notably, cannot be redefined this way. `?S4groupGeneric` lists the redefinable generics; the S3 section below explains why.
:::

## Inheritance: WeightHistory

A weight history is a time series plus a person's name and height. `contains` declares the superclass; the new class adds only its extra slots:

```{r,echo=TRUE}
setClass("WeightHistory",
  representation(height = "numeric", name = "character"),
  contains = "TimeSeries")
john.doe <- new("WeightHistory",
  data=c(170, 169, 171, 168, 170, 169),
  start=as.POSIXct("02/14/2009 0:00:00", tz="GMT",
                   format="%m/%d/%Y %H:%M:%S"),
  end=as.POSIXct("03/28/2009 0:00:00", tz="GMT",
                 format="%m/%d/%Y %H:%M:%S"),
  height=72, name="John Doe")
```

R validates the embedded TimeSeries automatically (try an invalid one yourself).

## Multiple Inheritance: Person + TimeSeries

A cleaner composition: define a `Person` class, then inherit from **both**:

```{r,echo=TRUE}
setClass("Person",
  representation(height = "numeric", name = "character"))
setClass("AltWeightHistory",
  contains = c("TimeSeries", "Person"))
```

Identical behavior, slightly cleaner design — and `AltWeightHistory` now inherits methods from both parents.

## Virtual Classes: setClassUnion

Add a `Cat` class — cats have names too. To write one `is.fluffy` method covering people *and* cats, create a **virtual class** above both with `setClassUnion`:

```{r,echo=TRUE}
setClass("Cat",
  representation(breed = "character", name = "character"))
setClassUnion("NamedThing", c("Person", "Cat"))
```

Methods can target `NamedThing`; objects cannot be created from it (its representation is ambiguous). A subclass method (say, `is.fluffy` for `Person`) would override the parent's. Bonus — class membership is testable:

```{r,echo=TRUE}
jane.doe <- new("AltWeightHistory",
  data=c(130, 129, 131, 128, 130, 129),
  start=as.POSIXct("02/14/2009 0:00:00", tz="GMT",
                   format="%m/%d/%Y %H:%M:%S"),
  end=as.POSIXct("03/28/2009 0:00:00", tz="GMT",
                 format="%m/%d/%Y %H:%M:%S"),
  height=67, name="Jane Doe")
is(jane.doe, "NamedThing")
is(john.doe, "TimeSeries")
```

# S4 in Depth

## setClass: the Full Signature

```{r,eval=FALSE,echo=TRUE}
setClass(Class, representation, prototype, contains=character(),
         validity, access, where, version, sealed, package,
         S3methods = FALSE)
```

| Argument | Description |
|---|---|
| `Class` | name of the new class (the only required argument) |
| `representation` | named list of slots and their classes (`"ANY"` allows anything) |
| `prototype` | object holding default slot values |
| `contains` | names of superclasses this class extends |
| `validity` | validity-checking function (default: none; changeable via `setValidity`) |
| `where` | environment to store the definition (default: where `setClass` was called) |
| `sealed` | may the class be redefined by another `setClass`? |
| `package` | package name for the class |
| `S3methods` | may S3 methods be written for this class? (default `FALSE`) |
| `access`, `version` | unused; S-PLUS compatibility |

Helpers `representation()` and `prototype()` (in `methods`) ease definitions that extend a basic type, have several superclasses, or mix both.

## Slot Rules, .Data, and Friends

- **Forbidden slot names** — reserved for attributes: `class`, `comment`, `dim`, `dimnames`, `names`, `row.names`, `tsp`. (Objects can have both slots *and* attributes.)
- Extending a **basic type** (the table later in this chapter) yields a `.Data` slot holding the basic part — code written for the built-in type keeps working, acting on `.Data`.
- `setIs(class1, class2, ...)` declares inheritance explicitly — an alternative to `contains`.
- On creation, R runs the class's **`initialize`** method if defined — typically to compute derived slots.
- `setClassUnion(name, members, where)` creates the virtual superclass seen earlier.

## New Objects and Slot Access

- `new(c, ...)` — the **constructor**: named arguments fill slots; an `initialize` method (if any) runs afterward.
- Reading slots: `object@slotname`, or equivalently `slot(object, "slotname")`.
- Writing slots uses ordinary assignment, with validity checked by default:

```{r,eval=FALSE,echo=TRUE}
birthdate@month <- "June"
slot(birthdate, "month") <- "June"
slot(birthdate, "month", check=FALSE) <- "June"   # skip validation: usually unwise
```

## Working with Objects, and Coercion

Object inspection vocabulary:

- `is(o, c)` — is object `o` a member of class `c`? `extends(c1, c2)` — does class `c1` extend `c2`?
- `slotNames(o)` / `slotNames(c)` — slot names of an object / class; `getSlots(c)` — the *classes* of each slot (nonintuitively).

Conversion: `as(o, c)` coerces object `o` to class `c`. For your own classes, register coercion methods with **`setAs`**:

| Argument | Description |
|---|---|
| `from`, `to` | class names of input and output |
| `def` | function performing the conversion |
| `replace` | function for the replacement form (when `as` is assigned into) |
| `where` | environment for the definition |

## setGeneric and setMethod: Full Signatures

`setGeneric(name, def=, group=list(), valueClass=character(), where=, package=, signature=, useAsDefault=, ...)` — key arguments:

- `name` / `def` — name and optional defining function; `group` — group generic (see `?S4groupGeneric`);
- `valueClass` — class(es) the return value must belong to; `signature` — formal-argument names and classes (`"ANY"` = anything);
- `useAsDefault` — which function becomes the default method.

`setMethod(f, signature=character(), definition, where=, valueClass=NULL, sealed=FALSE)`:

- `f` — generic (or its name); `signature` — argument classes to match; `definition` — the function to call; `sealed` — prevent redefinition.

## Managing Generics and Methods

Bookkeeping functions from the `methods` package:

| Function | Description |
|---|---|
| `isGeneric` / `isGroup` | does a generic / group generic of this name exist? |
| `removeGeneric` / `removeMethods` | remove a generic with its methods / just the methods |
| `getMethod`, `selectMethod` | fetch the method for a function + signature |
| `existsMethod`, `hasMethod` | does such a method exist? |
| `findMethod` | which package(s) contain it? |
| `showMethods` | display all methods of an S4 generic |
| `dumpMethod` / `dumpMethods` | dump one / all methods to code |
| `findFunction` | where on the search list a function is defined |

More help: `library(help="methods")`.

## Basic Classes

Classes of the built-in types — everything is built on these, and you may write methods for them:

| Category | Object type | Class |
|---|---|---|
| Vectors | `integer` / `double` / `complex` / `character` / `logical` / `raw` | `integer` / `numeric` / `complex` / `character` / `logical` / `raw` |
| Compound | `list` / `pairlist` / `environment` | `list` / `pairlist` / `environment` |
| Special | `NULL` | `NULL` |
| Language | `symbol` / `language` / `expression` | `name` / `call` / `expression` |
| Functions | `closure` / `special` / `builtin` | `function` |
| Other | `externalptr` | `externalptr` |

The six vector classes all extend the virtual class `vector`.

# Old-School OOP: S3

## S3 Classes Are Just Attributes

An S3 object is a primitive object wearing extra attributes — one of them named `class`. No formal definition exists; you can change the attributes by hand. (S3 resembles prototype-based languages like JavaScript.) Dissecting the built-in `ts` class:

```{r,echo=TRUE}
my.ts <- ts(data=c(1, 2, 3, 4, 5), start=c(2009, 2), frequency=12)
my.ts
attributes(my.ts)
typeof(my.ts)
unclass(my.ts)
```

A `ts` is a double vector plus `class` and `tsp` attributes. And S4 tools don't apply:

```{r,error=TRUE,echo=TRUE}
my.ts@tsp
```

S3 inheritance is informal — a class attribute with several elements means the first is the class, the rest are "inherited" classes, making inheritance a property of *objects*, not classes. Encapsulation is unenforced; parametric polymorphism is absent. Simple polymorphism, however, works — via S3 generics.

## S3 Methods: Naming Convention + UseMethod

S3 generics dispatch **by naming convention**, not registration:

1. pick a generic name, say `gname`;
2. define `gname` with body `UseMethod("gname")`;
3. for each class, define `gname.classname` whose first argument is such an object.

The real thing:

```{r,echo=TRUE}
plot
```

`UseMethod` walks the object's class vector looking for `plot.class`; failing everything, it tries `plot.default`. (`NextMethod`, callable inside a method, hands off to the next available method.)

## Adding an S3 Method to Our S4 Class

Give `TimeSeries` a plot method by *merely naming a function correctly*:

```{r,echo=TRUE}
plot.TimeSeries <- function(x, ...) {
  plot(x@data, ...)
}
plot(my.TimeSeries)
```

`plot(my.TimeSeries)` quietly becomes `plot.TimeSeries(my.TimeSeries)`.

## S3 Inside S4: setOldClass

S4 slots cannot hold an S3 class directly — first wrap it with `setOldClass(Classes, prototype, where, test=FALSE, S4Class)`:

- `Classes` — name(s) of the old-style class;
- `prototype` — default object for the S4 class;
- `test` — set `TRUE` if there can be multiple inheritance;
- `S4Class` — an S4 class definition to base the new class on.

## Finding Hidden S3 Methods

Package authors may **hide** individual methods to push you toward the generic — encapsulation by obscurity:

```{r,error=TRUE,echo=TRUE}
library(lattice)
methods(histogram)
histogram.factor()
```

To read the hidden code anyway:

```{r,eval=FALSE,echo=TRUE}
getS3method(f="histogram", class="formula")
getAnywhere(histogram.formula)
```

::: {.callout-tip}
**Modern footnote.** Today's ecosystem adds further systems — `R6` for mutable reference objects and `S7` as the planned successor unifying S3/S4. The concepts you just learned (classes, methods, dispatch, inheritance) transfer directly.
:::

## Copyright Notice

- **Copyright.** These slides are adapted from *R in a Nutshell: A Desktop Quick Reference* (2nd ed.) by Joseph Adler, O'Reilly Media. All rights reserved by the original author and publisher.

- **Non-commercial use only.** These materials are strictly for educational purposes and may not be used for commercial gain.

- **Attribution.** Any reproduction, distribution, or use of these materials must properly credit the original source.
