---
title: "Programming for Applications"
subtitle: "Chapter 15: ggplot2"
author: "Yu-You Liou (NTU)"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "R in a Nutshell: A Desktop Quick Reference"
logo: "logo.svg"
editor: source
bibliography: bib.bib
suppress-bibliography: true
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch15-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center")
library(ggplot2)
```

# **A Grammar of Graphics**

## The Third System

`ggplot2` [@wickham2016ggplot2] implements Wilkinson's **grammar of graphics** [@wilkinson2011grammar]: instead of choosing from a menu of chart types, you *compose* a chart from independent parts —

- **data** and **aesthetic mappings** (which variable drives position, color, size, shape);
- **geoms** (points, lines, bars — the visible marks);
- **stats** (transformations like binning or smoothing);
- **scales**, **coordinates**, **facets**, and **themes**.

This is the system our companion textbook (and the `da` course) is devoted to; here we meet its essentials.

```{r,echo=TRUE,cache=TRUE}
temp <- tempfile(fileext = ".rda")
download.file("https://raw.githubusercontent.com/cran/nutshell/master/data/batting.2008.rda", temp, mode = "wb")
load(temp)
hitters <- subset(batting.2008, AB > 100)
```

## Anatomy of a ggplot Call

```{r,echo=TRUE,out.width="62%"}
ggplot(hitters, aes(x = AB, y = H)) +
  geom_point(alpha = 0.4)
```

- `ggplot(data, aes(...))` establishes the data and mappings;
- `+ geom_*()` adds a layer that draws them;
- further `+` terms refine scales, labels, facets, and themes.

## Aesthetic Mappings

Map a third variable to color and the legend writes itself:

```{r,echo=TRUE,out.width="62%"}
ggplot(hitters, aes(x = AB, y = H, color = HR)) +
  geom_point() +
  labs(x = "At-bats", y = "Hits", color = "Home runs")
```

::: {.callout-note}
**Mapping vs. setting.** `aes(color = HR)` *maps* data to color (legend, scale); `geom_point(color = "red")` *sets* a constant. Mixing these up is the classic ggplot beginner error.
:::

## Histograms and Densities

Each geom implies a default stat — `geom_histogram()` bins for you:

```{r,echo=TRUE,out.width="62%",message=FALSE}
ggplot(hitters, aes(x = H / AB)) +
  geom_histogram(bins = 30)
```

```{r,echo=TRUE,out.width="62%"}
ggplot(hitters, aes(x = H / AB)) +
  geom_density(fill = "steelblue", alpha = 0.3)
```

## Smoothing Layers

Layers stack — raw data underneath, a model on top:

```{r,echo=TRUE,out.width="65%",message=FALSE}
ggplot(hitters, aes(x = AB, y = H)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm")
```

`geom_smooth()` defaults to loess with a confidence ribbon; `method = "lm"` fits a straight line.

## Bars and Boxes

```{r,echo=TRUE,out.width="65%"}
team.hr <- aggregate(HR ~ teamID, data = batting.2008, sum)
top8 <- head(team.hr[order(-team.hr$HR), ], 8)
ggplot(top8, aes(x = reorder(teamID, -HR), y = HR)) +
  geom_col() +
  labs(x = NULL, y = "Team home runs (top 8)")
```

`geom_boxplot()`, `geom_violin()`, and `geom_jitter()` cover the distribution-by-group cases of Chapter 13 in grammar form.

## Faceting

The lattice trick, grammar style:

```{r,echo=TRUE,out.width="70%",message=FALSE}
ggplot(hitters, aes(x = AB, y = H)) +
  geom_point(alpha = 0.3) +
  facet_wrap(~ bats) +
  labs(title = "Hits vs. at-bats by batting side")
```

`facet_wrap(~ g)` wraps panels into a grid; `facet_grid(rows ~ cols)` crosses two factors.

## Scales and Themes

Every mapping has a scale you may replace; themes control the non-data ink:

```{r,echo=TRUE,out.width="65%"}
ggplot(hitters, aes(x = AB, y = H, color = HR)) +
  geom_point() +
  scale_color_gradient(low = "grey80", high = "red") +
  scale_x_log10() +
  theme_minimal(base_size = 13) +
  labs(title = "Hits vs. at-bats", subtitle = "color = home runs")
```

`theme_bw()`, `theme_classic()`, and the `theme()` function (for surgical adjustments) round out the toolkit; `ggsave("fig.pdf")` writes the last plot to disk.

## Three Systems Side by Side

| | Base | Lattice | ggplot2 |
|:---|:---|:---|:---|
| Model | ink on paper | conditioned panels | grammar of layers |
| Composition | sequential calls | one formula | `+` pipeline |
| Legends | manual | mostly automatic | automatic |
| Extensibility | low-level | panel functions | new geoms/stats |

All three live happily in one analysis. For this course: base for quick looks, lattice for conditioning, `ggplot2` for publication figures — and the dedicated Data Analysis course for everything deeper.

## Copyright Notice

- **Copyright.** These slides are adapted from *R in a Nutshell: A Desktop Quick Reference* (2nd ed.) by Joseph Adler, O'Reilly Media. All rights reserved by the original author and publisher.

- **Non-commercial use only.** These materials are strictly for educational purposes and may not be used for commercial gain.

- **Attribution.** Any reproduction, distribution, or use of these materials must properly credit the original source.
