---
title: "Programming for Applications"
subtitle: "Chapter 15: ggplot2"
author: "Yu-You Liou (NTU)"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "R in a Nutshell: A Desktop Quick Reference"
logo: "logo.svg"
editor: source
bibliography: bib.bib
suppress-bibliography: true
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch15-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center", fig.height = 4, message = FALSE, warning = FALSE)
library(ggplot2)
```

# **The Grammar of Graphics**

## A Short Introduction

Hadley Wickham's **ggplot2** became one of R's most popular packages — not just for pretty output, but because it speaks a *language* for describing plots: the **grammar of graphics**. You describe **what** you want to show, not **how** to draw it. The toy data (lattice's, with renamed variables):

```{r,echo=TRUE}
d <- data.frame(a=c(0:9), b=c(1:10), c=c(rep(c("Odd", "Even"), times=5)))
qplot(x=a, y=b, data=d)
```

We specified an x, a y, and a data set — never the words "scatter plot." (`qplot` = quick plot; deprecated in recent ggplot2 but still working — the full `ggplot()` style follows shortly.)

## Plot Objects

Like lattice, ggplot2 functions **return an object**; printing draws it. Inside functions and scripts: `print` it yourself.

```{r,echo=TRUE}
first.ggplot2.example <- qplot(x=a, y=b, data=d)
summary(first.ggplot2.example)
```

The summary lays the plot's anatomy bare: the data, the **mapping** (x = a, y = b), the faceting, and the layer (`geom_point` + `stat_identity` + identity position).

## Facets and Colors

Panels (lattice's conditioning) are **facets**, specified by formula; group-by-color is just another mapping:

```{r,echo=TRUE}
qplot(x=a, y=b, data=d, facets=~c)     # panels left-right
qplot(x=a, y=b, data=d, facets=c~.)    # panels top-bottom — easy to flip!
qplot(x=a, y=b, data=d, color=c)       # color instead of panels
```

(Hadley's preferred spelling of the first: `qplot(x=a, y=b, data=d) + facet_wrap(~c)`.)

## One Variable: Histogram and Density

With only `x` supplied, `qplot` defaults to a histogram; ask for another **geom** by name:

```{r,echo=TRUE}
set.seed(123456789)
e <- data.frame(f=rnorm(1000))
qplot(x=f, data=e)
qplot(x=f, data=e, geom="density")
```

## The Grammar's Components

Drawing a chart means doing many things at once; the grammar names each one, and ggplot2's functions mirror the names:

- **Data** — what is visualized;
- **Mappings** — variables → chart components;
- **Geometric objects (geom)** — points, bars, lines... (`geom_point`, `geom_bar`, ...);
- **Aesthetic properties (aes)** — how it looks: sizes, labels, ticks;
- **Scales** — how variables map to aesthetics;
- **Coordinates** — Cartesian (`coord_cartesian`), polar (`coord_polar`), map projections (`coord_map`);
- **Statistical transformations (stat)** — summaries: `stat_bin` for histograms, `stat_boxplot`...;
- **Facets** — partitioning into subplots;
- **Positional adjustments** — fine placement (stack, dodge, jitter...).

`summary` of any plot object reports every component — your map when tuning a chart.

# A More Complex Example: Medicare Data

## The Data

Real, complicated data: US **Medicare** cost and outcome measures, packaged three ways — `outcome.of.care.measures.national` (national mortality/readmission rates for heart attack, heart failure, pneumonia), `medicare.payments` (average payment per hospital × 70 conditions; 140,722 records), `medicare.payments.by.state`:

```{r,echo=TRUE}
temp <- tempfile(fileext = ".rda")
download.file("https://raw.githubusercontent.com/cran/nutshell/master/data/outcome.of.care.measures.national.rda",
              temp, mode = "wb")
load(temp)
outcome.of.care.measures.national
```

## Bar Charts: Facets...

One categorical x, so a univariate plot with bar heights from `weight=Rate`; facet by Measure, fill by Measure:

```{r,echo=TRUE}
qplot(x=Condition, data=outcome.of.care.measures.national,
      geom="bar", weight=Rate, facets=Measure~., fill=Measure)
```

## ...or Dodging

Drop the facets, set `position="dodge"` to place bars side by side. The book wrote `qplot(..., position="dodge")`, but modern ggplot2 has removed `qplot`'s `position` argument — position adjustments now belong to the *layer*, so we write it the full-grammar way:

```{r,echo=TRUE}
ggplot(outcome.of.care.measures.national,
       aes(x=Condition, weight=Rate, fill=Measure)) +
  geom_bar(position="dodge")
```

Both work — *differently*: faceting invites comparing conditions within each measure; dodging invites comparing measures within each condition. The display is an argument.

## Payments vs. Volume

Question: do hospitals treating more cases charge Medicare more or less? Restrict to the three heart-failure diagnosis groups and plot payment vs. case count, colored by diagnosis:

```{r,echo=TRUE}
temp <- tempfile(fileext = ".rda")
download.file("https://raw.githubusercontent.com/cran/nutshell/master/data/medicare.payments.rda",
              temp, mode = "wb")
load(temp)
heart.failure <- c("Heart failure and shock w/o CC/MCC",
                   "Heart failure and shock w MCC",
                   "Heart failure and shock w CC")
payment.plot <- qplot(x=Number.Of.Cases, y=Medicare.Average.Payment,
  data=subset(medicare.payments, Diagnosis.Related.Group %in% heart.failure),
  color=Diagnosis.Related.Group)
payment.plot
```

Unreadable — everything clumps left. (Small counts are censored by HIPAA.)

## Fixing the Plot by Adding Layers

Recipe: log x-scale, semi-opaque points (`alpha=I(1/10)`), a smoothing line, trimmed y. The elegant route **adds components with `+`**:

```{r,echo=TRUE}
payment.plot.alpha <- qplot(x=Number.Of.Cases, y=Medicare.Average.Payment,
  data=subset(medicare.payments, Diagnosis.Related.Group %in% heart.failure),
  color=Diagnosis.Related.Group, alpha=I(1/10), ylim=c(0, 20000))
payment.plot.scaled <- payment.plot.alpha + scale_x_log10() + geom_smooth()
payment.plot.scaled
```

`summary` of this object now shows **two** geom/stat/position triples — one per layer — with the alpha aesthetic threaded through.

## Reordering Factors for Dot Plots

Costs rise with volume — geography, perhaps? State-level data, with `State` levels **reordered by payment** (sorting the *levels*, not the rows) before a dot plot:

```{r,echo=TRUE,fig.height=5}
temp <- tempfile(fileext = ".rda")
download.file("https://raw.githubusercontent.com/cran/nutshell/master/data/medicare.payments.by.state.rda",
              temp, mode = "wb")
load(temp)
medicare.payments.by.state.hf <- subset(medicare.payments.by.state,
  Diagnosis.Related.Group %in% heart.failure)
medicare.payments.by.state.hf$State <- with(medicare.payments.by.state.hf,
  reorder(State, Medicare.Average.Payment.Maximum, mean))
qplot(x=Medicare.Average.Payment.Maximum, y=State,
      data=medicare.payments.by.state.hf,
      color=Diagnosis.Related.Group)
```

Top of the list: Northern Mariana Islands, Alaska, Virgin Islands — remote and expensive — then NY, MD, CA: high cost of living *and* big hospitals. Cheapest: Puerto Rico, American Samoa. Cost and volume are both correlated **with geography**.

## A Choropleth Map

Are adjacent states similar? Merge payments with state polygons (`maps` package) and fill by payment — `geom="polygon"` plus a repositioned legend:

```{r,echo=TRUE}
library(maps)
states <- map_data("state")
state.name.map <- data.frame(abb=state.abb, region=tolower(state.name),
                             stringsAsFactors=FALSE)
states <- merge(states, state.name.map, by="region")
toplot <- merge(states, medicare.payments.by.state,
                by.x="abb", by.y="State")
toplot <- toplot[order(toplot$order), ]
qplot(long, lat,
      data=subset(toplot,
        Diagnosis.Related.Group=="Heart failure and shock w/o CC/MCC"),
      group=group, fill=Medicare.Average.Payment.Maximum,
      geom="polygon") +
  theme(legend.position="bottom", legend.direction="vertical")
```

(The book wrote `opts(...)` — long since replaced by `theme()`.)

# Quick Plot and the Full Grammar

## qplot: Arguments

```{r,eval=FALSE,echo=TRUE}
qplot(x, y, ..., data, facets, margins, geom, stat,
      position, xlim, ylim, log, main, xlab, ylab, asp)
```

| Argument | Description | Default |
|---|---|---|
| `x`, `y`, `data` | values and (optional) data frame | |
| `facets` | facet formula: one-sided → `facet_wrap`, two-sided → `facet_grid` | `NULL` |
| `geom` | geoms as character vector | `"auto"`: point (x,y) / histogram (x) |
| `stat` / `position` | statistics / position adjustments | `list(NULL)` |
| `xlim`, `ylim`, `log` | limits; `log` = `""`/`"x"`/`"y"`/`"xy"` | |
| `main`, `xlab`, `ylab`, `asp` | titles, labels, aspect | deparsed expressions |
| `...` | other aesthetics passed to layers | |

## ggplot + Layers: the Verbose, Flexible Way

Beyond `qplot`: create the object with `ggplot(data, mapping=aes(...))`, then **add layers and options with `+`**:

```{r,echo=TRUE}
plt <- ggplot(data=d, mapping=aes(x=a, y=b)) + geom_point()
summary(plt)
```

- `aes(x, y, ...)` declares the mappings (relatives: `aes_string`, `aes_auto` in the book's era).
- A bare `ggplot(...)` object has no layers — printing it errors ("No layers in plot").
- The generic `layer(...)` adds a layer by geom short-name (`layer("point")`); in practice the `geom_*` convenience functions do this.

## The geom Vocabulary

| | | |
|---|---|---|
| `geom_abline` line by slope+intercept | `geom_errorbar(h)` error bars | `geom_polygon` polygons |
| `geom_area` area plot | `geom_freqpoly` frequency polygon | `geom_quantile` quantile-regression lines |
| `geom_bar` bars | `geom_hex` hexagonal binning | `geom_rect` rectangles |
| `geom_bin2d` 2-D bin heatmap | `geom_histogram` histogram | `geom_ribbon` ribbons (à la Minard) |
| `geom_blank` nothing | `geom_hline` / `geom_vline` h/v lines | `geom_rug` rug |
| `geom_boxplot` box plots | `geom_jitter` jittered points | `geom_segment` segments |
| `geom_contour` contours | `geom_line` / `geom_path` lines/paths | `geom_smooth` smoothed mean |
| `geom_crossbar` boxless boxes | `geom_linerange` vertical intervals | `geom_step` steps |
| `geom_density(2d)` densities | `geom_point` / `geom_pointrange` | `geom_text` text |
| | | `geom_tile` tiles |

## stats, scales, coords, facets, positions

- **stat_**: `bin`, `bin2d`, `binhex`, `boxplot`, `contour`, `density`, `density2d`, `function`, `identity`, `qq`, `quantile`, `smooth`, `spoke`, `sum`, `summary`, `unique`, plus line helpers `abline`/`hline`/`vline` (modern ggplot2 folds these into geoms).
- **scale_**: `alpha`, `brewer` (ColorBrewer), `continuous`, `date`/`datetime`, `discrete`, `gradient`/`gradient2`/`gradientn`, `grey`, `hue`, `identity`, `linetype`, `manual`, `shape`, `size` (today spelled e.g. `scale_color_brewer`, `scale_x_continuous`).
- **coord_**: `cartesian`, `equal`, `flip`, `map`, `polar`, `trans`.
- **facet_**: `facet_grid` (grid layout) and `facet_wrap` (wrapped 1-D list).
- **position_**: `dodge`, `fill`, `identity`, `jitter`, `stack`.

## Learning More

- Hadley Wickham's book *ggplot2: Elegant Graphics for Data Analysis* — the definitive treatment.
- The official documentation site (now <https://ggplot2.tidyverse.org>) and the *R Graphics Cookbook*.

::: {.callout-tip}
**Course note on API drift.** Since the book: `opts()` → `theme()`; `qplot` is soft-deprecated in favor of `ggplot() + geom_*()`; scales are written `scale_<aesthetic>_<type>`. The grammar itself — data, mapping, geom, stat, scale, coord, facet, position — is unchanged, which is exactly the point of having a grammar.
:::

## Copyright Notice

- **Copyright.** These slides are adapted from *R in a Nutshell: A Desktop Quick Reference* (2nd ed.) by Joseph Adler, O'Reilly Media. All rights reserved by the original author and publisher.

- **Non-commercial use only.** These materials are strictly for educational purposes and may not be used for commercial gain.

- **Attribution.** Any reproduction, distribution, or use of these materials must properly credit the original source.
