---
title: "Programming for Applications"
subtitle: "Chapter 3: A Short R Tutorial"
author: "Yu-You Liou (NTU)"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "R in a Nutshell: A Desktop Quick Reference"
logo: "logo.svg"
editor: source
bibliography: bib.bib
suppress-bibliography: true
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch3-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center")
```

# **Learning by Typing**

## How to Use This Chapter

This chapter is a hands-on tour of R, packed with examples. The best way to learn is to **start R and type along**:

- enter exactly what the slides show — or change it slightly and see what happens (the book's suggestion: if the example says `3 + 4`, try `3 - 4`);
- everything happens at the console, so if terms like *prompt* or *expression* feel foreign, revisit Chapter 2 first.

By the end you will have touched every major idea in the course: vectors, functions, variables, data structures, classes, models, charts, and the help system.

# Basic Operations in R

## Expressions and Evaluation

Type an expression at the console, press Enter, and R **evaluates** it and prints the result (if there is one). Simple math works exactly as expected, including operator precedence and parentheses:

```{r,echo=TRUE}
1 + 2 + 3
1 + 2 * 3
(1 + 2) * 3
```

The interactive interpreter automatically prints whatever object an expression returns — that is why you see the answers without asking for them.

## The Mysterious `[1]`: Everything Is a Vector

Why does every answer start with `[1]`? Because **any number you type is a vector** — an ordered collection of numbers. The bracketed number is the *index of the first element shown on that row*. Each result above is a vector with a single element, hence `[1]`.

Longer vectors are built with the `c(...)` function (*c* for **combine**):

```{r,echo=TRUE}
c(0, 1, 1, 2, 3, 5, 8)   # first seven Fibonacci numbers
```

## Indices Across Multiple Lines

The **sequence operator** `:` produces consecutive integers — handy for seeing how row labels work when output wraps:

```{r,echo=TRUE}
1:50
```

Each row begins with the index of its first element (e.g., `[23]`, `[45]`), so you can locate any element at a glance.

## Vector Arithmetic Is Pairwise

Operations on two vectors are matched **element by element**, returning a new vector:

```{r,echo=TRUE}
c(1, 2, 3, 4) + c(10, 20, 30, 40)
c(1, 2, 3, 4) * c(10, 20, 30, 40)
c(1, 2, 3, 4) - c(1, 1, 1, 1)
```

This is the single most characteristic habit of R: you rarely loop over elements — you operate on whole vectors at once.

## Recycling: Unequal Lengths

When the vectors differ in length, R **recycles** the shorter one, repeating it as many times as needed:

```{r,echo=TRUE,warning=TRUE}
c(1, 2, 3, 4) + 1
1 / c(1, 2, 3, 4, 5)
c(1, 2, 3, 4) + c(10, 100)
c(1, 2, 3, 4, 5) + c(10, 100)
```

::: {.callout-warning}
**Watch the warning.** If the longer length is not a multiple of the shorter, R still computes an answer but warns you — usually a sign of a mistake in your code.
:::

## Character Vectors

Expressions can involve text too:

```{r,echo=TRUE}
"Hello world."
c("Hello world", "Hello R interpreter")
```

- A quoted string is a **character vector** of length 1; combining two gives length 2.
- Terminology alert for C programmers: in C, a "character" is a single letter and an ordered set is a "string"; in R, what C calls a string is a single **character value**.

## Comments

Anything after a pound sign (`#`) on a line is ignored — even mid-expression:

```{r,echo=TRUE}
# a comment at the beginning of a line
1 + 2 + # and one in the middle
  3
```

Comment generously: your future self is the most frequent reader of your code.

# Functions

## Calling Functions

In R, the operations that do the work are **functions** — just like math class. Most calls take the form `f(argument1, argument2, ...)`:

```{r,echo=TRUE}
exp(1)
cos(3.141593)
log2(1)
```

Each of these takes a single argument. Many functions take several.

## Named and Positional Arguments

Arguments can be supplied **by name**, or **by position** when given in the default order:

```{r,echo=TRUE}
log(x = 64, base = 4)
log(64, 4)
```

Both calls are identical. Named arguments shine when a function has many parameters and you want to set only a few.

## Operators Are Functions Too

Not every function looks like `f(...)`. Some appear as **operators**:

```{r,echo=TRUE}
17 + 2
2 ^ 10
3 == 4
```

- `+` is ordinary addition; `^` is exponentiation (note: *not* commutative); `==` tests equality and returns a **Boolean** (`TRUE`/`FALSE`) — yes, R has a logical data type.
- Under the hood, the interpreter translates every operator into an equivalent function call — a fact we exploit in Chapter 5.

# Variables

## Assignment: `x <- 1`, Read "x gets 1"

R lets you attach names to values. The assignment operator is `<-`, pronounced **"gets"** — pleasingly close to algorithm pseudocode:

```{r,echo=TRUE}
x <- 1
y <- 2
z <- c(x, y)
z
```

## Substitution Happens at Assignment Time

The value is substituted **when the assignment is made**, not when the variable is later evaluated. Change `y` after building `z`, and `z` is unmoved:

```{r,echo=TRUE}
y <- 4
z
```

The subtleties of how variables are evaluated (environments, lazy evaluation) wait until Chapter 8.

## Indexing a Vector by Position

R offers several ways to refer to members of a vector:

```{r,echo=TRUE}
b <- c(1,2,3,4,5,6,7,8,9,10,11,12)
b[7]          # the 7th item
b[1:6]        # items 1 through 6
b[c(1,6,11)]  # items 1, 6, and 11
b[c(8,4,9)]   # out of order: returned in the order referenced
```

The index vector can be any integer vector — contiguous, scattered, even shuffled.

## Indexing with a Logical Vector

You can also select elements with a vector of `TRUE`/`FALSE` values. Example: keep only multiples of 3, i.e., elements congruent to 0 (mod 3):

```{r,echo=TRUE}
b %% 3 == 0          # the logical mask
b[b %% 3 == 0]       # apply it
```

This *filter by condition* idiom is the workhorse of data analysis in R — you will use it constantly.

## Two More Assignment Operators

- **`=` also assigns**, left to right, like most languages. Use it if you prefer — but the book (and these slides) stick with `<-` for readability. Never confuse `=` (assign) with `==` (test equality):

```{r,echo=TRUE}
one <- 1; two <- 2
one = two    # assigns! one is now 2
one
one <- 1; two <- 2
one == two   # compares
```

- **`->` assigns to the right** — occasionally handy when you typed a long expression and only then remembered to save it:

```{r,echo=TRUE}
3 -> three
three
```

::: {.callout-warning}
Inside a function call, `f(x <- 3)` assigns `x` in your workspace *and then* passes the value — almost never what you intend. Use `=` to bind arguments: `f(x = 3)`.
:::

## Functions Are Objects with Names

A function is just another object bound to a symbol. Define your own and call it like a built-in:

```{r,echo=TRUE}
f <- function(x, y) { c(x + 1, y + 1) }
f(1, 2)
```

A useful trick follows: typing the *name alone* (no parentheses) prints the function's source code:

```{r,echo=TRUE}
f
```

This works on most built-in functions too — R is open source all the way down.

# Introduction to Data Structures

## Arrays

An **array** is a multidimensional vector. Internally, vectors and arrays are stored identically — an array is simply a vector wearing a dimension attribute — but arrays display and index differently:

```{r,echo=TRUE}
a <- array(c(1,2,3,4,5,6,7,8,9,10,11,12), dim = c(3, 4))
a
a[2, 2]      # row 2, column 2
```

Compare the same contents as a plain vector:

```{r,echo=TRUE}
v <- c(1,2,3,4,5,6,7,8,9,10,11,12)
v
```

## Matrices and Higher Dimensions

A **matrix** is just a two-dimensional array:

```{r,echo=TRUE}
m <- matrix(data = c(1,2,3,4,5,6,7,8,9,10,11,12), nrow = 3, ncol = 4)
m
```

Arrays may have more than two dimensions:

```{r,echo=TRUE}
w <- array(1:18, dim = c(3, 3, 2))
w[1, 1, 1]
```

(Printed in full, `w` appears as two 3×3 slices labeled `, , 1` and `, , 2`.)

## Slicing Arrays

R's indexing syntax for parts of an array is clean: one index per dimension, separated by commas — and **omitting an index selects everything** in that dimension:

```{r,echo=TRUE}
a[1, 2]        # single cell
a[1:2, 1:2]    # submatrix
a[1, ]         # first row only
a[, 1]         # first column only
a[1:2, ]       # a range of rows
a[c(1, 3), ]   # even a noncontiguous set of rows
```

## Lists: Mixing Types

Everything so far held a single data type. **Lists** hold a heterogeneous collection of objects, with optionally named components — subtly different from "lists" in other languages:

```{r,echo=TRUE}
e <- list(thing = "hat", size = "8.25")
e
```

Items are accessible by name or by position — note the three distinct forms:

```{r,echo=TRUE}
e$thing    # by name: the element itself
e[1]       # single bracket: a sublist
e[[1]]     # double bracket: the element itself
```

## Lists Inside Lists

A list can even contain other lists:

```{r,echo=TRUE}
g <- list("this list references another list", e)
g
```

This nesting ability is what lets R represent arbitrarily complex objects — fitted models, for instance, are big nested lists.

## Data Frames

A **data frame** is a list of named vectors, all the same length — think *spreadsheet* or *database table*. It is the natural container for experimental data. The book's example: 2008 win/loss records of the National League East:

```{r,echo=TRUE}
teams <- c("PHI","NYM","FLA","ATL","WSN")
w <- c(92, 89, 94, 72, 59)
l <- c(70, 73, 77, 90, 102)
nleast <- data.frame(teams, w, l)
nleast
```

## Picking Values Out of a Data Frame

Columns are reached with the `$` operator (just like list components):

```{r,echo=TRUE}
nleast$w
```

To find a specific value — say, losses by the Florida Marlins — build a logical vector and use it as a filter:

```{r,echo=TRUE}
nleast$teams == "FLA"
nleast$l[nleast$teams == "FLA"]
```

Chapter 11 shows how to import data frames from files and databases; beyond lists, R also offers formal class definitions via **S4 objects** for heterogeneous data.

# Objects and Classes

## Every Object Has a Class

R is an object-oriented language: every object has a type and belongs to a **class**. We have met several classes already — query them with `class()`:

```{r,echo=TRUE}
class(teams)
class(w)
class(nleast)
class(class)
```

Note the last line: a *function* is an object of class `function`.

## Methods and Generic Functions

- Functions tied to a specific class are called **methods**. (R's class system is far less formal than Java's — not every function belongs to a class.)
- Methods for different classes may share one name: such names are **generic functions**. They buy you two things:
  1. easy guessing — the function name you know probably works on the new class;
  2. code reuse — one piece of code can handle many types.

Example: `+` is generic. It adds numbers, but also knows what to do with a date and a number:

```{r,echo=TRUE}
17 + 6
as.Date("2009-09-08") + 7
```

## The Hidden `print` Call

When you evaluate an expression at the console, the interpreter silently calls the generic function `print` on the result:

```{r,echo=TRUE}
x <- 1 + 2 + 3 + 4
x          # actually runs print(x)
```

- Define a new class, define a `print` method for it, and you control how its objects display.
- Some packages exploit this: **lattice** plotting functions *return* objects without drawing; the console's automatic `print` is what plots them. Call a lattice function inside another function or script and nothing appears unless you `print` it explicitly — a classic gotcha.

Objects return in depth in Chapter 7; classes in Chapter 10.

# Models and Formulas

## What Is a Model?

To statisticians, a **model** concisely describes a set of data, usually via a mathematical formula. Two typical goals:

- **predictive** — train on data, predict new values;
- **descriptive** — understand the data better.

For a linear model predicting $y$ from $x_1, x_2, \ldots, x_n$ (the **dependent** and **independent** variables):

$$y = c_0 + c_1 x_1 + c_2 x_2 + \cdots + c_n x_n + \varepsilon$$

R expresses the relationship as `y ~ x1 + x2 + ... + xn` — a **formula object**.

## Fitting a Linear Model: `lm`

The `cars` data set (in the base distribution, collected in the 1920s) records car speeds and stopping distances. Assume stopping distance is a linear function of speed; the formula is `dist ~ speed`, and `lm` estimates the parameters:

```{r,echo=TRUE}
cars.lm <- lm(formula = dist ~ speed, data = cars)
cars.lm
```

Printing an `lm` object shows the original call (so you can see data and formula) and the estimated coefficients.

## More Detail: `summary`

```{r,echo=TRUE}
summary(cars.lm)
```

The summary reports the call, the residual distribution, the coefficient table (estimates, standard errors, $t$-values, $p$-values), and fit statistics ($R^2$, $F$-statistic).

## To Name or Not to Name?

You may also fit-and-inspect in one breath, without saving the model:

```{r,eval=FALSE,echo=TRUE}
lm(dist ~ speed, data = cars)
summary(lm(dist ~ speed, data = cars))
```

Convenient — but naming the model is usually wiser:

- follow-up analyses (residual plots, extra statistics, adding/dropping variables) need the object;
- refitting a complex model on big data is slow; a saved object spares you;
- named objects make code easier to read and modify.

# Charts and Graphics

## The Graphics Toolbox

R ships several visualization systems: **graphics** (the classic base system), **grid**, and **lattice** — with `graphics` and `lattice` the most used in the book's era (and **ggplot2**, Chapter 15, dominating today).

- Everything Excel can draw — column, bar, line, pie, scatter — R draws too, with far better automation and customization.
- And then hundreds of chart types Excel never dreamed of.

To make this concrete, the book uses real data: every field goal attempted in the NFL in 2005. (A field goal: kicking the ball between the goalposts for 3 points; a miss hands possession to the other team at the spot of the kick.)

## Loading the Example Data

The book loads data with `library(nutshell)`; that package has left CRAN, so this course fetches the same data sets from the CRAN GitHub mirror:

```{r,echo=TRUE}
temp <- tempfile(fileext = ".rda")
download.file("https://raw.githubusercontent.com/cran/nutshell/master/data/field.goals.rda",
              temp, mode = "wb")
load(temp)
names(field.goals)
```

`names()` lists the columns: home and away team, week, quarter, offense, defense, play type, yards, stadium type, player.

## A First Histogram

The `hist` function shows a distribution in one line:

```{r,echo=TRUE}
hist(field.goals$yards)
```

(Your colors may differ from the book's — the author tweaked graphical parameters for print.)

## Tuning the Histogram: `breaks`

Wanting more detail at different distances, increase the number of bins via the `breaks` argument:

```{r,echo=TRUE}
hist(field.goals$yards, breaks = 35)
```

A single argument transforms the picture — typical of base graphics: sensible defaults, endless knobs.

## Tabulating and Strip Charts

How many kicks were blocked? The `table` function tabulates a categorical variable:

```{r,echo=TRUE}
table(field.goals$play.type)
```

A **strip chart** plots one point per observation along an axis. Select only blocked kicks, jitter the points so they don't overprint, and restyle them with `pch`:

```{r,echo=TRUE}
stripchart(field.goals[field.goals$play.type == "FG blocked", ]$yards,
           pch = 19, method = "jitter")
```

## Scatterplots: the `cars` Data Again

The `cars` data set has 50 observations of speed and stopping distance:

```{r,echo=TRUE}
dim(cars)
names(cars)
summary(cars)
```

Plot the relationship, with labeled axes:

```{r,echo=TRUE}
plot(cars, xlab = "Speed (mph)", ylab = "Stopping distance (ft)",
     las = 1, xlim = c(0, 25))
```

At a glance: stopping distance grows roughly in proportion to speed.

## Lattice Graphics and Factors

**Lattice** excels at comparing groups. It is not loaded by default — calling a lattice function first requires:

```{r,echo=TRUE}
library(lattice)
```

The example data: American food consumption, 1980–2005, in the `consumption` data set (48 observations: `Amount` consumed, `Food` type, `Year`, in 5-year steps).

```{r,echo=TRUE}
temp <- tempfile(fileext = ".rda")
download.file("https://raw.githubusercontent.com/cran/nutshell/master/data/consumption.rda",
              temp, mode = "wb")
load(temp)
```

Two columns are numeric (`Amount`, `Year`); two are a new type: **factors** — R's compact representation of categorical values, created with `factor()` and central to modeling functions.

## A Conditioned Dot Plot

We want `Amount` against `Year`, drawn *separately for each* `Food` — lattice expresses this with the formula `Amount ~ Year | Food`:

```{r,eval=FALSE,echo=TRUE}
dotplot(Amount ~ Year | Food, consumption)   # default settings
```

The book's author found the defaults hard to read (labels too big, shared scales, awkward stacking) and tuned two options:

```{r,echo=TRUE}
dotplot(Amount ~ Year | Food, data = consumption,
        aspect = "xy", scales = list(relation = "sliced", cex = .4))
```

- `aspect` adjusts panel aspect ratios so changes bank toward 45° (easiest to perceive);
- `scales` controls axis drawing — `"sliced"` gives each panel its own slice of the range.

Chapter 14 treats lattice in full.

# Getting Help

## The Help System

R documents every function in every installed package:

```{r,eval=FALSE,echo=TRUE}
help(glm)      # full help page for glm
?glm           # exactly equivalent shorthand
?`+`           # operators need backquotes
example(glm)   # auto-run the examples from a help file
```

## Searching for Help

Can't remember the function's name? Search by topic:

```{r,eval=FALSE,echo=TRUE}
help.search("regression")   # list relevant topics
??regression                # shorthand for the same
```

For a whole package's documentation, the `help` option of `library` is more complete than a single page:

```{r,eval=FALSE,echo=TRUE}
library(help = "grDevices")
```

## Vignettes

Some packages — especially from Bioconductor — include **vignettes**: short documents that walk through using the package, examples included.

```{r,eval=FALSE,echo=TRUE}
vignette("affy")        # view a specific vignette (if installed)
vignette(all = FALSE)   # vignettes in attached packages
vignette(all = TRUE)    # vignettes in all installed packages
```

::: {.callout-tip}
**Habit to build now:** when a function surprises you, read `?fun` *before* searching the web — R's built-in documentation is unusually good, and the examples actually run.
:::

## Copyright Notice

- **Copyright.** These slides are adapted from *R in a Nutshell: A Desktop Quick Reference* (2nd ed.) by Joseph Adler, O'Reilly Media. All rights reserved by the original author and publisher.

- **Non-commercial use only.** These materials are strictly for educational purposes and may not be used for commercial gain.

- **Attribution.** Any reproduction, distribution, or use of these materials must properly credit the original source.
               