---
title: "Programming for Applications"
subtitle: "Chapter 3: A Short R Tutorial"
author: "Yu-You Liou (NTU)"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "R in a Nutshell: A Desktop Quick Reference"
logo: "logo.svg"
editor: source
bibliography: bib.bib
suppress-bibliography: true
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch3-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center")
```

# **Your First Half Hour in R**

## Basic Operations

Type an expression at the prompt; R evaluates it and prints the value. Arithmetic works exactly as you would hope, with the usual precedence rules:

```{r,echo=TRUE}
1 + 2 + 3
1 + 2 * 3
(1 + 2) * 3
```

The `[1]` in front of each answer is a hint about R's deepest design decision: **every number is a vector**. The bracketed number is the index of the first element printed on that row — here each result is a vector of length one.

## Vectors

A vector is an ordered collection of values. Build one with `c()` ("combine"), or generate integer sequences with the colon operator:

```{r,echo=TRUE}
c(0, 1, 1, 2, 3, 5, 8)
1:50
```

With a long vector, the leading `[23]`-style labels show the index of the first element of each printed row.

## Vector Arithmetic

Operations between two vectors are applied **element by element**:

```{r,echo=TRUE}
c(1, 2, 3, 4) + c(10, 20, 30, 40)
c(1, 2, 3, 4) * c(10, 20, 30, 40)
```

If the lengths differ, the shorter vector is **recycled** — repeated until it covers the longer one:

```{r,echo=TRUE,warning=TRUE}
c(1, 2, 3, 4) + 1
c(1, 2, 3, 4) + c(10, 100)
c(1, 2, 3, 4, 5) + c(10, 100)
```

R warns you when the longer length is not a multiple of the shorter — usually the sign of a bug.

## Character Vectors and Comments

Text values are **character vectors**; what other languages call a "string" is a single character value in R:

```{r,echo=TRUE}
"Hello world."
c("Hello world", "Hello R interpreter")
```

Anything after a `#` on a line is a comment and is ignored by the interpreter:

```{r,echo=TRUE}
1 + 2 + # comments can even sit mid-expression
  3
```

## Functions

Work in R is done by **functions**, called in the familiar mathematical form `f(argument1, argument2, ...)`:

```{r,echo=TRUE}
exp(1)
cos(pi)
log2(1)
```

Arguments can be matched **by name** or **by position**:

```{r,echo=TRUE}
log(x = 64, base = 4)
log(64, 4)
```

::: {.callout-note}
**Operators are functions too.** Symbols like `+`, `^`, and `==` are translated into function calls by the interpreter. `2 ^ 10` and `"^"(2, 10)` are the same computation.
:::

```{r,echo=TRUE}
2 ^ 10
3 == 4
```

## Variables

Assign a value to a name with the arrow operator `<-`, then reuse it:

```{r,echo=TRUE}
x <- 1
y <- c(2, 4, 6)
x + y
```

- Names are case-sensitive: `x` and `X` are different variables.
- Variables can hold any object — vectors, functions, model results, entire data sets.
- `ls()` lists the variables currently defined; `rm(x)` removes one.
- `=` also assigns at the top level, but `<-` is the universal R convention and the one we use.

# Data Structures

## Arrays and Matrices

An **array** is a vector that carries a `dim` attribute describing its shape; a **matrix** is the two-dimensional special case:

```{r,echo=TRUE}
a <- array(1:12, dim = c(3, 4))
a
m <- matrix(1:12, nrow = 3, ncol = 4)
a[2, 2]
```

Arrays may have three or more dimensions:

```{r,echo=TRUE}
w <- array(1:18, dim = c(3, 3, 2))
w[1, 1, 2]
```

## Indexing

R's subscript notation is concise: one index per dimension, separated by commas, with an empty slot meaning "everything":

```{r,echo=TRUE}
a[1, ]          # first row
a[, 1]          # first column
a[1:2, 1:2]     # a sub-block
a[c(1, 3), ]    # noncontiguous rows
```

## Lists

Vectors and arrays hold a single type. A **list** can mix types freely, with optionally named components reachable by name or position:

```{r,echo=TRUE}
e <- list(thing = "hat", size = 8.25)
e$thing
e[[2]]
```

Lists are the workhorse container of R: model results, parsed files, and most complex objects are lists underneath.

## Data Frames

A **data frame** is a list of equal-length vectors displayed as a table — the standard representation of a data set, with one row per observation and one column per variable:

```{r,echo=TRUE}
teams <- data.frame(
  team   = c("PHI", "NYM", "FLA", "ATL", "WSN"),
  wins   = c(92, 88, 84, 72, 59),
  losses = c(70, 74, 78, 90, 102)
)
teams
teams$wins
```

## Objects and Classes

Every value in R is an **object**, and every object has a **class** that describes what it is:

```{r,echo=TRUE}
class(c(1, 2, 3))
class("hello")
class(teams)
class(sin)
```

Generic functions such as `print()` and `summary()` inspect the class of their argument and dispatch to an appropriate method — which is why `summary()` of a vector and `summary()` of a regression look completely different. Chapter 10 examines this machinery in depth.

# Models and Graphics

## Models and Formulas

Statistical relationships are written in R's **formula notation**: `y ~ x1 + x2` reads "model `y` as a function of `x1` and `x2`." For a linear model

$$y = c_0 + c_1 x_1 + c_2 x_2 + \cdots + c_n x_n + \varepsilon,$$

we fit with `lm()`. The classic `cars` data set (1920s stopping distances) is built in:

```{r,echo=TRUE}
cars.lm <- lm(formula = dist ~ speed, data = cars)
cars.lm
```

## Inspecting a Model

Printing shows only the call and coefficients; `summary()` reports residuals, standard errors, $t$-statistics, $p$-values, and $R^2$:

```{r,echo=TRUE}
summary(cars.lm)
```

The fitted object is itself data — you can extract coefficients, residuals, and predictions from it. Part V of the course is devoted to models like this.

## Loading the Course Data Sets

The textbook ships its example data in a `nutshell` package and loads it with `library(nutshell); data(...)`. That package is no longer on CRAN, so **this course uses a simpler convention**: download each data set from the CRAN GitHub mirror to a temporary file, then load it.

```{r,echo=TRUE}
# book style (do not use):  library(nutshell); data(field.goals)
# course style:
temp <- tempfile(fileext = ".rda")
download.file("https://raw.githubusercontent.com/cran/nutshell/master/data/field.goals.rda", temp, mode = "wb")
load(temp)
names(field.goals)
```

::: {.callout-tip}
**Pattern to remember.** Replace `xxx` with the data set name:
`temp <- tempfile(fileext = ".rda")`
`download.file("https://raw.githubusercontent.com/cran/nutshell/master/data/xxx.rda", temp, mode = "wb")`
`load(temp)`
:::

## Charts and Graphics

The `field.goals` data records every NFL field-goal attempt in 2005 (a kick worth 3 points; `yards` is the attempt distance). A histogram takes one line:

```{r,echo=TRUE,out.width="70%"}
hist(field.goals$yards)
```

## Refining a Chart

More bins reveal more structure — most attempts cluster between 20 and 50 yards:

```{r,echo=TRUE,out.width="70%"}
hist(field.goals$yards, breaks = 35)
```

## Tables and Strip Charts

`table()` tabulates a categorical variable; a strip chart draws one point per observation — here, the distances of blocked kicks:

```{r,echo=TRUE,out.width="65%"}
table(field.goals$play.type)
stripchart(field.goals[field.goals$play.type == "FG blocked", "yards"],
           pch = 19, method = "jitter")
```

## Getting Help

Three tools answer most questions without leaving R:

```{r,eval=FALSE,echo=TRUE}
?hist              # help page for a known function
help.search("regression")   # search installed documentation
example(hist)      # run a function's examples
```

- Help pages document every argument and end with runnable examples.
- `??topic` is shorthand for `help.search("topic")`.
- Online: the R manuals on CRAN, and the enormous archive of answered questions on Stack Overflow.

## Copyright Notice

- **Copyright.** These slides are adapted from *R in a Nutshell: A Desktop Quick Reference* (2nd ed.) by Joseph Adler, O'Reilly Media. All rights reserved by the original author and publisher.

- **Non-commercial use only.** These materials are strictly for educational purposes and may not be used for commercial gain.

- **Attribution.** Any reproduction, distribution, or use of these materials must properly credit the original source.
