---
title: "Programming for Applications"
subtitle: "Chapter 17: Probability Distributions"
author: "Yu-You Liou (NTU)"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "R in a Nutshell: A Desktop Quick Reference"
logo: "logo.svg"
editor: source
bibliography: bib.bib
suppress-bibliography: true
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch17-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center")
```

# **The d-p-q-r System**

## Four Letters, Every Distribution

For each distribution `xxx`, R supplies four functions:

| Prefix | Function | Question answered |
|:---|:---|:------|
| `d` | density / mass | how likely is the value $x$? |
| `p` | cumulative distribution | $P(X \le q)$? |
| `q` | quantile | which value has probability $p$ below it? |
| `r` | random generation | give me $n$ draws |

```{r,echo=TRUE}
dnorm(0)              # standard normal density at 0
pnorm(1.96)           # P(Z <= 1.96)
qnorm(0.975)          # the famous 1.96, recovered
set.seed(1); rnorm(3) # three random normals
```

## The Normal Distribution

$$f(x) = \frac{1}{\sigma\sqrt{2\pi}}\, e^{-\frac{(x-\mu)^2}{2\sigma^2}}$$

```{r,echo=TRUE,out.width="65%"}
x <- seq(-4, 4, length.out = 200)
plot(x, dnorm(x), type = "l", lwd = 2, ylab = "density")
lines(x, dnorm(x, sd = 1.5), lty = 2)
legend("topright", c("sd = 1", "sd = 1.5"), lty = 1:2)
```

`mean=` and `sd=` shift and stretch; defaults give the standard normal.

## Visualizing p and q

The shaded area *is* `pnorm()`; `qnorm()` runs the lookup backwards:

```{r,echo=TRUE,out.width="65%"}
x <- seq(-4, 4, length.out = 200)
plot(x, dnorm(x), type = "l", lwd = 2, ylab = "density")
xs <- seq(-4, qnorm(0.975), length.out = 120)
polygon(c(xs, rev(xs)), c(dnorm(xs), rep(0, 120)),
        col = "lightsteelblue", border = NA)
abline(v = qnorm(0.975), lty = 2)
text(2.6, 0.1, "97.5%")
```

## Continuous Families

The same four-letter pattern covers every classic continuous family:

```{r,echo=TRUE}
punif(0.5)                  # uniform on [0,1]
pexp(1, rate = 1)           # exponential
pt(2.086, df = 20)          # Student's t
pchisq(3.84, df = 1)        # chi-squared
pf(4.35, df1 = 2, df2 = 20) # F
```

- `t`, `chisq`, and `f` are the workhorses of Chapter 18's tests — their `q` functions generate the critical-value tables you once looked up in appendices.
- Also available: `lnorm`, `gamma`, `beta`, `weibull`, `cauchy`, and more.

## Discrete Families

```{r,echo=TRUE}
dbinom(7, size = 10, prob = 0.5)   # exactly 7 heads in 10 tosses
pbinom(7, size = 10, prob = 0.5)   # at most 7
dpois(3, lambda = 2)               # Poisson: 3 events when mean is 2
dgeom(2, prob = 0.2)               # first success on 3rd trial
dhyper(2, m = 4, n = 8, k = 3)     # sampling without replacement
```

```{r,echo=TRUE,out.width="60%"}
barplot(dbinom(0:10, 10, 0.5), names.arg = 0:10,
        main = "Binomial(10, 0.5)")
```

## Random Numbers and Reproducibility

```{r,echo=TRUE}
set.seed(2026)
rbinom(5, size = 10, prob = 0.3)
sample(c("H", "T"), 5, replace = TRUE)
```

::: {.callout-tip}
**Always `set.seed()` in homework and reports.** Pseudo-random sequences restart identically from a seed, so your graders — and your future self — can reproduce every simulation exactly.
:::

## Simulation as a Tool

When theory gets hard, simulate. What is the chance that the maximum of five dice is 6?

```{r,echo=TRUE}
set.seed(7)
sims <- replicate(10000, max(sample(1:6, 5, replace = TRUE)))
mean(sims == 6)
1 - (5/6)^5          # exact answer, for comparison
```

`replicate()` plus the `r` functions turn probability puzzles into ten-second experiments — the same pattern that powers the bootstrap (Chapter 16) and power analysis (Chapter 19).

## Copyright Notice

- **Copyright.** These slides are adapted from *R in a Nutshell: A Desktop Quick Reference* (2nd ed.) by Joseph Adler, O'Reilly Media. All rights reserved by the original author and publisher.

- **Non-commercial use only.** These materials are strictly for educational purposes and may not be used for commercial gain.

- **Attribution.** Any reproduction, distribution, or use of these materials must properly credit the original source.
