---
title: "Programming for Applications"
subtitle: "Chapter 17: Probability Distributions"
author: "Yu-You Liou (NTU)"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "R in a Nutshell: A Desktop Quick Reference"
logo: "logo.svg"
editor: source
bibliography: bib.bib
suppress-bibliography: true
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch17-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center", fig.height = 4.2)
```

# **d, p, q, r**

## Why Distributions Matter Here

Many statistical tests compute a test statistic, then compare it to a value from a **theoretical distribution** — exactly what Chapter 18 will do. R supplies, for every common distribution, functions to calculate **densities**, **distribution functions** (cumulative probabilities), **quantiles**, and **random draws**. This chapter walks through the normal distribution in detail, then catalogs the whole family.

# The Normal Distribution

## dnorm: the Density

$$f(x) = \frac{1}{\sigma\sqrt{2\pi}}\, e^{-\frac{(x-\mu)^2}{2\sigma^2}}$$

`dnorm(x, mean=0, sd=1, log=FALSE)` evaluates the density at `x` (`log=TRUE` returns its logarithm). Plotting it is one line:

```{r,echo=TRUE}
plot(dnorm, -3, 3, main = "Normal Distribution")
```

## pnorm: the Distribution Function

`pnorm(q, mean=0, sd=1, lower.tail=TRUE, log.p=FALSE)` returns $p = \Pr(x \le q)$ — the probability that a random value falls at or below `q`. `lower.tail=FALSE` flips to $\Pr(x > q)$; `log.p=TRUE` returns $\log(p)$:

```{r,echo=TRUE}
pnorm(0)          # symmetric around 0, so exactly .5
pnorm(-1)         # below one sd under the mean
pnorm(1.96) - pnorm(-1.96)   # within 1.96 sd: the famous 95%
plot(pnorm, -3, 3, main = "Cumulative Normal Distribution")
```

## qnorm: the Quantile Function

`qnorm(p, mean=0, sd=1, lower.tail=TRUE, log.p=FALSE)` is `pnorm`'s inverse: it returns the `q` for which $p = \Pr(x \le q)$:

```{r,echo=TRUE}
qnorm(0.5)                     # the median
qnorm(log(0.5), log.p=TRUE)
qnorm(pnorm(-1))               # inverse, demonstrated
c(qnorm(.025), qnorm(.975))    # both ends of a 95% confidence interval
```

## rnorm: Random Draws

`rnorm(n, mean=0, sd=1)` generates `n` random values — for testing functions, simulations, sampling methods:

```{r,echo=TRUE}
hist(rnorm(10000), breaks=50)
```

# The Naming System

## Four Prefixes, Shared Arguments

Nearly every distribution follows one convention:

- **`d`** — probability density function (for discrete distributions, technically the probability *mass* function);
- **`p`** — distribution function;
- **`q`** — quantile function;
- **`r`** — random number generator.

And shared arguments by type: density functions take `x`, `log`; distribution functions take `q`, `lower.tail`, `log.p`; quantile functions take `p`, `lower.tail`, `log.p`; random generators take `n` (renamed `nn` for the hypergeometric, whose parameters claimed the letter). Knowing the system, you can usually *guess* the function you need.

# Distribution Function Families

## Continuous Distributions

| Family | R functions | Family arguments |
|---|---|---|
| Beta | `dbeta` `pbeta` `qbeta` `rbeta` | `shape1`, `shape2`, `ncp=0` |
| Cauchy | `dcauchy` `pcauchy` `qcauchy` `rcauchy` | `location`, `scale` |
| Chi-squared | `dchisq` `pchisq` `qchisq` `rchisq` | `df`, `ncp=0` |
| Exponential | `dexp` `pexp` `qexp` `rexp` | `rate` |
| F | `df` `pf` `qf` `rf` | `df1`, `df2`, `ncp` |
| Gamma | `dgamma` `pgamma` `qgamma` `rgamma` | `shape`, `rate=1`, `scale=1/rate` |
| Log-normal | `dlnorm` `plnorm` `qlnorm` `rlnorm` | `meanlog`, `sdlog` |
| Logistic | `dlogis` `plogis` `qlogis` `rlogis` | `location`, `scale` |
| Normal | `dnorm` `pnorm` `qnorm` `rnorm` | `mean`, `sd` |
| Student's t | `dt` `pt` `qt` `rt` | `df`, `ncp` |
| Uniform | `dunif` `punif` `qunif` `runif` | `min`, `max` |
| Weibull | `dweibull` `pweibull` `qweibull` `rweibull` | `shape`, `scale` |

## Discrete and Special Distributions

| Family | R functions | Family arguments |
|---|---|---|
| Binomial | `dbinom` `pbinom` `qbinom` `rbinom` | `size`, `prob` |
| Birthday (coincidences) | `pbirthday` `qbirthday` | `classes`, `coincident` |
| Geometric | `dgeom` `pgeom` `qgeom` `rgeom` | `prob` |
| Hypergeometric | `dhyper` `phyper` `qhyper` `rhyper` | `m`, `n`, `k` (random's count is `nn`!) |
| Multinomial | `dmultinom` `rmultinom` | `size`, `prob` |
| Negative binomial | `dnbinom` `pnbinom` `qnbinom` `rnbinom` | `size`, `prob`, `mu` |
| Poisson | `dpois` `ppois` `qpois` `rpois` | `lambda` |
| Studentized range | `ptukey` `qtukey` | `nmeans`, `df`, `nranges` |
| Wilcoxon rank sum | `dwilcox` `pwilcox` `qwilcox` `rwilcox` | `m`, `n` |
| Wilcoxon signed rank | `dsignrank` `psignrank` `qsignrank` `rsignrank` | `n` |

```{r,echo=TRUE}
dbinom(3, size=10, prob=0.5)    # exactly 3 heads in 10 fair flips
ppois(2, lambda=4)              # at most 2 events when 4 are expected
qbirthday(0.5)                  # people needed for a 50% shared birthday
```

::: {.callout-tip}
**The d-p-q-r grid is your map.** Test statistics (Chapter 18) live in `pt`/`pf`/`pchisq`; confidence intervals in `qt`/`qnorm`; simulations in `r*`. Whenever a formula says "compare to the χ² distribution with k degrees of freedom," the R translation is one `pchisq` call.
:::

## Copyright Notice

- **Copyright.** These slides are adapted from *R in a Nutshell: A Desktop Quick Reference* (2nd ed.) by Joseph Adler, O'Reilly Media. All rights reserved by the original author and publisher.

- **Non-commercial use only.** These materials are strictly for educational purposes and may not be used for commercial gain.

- **Attribution.** Any reproduction, distribution, or use of these materials must properly credit the original source.
