---
title: "Programming for Applications"
subtitle: "Chapter 16: Analyzing Data"
author: "Yu-You Liou (NTU)"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "R in a Nutshell: A Desktop Quick Reference"
logo: "logo.svg"
editor: source
bibliography: bib.bib
suppress-bibliography: true
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch16-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center", fig.height = 4.2)
```

# **First Analyses**

## Summary Statistics: mean, min, max

The building blocks for everything that follows. With the `dow30` quote data from Chapter 12 (packaged in `nutshell`):

```{r,echo=TRUE}
temp <- tempfile(fileext = ".rda")
download.file("https://raw.githubusercontent.com/cran/nutshell/master/data/dow30.rda",
              temp, mode = "wb")
load(temp)
mean(dow30$Open); min(dow30$Open); max(dow30$Open)
```

Two arguments worth memorizing:

```{r,echo=TRUE}
mean(c(1, 2, 3, 4, 5, NA))                  # any NA poisons the result...
mean(c(1, 2, 3, 4, 5, NA), na.rm=TRUE)      # ...unless you remove it
mean(c(-1, 0:100, 2000))                    # outliers drag the mean...
mean(c(-1, 0:100, 2000), trim=0.1)          # ...trim filters that fraction
```

## range, quantile, fivenum, IQR

```{r,echo=TRUE}
range(dow30$Open)
quantile(dow30$Open, probs=c(0, 0.25, 0.5, 0.75, 1.0))
fivenum(dow30$Open)
IQR(dow30$Open)
```

All of these combine naturally with `apply`/`tapply` for per-group statistics (Chapter 12).

## summary and str

The most convenient overview is generic `summary` — quartiles + mean for numerics, top counts for factors (nothing meaningful for plain character vectors):

```{r,echo=TRUE}
summary(dow30[, c("Open", "Volume")])
```

The popular alternative `str` displays an object's *structure*:

```{r,echo=TRUE}
str(dow30)
```

## stem: a Text Histogram

`stem(x, scale=1, width=80, atom=1e-08)` draws a stem-and-leaf plot right in the console. Distances of *missed* NFL field goals, 2005:

```{r,echo=TRUE}
temp <- tempfile(fileext = ".rda")
download.file("https://raw.githubusercontent.com/cran/nutshell/master/data/field.goals.rda",
              temp, mode = "wb")
load(temp)
stem(subset(field.goals, play.type=="FG no")$yards)
```

# Correlation and Covariance

## Three Correlation Measures

Correlation answers "when x rises, does y rise — and how tightly?", measuring **linear dependence** on a −1...1 scale. Three statistics:

- **Pearson** — Excel's `CORREL`; rooted in normal-distribution theory, best for normal-ish data;
- **Spearman** — nonparametric, no distributional assumptions (correlates the *ranks*);
- **Kendall's tau** — compares concordant vs. discordant *pairs of rankings*.

```{r,eval=FALSE,echo=TRUE}
cor(x, y = NULL, use = "everything",
    method = c("pearson", "kendall", "spearman"))
```

`x`,`y` may be two vectors, or one data frame/matrix (→ a correlation *matrix*). `use` handles `NA`s: `"all.obs"` (error), `"everything"` (NA result), `"complete.obs"` / `"na.or.complete"` (drop incomplete rows), `"pairwise.complete.obs"` (drop per pair).

## Example: Weight Gain and Birth Weight

Does the mother's weight gain correlate with the baby's weight? Clean first (valid values, single births, gestation > 35 weeks), then look — 100k+ points need `smoothScatter`, not `plot`:

```{r,echo=TRUE}
temp <- tempfile(fileext = ".rda")
download.file("https://raw.githubusercontent.com/cran/nutshell/master/data/births2006.smpl.rda",
              temp, mode = "wb")
load(temp)
births2006.cln <- births2006.smpl[
  !is.na(births2006.smpl$WTGAIN) & !is.na(births2006.smpl$DBWT) &
  births2006.smpl$DPLURAL == "1 Single" & births2006.smpl$ESTGEST>35,]
smoothScatter(births2006.cln$WTGAIN, births2006.cln$DBWT)
cor(births2006.cln$WTGAIN, births2006.cln$DBWT)
cor(births2006.cln$WTGAIN, births2006.cln$DBWT, method="spearman")
```

A modest correlation by either measure — the blob is angled, just barely. (The book used the full 3.2M-record file; our 10% sample gives near-identical values.)

## Covariance

**Covariance** is the Pearson numerator — same arguments as `cor`:

```{r,echo=TRUE}
cov(births2006.cln$WTGAIN, births2006.cln$DBWT)
```

Relatives: `cov2cor` converts a covariance matrix to a correlation matrix; `cov.wt(x, wt=, cor=, center=, method=c("unbiased","ML"))` computes *weighted* covariance.

# Principal Components Analysis

## prcomp and princomp

**PCA** rewrites possibly-correlated variables as uncorrelated components. Two implementations: `prcomp` (preferred; formula method `prcomp(~..., data=, subset=, na.action=)`, default method with `retx`, `center=TRUE`, `scale.`, `tol`) and the S-PLUS-compatible `princomp` (eigendecomposition of the correlation/covariance matrix) — the book demonstrates the latter.

Data: team batting statistics 2000–2008, extracted from the Baseball Databank by SQL (renamed to readable play names) and packaged as `team.batting.00to08`:

```{r,echo=TRUE}
temp <- tempfile(fileext = ".rda")
download.file("https://raw.githubusercontent.com/cran/nutshell/master/data/team.batting.00to08.rda",
              temp, mode = "wb")
load(temp)
batting.pca <- princomp(~singles+doubles+triples+homeruns+
                          walks+hitbypitch+sacrificeflies+
                          stolenbases+caughtstealing,
                        data=team.batting.00to08)
summary(batting.pca)
```

Two components already carry ~78% of the variance; four carry ~92%.

## loadings, Scree Plot, Biplot

`loadings` shows each variable's contribution to each component:

```{r,echo=TRUE}
loadings(batting.pca)
```

```{r,echo=TRUE}
plot(batting.pca)                                      # "scree" plot
biplot(batting.pca, cex=0.5, col=c("gray50", "black")) # variables + observations
```

The biplot draws variable contributions to components 1–2 *and* the observations on one scale: singles and walks dominate the first two components. (This data returns for linear modeling in Chapter 20.)

# Factor Analysis

## factanal

Some quantities are observable (test scores); some are not (intelligence). **Factor analysis** infers hidden factors from observed variables — `factanal` in `stats`:

```{r,eval=FALSE,echo=TRUE}
factanal(x, factors, data = NULL, covmat = NULL, n.obs = NA,
         subset, na.action, start = NULL,
         scores = c("none", "regression", "Bartlett"),
         rotation = "varimax", control = NULL, ...)
```

| Argument | Description |
|---|---|
| `x` / `data` | formula or numeric matrix (+ data frame for the formula) |
| `factors` | how many factors to fit |
| `covmat` / `n.obs` | alternatively: a covariance matrix (or `cov.wt` output) + observation count |
| `subset` / `na.action` / `start` | the usual suspects |
| `scores` | `"none"`, `"regression"` (Thompson), `"Bartlett"` (weighted least squares) |
| `rotation` | factor rotation function, default `"varimax"` |

# Bootstrap Resampling

## The Idea, and boot

Is your statistic sensitive to a few outliers? What's its sampling range? **Bootstrapping** answers for *any* statistic: repeatedly resample observations **with replacement**, recompute the statistic, study the spread — formally, an estimate of an estimator's **bias**.

```{r,eval=FALSE,echo=TRUE}
library(boot)
boot(data, statistic, R, sim="ordinary", stype="i",
     strata=rep(1,n), L=NULL, m=0, weights=NULL,
     ran.gen=function(d, p) d, mle=NULL, simple=FALSE, ...)
```

Key arguments: `data`; `statistic` — a *function of (data, selection)* where `stype` defines the second argument (`"i"` indices, `"f"` frequencies, `"w"` weights); `R` — number of replicates; `sim` — `"ordinary"`, `"parametric"`, `"balanced"`, `"permutation"`, `"antithetic"`; `strata` for multisample problems; `ran.gen`/`mle` for parametric simulation; `simple=TRUE` trades time for memory.

## Example: How Biased Is the Median?

Media report *median* home prices. Bootstrapping June 2008 San Francisco sale prices:

```{r,echo=TRUE}
temp <- tempfile(fileext = ".rda")
download.file("https://raw.githubusercontent.com/cran/nutshell/master/data/sanfrancisco.home.sales.rda",
              temp, mode = "wb")
load(temp)
home.sale.prices.june2008 <- subset(sanfrancisco.home.sales,
  format(saledate, "%Y-%m")=="2008-06")$price
library(boot)
b <- boot(data=home.sale.prices.june2008,
          statistic=function(d, i) { median(d[i]) },
          R=1000)
b
```

Verdict: the median is a very slightly biased estimator here — and now we also know its standard error, free of any normality assumption.

## Copyright Notice

- **Copyright.** These slides are adapted from *R in a Nutshell: A Desktop Quick Reference* (2nd ed.) by Joseph Adler, O'Reilly Media. All rights reserved by the original author and publisher.

- **Non-commercial use only.** These materials are strictly for educational purposes and may not be used for commercial gain.

- **Attribution.** Any reproduction, distribution, or use of these materials must properly credit the original source.
