---
title: "Programming for Applications"
subtitle: "Chapter 16: Analyzing Data"
author: "Yu-You Liou (NTU)"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "R in a Nutshell: A Desktop Quick Reference"
logo: "logo.svg"
editor: source
bibliography: bib.bib
suppress-bibliography: true
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch16-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center")
```

# **First Numbers from the Data**

## Summary Statistics

The basic vocabulary of description, each one function away:

```{r,echo=TRUE,cache=TRUE}
temp <- tempfile(fileext = ".rda")
download.file("https://raw.githubusercontent.com/cran/nutshell/master/data/dow30.rda", temp, mode = "wb")
load(temp)
x <- dow30$Open
mean(x); median(x)
var(x); sd(x)
range(x); quantile(x, c(0.25, 0.5, 0.75))
```

- `mean(x, trim = 0.1)` resists outliers; every function here accepts `na.rm=TRUE`.
- `fivenum()` gives Tukey's five-number summary; `IQR()` the interquartile range.

## summary() Everywhere

`summary()` adapts to its argument — numeric vectors, factors, data frames, models:

```{r,echo=TRUE}
summary(dow30[, c("Open", "Close", "Volume")])
```

For a fast all-in-one look at any new data set: `summary(df)` plus `str(df)` answers "what am I holding?" in two lines.

## Correlation and Covariance

How do two series move together?

```{r,echo=TRUE}
cov(dow30$Open, dow30$Close)
cor(dow30$Open, dow30$Close)
cor(dow30[, c("Open", "High", "Low", "Close")])[1:4, 1:4]
```

- `cor()` defaults to Pearson; `method = "spearman"` or `"kendall"` ranks instead — robust to outliers and nonlinearity.
- A correlation matrix is the standard reconnaissance before regression (Chapter 20).

::: {.callout-warning}
**Correlation is not causation.** It is also blind to nonlinear structure — always look at the scatter plot too.
:::

## Principal Components Analysis

PCA re-expresses correlated variables as orthogonal **components** ordered by explained variance:

```{r,echo=TRUE}
pca <- prcomp(USArrests, scale. = TRUE)
summary(pca)
```

```{r,echo=TRUE,out.width="60%"}
biplot(pca, cex = 0.6)
```

- `scale.=TRUE` standardizes first — essential when units differ.
- The first two components often capture most of the structure: a built-in dimensionality reduction for plotting and modeling.

## Factor Analysis

Where PCA is pure variance bookkeeping, **factor analysis** posits latent variables that explain observed correlations:

```{r,echo=TRUE}
fa <- factanal(mtcars[, c("mpg", "disp", "hp", "wt", "qsec", "drat")],
               factors = 2)
fa$loadings
```

- Loadings show how strongly each observed variable reflects each hidden factor.
- Common in psychometrics and survey research: "do these 20 questions measure 2 underlying traits?"

## Bootstrap Resampling

When no formula exists for a statistic's uncertainty, **resample**: draw observations with replacement, recompute, repeat:

```{r,echo=TRUE,message=FALSE}
library(boot)
med.fun <- function(d, i) median(d[i])
set.seed(1)
b <- boot(dow30$Open, med.fun, R = 1000)
b
boot.ci(b, type = "perc")$percent
```

The spread of the 1,000 recomputed medians estimates the sampling distribution — no normality assumption required.

## A Workflow

1. `str()` and `summary()` — types, ranges, missingness.
2. Plots — histograms and scatter plots (Chapters 13–15).
3. Correlations — what moves together?
4. Reduce — PCA/factors if there are many variables.
5. Quantify uncertainty — bootstrap when in doubt.

Then, and only then, the formal tests of Chapter 18 and the models of Chapters 20–21.

## Copyright Notice

- **Copyright.** These slides are adapted from *R in a Nutshell: A Desktop Quick Reference* (2nd ed.) by Joseph Adler, O'Reilly Media. All rights reserved by the original author and publisher.

- **Non-commercial use only.** These materials are strictly for educational purposes and may not be used for commercial gain.

- **Attribution.** Any reproduction, distribution, or use of these materials must properly credit the original source.
