---
title: "Programming for Applications"
subtitle: "Chapter 26: R and Hadoop"
author: "Yu-You Liou (NTU)"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "R in a Nutshell: A Desktop Quick Reference"
logo: "logo.svg"
editor: source
bibliography: bib.bib
suppress-bibliography: true
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch26-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center")
```

# **R at Scale**

## When Data Outgrows One Machine

Everything so far assumed your data fits in one computer's memory. When it does not — terabytes of logs, billions of records — computation must be **distributed** across a cluster. **Hadoop** is the classic open-source platform for this, built on two pieces:

- **HDFS** — a distributed file system that stores data in blocks replicated across many machines;
- **MapReduce** — a programming model for processing that data in parallel, close to where it lives.

This chapter explains the model and how R connects to it; the code illustrates concepts rather than running locally.

## The MapReduce Model

Many computations factor into two stages:

::: {.callout-note}
**Map** applies a function to each record independently, emitting **key–value** pairs. **Reduce** gathers all values sharing a key and combines them. The framework handles the hard parts in between — distributing data, sorting by key, and recovering from failures.
:::

The canonical example — word count — captures the whole idea: map each line to `(word, 1)` pairs; reduce by summing the 1s for each word.

## Word Count in Pseudo-R

```{r,eval=FALSE,echo=TRUE}
# MAP: emit (word, 1) for every word in a line
map_fn <- function(key, line) {
  words <- strsplit(tolower(line), "\\W+")[[1]]
  lapply(words, function(w) keyval(w, 1))
}

# REDUCE: sum the counts for each distinct word
reduce_fn <- function(word, counts) {
  keyval(word, sum(unlist(counts)))
}
```

Because each map runs on one record and each reduce on one key, the framework can run thousands of them in parallel across the cluster — the programmer never writes scheduling or networking code.

## RHadoop

**RHadoop** is a family of packages letting you write MapReduce jobs in R instead of Java:

| Package | Role |
|:---|:---|
| `rhdfs` | read/write HDFS from R |
| `rmr2` | write MapReduce jobs in R (`mapreduce()`) |
| `rhbase` | access the HBase database |
| `plyrmr` | dplyr-style data manipulation over Hadoop |

```{r,eval=FALSE,echo=TRUE}
library(rmr2)
wordcount <- mapreduce(
  input    = "hdfs://input/books",
  map      = map_fn,
  reduce   = reduce_fn)
from.dfs(wordcount)        # pull the (small) result back into R
```

## The Pattern: Reduce, Then Pull

The recurring strategy across big-data tools:

1. **Push computation to the data** — run map/reduce where the blocks live, avoiding network transfer of terabytes.
2. **Shrink on the cluster** — aggregation, filtering, and summarization collapse big data into small results.
3. **Pull the summary into R** — once the data is small, all of Chapters 13–23 apply normally.

`from.dfs()` and `to.dfs()` move data between R's memory and HDFS; you bring back only what fits.

## The Modern Landscape

Hadoop MapReduce was the pioneer; today's ecosystem has moved on, but the *principles* are unchanged:

- **Apache Spark** with the `sparklyr` package — in-memory distributed computing, far faster than disk-based MapReduce, with a `dplyr` front end and built-in machine learning.
- **`arrow` / `duckdb`** — query larger-than-memory data efficiently on a single machine, often removing the need for a cluster at all.
- **Cloud data warehouses** (BigQuery, Snowflake) queried from R via DBI — push SQL to the warehouse, retrieve summaries.

The decision rule is the same as in Chapter 11: do the heavy lifting where the data lives, bring back only the answer.

## Course Wrap-Up

Over twenty-six chapters this course traveled from `1 + 1` to distributed computing:

- **Parts I–II** — installing R and the structure of the language;
- **Part III** — getting real data in and cleaning it;
- **Part IV** — three graphics systems;
- **Part V** — the statistical toolkit, from t-tests to random forests to ARIMA;
- **Part VI** — performance and scale.

The same language spans all of it. From here, depth comes from practice and from the focused follow-on courses — and every data set in these slides is one `download.file()` + `load()` away whenever you want to experiment.

## Copyright Notice

- **Copyright.** These slides are adapted from *R in a Nutshell: A Desktop Quick Reference* (2nd ed.) by Joseph Adler, O'Reilly Media. All rights reserved by the original author and publisher.

- **Non-commercial use only.** These materials are strictly for educational purposes and may not be used for commercial gain.

- **Attribution.** Any reproduction, distribution, or use of these materials must properly credit the original source.
