---
title: "Programming for Applications"
subtitle: "Chapter 24: Optimizing R Programs"
author: "Yu-You Liou (NTU)"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "R in a Nutshell: A Desktop Quick Reference"
logo: "logo.svg"
editor: source
bibliography: bib.bib
suppress-bibliography: true
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch24-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center", fig.height = 4.2)
```

# **Making R Faster**

## Measure First, Then Optimize

The golden rule: **never optimize blindly**. Find where time and memory actually go, then fix *that*. R gives you tools for both.

# Measuring Performance

## Timing: system.time

The simplest measurement — wrap any expression:

```{r,echo=TRUE}
system.time(for (i in 1:1e6) sqrt(i))
```

Three numbers: **user time** (work R itself did), **system time** (work the OS did on R's behalf), **elapsed time** (wall clock). `gcFirst=TRUE` (the default) runs garbage collection first for a cleaner reading. (For careful microbenchmarks the modern `microbenchmark` and `bench` packages repeat and compare.)

## Profiling: Rprof and summaryRprof

To see *where* the time goes across a whole program, **profile** it:

```{r,eval=FALSE,echo=TRUE}
Rprof("Rprof.out", interval = 0.02, memory.profiling = FALSE)
do.stuff()                       # ... your code ...
Rprof(NULL)                      # stop profiling
summaryRprof("Rprof.out")        # which functions ate the time
```

`Rprof` samples the call stack every `interval` seconds and writes to a file; `Rprof(NULL)` stops it. `summaryRprof` ranks functions by self- and total time (and, with `memory=`, by allocation). This is how you find the *one* line consuming most of the runtime — often a surprise.

## Monitoring Memory

```{r,echo=TRUE}
gc()                             # garbage-collect AND report usage
object.size("Hello world!")      # approximate size of one object
```

- **`gc()`** forces garbage collection *and* reports memory in use.
- **`object.size(x)`** estimates an object's footprint — invaluable when a "small" data frame is secretly huge.
- **`memory.profile()`** breaks usage down by object type; on Windows, `memory.size()`/`memory.limit()` report and cap total usage.
- **`Rprofmem(filename, threshold=)`** logs large allocations (R must be built with memory profiling).

# Optimizing Your Code

## Use Vector Operations

R's single biggest speed lever: replace **iterative element-by-element code with whole-vector operations**. The Chapter 8 example, dramatized — growing a vector in a loop versus working on it at once:

```{r,echo=TRUE}
slow <- function(n) { v <- NA; for (i in 1:n) v[i] <- i^2; v }
fast <- function(n) { (1:n)^2 }
system.time(slow(50000))["elapsed"]
system.time(fast(50000))["elapsed"]
```

The vectorized version is dramatically faster *and* clearer. Two reasons: the loop's body re-dispatches R functions tens of thousands of times, and **growing a vector reallocates it on every step**. Pre-allocating (`v <- numeric(n)`) helps; vectorizing helps far more.

## Transform Problems to Use Built-in Functions

Many loops can be rephrased as calls to functions implemented in C:

- the **apply family** — `sapply`, `lapply`, `vapply`, `mapply`, `tapply` (Chapters 9, 12);
- **matrix algebra** — `%*%`, `crossprod`, `rowSums`, `colMeans` instead of nested loops;
- **`outer`**, `cumsum`, `diff`, `tabulate`, `findInterval` for common patterns.

```{r,echo=TRUE}
m <- matrix(1:1e6, ncol=1000)
system.time(apply(m, 2, sum))["elapsed"]      # readable
system.time(colSums(m))["elapsed"]            # built-in, far faster
```

The lesson: a built-in that does the loop *inside C* beats any R-level loop.

## Lookup Performance: Environments as Hash Tables

Repeatedly looking values up by name? A **named vector or list** does a linear scan; an **environment** is a hash table with near-constant-time lookup. For thousands of keyed lookups, store the data in an environment:

```{r,echo=TRUE}
e <- new.env()
assign("BG", 42, envir=e)        # or e[["BG"]] <- 42
get("BG", envir=e)               # O(1) lookup
```

The book benchmarks this against named-vector lookup and finds environments dramatically faster as the table grows — the trick behind many high-performance R packages.

## The Byte-Code Compiler, and Beyond

R ships a **byte-code compiler** (`compiler` package) that speeds up pure-R functions — and since R 2.14, base and recommended packages are pre-compiled; since R 3.4, functions are JIT-compiled on use:

```{r,echo=TRUE}
library(compiler)
slow.c <- cmpfun(slow)           # compile to byte code
system.time(slow.c(50000))["elapsed"]
```

::: {.callout-tip}
**The optimization ladder.** (1) Vectorize. (2) Use built-in/C functions and matrix algebra. (3) Pre-allocate; use environments for lookups. (4) Compile hot functions. (5) Only then go parallel (Chapter 26) or rewrite the bottleneck in C++ (`Rcpp`). Always re-profile after each step — and stop when it's fast *enough*.
:::

## Copyright Notice

- **Copyright.** These slides are adapted from *R in a Nutshell: A Desktop Quick Reference* (2nd ed.) by Joseph Adler, O'Reilly Media. All rights reserved by the original author and publisher.

- **Non-commercial use only.** These materials are strictly for educational purposes and may not be used for commercial gain.

- **Attribution.** Any reproduction, distribution, or use of these materials must properly credit the original source.
