---
title: "Programming for Applications"
subtitle: "Chapter 24: Optimizing R Programs"
author: "Yu-You Liou (NTU)"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "R in a Nutshell: A Desktop Quick Reference"
logo: "logo.svg"
editor: source
bibliography: bib.bib
suppress-bibliography: true
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch24-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center")
```

# **Making R Fast**

## Measure Before You Optimize

The cardinal rule: never guess where time goes — measure it.

```{r,echo=TRUE}
system.time({
  x <- numeric(0)
  for (i in 1:20000) x <- c(x, i)   # the growing-vector sin
})
system.time({
  y <- numeric(20000)
  for (i in 1:20000) y[i] <- i      # pre-allocated
})
```

Same loop, wildly different cost — because `c()` re-copies the whole vector every iteration.

## Profiling

For whole programs, the profiler attributes time to functions:

```{r,echo=TRUE}
slow_demo <- function(n) {
  m <- matrix(rnorm(n * 50), ncol = 50)
  out <- numeric(ncol(m))
  for (j in seq_len(ncol(m))) out[j] <- mean(sort(m[, j]))
  out
}
Rprof(tmp <- tempfile())
invisible(slow_demo(20000))
Rprof(NULL)
head(summaryRprof(tmp)$by.self, 5)
```

- `Rprof()` samples the call stack; `summaryRprof()` tabulates it.
- `Rprof(memory.profiling = TRUE)` and `tracemem()` expose hidden copying.

## Vectorize

The single biggest R speedup: replace element-by-element loops with whole-vector operations, which run in compiled C:

```{r,echo=TRUE}
v <- rnorm(1e6)
system.time({ s <- 0; for (xi in v) s <- s + xi^2 })
system.time(sum(v^2))
```

- `rowSums()`, `colMeans()`, `cumsum()`, `pmax()`, `ifelse()`-free logical math (`(x > 0) * x`) all vectorize common patterns.
- The apply family is *clearer* than loops but not inherently faster — true vector ops are.

## Lookup Performance

How you find things matters as data grows:

```{r,echo=TRUE}
n <- 1e5
named.vec <- setNames(rnorm(n), paste0("k", 1:n))
e <- new.env(hash = TRUE)
for (k in c("k50000")) assign(k, named.vec[[k]], envir = e)
system.time(for (i in 1:200) named.vec["k50000"])   # name lookup: linear
system.time(for (i in 1:200) get("k50000", envir = e))  # hash: constant
```

- Name-based vector indexing scans; hashed environments (and the `fastmatch`/`data.table` packages) jump straight there.
- `match()` beats repeated `%in%` in loops; sorted keys enable `findInterval()`.

## Memory Matters

R keeps everything in RAM, and copying is the silent killer:

```{r,echo=TRUE}
gc()[, c("used", "max used")] |> head(2)
object.size(rnorm(1e6))
```

- `rm()` big objects you are done with; `gc()` reports (and triggers) garbage collection.
- Modify columns rather than whole data frames inside loops; each "small" change of a big object may copy all of it.
- For data bigger than RAM: `data.table` (in-memory but frugal), `ff`/`bigmemory` (disk-backed), databases (Chapter 11), or Hadoop (Chapter 26).

## The Byte-Code Compiler

R compiles functions to byte code; ensure hot functions benefit:

```{r,echo=TRUE}
library(compiler)
f <- function(n) { s <- 0; for (i in 1:n) s <- s + i; s }
fc <- cmpfun(f)
system.time(f(2e6)); system.time(fc(2e6))
```

Since R 3.4 the JIT compiles most functions automatically — but the gap shows what the interpreter costs, and why C++ via the `Rcpp` package remains the heavy artillery for unavoidable loops.

## Parallel Computation

Many tasks are embarrassingly parallel — independent pieces, no communication:

```{r,echo=TRUE}
library(parallel)
detectCores()
system.time(res1 <- lapply(1:8, function(i) mean(rnorm(4e5))))
cl <- makeCluster(2)
system.time(res2 <- parLapply(cl, 1:8, function(i) mean(rnorm(4e5))))
stopCluster(cl)
```

- `mclapply()` (Unix/macOS forking) is the one-line version; `parLapply()` works everywhere.
- Simulations, bootstraps, cross-validation folds, and grid searches parallelize perfectly; overhead means small tasks may not.

## An Optimization Checklist

1. **Measure** (`system.time`, `Rprof`) — find the real bottleneck.
2. **Vectorize** the hot loop; pre-allocate results.
3. **Avoid copies** — beware growing objects and repeated subsetting of big frames.
4. **Better algorithms/structures** — hashing, sorting, `data.table`.
5. **Parallelize** independent work.
6. **Drop to C++** (`Rcpp`) only when all of the above still isn't enough.

Readable-but-fast beats clever-but-fragile: optimize the 5% that matters and leave the rest clear.

## Copyright Notice

- **Copyright.** These slides are adapted from *R in a Nutshell: A Desktop Quick Reference* (2nd ed.) by Joseph Adler, O'Reilly Media. All rights reserved by the original author and publisher.

- **Non-commercial use only.** These materials are strictly for educational purposes and may not be used for commercial gain.

- **Attribution.** Any reproduction, distribution, or use of these materials must properly credit the original source.
