---
title: "Programming for Applications"
subtitle: "Chapter 11: Saving, Loading, and Editing Data"
author: "Yu-You Liou (NTU)"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "R in a Nutshell: A Desktop Quick Reference"
logo: "logo.svg"
editor: source
bibliography: bib.bib
suppress-bibliography: true
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch11-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center")
```

# **Data In, Data Out**

## Entering Data Within R

For small data sets, the constructors from Part II are all you need:

```{r,echo=TRUE}
salary <- c(18700000, 14626720, 14137500)
position <- c("QB", "QB", "DE")
top.contracts <- data.frame(position, salary)
top.contracts
```

- `rbind()` appends rows; `cbind()` appends columns.
- `scan()` reads values typed interactively — handy for quick experiments.

## The Edit GUI

R ships a minimal spreadsheet-style editor:

```{r,eval=FALSE,echo=TRUE}
top.contracts <- edit(top.contracts)  # edit returns the NEW object
fix(top.contracts)                    # fix saves the result back itself
```

::: {.callout-warning}
**`edit()` does not modify in place.** Forgetting to assign its result is a classic data-loss moment; `fix()` exists precisely to avoid that.
:::

- The editor's look differs by platform (Windows grid, macOS sheet, X11 window), but the function is identical.
- For real data entry, a spreadsheet program is more comfortable — the editor is best for quick inspection and one-cell fixes.

## Saving Objects

`save()` writes any set of R objects to a portable binary file; `load()` restores them, names and all:

```{r,echo=TRUE}
tmp <- tempfile(fileext = ".RData")
save(top.contracts, file = tmp)       # `file =` must be NAMED
rm(top.contracts)
load(tmp)
top.contracts[1, ]
```

- Files are cross-platform; compression is on by default.
- `save.image()` snapshots the whole workspace (what "Save workspace?" does on exit).
- `saveRDS()`/`readRDS()` save a *single* object without its name — often cleaner for programmatic use.

## Loading from the Web: the Course Convention

`download.file()` fetches any file from a URL to disk (`mode = "wb"` for binary files); `load()` then restores it. This is exactly how all course data sets are loaded:

```{r,echo=TRUE}
temp <- tempfile(fileext = ".rda")
download.file("https://raw.githubusercontent.com/cran/nutshell/master/data/top.bacon.searching.cities.rda", temp, mode = "wb")
load(temp)
top.bacon.searching.cities
```

The textbook's `library(nutshell); data(...)` is always replaced by this three-line pattern.

## Importing Text Files

The `read.table()` family parses delimited text into data frames:

```{r,eval=FALSE,echo=TRUE}
df  <- read.table("file.txt", header = TRUE, sep = "\t")
csv <- read.csv("file.csv")             # comma-separated
eu  <- read.csv2("file.csv")            # ; separator, , decimal
fwf <- read.fwf("file.dat", widths = c(8, 4, 4))
```

Key arguments worth memorizing: `header`, `sep`, `na.strings`, `stringsAsFactors`, `skip`, `colClasses` (huge speed win on big files).

```{r,echo=TRUE}
# readLines shows raw lines — always look at a file before parsing it
readLines(textConnection("a,b\n1,2\n3,4"))
```

## Other Software and Spreadsheets

- The `foreign` package reads legacy statistics formats: SAS (`read.xport`), SPSS (`read.spss`), Stata (`read.dta`), and more.
- Excel files: the `readxl` package (`read_excel("book.xlsx")`) needs no Excel installation; `openxlsx` also writes.
- The quick-and-dirty route still works: export a sheet to CSV, then `read.csv()`.

```{r,eval=FALSE,echo=TRUE}
library(readxl)
roster <- read_excel("stu/roster.xlsx", sheet = 1)
```

## Exporting Data

Every importer has a mirror image:

```{r,eval=FALSE,echo=TRUE}
write.csv(top.contracts, "contracts.csv", row.names = FALSE)
write.table(top.contracts, "contracts.txt", sep = "\t")
saveRDS(top.contracts, "contracts.rds")
```

- `row.names = FALSE` avoids the mysterious extra first column that surprises spreadsheet users.
- For data interchange with other languages, CSV remains the lingua franca; for R-to-R, use `.rds`.

## Importing from Databases

Three strategies, in increasing order of integration:

1. **Export then import**: dump a query result to CSV from the database tool, then `read.csv()` it. Crude but dependable.
2. **RODBC**: connect through ODBC drivers — works with virtually any database that has a driver.
3. **DBI**: R's native database interface, with backend packages per engine (`RSQLite`, `RPostgres`, `RMariaDB`, ...).

## RODBC in Brief

```{r,eval=FALSE,echo=TRUE}
library(RODBC)
ch <- odbcConnect("dsn_name", uid = "user", pwd = "pass")
df <- sqlQuery(ch, "SELECT player, yards FROM kicks WHERE good = 1")
sqlTables(ch)              # list tables
close(ch)
```

- `sqlFetch(ch, "tbl")` grabs a whole table; `sqlSave()` writes a data frame back.
- ODBC setup happens *outside* R (driver + data source name), which is most of the work in practice.

## DBI in Brief

```{r,echo=TRUE,message=FALSE}
library(DBI)
con <- dbConnect(RSQLite::SQLite(), ":memory:")   # a throwaway database
dbWriteTable(con, "contracts", top.contracts)
dbGetQuery(con, "SELECT position, MAX(salary) AS top FROM contracts GROUP BY position")
dbDisconnect(con)
```

- One generic API — `dbConnect`, `dbGetQuery`, `dbWriteTable`, `dbDisconnect` — over interchangeable backends.
- Prepared statements and incremental fetching (`dbSendQuery` + `dbFetch`) handle result sets too large for memory.

## Bigger Pipelines

- **TSDBI** offers a database interface specialized for time series.
- Data living in **Hadoop**/distributed storage is the subject of Chapter 26 — the usual pattern is to reduce it first, then pull the summary into R.
- Rule of thumb: push heavy filtering and joins into the database (`WHERE`, `GROUP BY`), and bring into R only what analysis actually needs.

## Copyright Notice

- **Copyright.** These slides are adapted from *R in a Nutshell: A Desktop Quick Reference* (2nd ed.) by Joseph Adler, O'Reilly Media. All rights reserved by the original author and publisher.

- **Non-commercial use only.** These materials are strictly for educational purposes and may not be used for commercial gain.

- **Attribution.** Any reproduction, distribution, or use of these materials must properly credit the original source.
