---
title: "Programming for Applications"
subtitle: "Chapter 6: R Syntax"
author: "Yu-You Liou (NTU)"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "R in a Nutshell: A Desktop Quick Reference"
logo: "logo.svg"
editor: source
bibliography: bib.bib
suppress-bibliography: true
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch6-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center")
```

# **Writing Valid R**

## Sugar over Function Calls

Chapter 5 showed that *every* R expression can be rewritten as a function call. Yet R provides special syntax — **syntactic sugar** — so that assignments, lookups, and arithmetic read naturally. (Without it, R code would look like LISP, all parentheses; no coincidence, since S was inspired by LISP.)

This chapter is a readable, not exhaustive, guide to writing valid R expressions:

- constants: numbers, characters, symbols;
- operators and their precedence;
- assignments, grouping, control structures;
- the indexing notation for data structures;
- and a few words on code style.

# Constants

## Numeric Vectors

Numbers are interpreted literally — and hexadecimal works with the `0x` prefix:

```{r,echo=TRUE}
1.1
2
2^1023
0x1
0xFFFF
```

By default every number — even a plain integer — is stored as a **double-precision floating-point** value. To get a true integer, use sequence notation or `as`:

```{r,echo=TRUE}
typeof(1)
typeof(1:1)
typeof(as(1, "integer"))
```

## Precision and Size Limits

`a:b` yields the integers from `a` to `b`; arbitrary sets combine with `c`:

```{r,echo=TRUE}
v <- c(173, 12, 1.12312, -93)
```

Flexibility has limits — doubles cannot distinguish or represent everything:

```{r,echo=TRUE}
(2^1023 + 1) == 2^1023   # limits of precision
2^1024                   # limits of size
```

In practice this rarely matters: the databases your data comes from can't represent such numbers either.

## Complex Numbers

Complex values are written as *real part* `+` *imaginary part* `i`:

```{r,echo=TRUE,warning=TRUE}
0+1i ^ 2
sqrt(-1+0i)
exp(0+1i * pi)
sqrt(-1)        # NaN! sqrt returns the same type it was given
```

Note the last line: handed the *numeric* `-1`, `sqrt` answers in numerics — and the numeric answer to √−1 is `NaN`, with a warning. Handed a complex number, it answers in complex.

## Character Vectors

A character object is the text between a pair of quotes — double or single, your choice, or escaped with backslashes:

```{r,echo=TRUE}
"hello"
'hello'
identical("\"hello\"", '"hello"')
identical('\'hello\'', "'hello'")
```

Single quotes are convenient when the text itself contains double quotes (and vice versa). Longer vectors, as always, come from `c`:

```{r,echo=TRUE}
numbers <- c("one", "two", "three", "four", "five")
numbers
```

## Symbols

A **symbol** is a constant that refers to another object — the name of a variable. `x <- 1` means "map the symbol `x` to the value 1 in the current environment."

Names that start with a letter and contain letters, digits, periods, and underscores can be typed directly — and **case matters**:

```{r,echo=TRUE}
x1 <- 1
X1 <- 2
x1
X1
x1.1 <- 1
x1.1_1 <- 1
```

## Backquotes and Reserved Words

Symbols with special syntax are reached by enclosing them in **backquotes** — that is how you get help on the assignment operator (`` ?`<-` ``), and, if you insist, how you create perverse names:

```{r,echo=TRUE}
`1+2=3` <- "hello"
`1+2=3`
```

Some words are **reserved** and cannot be used as symbols: `if`, `else`, `repeat`, `while`, `function`, `for`, `in`, `next`, `break`, `TRUE`, `FALSE`, `NULL`, `Inf`, `NaN`, `NA`, `NA_integer_`, `NA_real_`, `NA_complex_`, `NA_character_`, `...`, and `..1` through `..9`.

## Redefining Built-ins

Primitive functions *not* on the reserved list can be redefined — even `c`:

```{r,echo=TRUE}
c
c <- 1
c
v <- c(1, 2, 3)   # and yet the combine function still works!
v
```

```{r,include=FALSE}
rm(c)
```

R can tell from context that `c(1, 2, 3)` needs a *function* named `c`, and finds the original. Clever — but shadowing built-in names is still a recipe for confusion; avoid it in real code.

# Operators

## Binary and Unary Operators

An **operator** is a function of one or two arguments that can be written without parentheses:

```{r,echo=TRUE}
1 + 19      # addition
5 * 4       # multiplication
41 %% 21    # modulus
20 ^ 1      # exponent
21 %/% 2    # integer division
```

Unary operators take a single operand — negation (`-7`) and help (`?`) are familiar examples.

## Defining Your Own Operator

A user-defined binary operator is any function of two variables assigned to a name of the form `%text%`:

```{r,echo=TRUE}
`%myop%` <- function(a, b) { 2*a + 2*b }
1 %myop% 1
1 %myop% 2
```

Some core language constructs are themselves binary operators: **assignment** (symbol on the left, value on the right), **indexing** (symbol left, index right), and even **function calls** (function left, arguments right — don't be fooled by the closing bracket or parenthesis):

```{r,echo=TRUE}
x <- c(1, 2, 3, 4, 5)
x[3]
max(1, 2)
```

## Order of Operations

Just as in school math (`1 + 2 × 5` multiplies first), R resolves ambiguity with fixed **precedence** rules. From highest to lowest, roughly: function calls and grouping; indexing and lookup; arithmetic; comparison; formulas; assignment; help.

| Operators (by priority) | Description |
|---|---|
| `( {` | Function calls, grouping |
| `[ [[` | Indexing |
| `:: :::` | Namespace access |
| `$ @` | Component / slot extraction |
| `^` | Exponentiation (right to left) |
| `- +` | Unary minus, plus |
| `:` | Sequence |
| `%any%` | Special operators |
| `* /` | Multiply, divide |
| `+ -` | Binary add, subtract |
| `< > <= >= == !=` | Ordering, comparison |
| `!` | Negation |
| `& &&` | And |
| `| ||` | Or |
| `~` | Formulas |
| `-> ->>` | Rightward assignment |
| `=` | Assignment (right to left) |
| `<- <<-` | Assignment (right to left) |
| `?` | Help |

The authoritative, current list: `help(syntax)`.

# Assignments

## Plain and Property Assignments

Most assignments simply bind an object to a symbol:

```{r,echo=TRUE}
x <- 1
y <- list(shoes="loafers", hat="Yankees cap", shirt="white")
z <- function(a, b, c) { a ^ b / c }
v <- c(1, 2, 3, 4, 5, 6, 7, 8)
```

But R also allows **a function on the left-hand side** — replacing an object with a modified version of itself:

```{r,echo=TRUE}
dim(v) <- c(2, 4)        # v is now a 2x4 matrix
v[2, 2] <- 10
formals(z) <- alist(a=1, b=2, c=3)
```

## Replacement Functions

The magic behind the scenes: a statement of the form `fun(sym) <- val` is syntactic sugar for

```{r,eval=FALSE,echo=TRUE}
`fun<-`(sym, val)
```

which replaces the object bound to `sym` in the current environment. By convention `fun` names a *property* of the object. Define your own method named `myproperty<-`, and R lets you write `myproperty(x) <- value` — your own left-hand-side function.

# Grouping Expressions

## Semicolons and Parentheses

Expressions go on separate lines, or on one line separated by semicolons:

```{r,echo=TRUE}
x <- 1; y <- 2; z <- 3
```

**Parentheses** return the result of the expression inside — precedence-wise the same as a function call. Indeed, grouping is equivalent to calling an identity function:

```{r,echo=TRUE}
2 * (5 + 1)
f <- function(x) x
2 * f(5 + 1)        # the same thing
2 * 5 + 1           # versus the default order of operations
```

## Curly Braces

**Curly braces** evaluate a series of expressions (newline- or semicolon-separated) and return only the last:

```{r,echo=TRUE}
f <- function() { x <- 1; y <- 2; x + y }
f()
{ x <- 1; y <- 2; x + y }    # braces work outside functions too
```

## Braces Do Not Create Environments

A crucial difference: a **function call** creates a new environment; **braces do not** — their contents evaluate in the current environment:

```{r,error=TRUE,echo=TRUE}
f <- function() { u <- 1; v <- 2; u + v }
f()
u                       # not defined out here!
```

```{r,echo=TRUE}
{ u <- 1; v <- 2; u + v }
u                       # defined: braces ran in the current environment
v
```

Internally, brace notation is a call to the `` `{` `` function — though its arguments are not evaluated like a standard function's. (Scope and environments: Chapter 8.)

# Control Structures

## Conditional Statements

The forms are `if (condition) true_expression else false_expression` and `if (condition) expression`. Because the branches are *not always evaluated*, `if` has type **special**:

```{r,echo=TRUE}
typeof(`if`)
if (FALSE) "this will not be printed"
if (FALSE) "this will not be printed" else "this will be printed"
x <- 10
if (is(x, "numeric")) x/2 else print("x is not numeric")
```

## Conditions Must Be Scalar; ifelse Is the Vector Version

An `if` condition is **not** a vector operation. In the book's era a longer condition used the first element with a warning; **modern R (≥ 4.2) makes it an error**:

```{r,error=TRUE,echo=TRUE}
x <- 10
y <- c(8, 10, 12, 3, 17)
if (x < y) x else y
```

For an element-wise choice, use `ifelse`:

```{r,echo=TRUE}
a <- c("a", "a", "a", "a", "a")
b <- c("b", "b", "b", "b", "b")
ifelse(c(TRUE, FALSE, TRUE, FALSE, TRUE), a, b)
```

## switch

Returning different values for different inputs via chained `else if`s is verbose:

```{r,echo=TRUE}
switcheroo.if.then <- function(x) {
  if (x == "a")      "alligator"
  else if (x == "b") "bear"
  else if (x == "c") "camel"
  else               "moose"
}
```

`switch` does the same job tidily — named arguments map inputs to results; the unnamed argument is the default:

```{r,echo=TRUE}
switcheroo.switch <- function(x) {
  switch(x, a="alligator", b="bear", c="camel", "moose")
}
switcheroo.if.then("a"); switcheroo.switch("a")
switcheroo.if.then("f"); switcheroo.switch("f")
```

## Loops: repeat

R has three looping constructs. The simplest, `repeat`, just repeats an expression — forever, unless you `break` (and `next` skips to the next iteration). Multiples of 5 up to 25:

```{r,echo=TRUE}
i <- 5
repeat { if (i > 25) break else { print(i); i <- i + 5 } }
```

Forget the `break` and you have an infinite loop — occasionally useful for interactive applications, otherwise a bug.

## Loops: while

`while (condition) expression` repeats as long as the condition holds; `break` and `next` work here too:

```{r,echo=TRUE}
i <- 5
while (i <= 25) { print(i); i <- i + 5 }
```

## Loops: for

`for (var in list) expression` iterates over the elements of a vector or list:

```{r,echo=TRUE}
for (i in seq(from=5, to=25, by=5)) print(i)
```

Two properties to remember:

```{r,echo=TRUE}
for (i in seq(from=5, to=25, by=5)) i   # 1. no printing unless you print()
i                                        # 2. the loop variable leaks into the calling environment
```

Like conditionals, `repeat`, `while`, and `for` have type **special** — their body is not necessarily evaluated.

## Looping Extensions: iterators and foreach

::: {.callout-note}
Missing Java-style iterators or foreach? CRAN add-on packages supply both — and they make code easier to parallelize.
:::

**iterators**: `iter(obj, checkFunc=..., recycle=...)` builds an iterator; `nextElem` fetches values (filtered by `checkFunc`) until it stops with "StopIteration":

```{r,error=TRUE,echo=TRUE}
library(iterators)
onetofive <- iter(1:5)
nextElem(onetofive); nextElem(onetofive); nextElem(onetofive)
nextElem(onetofive); nextElem(onetofive)
nextElem(onetofive)    # exhausted
```

**foreach**: builds a loop object, evaluated by the `%do%` (serial) or `%dopar%` (parallel — Chapter 26) operators:

```{r,echo=TRUE,message=FALSE}
library(foreach)
sqrts.1to5 <- foreach(i=1:5) %do% sqrt(i)
sqrts.1to5[1:3]
```

# Accessing Data Structures

## The Four Access Operators

| Syntax | Objects | Description |
|---|---|---|
| `x[i]` | vectors, lists | Elements described by `i` — an integer, character (names), or logical vector. No partial matching. On a list returns a **list**; on a vector, a vector. |
| `x[[i]]` | vectors, lists | A **single** element matching `i` (integer or character of length 1). Partial name matching with `exact=FALSE`. |
| `x$n` | lists | The element named `n`. |
| `x@n` | S4 objects | The value in slot `n`. |

Single vs. double brackets, three differences: `[[` returns exactly one element while `[` may return several; by *name*, `[` matches exactly while `[[` allows partial matches; on lists, `[` returns a list but `[[` returns the element itself.

## Indexing by Integer Vector

```{r,echo=TRUE}
v <- 100:119
v[5]
v[1:5]
v[c(1, 6, 11, 16)]
v[[3]]              # double brackets: single element
```

**Negative integers exclude** the named positions:

```{r,echo=TRUE}
v[-15:-1]           # everything except elements 1..15
```

## Integer Indexing of Lists

The same notation applies to lists:

```{r,echo=TRUE}
l <- list(a=1, b=2, c=3, d=4, e=5, f=6, g=7, h=8, i=9, j=10)
l[1:3]
l[-7:-1]
```

## Integer Indexing of Matrices

For multidimensional structures, give one index per dimension — or omit one to take everything along it:

```{r,echo=TRUE}
m <- matrix(data=c(101:112), nrow=3, ncol=4)
m
m[3]        # single index: positions in the underlying vector
m[3, 4]
m[1:2, 1:2]
m[1:2, ]
m[, 3:4]
```

## Dimension Dropping and drop=FALSE

Subsets are automatically **coerced to the fewest sensible dimensions**: a matrix-shaped subset returns a matrix, a vector-shaped one a vector. Disable with `drop=FALSE`:

```{r,echo=TRUE}
a <- array(data=c(101:124), dim=c(2, 3, 4))
class(a[1, 1, ])
class(a[1, , ])
class(a[1:2, 1:2, 1:2])
class(a[1, 1, 1, drop=FALSE])
```

## Replacing and Extending by Index

The same notation **assigns** — single cells, blocks, even positions beyond the current end (gaps become `NA`):

```{r,echo=TRUE}
m[1] <- 1000
m[1:2, 1:2] <- matrix(c(1001:1004), nrow=2, ncol=2)
m
v <- 1:12
v[15] <- 15
v
```

A factor used as an index is interpreted as an integer vector.

## Indexing by Logical Vector

A logical vector selects the `TRUE` positions — and is **recycled** if shorter than the target:

```{r,echo=TRUE}
v <- 100:119
v[rep(c(TRUE, FALSE), 10)]
v[(v == 103)]          # compute the mask from the vector itself
v[(v %% 3 == 0)]       # multiples of three
v[c(TRUE, FALSE, FALSE)]   # recycled mask
```

Lists too:

```{r,echo=TRUE}
l[(l > 7)]
```

## Indexing by Name

List elements may be named; index them with `$`, or by name vectors in single brackets:

```{r,echo=TRUE}
l <- list(a=1, b=2, c=3, d=4, e=5, f=6, g=7, h=8, i=9, j=10)
l$j
l[c("a", "b", "c")]
```

Double brackets take a single name — and uniquely allow **partial matching** with `exact=FALSE`:

```{r,echo=TRUE}
dairy <- list(milk="1 gallon", butter="1 pound", eggs=12)
dairy$milk
dairy[["milk"]]
dairy[["mil"]]                  # no exact match: NULL
dairy[["mil", exact=FALSE]]     # partial match allowed
```

## Indexing Nested Lists

For lists of lists, `[[` accepts a *vector* and walks down the levels:

```{r,echo=TRUE}
fruit <- list(apples=6, oranges=3, bananas=10)
shopping.list <- list(dairy=dairy, fruit=fruit)
shopping.list[[c("dairy", "milk")]]
shopping.list[[c(1, 2)]]
```

# R Code Style Standards

## Readable Code Is Maintainable Code

Style isn't syntax, but it decides whether others (and future you) can read your code. The book follows Google's R Style Guide; the highlights:

- **Indentation:** two spaces, never tabs; inside parentheses, align to the innermost.
- **Spacing:** single spaces around binary operators and after commas; none between a function name and its argument list.
- **Blocks:** opening brace `{` not on its own line; closing brace `}` on its own line; indent inner blocks two spaces.
- **Semicolons:** omit when optional.
- **Naming:** objects in lowercase words separated by periods; function names in CapWords and preferably verbs.
- **Assignment:** use `<-`, not `=`.

And no — you are not obliged to name your objects `field.goals` or `top.bacon.searching.cities`. It's just convention. (The modern community mostly follows the *tidyverse style guide*, which prefers `snake_case` — pick one style and stay consistent.)

## Copyright Notice

- **Copyright.** These slides are adapted from *R in a Nutshell: A Desktop Quick Reference* (2nd ed.) by Joseph Adler, O'Reilly Media. All rights reserved by the original author and publisher.

- **Non-commercial use only.** These materials are strictly for educational purposes and may not be used for commercial gain.

- **Attribution.** Any reproduction, distribution, or use of these materials must properly credit the original source.
