---
title: "Programming for Applications"
subtitle: "Chapter 6: R Syntax"
author: "Yu-You Liou (NTU)"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "R in a Nutshell: A Desktop Quick Reference"
logo: "logo.svg"
editor: source
bibliography: bib.bib
suppress-bibliography: true
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch6-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center")
```

# **Constants**

## Numeric Vectors

Literal numbers default to double-precision floating point; suffixes and prefixes change the interpretation:

```{r,echo=TRUE}
1.1                 # double
123L                # the L suffix forces an integer
1e3                 # scientific notation
0xFF                # hexadecimal
1+2i                # complex
```

- `class()` distinguishes `numeric` (double), `integer`, and `complex`.
- Very large or very small magnitudes print in scientific notation automatically.

## Character Vectors

Strings may be quoted with double or single quotes; special characters are written with backslash escapes:

```{r,echo=TRUE}
"hello"
'also a string'
cat("tab:\tnewline:\nquote:\"done\"\n")
```

- `\t`, `\n`, `\\`, `\"` are the escapes you will actually use.
- `nchar()` counts characters; `paste()` glues strings together.

## Logical Constants and Symbols

```{r,echo=TRUE}
TRUE; FALSE
```

- Always spell out `TRUE`/`FALSE`. The shortcuts `T`/`F` are ordinary variables that someone can reassign — a classic trap.
- A **syntactic name** starts with a letter (or a dot not followed by a digit) and contains letters, digits, dots, and underscores.
- Any other name can still be used if wrapped in backquotes:

```{r,echo=TRUE}
`my odd name` <- 42
`my odd name`
```

# Operators

## The Operator Zoo

```{r,echo=TRUE}
7 %% 3              # modulus
7 %/% 3             # integer division
2 ^ 5               # exponentiation
1:3 %in% c(2, 3, 5) # set membership
```

Comparison (`<`, `<=`, `==`, `!=`) and logical operators (`!`, `&`, `|`, `&&`, `||`) return logical vectors:

```{r,echo=TRUE}
c(1, 5, 10) > 4
TRUE & c(TRUE, FALSE)
```

- Single `&`/`|` are vectorized; double `&&`/`||` examine one condition only and are reserved for `if()` tests.
- Users can define new binary operators named `%anything%`.

## Order of Operations

Operator precedence mostly matches school algebra — `^` before `*`/`/` before `+`/`-`, arithmetic before comparison before logic, assignment last:

```{r,echo=TRUE}
2 + 3 * 4 ^ 2
-2 ^ 2              # unary minus binds AFTER ^: this is -(2^2)
(-2) ^ 2
```

::: {.callout-tip}
**When in doubt, parenthesize.** Parentheses cost nothing and document intent; see `?Syntax` for the full precedence table.
:::

## Assignments

Several assignment forms exist:

```{r,echo=TRUE}
x <- 5              # the standard
6 -> y              # rightward (rare)
z = 7               # allowed at top level
assign("w", 8)      # assignment as a function call
c(x, y, z, w)
```

- Inside function calls, `=` names an argument, it does not assign — one more reason to standardize on `<-`.
- The superassignment operator `<<-` modifies a variable in an enclosing environment (Chapter 8); use sparingly.
- Replacement forms like `names(x) <- ...` and `dim(x) <- ...` look like magic but are ordinary function calls (`"names<-"`).

# Expressions

## Separating and Grouping Expressions

```{r,echo=TRUE}
a <- 1; b <- 2          # semicolons separate expressions on one line
{                       # braces group; value = last expression
  s <- a + b
  s * 10
}
```

- Parentheses both override precedence and return their contents visibly:

```{r,echo=TRUE}
(v <- c(1, 2, 3))       # assign AND print in one stroke
```

# Control Structures

## Conditional Statements

```{r,echo=TRUE}
x <- 7
if (x > 5) "big" else "small"

# vectorized version: one test per element
ifelse(c(1, 9, 3) > 5, "big", "small")
```

`switch()` selects among named alternatives:

```{r,echo=TRUE}
grade <- function(letter)
  switch(letter, A = "excellent", B = "good", "see me")
grade("A"); grade("C")
```

The unnamed final argument is the default case.

## Loops: for, while, repeat

```{r,echo=TRUE}
total <- 0
for (i in 1:5) total <- total + i
total

n <- 1
while (n < 100) n <- n * 2
n
```

- `repeat { }` loops forever until `break`; `next` skips to the following iteration.
- `for` iterates over any vector or list, not just numbers.

::: {.callout-note}
**Loops are often unnecessary.** Vectorized arithmetic (`x * 2`) and the apply family (`sapply`, `lapply`, Chapter 9) express most iterations more clearly — and usually faster.
:::

## Loop Accumulators Done Right

Growing an object inside a loop re-allocates memory at every step. Pre-allocate instead:

```{r,echo=TRUE}
squares <- numeric(10)       # pre-allocate
for (i in 1:10) squares[i] <- i^2
squares

# the idiomatic one-liner
sapply(1:10, function(i) i^2)
```

This habit matters enormously for performance (Chapter 24).

## Copyright Notice

- **Copyright.** These slides are adapted from *R in a Nutshell: A Desktop Quick Reference* (2nd ed.) by Joseph Adler, O'Reilly Media. All rights reserved by the original author and publisher.

- **Non-commercial use only.** These materials are strictly for educational purposes and may not be used for commercial gain.

- **Attribution.** Any reproduction, distribution, or use of these materials must properly credit the original source.
