---
title: "Programming for Applications"
subtitle: "Chapter 9: Functions"
author: "Yu-You Liou (NTU)"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "R in a Nutshell: A Desktop Quick Reference"
logo: "logo.svg"
editor: source
bibliography: bib.bib
suppress-bibliography: true
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch9-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center")
```

# **The Workhorses**

## The Function Keyword

Functions evaluate input arguments and return an output value. The definition syntax is

```{r,eval=FALSE,echo=TRUE}
function(arguments) body
```

where *arguments* is a set of symbol names (optionally with default values) available inside the body, and *body* is an R expression. Braces are customary but optional for a single expression — these two are equivalent:

```{r,echo=TRUE}
f <- function(x, y) x + y
f <- function(x, y) { x + y }
```

# Arguments

## Defaults Make Arguments Optional

Give an argument a default value and callers may omit it; omit a *required* argument and the error arrives only when the function tries to **use** it:

```{r,error=TRUE,echo=TRUE}
f <- function(x, y) { x + y }
f(1, 2)
g <- function(x, y = 10) { x + y }
g(1)
g(1, 2)     # overriding the default
f(1)        # missing y: error when `+` needs it
```

::: {.callout-note}
A function that simply never touches an uninitialized argument runs fine. You can even detect missing arguments yourself — e.g., inspecting `as.list(match.call())` for a `NULL` entry — but in practice, **put defaults in the signature**: it keeps functions clear and readable.
:::

## Variable-Length Argument Lists: ...

The ellipsis `...` means "all the other arguments." Two main uses. First, **passing extras through** to another function — here everything after `x` flows into `summary`:

```{r,echo=TRUE}
v <- c(sqrt(1:100))
f <- function(x, ...) { print(x); summary(...) }
f("Here is the summary for v.", v, digits = 2)
```

(Recall from Chapter 7 that `...` is a special object type, manipulable *only* inside a function body.)

## Reading the Arguments in ...

Second, you can **consume** the variable arguments yourself, by converting `...` to a list. A function that sums all its arguments:

```{r,echo=TRUE}
addemup <- function(x, ...) {
  args <- list(...)
  for (a in args) x <- x + a
  x
}
addemup(1, 1)
addemup(1, 2, 3, 4, 5)
```

Individual items are also reachable directly as `..1`, `..2`, ... up to `..9`. And named arguments remain valid symbols inside the body (scoping: Chapter 8).

## Return Values

An explicit `return` works as in other languages — but R simply returns the **last evaluated expression**, so `return` is commonly omitted:

```{r,echo=TRUE}
f <- function(x) { return(x^2 + 3) }
f(3)
f <- function(x) { x^2 + 3 }
f(3)
```

Use the explicit form when it makes the code cleaner — e.g., early exits.

# Functions as Arguments

## Higher-Order Functions: sapply

Many R functions accept *other functions* as arguments (modeling functions, for instance, take a function describing how to treat missing values). The classic example is `sapply`, which applies a function to every element of a vector:

```{r,echo=TRUE}
a <- 1:7
sapply(a, sqrt)
```

A toy example — `sqrt(1:7)` does the same — but many useful functions don't accept multi-element vectors, and `sapply` extends them gracefully. (The whole family of summarizing relatives: Chapter 12.)

## Anonymous Functions

Functions are objects, so they don't need names. A function passed without ever being named is **anonymous**. A minimal demonstration — a function that applies its argument to the number 3:

```{r,echo=TRUE}
apply.to.three <- function(f) { f(3) }
apply.to.three(function(x) { x * 7 })
```

How the interpreter reads it: `f` is bound to the anonymous `function(x) {x * 7}`; evaluating `f(3)` binds `x` to 3; the body returns `3 * 7 = 21`.

## Anonymous Functions with sapply

The everyday use: passing one-off functions to apply-style functions:

```{r,echo=TRUE}
a <- c(1, 2, 3, 4, 5)
sapply(a, function(x) { x + 1 })
```

This family is a real **alternative to control structures**. Squaring each element with a loop versus an apply:

```{r,echo=TRUE}
v <- 1:20
w <- NULL
for (i in 1:length(v)) { w[i] <- v[i]^2 }
w
w <- sapply(v, function(i) { i^2 })
w
```

The second version states its intent — apply this function to each element — and is faster, too (Chapter 24).

## Applying Anonymous Functions Directly

You can define an anonymous function and call it on the spot — but the function object **must be wrapped in parentheses**:

```{r,echo=TRUE}
(function(x) { x + 1 })(1)
```

Why? Function calls `f(arguments)` bind very tightly. Without the parentheses, `function(x) {x+1}(1)` quietly defines a function whose *body* is `{x+1}(1)` — no error appears, because the body isn't evaluated yet. Valid R, almost certainly not what you meant.

# Properties of Functions

## args and formals

To see what arguments a function accepts, `args` returns a function object with a `NULL` body:

```{r,echo=TRUE}
args(sin)
args(args)
args(lm)
```

To *manipulate* the argument list from code, use `formals`, which returns a **pairlist** — one pair per argument, value `NULL` where no default exists. (`formals` works only on closures, not built-ins.)

```{r,echo=TRUE}
f <- function(x, y=1, z=2) { x + y + z }
f.formals <- formals(f)
f.formals
f.formals$y
```

## Changing Arguments: formals<- and alist

`formals` also works on the **left-hand side** of an assignment — change a default in place:

```{r,echo=TRUE}
f.formals$y <- 3
formals(f) <- f.formals
args(f)
```

The convenience function `alist` builds an argument list using definition syntax (note: an argument with no default still needs its equals sign):

```{r,echo=TRUE}
formals(f) <- alist(x=, y=100, z=200)
f
```

## body

The body, too, is readable and writable — its type is `expression`, so assign expressions:

```{r,echo=TRUE}
body(f)
body(f) <- expression({x * y * z})
f
```

Functions are data, all the way down: arguments and body alike can be inspected and rewritten by ordinary R code.

# Argument Order and Named Arguments

## Three Ways to Supply Arguments

Inside a function, arguments are accessed by the names from the definition:

```{r,echo=TRUE}
addTheLog <- function(first, second) { first + log(second) }
```

Callers may specify arguments three ways, matched in priority order:

```{r,echo=TRUE}
addTheLog(second=exp(4), first=1)   # 1. exact names (matched first)
addTheLog(s=exp(4), f=1)            # 2. partially matching names
addTheLog(1, exp(4))                # 3. positional order
```

One restriction: with generic functions you cannot name the argument carrying the dispatching object — other arguments can still be named.

## Prefer Exact Names; Partial Matching Bites

Exact names cost keystrokes but buy readability and remove ambiguity. Partial matching is **deprecated** for good reason:

```{r,error=TRUE,echo=TRUE}
f <- function(arg1=10, arg2=20) {
  print(paste("arg1:", arg1))
  print(paste("arg2:", arg2))
}
f(arg=1)            # ambiguous: error
f(arg=1, arg2=2)    # now "arg" can only mean arg1...
f(arg=1, arg1=2)    # ...or suddenly it means arg2!
```

The same prefix lands on different arguments depending on what else is supplied — a recipe for silent bugs.

# Side Effects

## Beyond the Return Value

Every function returns a value; some *also* change variables in other environments, draw graphics, read or write files, or talk to the network. These extra actions are **side effects**.

The key side-effecting operator is `<<-`. The statement `var <<- value` searches the current environment for `var`, then the parent, recursively upward; if the global environment is reached without a match, the assignment happens **there**:

```{r,error=TRUE,echo=TRUE}
rm(list = ls())
doesnt.assign.x <- function(i) { x <- i }
doesnt.assign.x(4)
x                                   # <- stayed local: not found
assigns.x <- function(i) { x <<- i }
assigns.x(4)
x                                   # <<- escaped to the global environment
```

## I/O and Graphics

Two more families of side effects round out the picture:

- **Input/output.** Loading data from files or the Internet, saving data to disk — actions beyond returning an object. The full treatment is Chapter 11.
- **Graphics.** Plotting functions may return objects, but their point is the drawing — on screen or to a file. Chapters 13 and 14 cover them.

::: {.callout-tip}
**Design habit:** keep most functions *pure* (no side effects) and concentrate the side effects in a few clearly named places. Pure functions are easier to test, reuse, and parallelize.
:::

## Copyright Notice

- **Copyright.** These slides are adapted from *R in a Nutshell: A Desktop Quick Reference* (2nd ed.) by Joseph Adler, O'Reilly Media. All rights reserved by the original author and publisher.

- **Non-commercial use only.** These materials are strictly for educational purposes and may not be used for commercial gain.

- **Attribution.** Any reproduction, distribution, or use of these materials must properly credit the original source.
