---
title: "Programming for Applications"
subtitle: "Chapter 9: Functions"
author: "Yu-You Liou (NTU)"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "R in a Nutshell: A Desktop Quick Reference"
logo: "logo.svg"
editor: source
bibliography: bib.bib
suppress-bibliography: true
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch9-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center")
```

# **Building Your Own Tools**

## The Function Keyword

The `function` keyword creates a function object; assigning it to a name is optional:

```{r,echo=TRUE}
square <- function(x) x ^ 2
square(12)
```

A multi-expression body needs braces; the function's value is the last expression evaluated:

```{r,echo=TRUE}
zscore <- function(v) {
  m <- mean(v)
  s <- sd(v)
  (v - m) / s
}
round(zscore(c(2, 4, 6, 8)), 2)
```

## Arguments and Defaults

```{r,echo=TRUE}
trim_mean <- function(v, trim = 0.1, digits = 2) {
  round(mean(v, trim = trim), digits)
}
trim_mean(c(1, 2, 3, 100))            # defaults used
trim_mean(c(1, 2, 3, 100), trim = 0.25)
```

- Arguments match first by exact name, then by partial name, then by position.
- An argument with no default and no supplied value triggers an error **only if the body actually uses it** — lazy evaluation again.
- Defaults may reference other arguments: `function(x, n = length(x))`.

## The ... Argument

The ellipsis `...` absorbs any extra arguments and can forward them onward:

```{r,echo=TRUE}
loud_plot_summary <- function(v, ...) {
  cat("n =", length(v), "\n")
  summary(v, ...)
}
loud_plot_summary(c(1, 5, 9, NA), na.rm = TRUE)
```

This is how wrapper functions pass options through to `plot()`, `mean()`, and friends without naming every possibility.

## Return Values

```{r,echo=TRUE}
sign_word <- function(x) {
  if (x > 0) return("positive")     # explicit, early return
  if (x < 0) return("negative")
  "zero"                            # implicit: last value
}
sign_word(-3); sign_word(0)
```

- `return()` exits immediately — useful for guard clauses.
- Otherwise the last evaluated expression is the value; idiomatic R leans on this.
- `invisible(x)` returns a value without printing it at the console — what `assign()` and plotting functions do.

## Functions as Arguments

Higher-order functions take functions as inputs. The `sapply`/`lapply` family applies a function over a vector or list:

```{r,echo=TRUE}
sapply(1:5, sqrt)
lapply(list(a = 1:3, b = 4:6), mean)
```

`apply()` works over matrix margins (1 = rows, 2 = columns):

```{r,echo=TRUE}
m <- matrix(1:6, nrow = 2)
apply(m, 1, sum)     # row sums
apply(m, 2, max)     # column maxima
```

`mapply()` zips several vectors through one function in parallel.

## Anonymous Functions

When a helper is needed exactly once, skip the name:

```{r,echo=TRUE}
sapply(1:5, function(i) i * (i + 1) / 2)

# applied directly at the point of definition
(function(x) x ^ x)(3)
```

Since R 4.1 there is a compact backslash shorthand:

```{r,echo=TRUE}
sapply(1:5, \(i) i ^ 2)
```

Anonymous functions keep one-off logic exactly where it is used.

## Properties of Functions

A function object dissects into three parts:

```{r,echo=TRUE}
args(trim_mean)        # human-readable signature
formals(trim_mean)     # arguments + defaults, as a list
body(zscore)           # the code
environment(zscore)    # where it was defined
```

You can even *modify* these pieces programmatically — `body(f) <- ...` — which is occasionally useful and always impressive at parties.

## Argument Verification

Robust functions check their inputs early and fail informatively:

```{r,echo=TRUE,error=TRUE}
safe_log <- function(x) {
  stopifnot(is.numeric(x), all(x > 0))
  log(x)
}
safe_log(c(1, 10))
safe_log(-5)
```

- `stop("message")` raises an error; `warning()` raises a warning but continues.
- `missing(arg)` tests whether the caller supplied an argument.
- `match.arg()` validates an argument against a fixed set of choices.

## Copyright Notice

- **Copyright.** These slides are adapted from *R in a Nutshell: A Desktop Quick Reference* (2nd ed.) by Joseph Adler, O'Reilly Media. All rights reserved by the original author and publisher.

- **Non-commercial use only.** These materials are strictly for educational purposes and may not be used for commercial gain.

- **Attribution.** Any reproduction, distribution, or use of these materials must properly credit the original source.
