---
title: "Programming for Applications"
subtitle: "Chapter 8: Symbols and Environments"
author: "Yu-You Liou (NTU)"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "R in a Nutshell: A Desktop Quick Reference"
logo: "logo.svg"
editor: source
bibliography: bib.bib
suppress-bibliography: true
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch8-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center")
```

# **Where Names Live**

## Environments, Finally Defined

We have danced around environments since Chapter 3; time to define them. An **environment** is an R object containing:

- the set of **symbols** available in a given context,
- the **objects** associated with those symbols (together: the **frame**), and
- a pointer to a **parent environment**.

Every evaluation context has an environment. To resolve a symbol, R searches the current environment first, then **recursively searches parent environments** until it finds a match.

# Symbols

## Resolution Happens at Construction Time

Defining a variable assigns a symbol to a value *in an environment*. When an object is built from symbols, R resolves them **at construction time**:

```{r,echo=TRUE}
x <- 1; y <- 2; z <- 3
v <- c(x, y, z)
v
x <- 10      # v was already built; changing x changes nothing
v
```

## Delaying Evaluation: quote and eval

You can postpone symbol resolution by quoting the expression and evaluating it later — each `eval` resolves symbols *as of now*:

```{r,echo=TRUE}
x <- 1; y <- 2; z <- 3
v <- quote(c(x, y, z))
eval(v)
x <- 5
eval(v)
```

## Promises: delayedAssign

A **promise** delays evaluation until the variable is *first used*:

```{r,echo=TRUE}
x <- 1; y <- 2; z <- 3
delayedAssign("v", c(x, y, z))
x <- 5
v          # evaluated now — with the current x
```

Packages use promises to expose objects without loading them into memory. Two caveats: you cannot test whether an object *is* a promise, nor recover the environment where one was created.

# Working with Environments

## The Environment Toolbox

Environments are objects (internally, hash tables of symbol mappings — an efficiency trick exploited in Chapter 24). The vocabulary:

| Function | Description |
|---|---|
| `assign` / `get` / `exists` | bind, fetch, or test a name in an environment |
| `objects` | all names defined in an environment |
| `remove` (`rm`) | drop objects from an environment |
| `search` | names of attached packages — the chain of parent environments |
| `searchpaths` | paths of attached packages |
| `attach` / `detach` | add/remove a list, data frame, or data file to/from the search path |
| `emptyenv` | the empty environment — every chain ends here |
| `parent.env` | the parent of an environment |
| `baseenv` | the environment of the base package |
| `globalenv` / `.GlobalEnv` | the user's workspace |
| `environment` | a function's environment (no argument: the current one) |
| `new.env` | a fresh environment object |

## objects and rm

The two everyday calls — what's here, and clean it up:

```{r,echo=TRUE}
rm(list = ls())      # start with a clean slate
x <- 1; y <- 2; z <- 3
objects()
rm(x)
objects()
```

## The Global Environment

Each session gets a fresh environment for your objects: the **global environment**. Surprisingly, it is not the root of the tree — it is the *last* environment in the search path's chain. Walking the parents from `.GlobalEnv`:

```{r,echo=TRUE}
e <- .GlobalEnv
while (environmentName(e) != environmentName(emptyenv())) {
  print(environmentName(parent.env(e))); e <- parent.env(e)
}
```

Attached packages line up as parents, ending at `base` and finally the **empty environment** — the one environment with no parent, to which all chains lead.

# Environments and Functions

## A Function Call Creates an Environment

Calling a function creates a new environment; the arguments are bound to symbols there. (In language lingo: R is **lexically scoped**.)

```{r,echo=TRUE}
env.demo <- function(a, b, c, d) { print(objects()) }
env.demo(1, "truck", c(1,2,3,4,5), pi)
```

Only the arguments appear — everything else lives in parent environments. And note carefully: a function's parent environment is **where it was created**, not where it is called. For functions you define at the console the two coincide; for functions from packages they differ.

## The Call Stack

R also maintains a **stack** of calling environments (a stack: add and remove only at the top — push and pop, like cafeteria trays). Each call pushes an environment; each return pops one.

| Function | Description |
|---|---|
| `sys.call` | the current function call, as a language object |
| `sys.frame` | the calling environment |
| `sys.nframe` | number of the current frame (0 at the console) |
| `sys.function` | the function being evaluated |
| `sys.parent` | number of the parent frame |
| `sys.calls` / `sys.frames` / `sys.parents` | calls, environments, parents for the whole stack |
| `sys.on.exit` | the `on.exit` expression of the current frame |
| `sys.status` | `sys.calls` + `sys.parents` + `sys.frames` in one list |
| `parent.frame` | `sys.frame(sys.parent(n))` — the parent frame |

Why care? A package function sometimes needs a symbol's meaning **in the calling context**. Modeling functions do exactly this: a formula contains symbol names but no data, and when you omit the `data` argument, the function hunts for the variables in the calling environment.

## Evaluating in Another Environment: eval

`eval` evaluates an expression in an arbitrary environment:

```{r,eval=FALSE,echo=TRUE}
eval(expr, envir = parent.frame(),
     enclos = if (is.list(envir) || is.pairlist(envir))
                parent.frame() else baseenv())
```

The book's example: a timing wrapper that records the start time, evaluates its argument **in the caller's frame**, and prints the elapsed time:

```{r,echo=TRUE}
timethis <- function(...) {
  start.time <- Sys.time()
  eval(..., sys.frame(sys.parent(sys.parent())))
  end.time <- Sys.time()
  print(end.time - start.time)
}
```

## The Timing Demo

Time an inefficient function that grows a vector element by element (console transcript):

```{r,eval=FALSE,echo=TRUE}
create.vector.of.ones <- function(n) {
  return.vector <- NA
  for (i in 1:n) {
    return.vector[i] <- 1
  }
  return.vector
}
timethis(returned.vector <- create.vector.of.ones(10000))
## Time difference of 1.485959 secs
length(returned.vector)
## [1] 10000
```

The timing is nice; the *point* is the last line — the assignment happened **in the calling environment**, so `returned.vector` exists there afterward. (Aside: pre-allocating with `length(return.vector) <- n` before the loop cuts the time from ~1.5 s to ~0.04 s — a preview of Chapter 24.)

## Shorthands: evalq, eval.parent, local

Three convenient wrappers around `eval`:

- `evalq(expr, ...)` ≡ `eval(quote(expr), ...)` — quote for me;
- `eval.parent(expr, n)` ≡ `eval(expr, parent.frame(n))` — evaluate in the caller;
- `local(expr)` ≡ `eval(quote(expr), envir=new.env())` — evaluate in a fresh environment.

`eval.parent` shortens the timing function nicely:

```{r,echo=TRUE}
timethis.b <- function(...) {
  start.time <- Sys.time()
  eval.parent(...)
  end.time <- Sys.time()
  print(end.time - start.time)
}
```

## Data as Environment: with and within

Treating a data frame or list as an environment lets you use its element names like symbols:

```{r,error=TRUE,echo=TRUE}
example.list <- list(a=1, b=2, c=3)
a + b + c                          # not in the workspace...
with(example.list, a + b + c)      # ...but with() finds them
within(example.list, d <- a + b + c)
```

`with(data, expr)` evaluates and returns the **result**; `within(data, expr)` modifies and returns the **data**.

## attach and detach

`attach` adds the objects of a "database" — a list, data frame, or saved data file — to the search path; `detach` removes them:

```{r,eval=FALSE,echo=TRUE}
attach(what, pos = 2, name = deparse(substitute(what)),
       warn.conflicts = TRUE)
detach(name, pos = 2, unload = FALSE)
```

`pos` is the search-path position, `name` labels the attached database (used again by `detach`), `warn.conflicts` warns about masked names, and `unload` controls namespace unloading.

::: {.callout-warning}
**Use attach sparingly.** With several data frames sharing column names, it becomes genuinely hard to tell which object a name refers to. Prefer `transform` to change columns and `with` to evaluate expressions against a data frame.
:::

# Exceptions

## Errors and Warnings

Invalid expressions produce **errors**; questionable ones produce **warnings**:

```{r,error=TRUE,echo=TRUE}
12 / "hat"
if (c(TRUE, FALSE)) TRUE else FALSE
```

(The second was merely a warning in the book's era; modern R promotes it to an error.) Why discuss exceptions in the environments chapter? Because when an exception occurs, the interpreter may abandon the current function and **signal the exception in the calling environment** — exception handling and environments are intertwined.

## Signaling: stop, warning, message

Three levels of telling the user something, from fatal to informative:

```{r,error=TRUE,echo=TRUE}
doWork <- function(filename) {
  if (file.exists(filename)) read.delim(filename)
  else stop("Could not open the file: ", filename)
}
doWork("file that doesn't exist")
```

```{r,warning=TRUE,echo=TRUE}
doNoWork <- function(filename) {
  if (file.exists(filename)) "la la la"
  else warning("File does not exist: ", filename)
}
doNoWork("another file that doesn't exist")
```

```{r,message=TRUE,echo=TRUE}
doNothing <- function(x) {
  message("This function does nothing.")
}
doNothing("another input value")
```

In your own programs: `stop` on errors that invalidate the work; `warning` for problems the caller should know about; `message` for chatter.

## Catching: try

Suppose `foo` calls `bar`, `bar` sometimes fails (can't open a file, say), and `foo` wants to *survive* and try something else. The simple tool is `try`, which hides most of the machinery:

```{r,echo=TRUE}
res <- try({x <- 1}, silent=TRUE)
res
res <- try({open("file that doesn't exist")}, silent=TRUE)
res
class(res)
```

`try(expr, silent)` evaluates `expr`; on failure it returns an object of class `"try-error"` (and prints the message unless `silent=TRUE`).

## Catching: tryCatch

The capable tool is `tryCatch`, taking an expression, a set of **handlers**, and a final expression:

```{r,eval=FALSE,echo=TRUE}
tryCatch(expression, handler1, handler2, ..., finally=finalexpr)
```

Evaluation order: R evaluates `expression`; if a **condition** (error or warning) arises, R picks the handler whose argument matches the condition's class; afterward `finally` is evaluated — with the handlers no longer active. A taste:

```{r,echo=TRUE}
tryCatch({
  log(-1)
}, warning = function(w) "caught a warning",
   finally = message("done either way"))
```

## Copyright Notice

- **Copyright.** These slides are adapted from *R in a Nutshell: A Desktop Quick Reference* (2nd ed.) by Joseph Adler, O'Reilly Media. All rights reserved by the original author and publisher.

- **Non-commercial use only.** These materials are strictly for educational purposes and may not be used for commercial gain.

- **Attribution.** Any reproduction, distribution, or use of these materials must properly credit the original source.
