---
title: "Programming for Applications"
subtitle: "Chapter 8: Symbols and Environments"
author: "Yu-You Liou (NTU)"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "R in a Nutshell: A Desktop Quick Reference"
logo: "logo.svg"
editor: source
bibliography: bib.bib
suppress-bibliography: true
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch8-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center")
```

# **Where Names Live**

## Symbols and Bindings

Defining a variable really means **binding a symbol to a value inside an environment**:

```{r,echo=TRUE}
x <- 1            # bind symbol `x` to 1 in the current environment
y <- quote(x + 1) # an unevaluated expression mentioning `x`
eval(y)           # evaluate it: `x` is looked up NOW
x <- 100
eval(y)           # same expression, new answer
```

An expression is data until evaluated; the value of its symbols is decided at evaluation time, in some environment.

## Delayed Evaluation and Promises

Evaluation can be postponed deliberately:

```{r,echo=TRUE}
delayedAssign("lazy", {message("computing now!"); 42})
# nothing printed yet ...
lazy             # first touch forces the promise
```

::: {.callout-note}
**Promises.** A promise packages an expression plus the environment to evaluate it in, and computes its value only when first needed. R passes *all* function arguments as promises — that is "lazy evaluation," and it is why an unused argument can contain an error without crashing the call.
:::

## What Is an Environment?

An **environment** is a set of symbol-to-value bindings plus a pointer to a **parent environment**. Lookup walks the chain: current environment first, then parent, then grandparent, until `emptyenv()`.

```{r,echo=TRUE}
e <- new.env()
assign("z", 99, envir = e)
get("z", envir = e)
exists("z", envir = e)
ls(envir = e)
```

Other tools: `rm(..., envir=)` removes bindings; `environmentName()` labels well-known environments.

## The Global Environment

Each session starts in the **global environment** — your workspace:

```{r,echo=TRUE}
environment()             # where am I?
ls()[1:5]                 # what have I defined?
```

The chain above the global environment consists of the attached packages — exactly the **search path**:

```{r,echo=TRUE}
search()
```

This is why loading a package makes its functions visible: it inserts an environment into the lookup chain.

## Environments and Functions

Calling a function creates a *fresh* environment for its local variables, whose parent is the environment where the function was **defined** (not where it was called):

```{r,echo=TRUE}
make_counter <- function() {
  count <- 0
  function() {
    count <<- count + 1   # write to the enclosing environment
    count
  }
}
tick <- make_counter()
tick(); tick(); tick()
```

The inner function *remembers* `count` between calls — a **closure**. This lexical scoping rule is the heart of R's function semantics.

## Global vs. Lexical Assignment

```{r,echo=TRUE}
v <- "global"
f <- function() { v <- "local"; v }
g <- function() { v <<- "overwritten"; v }
f(); v          # local assignment leaves global alone
g(); v          # <<- climbed up and replaced it
```

::: {.callout-warning}
**Use `<<-` sparingly.** Functions that silently modify state outside themselves are hard to test and debug. Closures (previous slide) are the disciplined use case.
:::

## Working with the Call Stack

The environment where a function was *called* is reachable too — essential for functions that interpret expressions in their caller's context:

```{r,echo=TRUE}
whodunit <- function() sys.function(-1)   # who called me?
show_caller_x <- function() {
  eval(quote(x), envir = parent.frame())
}
x <- "caller's x"
show_caller_x()
```

- `sys.call()` — the current call; `match.call()` — with argument names filled in.
- `parent.frame()` — the calling environment. Functions like `lm()` use exactly this trick to evaluate formulas against your data.

## Environments in Packages

- Each loaded package contributes an environment to the search path; lookup explains shadowing — a name in your workspace hides one in a package.
- Disambiguate explicitly with the `::` operator:

```{r,echo=TRUE}
stats::sd(c(1, 2, 3, 4))
```

- Packages use **namespaces** so their internal helpers stay invisible, and *promises* (via lazy data loading) so big data sets cost nothing until touched.

## Copyright Notice

- **Copyright.** These slides are adapted from *R in a Nutshell: A Desktop Quick Reference* (2nd ed.) by Joseph Adler, O'Reilly Media. All rights reserved by the original author and publisher.

- **Non-commercial use only.** These materials are strictly for educational purposes and may not be used for commercial gain.

- **Attribution.** Any reproduction, distribution, or use of these materials must properly credit the original source.
