---
title: "Programming for Applications"
subtitle: "Chapter 18: Statistical Tests"
author: "Yu-You Liou (NTU)"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "R in a Nutshell: A Desktop Quick Reference"
logo: "logo.svg"
editor: source
bibliography: bib.bib
suppress-bibliography: true
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch18-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center")
```

# **Asking Yes/No Questions of Data**

## The Logic of a Test

Every classical test follows one script: state a **null hypothesis** $H_0$, compute a statistic whose distribution under $H_0$ is known (Chapter 17), and report a **p-value** — the probability of a result at least this extreme if $H_0$ were true.

::: {.callout-note}
**Reading a p-value.** Small $p$ (conventionally $< 0.05$) means the data are surprising under $H_0$, so we reject it. A large $p$ never *proves* $H_0$; it only says the data are compatible with it.
:::

Every test below returns an object of class `htest` — same structure, same reading skills.

## One-Sample t-Test

Is the mean tire-test score different from a benchmark?

```{r,echo=TRUE,cache=TRUE}
temp <- tempfile(fileext = ".rda")
download.file("https://raw.githubusercontent.com/cran/nutshell/master/data/tires.sus.rda", temp, mode = "wb")
load(temp)
spd <- tires.sus$Time_To_Failure[tires.sus$Tire_Type == "E"]
t.test(spd, mu = 8)
```

The output bundles the statistic, df, p-value, confidence interval, and sample mean — extract pieces with `$p.value`, `$conf.int`.

## Two-Sample and Paired t-Tests

```{r,echo=TRUE}
b <- tires.sus$Time_To_Failure[tires.sus$Tire_Type == "B"]
d <- tires.sus$Time_To_Failure[tires.sus$Tire_Type == "D"]
t.test(b, d)                      # Welch two-sample by default
```

```{r,echo=TRUE}
before <- c(198, 201, 210, 185, 192)
after  <- before - c(3, 7, 2, 9, 1)
t.test(before, after, paired = TRUE)
```

Pairing removes between-subject variation — when measurements come in natural pairs, always say so.

## Comparing Variances and Many Means

```{r,echo=TRUE}
var.test(b, d)                    # F test for equal variances
```

For more than two groups, one-way **ANOVA** generalizes the t-test:

```{r,echo=TRUE}
fit <- aov(Time_To_Failure ~ Tire_Type, data = tires.sus)
summary(fit)
```

A small p-value says *some* means differ; `TukeyHSD(fit)` then reports *which* pairs.

## Normality and Distribution Tests

```{r,echo=TRUE}
set.seed(3); z <- rnorm(100)
shapiro.test(z)                       # H0: data are normal
ks.test(z, "pnorm")                   # against any named distribution
```

```{r,echo=TRUE,out.width="55%"}
qqnorm(z); qqline(z)
```

The Q–Q plot is the visual companion: points hugging the line support normality — and it shows *where* departures occur.

## Non-Parametric Tests

When normality is doubtful or data are ranks, swap in distribution-free analogues:

| Parametric | Non-parametric |
|:---|:---|
| one/two-sample t | `wilcox.test()` |
| one-way ANOVA | `kruskal.test()` |
| paired t | `wilcox.test(..., paired = TRUE)` |

```{r,echo=TRUE}
wilcox.test(b, d)
kruskal.test(Time_To_Failure ~ Tire_Type, data = tires.sus)
```

They trade a little power for a lot of robustness.

## Proportion and Binomial Tests

Field-goal kickers succeeded 787 times in 974 attempts. Is the true success rate 80%?

```{r,echo=TRUE}
prop.test(787, 974, p = 0.80)
binom.test(787, 974, p = 0.80)     # exact version
```

`prop.test()` with two counts compares two proportions — the A/B test in one line.

## Tabular Data: Chi-Squared and Friends

Do two categorical variables associate?

```{r,echo=TRUE}
tbl <- matrix(c(45, 15, 30, 40), nrow = 2,
              dimnames = list(group = c("treat", "control"),
                              outcome = c("yes", "no")))
tbl
chisq.test(tbl)
fisher.test(tbl)        # exact; safe for small cells
```

- Rule of thumb: expected counts below 5 → prefer `fisher.test()`.
- `mcnemar.test()` handles paired categorical data (before/after on the same subjects).

## Choosing a Test: a Map

1. **One numeric sample** vs. a reference: `t.test(x, mu=)`; non-normal → `wilcox.test`.
2. **Two numeric samples**: independent → `t.test(x, y)`; paired → `paired=TRUE`; non-normal → `wilcox.test`.
3. **3+ groups**: `aov()` → `TukeyHSD()`; non-normal → `kruskal.test`.
4. **Proportions**: `prop.test` / `binom.test`.
5. **Two categorical variables**: `chisq.test` / `fisher.test`.

Check assumptions (normality, equal variance, independence) *before* trusting p-values — and remember that statistical significance is not practical importance.

## Copyright Notice

- **Copyright.** These slides are adapted from *R in a Nutshell: A Desktop Quick Reference* (2nd ed.) by Joseph Adler, O'Reilly Media. All rights reserved by the original author and publisher.

- **Non-commercial use only.** These materials are strictly for educational purposes and may not be used for commercial gain.

- **Attribution.** Any reproduction, distribution, or use of these materials must properly credit the original source.
