---
title: "Programming for Applications"
subtitle: "Chapter 18: Statistical Tests"
author: "Yu-You Liou (NTU)"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "R in a Nutshell: A Desktop Quick Reference"
logo: "logo.svg"
editor: source
bibliography: bib.bib
suppress-bibliography: true
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch18-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center", fig.height = 4.2)
```

# **Hypotheses on Trial**

## The Shape of the Chapter

"Does the new drug beat the placebo?" "Does the redesign sell more?" "Does this strategy beat the index?" — formulate a hypothesis, design an experiment, collect data, test. Two halves:

- **continuous data** — normal-theory tests, then nonparametric counterparts;
- **discrete data** — proportions, binomial trials, contingency tables.

::: {.callout-warning}
**The book's own disclaimer applies here too:** these slides remind you *when* and *how* to run each test in R — they are no substitute for a proper statistics course on where the formulas come from and when they're safe.
:::

```{r,echo=TRUE,results='hide'}
# data sets used throughout this chapter
for (d in c("tires.sus", "field.goals", "SPECint2006",
            "mort06.smpl", "births2006.smpl")) {
  temp <- tempfile(fileext = ".rda")
  download.file(paste0("https://raw.githubusercontent.com/cran/nutshell/",
                       "master/data/", d, ".rda"), temp, mode = "wb")
  load(temp, envir = globalenv())
}
```

# Continuous Data: Normal-Theory Tests

## t.test: One Sample Against a Hypothesized Mean

`t.test(x, y=NULL, alternative=c("two.sided","less","greater"), mu=0, paired=FALSE, var.equal=FALSE, conf.level=0.95)` — `y=NULL` compares one vector against `mu`; `var.equal` chooses pooled variance vs. the Welch method. Suppose type-H tires "should" last 9 hours:

```{r,echo=TRUE}
times.to.failure.h <- subset(tires.sus,
  Tire_Type=="H" & Speed_At_Failure_km_h==160)$Time_To_Failure
mean(times.to.failure.h)
t.test(times.to.failure.h, mu=9)
```

Reading it: statistic `t`, degrees of freedom, **p-value 0.4684** — the alternative ("true mean ≠ 9") is *not* supported; the evidence does not imply the mean differs from 9. The 95% CI and sample mean close the report.

## t.test: Two Samples

Three tire types shared speed rating S — equal lifetimes expected. E vs. D, then E vs. B:

```{r,echo=TRUE}
times.to.failure.e <- subset(tires.sus,
  Tire_Type=="E" & Speed_At_Failure_km_h==180)$Time_To_Failure
times.to.failure.d <- subset(tires.sus,
  Tire_Type=="D" & Speed_At_Failure_km_h==180)$Time_To_Failure
times.to.failure.b <- subset(tires.sus,
  Tire_Type=="B" & Speed_At_Failure_km_h==180)$Time_To_Failure
t.test(times.to.failure.e, times.to.failure.d)
t.test(times.to.failure.e, times.to.failure.b)$p.value
```

E vs. D: significant at 95% (p = 0.034). E vs. B: not (p = 0.164) — and note R never *says* "significant"; you read the p-value yourself.

## t.test: the Formula Interface

With data in a frame and groups in a factor: `t.test(formula, data, subset, na.action)`. TV pundits insist outdoor kicking is harder — do successful field goals differ in distance?

```{r,echo=TRUE}
good <- transform(
  field.goals[field.goals$play.type=="FG good", c("yards","stadium.type")],
  outside=(stadium.type=="Out"))
bad <- transform(
  field.goals[field.goals$play.type=="FG no", c("yards","stadium.type")],
  outside=(stadium.type=="Out"))
t.test(yards~outside, data=good)
```

A yard longer indoors on average — not significant (p = 0.26). Misses (`data=bad`): p = 0.23. Attempts overall: p = 0.12. The pundits remain unvindicated.

## Paired Data

Two observations per subject (before/after) call for a **paired t-test**: `paired=TRUE`. SPEC's CPU benchmarks report a *baseline* and an *optimized* result per system — paired by machine. Single-chip dual-core systems:

```{r,echo=TRUE}
t.test(subset(SPECint2006, Num.Chips==1 & Num.Cores==2)$Baseline,
       subset(SPECint2006, Num.Chips==1 & Num.Cores==2)$Result,
       paired=TRUE)
```

Resoundingly significant — no shock: tuning helps, and submitting optimized results is voluntary.

## Comparing Variances: var.test and bartlett.test

- **`var.test(x, y, ratio=1, alternative=, conf.level=)`** (or formula form) runs an **F-test** on two normal-population variances.
- **`bartlett.test(x, g)`** (or formula) tests homogeneity of variance across groups.

```{r,echo=TRUE}
field.goals.inout <- transform(field.goals, outside=(stadium.type=="Out"))
var.test(yards~outside, data=field.goals.inout)
bartlett.test(yards~outside, data=field.goals.inout)
```

Both p-values ≈ 0.03: the indoor/outdoor variance difference *is* significant — the means weren't, the spreads are.

## ANOVA: Comparing Many Means

**Analysis of variance** compares means across multiple groups (named for the method, not the result!). The quick tool is `aov(formula, data, projections=, qr=, contrasts=)`. Age at death by cause, US 2006 (`mort06.smpl`, recoded by the author from CDC files):

```{r,echo=TRUE}
aov(age~Cause, data=mort06.smpl)
```

`model.tables(aov(...))` prints per-level effects (Accidents −21.4 years vs. overall, Homicide −40.1, Alzheimer's +13.8...). A subtler example — weight gain by birth month:

```{r,echo=TRUE}
births2006.cln <- births2006.smpl[births2006.smpl$WTGAIN<99 &
                                  !is.na(births2006.smpl$WTGAIN),]
aov(WTGAIN~DOB_MM, births2006.cln)
```

## anova(lm(...)), oneway.test, and Friends

The often-better route: fit with `lm`, extract the table with `anova` — F-statistic and p-value included (and `update` modifies big models cheaply; Chapter 20):

```{r,echo=TRUE}
mort06.smpl.lm <- lm(age~Cause, data=mort06.smpl)
anova(mort06.smpl.lm)
```

- Variances unequal? `oneway.test(formula, data, var.equal=FALSE)` — the Welch idea, generalized.
- More `aov` accessories: `proj` (projections), `TukeyHSD` (CIs on pairwise level differences), `se.contrast` (standard errors of contrasts).

## Pairwise t-Tests

Want *which* groups differ, not just *whether*: `pairwise.t.test(x, g, p.adjust.method=, pool.sd=, paired=, alternative=)` runs all pairs with multiple-testing correction (default: Holm):

```{r,echo=TRUE}
pairwise.t.test(tires.sus$Time_To_Failure, tires.sus$Tire_Type)
```

No significant difference for C–L, D–L, D–E; very significant for B–H, C–H, E–L...

## Testing Normality: shapiro.test

Are field-goal distances normal? Eyeball first, then the **Shapiro-Wilk test**:

```{r,echo=TRUE}
par(mfcol=c(1, 2), ps=6.5)
hist(field.goals$yards, breaks=25)
qqnorm(field.goals$yards, pch=".")
par(mfcol=c(1, 1))
shapiro.test(field.goals$yards)
```

Bell-ish shape, roughly linear Q-Q — yet p ≈ 10⁻¹²: quite likely *not* normal. Eyes deceive; tests quantify.

## Arbitrary Distributions: ks.test

The **Kolmogorov-Smirnov test** compares a vector to *any* distribution (`y` = data vector, distribution name, or function), or two vectors to each other:

```{r,echo=TRUE,warning=TRUE}
ks.test(field.goals$yards, pnorm)
ks.test(jitter(subset(SPECint2006, Num.Chips==1&Num.Cores==2)$Baseline),
        jitter(subset(SPECint2006, Num.Chips==1&Num.Cores==2)$Result))
```

The ties warning is itself informative — true normal values almost never tie. (We `jitter` the second test to silence it.) Baseline and optimized results: unlikely to share a distribution (p = 0.012).

## Correlation Tests: cor.test

Chapter 16's `cor` gives a number; `cor.test(x, y, alternative=, method=, exact=, conf.level=)` (or formula form) gives **significance**:

```{r,echo=TRUE}
cor.test(c(1,2,3,4,5,6,7,8), c(0,2,4,6,8,10,11,14))$p.value
cor.test(c(1,2,3,4,5,6,7,8), c(5,3,8,1,7,0,0,3))$p.value
```

The Chapter 13 toxins-and-lung-cancer relationship, tested:

```{r,echo=TRUE}
temp <- tempfile(fileext = ".rda")
download.file("https://raw.githubusercontent.com/cran/nutshell/master/data/toxins.and.cancer.rda",
              temp, mode = "wb")
load(temp)
with(toxins.and.cancer, cor.test(air_on_site/Surface_Area,
                                 deaths_lung/Population))
```

Significant positive correlation (p = 0.0015) — but resist the causal leap: industrial states may differ in income, healthcare, smoking, or even death-certificate habits.

# Continuous Data: Nonparametric Tests

## wilcox.test

When normality is doubtful, **rank-based** tests trade some power for robustness. The Wilcoxon test is the t-test's counterpart: its statistic counts, over all pairs, how often y[j] < x[i] (about half if the distributions match) — no one-sample-vs-mean version exists:

```{r,echo=TRUE,warning=TRUE}
wilcox.test(times.to.failure.e, times.to.failure.d)
wilcox.test(yards~outside, data=good)
```

Note the contrast with the t-test: E vs. D *barely misses* significance here (p = 0.0505). Ties trigger a normal approximation (hence the warning).

## kruskal.test and fligner.test

- **Kruskal-Wallis rank-sum test** — nonparametric ANOVA: `kruskal.test(formula, data)`;
- **Fligner-Killeen (median) test** — nonparametric variance comparison: `fligner.test(formula, data)`.

```{r,echo=TRUE}
kruskal.test(age~Cause, data=mort06.smpl)
fligner.test(age~Cause, data=mort06.smpl)
```

- Scale-parameter differences: **`ansari.test(x, y, ...)`** (Ansari-Bradley) and **`mood.test(x, y, ...)`** (Mood) — both with formula variants.

# Discrete Data

## prop.test: Comparing Proportions

`prop.test(x, n, p=NULL, alternative=, conf.level=, correct=)` tests whether success proportions differ across groups (or equal given values). Is field-goal *success* equal indoors and out? Build a successes/failures table (dropping the 8 aborted and 24 blocked kicks), transpose, test:

```{r,echo=TRUE}
field.goals.goodbad <- field.goals[field.goals$play.type=="FG good" |
                                   field.goals$play.type=="FG no", ]
field.goals.table <- table(field.goals.goodbad$play.type,
                           field.goals.goodbad$stadium.type)
field.goals.table.t <- t(field.goals.table[3:4,])
field.goals.table.t
prop.test(field.goals.table.t)
```

p = 0.31: no significant difference in success rates among Both/In/Out.

## binom.test: Bernoulli Trials

A series of identical two-outcome trials (coin flips) follows the **binomial** distribution. `binom.test(x, n, p=0.5, alternative=, conf.level=)` tests an observed success count. David Ortiz hit .264 (110/416) in 2008 — if he were "truly" a .300 hitter, how likely is hitting .264 or worse?

```{r,echo=TRUE}
binom.test(x=110, n=416, p=0.3, alternative="less")
```

p = 0.062: about a 6% chance. Notable: this p-value is the probability of being *at least as far* in the stated direction — read `alternative` carefully.

## fisher.test: Exact Independence Testing

For contingency tables, the hypothesis is **independence** of the two variables. For small tables, **Fisher's exact test** is best. Arguments: `x` (matrix or factor), `y` (factor when `x` is), `workspace`, `hybrid` (approximate for >2×2), `or` (hypothesized odds ratio), `alternative`, `conf.int`/`conf.level`, `simulate.p.value`/`B` (Monte Carlo). Delivery method × sex, July 2006:

```{r,echo=TRUE}
births.july.2006 <- births2006.smpl[births2006.smpl$DMETH_REC!="Unknown" &
                                    births2006.smpl$DOB_MM==7, ]
method.and.sex <- table(births.july.2006$SEX,
  as.factor(as.character(births.july.2006$DMETH_REC)))
method.and.sex
fisher.test(method.and.sex)
```

Tiny p-value: reject independence — C-section rates differ by sex (46.7% vs. 49.2% of each sex's births... or rather, the odds ratio ≈ 0.91).

## fisher.test on Twins; chisq.test

For twins only, the story reverses:

```{r,echo=TRUE}
twins.2006 <- births2006.smpl[births2006.smpl$DPLURAL=="2 Twin" &
                              births2006.smpl$DMETH_REC != "Unknown",]
method.and.sex.twins <- table(twins.2006$SEX,
  as.factor(as.character(twins.2006$DMETH_REC)))
fisher.test(method.and.sex.twins)$p.value
```

p = 0.67: independence is plausible. Fisher's test gets expensive for large tables — the workhorse alternative is the **chi-squared test**, which compares the sample to hypothesized probabilities (`chisq.test(x, y=, correct=, p=, rescale.p=, simulate.p.value=, B=)`):

```{r,echo=TRUE}
chisq.test(method.and.sex.twins)
chisq.test(twins.2006$DMETH_REC, twins.2006$SEX)$p.value   # same, from factors
```

## chisq.test: Goodness of Fit, Higher Dimensions

Births by weekday — could the weekend deficit be chance, if every day were equally likely (the default `p`)?

```{r,echo=TRUE}
births2006.byday <- table(births2006.smpl$DOB_WK)
births2006.byday
chisq.test(births2006.byday)
```

Emphatically not chance (though with n this big, *everything* is significant). Multidimensional tables work too — day × month:

```{r,echo=TRUE}
chisq.test(table(births2006.smpl$DOB_WK, births2006.smpl$DOB_MM))
```

- Three-way interactions: **`mantelhaen.test(x, y, z, ...)`** (Cochran-Mantel-Haenszel).
- Symmetry of a 2-D table: **`mcnemar.test(x, y, correct=)`**.

## friedman.test

The **Friedman rank-sum test** is the nonparametric counterpart of two-way ANOVA — `friedman.test(y, groups, blocks)` or the formula/table forms:

```{r,echo=TRUE}
friedman.test(method.and.sex.twins)
```

Agreeing with the chi-squared verdict: the two distributions are very likely independent.

## Copyright Notice

- **Copyright.** These slides are adapted from *R in a Nutshell: A Desktop Quick Reference* (2nd ed.) by Joseph Adler, O'Reilly Media. All rights reserved by the original author and publisher.

- **Non-commercial use only.** These materials are strictly for educational purposes and may not be used for commercial gain.

- **Attribution.** Any reproduction, distribution, or use of these materials must properly credit the original source.
