---
title: "Programming for Applications"
subtitle: "Chapter 19: Power Tests"
author: "Yu-You Liou (NTU)"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "R in a Nutshell: A Desktop Quick Reference"
logo: "logo.svg"
editor: source
bibliography: bib.bib
suppress-bibliography: true
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch19-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center")
```

# **Designing Experiments**

## Power: the Missing Quantity

A test can fail two ways: rejecting a true $H_0$ (**Type I error**, probability $\alpha$) or missing a real effect (**Type II error**, probability $\beta$).

::: {.callout-note}
**Power** $= 1 - \beta$: the probability of detecting an effect that truly exists. Four quantities are locked together — sample size $n$, significance level $\alpha$, effect size, and power. Fix any three and the fourth is determined.
:::

Power analysis answers the planning question every experiment faces: **how many observations do I need?**

## power.t.test()

Sample size for detecting a 5-unit difference (sd = 10) with 80% power:

```{r,echo=TRUE}
power.t.test(delta = 5, sd = 10, sig.level = 0.05, power = 0.80)
```

About 64 subjects *per group*. Leave out any one argument and R solves for it:

```{r,echo=TRUE}
power.t.test(n = 30, delta = 5, sd = 10)$power
```

With only 30 per group, power falls below 50% — a coin flip's chance of finding the effect.

## power.prop.test()

Comparing two proportions — does a change lift conversion from 10% to 12%?

```{r,echo=TRUE}
power.prop.test(p1 = 0.10, p2 = 0.12, power = 0.80)
```

Thousands of observations per arm: small effects on small baselines are *expensive*. This single calculation prevents the most common A/B-testing mistake — calling a test after too few visitors.

## power.anova.test()

For comparing several group means:

```{r,echo=TRUE}
power.anova.test(groups = 4,
                 between.var = 1, within.var = 9,
                 power = 0.80)
```

- `between.var` / `within.var` encode the effect size: how far apart the group means are relative to the noise.
- More groups at fixed total $n$ means fewer observations per group — and less power.

## Effect Sizes and Curves

Effect size is the input you must *guess* — from pilot data, past studies, or the smallest difference that would matter. Plot power across the plausible range:

```{r,echo=TRUE,out.width="65%"}
deltas <- seq(1, 10, 0.5)
pw <- sapply(deltas, function(d)
  power.t.test(n = 50, delta = d, sd = 10)$power)
plot(deltas, pw, type = "l", lwd = 2,
     xlab = "True difference", ylab = "Power (n = 50/group)")
abline(h = 0.8, lty = 2)
```

## Power by Simulation

When no canned formula fits your design, simulate it (Chapter 17 toolkit):

```{r,echo=TRUE}
set.seed(9)
one_run <- function(n, delta) {
  t.test(rnorm(n), rnorm(n, mean = delta))$p.value < 0.05
}
mean(replicate(2000, one_run(64, 5/10 * 10)))   # matches power.t.test
```

Replace `rnorm()` with your actual design — unequal groups, skewed errors, clustered data — and the same loop still measures power.

## Planning Checklist

1. Decide $\alpha$ (usually 0.05) and target power (usually 0.80+).
2. Specify the **smallest effect worth detecting** — a scientific judgment, not a statistical one.
3. Estimate the noise (sd or baseline proportion).
4. Compute $n$ with `power.*.test()` or simulation.
5. Add margin for dropouts — then collect data *before* peeking at p-values.

Underpowered studies waste effort coming and going: they miss real effects, and their occasional "significant" findings exaggerate them.

## Copyright Notice

- **Copyright.** These slides are adapted from *R in a Nutshell: A Desktop Quick Reference* (2nd ed.) by Joseph Adler, O'Reilly Media. All rights reserved by the original author and publisher.

- **Non-commercial use only.** These materials are strictly for educational purposes and may not be used for commercial gain.

- **Attribution.** Any reproduction, distribution, or use of these materials must properly credit the original source.
