---
title: "Programming for Applications"
subtitle: "Chapter 19: Power Tests"
author: "Yu-You Liou (NTU)"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "R in a Nutshell: A Desktop Quick Reference"
logo: "logo.svg"
editor: source
bibliography: bib.bib
suppress-bibliography: true
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch19-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center")
```

# **Designing Experiments**

## How Much Data Do You Need?

Before collecting data, two questions decide an experiment's fate:

- how much data is needed for a statistically significant result?
- given the data you *can* collect, how strong a result is even detectable?

R's `power.*` family answers both. The vocabulary: **significance level** = Type I error probability (false positive); **power** = 1 − Type II error probability (1 − false negative). Each function computes whichever parameter you leave as `NULL`.

## A Worked Example: Designing a Drug Trial

Testing an antidepressant, with depression measured on the HAMD scale (0–48). Assume sd = 8.9 (from FDA materials), significance 0.05, desired power 0.95. With 25 subjects per arm, what group difference would be detectable?

```{r,echo=TRUE}
power.t.test(power=.95, sig.level=.05, sd=8.9, n=25)
```

A difference of at least **9.26** HAMD points. Double the subjects:

```{r,echo=TRUE}
power.t.test(power=.95, sig.level=.05, sd=8.9, n=50)
```

Now **6.48** suffices. This is experiment design in action: trade sample size against detectable effect *before* spending a dollar on data collection.

# t-Test Design

## power.t.test

```{r,eval=FALSE,echo=TRUE}
power.t.test(n = NULL, delta = NULL, sd = 1, sig.level = 0.05,
             power = NULL,
             type = c("two.sample", "one.sample", "paired"),
             alternative = c("two.sided", "one.sided"),
             strict = FALSE)
```

| Argument | Description |
|---|---|
| `n` | observations *per group* |
| `delta` | true difference in means |
| `sd` | true standard deviation |
| `sig.level` | significance level (Type I error probability) |
| `power` | 1 − Type II error probability |
| `type` | two-sample / one-sample / paired |
| `alternative` | two-sided or one-sided |
| `strict` | strict interpretation in the two-sided case |

Supply **four** of `n`, `delta`, `sd`, `sig.level`, `power`; leave the fifth `NULL` — that's what gets computed.

# Proportion Test Design

## power.prop.test

For experiments measured with `prop.test`:

```{r,eval=FALSE,echo=TRUE}
power.prop.test(n = NULL, p1 = NULL, p2 = NULL, sig.level = 0.05,
                power = NULL,
                alternative = c("two.sided", "one.sided"),
                strict = FALSE)
```

`p1`/`p2` are the success probabilities in the two groups; the rest as before. Again: four of `n`, `p1`, `p2`, `sig.level`, `power` given, the `NULL` one computed.

## Are Situational Baseball Stats Nonsense?

ESPN flashes stats like "3-for-10 with two on and two out." Suppose a batter hits .300 in some situation but .260 otherwise — how many situational at-bats make .300 significant (α = 0.05, power 0.95, one-sided)?

```{r,echo=TRUE}
power.prop.test(p1=.260, p2=.300, sig.level=0.05,
                power=.95, alternative="one.sided")
```

**Over 2,724 at-bats** — per group. Flip the question: with n = 10, what's achievable?

```{r,echo=TRUE}
power.prop.test(n=10, p1=.260, p2=.300, power=.95,
                sig.level=NULL, alternative="one.sided")$sig.level
power.prop.test(n=10, p1=.260, p2=.300, power=NULL,
                sig.level=.05, alternative="one.sided")$power
```

Significance level 0.93, or power 0.07. The book's verdict: most situational statistics are nonsense — now you can prove it.

# ANOVA Test Design

## power.anova.test

For multi-group designs analyzed with ANOVA:

```{r,eval=FALSE,echo=TRUE}
power.anova.test(groups = NULL, n = NULL,
                 between.var = NULL, within.var = NULL,
                 sig.level = 0.05, power = NULL)
```

| Argument | Description |
|---|---|
| `groups` | number of groups |
| `n` | observations per group |
| `between.var` / `within.var` | variance between / within groups |
| `sig.level` / `power` | as before |

Here **six** of the seven parameters must be specified; the remaining `NULL` argument is calculated.

```{r,echo=TRUE}
power.anova.test(groups=4, between.var=1, within.var=3, power=.80)
```

::: {.callout-tip}
**Habit for every empirical project:** run the power calculation *before* the experiment. It converts wishful thinking ("we'll see if it's significant") into design ("we need n per group to detect the effect we care about").
:::

## Copyright Notice

- **Copyright.** These slides are adapted from *R in a Nutshell: A Desktop Quick Reference* (2nd ed.) by Joseph Adler, O'Reilly Media. All rights reserved by the original author and publisher.

- **Non-commercial use only.** These materials are strictly for educational purposes and may not be used for commercial gain.

- **Attribution.** Any reproduction, distribution, or use of these materials must properly credit the original source.
