---
title: "Programming for Applications"
subtitle: "Chapter 1: Getting and Installing R"
author: "Yu-You Liou (NTU)"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "R in a Nutshell: A Desktop Quick Reference"
logo: "logo.svg"
editor: source
bibliography: bib.bib
suppress-bibliography: true
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch1-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center")
```

# **Welcome to R**

## What Is R?

R [@rcore2024] is a free, open-source environment for statistical computing and graphics, descended from the S language developed at Bell Laboratories [@becker1988new]. Throughout this course, the single name "R" actually bundles several things together:

- a **programming language** designed around data;
- an **interpreter** that executes programs written in that language;
- a **graphics system** for producing statistical figures;
- a desktop **application** for Windows, macOS, or Linux that wraps the interpreter, the standard packages, and a user interface into one install.

Our textbook is *R in a Nutshell* (2nd ed.) [@adler2012r]. This first chapter has one practical goal: get a working copy of R onto your machine.

## Why R?

::: {.callout-note}
**Three reasons R dominates applied data work.** It is free of license fees, it runs on every major platform, and it is surrounded by an enormous ecosystem of contributed packages covering nearly every statistical method ever published.
:::

- R is maintained by an international core team of developers and distributed through **CRAN** (the Comprehensive R Archive Network).
- Because the source code is open, anyone can inspect how a method is implemented — and anyone can port R to a new platform. That openness is precisely why binaries exist for every major desktop system.
- Everything you need for this course costs exactly zero dollars.

# R Versions

## The Release Cycle

- The R Core Team publishes official releases on a regular schedule: historically twice a year (spring and autumn in the book's era); today, one major `x.y.0` release each spring, followed by patch releases (`x.y.1`, `x.y.2`, ...) that fix bugs.
- Version numbers follow the pattern **major.minor.patch**, e.g., `4.4.2`.
- A typical release brings bug fixes, performance improvements, and a few new functions. Genuine changes to the *language* are rare and usually touch only obscure corners (the book's example: the type of `NA` in partially initialized arrays changed back in R 2.5).

## Version Stability: Why You Should Not Worry

- The textbook's code was checked against R **2.15.1** (2012). Because the language is so stable, essentially every example still behaves the same way today; where modern R differs, these slides use the modern behavior.
- The same logic applies to download links: filenames embed the version number (`R-4.4.2-win.exe`), so the exact name you see will always be newer than any screenshot or book. **Just take the latest one.**
- Code you write against one recent version will almost always run unchanged on another. Results across nearby versions should be virtually identical.

## Which Version Should You Use?

- For this course: **the latest released version** from CRAN.
- You can check what you are running at any time:

```{r,echo=TRUE}
R.version.string
```

- Do not worry if your patch number differs from your neighbor's — results will be identical for everything we do.

# Getting and Installing R Binaries

## Binaries, and Where to Find Them

A **binary** is a precompiled, ready-to-run build of R — the easiest installation route on every desktop platform. To get one:

1. Visit the official site: <https://cran.r-project.org> and follow the *Download* link.
2. The download link leads to a list of **mirror sites**, organized by country. CRAN is replicated worldwide; a geographically close mirror is usually also close on the network, hence fast. The modern "0-Cloud" mirror automatically routes you to a good server, so it is a safe default.
3. Pick the installer matching your operating system and run it.

A few platform-specific notes follow — Windows and macOS users normally download and run installers themselves, while Linux users are better served by a package manager.

## Windows

- Click *Download R for Windows* → *base* → run the `.exe` installer.
- Installation works like any other Windows program: easy **if you have the right permissions**, awkward if you do not (common on corporate or lab machines).
  - Simplest fix: install R as a standard user **into your own file space** (e.g., under your user directory).
  - Alternatively, install — and later install packages — as an Administrator account.
- The default location is under `%ProgramFiles%\R` (usually `C:\Program Files\R\`), with a Start-menu entry in the group *R*.
- In the book's day, CRAN shipped only 32-bit Windows builds; today both the installer and R itself are 64-bit on any modern machine, so no choices need to be made.

## macOS

- Click *Download R for macOS* and choose the build that matches your hardware:
  - **Apple Silicon** (M-series chips): the `arm64` package.
  - **Intel** Macs: the `x86_64` package.
- The `.pkg` installer places an application called **R** in your Applications folder and a command-line `R` on your path: download, double-click, follow the prompts.
- As with any application, you need adequate permissions on the machine — on your personal laptop that just means remembering your password; on a managed computer you may need the administrator's help.
- Older hardware or operating systems? CRAN keeps **legacy builds** of earlier R versions that may suit older systems better. (The book's era offered separate universal binaries for OS X 10.4/10.5 — the same idea, two decades earlier.)
- If you grab the wrong build for your chip, R will refuse to start — when in doubt, check *About This Mac*.

## Linux and Unix: Package Managers

On Linux, the cleanest route is the distribution's **package-management system**: it fetches R, resolves every other piece of software R needs, and makes later upgrades a one-line affair. Before you start, make sure you know the root password or have `sudo` privileges; otherwise you will need the system administrator.

```{bash,eval=FALSE,echo=TRUE}
# Debian / Ubuntu
sudo apt-get update
sudo apt-get install r-base r-base-dev

# Fedora / Red Hat (the book's era used: sudo yum install R.x86_64)
sudo dnf install R
```

- Updating later is equally simple: `sudo apt-get upgrade` or `sudo dnf update R` brings you to the newest packaged version.
- CRAN maintains its own repositories with newer builds than most distributions ship; instructions for adding them are on the CRAN Linux pages.
- `r-base-dev` (or the equivalent) is worth installing too — it provides the compilers needed to build contributed packages from source.

## Linux and Unix: Installing from Downloaded Files

If you prefer (or a package manager is unavailable), you can download the files and install manually:

- CRAN provides **precompiled packages** for several distributions — Debian, Ubuntu, Fedora/Red Hat, SUSE — plus builds for Solaris.
- On Red Hat–style systems the tool is RPM. The book's example: having downloaded `R-2.15.1.fc10.i386.rpm` into `~/Downloads`, install it with:

```{bash,eval=FALSE,echo=TRUE}
rpm -i ~/Downloads/R-2.15.1.fc10.i386.rpm
# Debian-style equivalent:
# sudo dpkg -i r-base-core_*.deb
```

- **Other Unix systems** often have their own channels — for instance, R is available through the FreeBSD Ports collection. Consult your system's documentation for installing third-party software.

# Building R from Source

## Should You Build Your Own R?

It is standard practice to build R from source on Linux/Unix — but **not** on Windows or macOS, where it is tricky and yields little benefit for ordinary users:

- it will not save **space** — the build pulls in large supporting tools (LaTeX among them);
- it will not save **time** — unless you already have every tool and a glacial network connection;
- the one real payoff is squeezing out extra **performance**, which in practice is rarely the bottleneck, even on large data sets.

The fact that you *can* build R anywhere is, however, exactly how R reaches unusual platforms — a direct benefit of open source.

## The Classic Build Sequence

For the curious, building on a Unix-like system follows the time-honored ritual:

```{bash,eval=FALSE,echo=TRUE}
tar -xzf R-4.4.2.tar.gz
cd R-4.4.2
./configure
make
sudo make install
```

- `configure` inspects your system and prepares the build; `make` compiles; `make install` copies the result into place.
- The textbook defers details to its later chapter on R internals; for the definitive treatment see the manual *R Installation and Administration* on CRAN.

# Wrapping Up

## Checking Your Installation

Launch R and confirm the console responds:

```{r,echo=TRUE}
1 + 1
sessionInfo()$R.version$version.string
```

If you see sensible output, your installation works and you are ready for Chapter 2, where we tour the user interface.

::: {.callout-tip}
**Keep R current.** When a new version appears (CRAN announces them on the front page), upgrading follows exactly the same steps as installing — and on Linux, your package manager does it for you.
:::

## Copyright Notice

- **Copyright.** These slides are adapted from *R in a Nutshell: A Desktop Quick Reference* (2nd ed.) by Joseph Adler, O'Reilly Media. All rights reserved by the original author and publisher.

- **Non-commercial use only.** These materials are strictly for educational purposes and may not be used for commercial gain.

- **Attribution.** Any reproduction, distribution, or use of these materials must properly credit the original source.
