---
title: "Programming for Applications"
subtitle: "Chapter 1: Getting and Installing R"
author: "Yu-You Liou (NTU)"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "R in a Nutshell: A Desktop Quick Reference"
logo: "logo.svg"
editor: source
bibliography: bib.bib
suppress-bibliography: true
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch1-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center")
```

# **Welcome to R**

## What Is R?

R [@rcore2024] is a free, open-source environment for statistical computing and graphics, descended from the S language developed at Bell Laboratories [@becker1988new]. Throughout this course, the single name "R" actually bundles several things together:

- a **programming language** designed around data;
- an **interpreter** that executes programs written in that language;
- a **graphics system** for producing statistical figures;
- a desktop **application** for Windows, macOS, or Linux that wraps the interpreter, the standard packages, and a user interface into one install.

Our textbook is *R in a Nutshell* (2nd ed.) [@adler2012r]. This first chapter has one practical goal: get a working copy of R onto your machine.

## Why R?

::: {.callout-note}
**Three reasons R dominates applied data work.** It is free of license fees, it runs on every major platform, and it is surrounded by an enormous ecosystem of contributed packages covering nearly every statistical method ever published.
:::

- R is maintained by an international core team of developers and distributed through **CRAN** (the Comprehensive R Archive Network).
- Because the source code is open, anyone can inspect how a method is implemented — and port R to new platforms.
- Everything you need for this course costs exactly zero dollars.

# R Versions

## The Release Cycle

- The R Core Team publishes official releases on a regular schedule: a major `x.y.0` release each spring, followed by patch releases (`x.y.1`, `x.y.2`, ...) that fix bugs.
- Version numbers follow the pattern **major.minor.patch**, e.g., `4.4.2`.
- The language itself is very stable. Release-to-release changes are mostly bug fixes, performance work, and a handful of new functions, so code written for one recent version almost always runs unchanged on another.
- The textbook was written against a much older release (2.15.1); essentially every example still behaves the same way today. Where modern R differs, the slides use the modern behavior.

## Which Version Should You Use?

- For this course: **the latest released version** from CRAN.
- You can check what you are running at any time:

```{r,echo=TRUE}
R.version.string
```

- Do not worry if your patch number differs from your neighbor's — results will be identical for everything we do.

# Getting and Installing R Binaries

## Where to Download

R has been ported to every major desktop platform, and prebuilt installers ("binaries") are the easiest way to get it:

1. Visit the official site: <https://cran.r-project.org>.
2. CRAN is replicated on **mirror sites** around the world. Any mirror works; a geographically close one is usually fastest. The default "0-Cloud" mirror automatically routes you to a good server.
3. Pick the download link matching your operating system and run the installer.

::: {.callout-tip}
**Filenames contain version numbers.** Links look like `R-4.4.2-win.exe` — the exact number will have moved on by the time you look. Always just take the newest one.
:::

## Windows

- Click *Download R for Windows* → *base* → run the `.exe` installer.
- The default settings are fine: R installs under `C:\Program Files\R\` and appears in the Start menu.
- Both the installer and R itself are 64-bit on any modern machine; no choices need to be made.

```{r,eval=FALSE,echo=TRUE}
# After installing, launch "R x.y.z" from the Start menu,
# or better: install RStudio (Chapter 2) and launch that instead.
```

## macOS

- Click *Download R for macOS* and choose the build that matches your hardware:
  - **Apple Silicon** (M-series chips): the `arm64` package.
  - **Intel** Macs: the `x86_64` package.
- The `.pkg` installer places an application called **R** in your Applications folder and a command-line `R` on your path.
- If you mix the two builds up, R will refuse to start — when in doubt, check *About This Mac*.

## Linux and Unix: Package Managers

On Linux, the cleanest route is your distribution's package-management system, which also keeps R updated automatically:

```{bash,eval=FALSE,echo=TRUE}
# Debian / Ubuntu
sudo apt-get update
sudo apt-get install r-base r-base-dev

# Fedora / Red Hat
sudo dnf install R
```

- CRAN maintains its own repositories with newer builds than most distributions ship; instructions for adding them are on the CRAN Linux pages.
- `r-base-dev` (or the equivalent) is worth installing too — it provides the compilers needed to build contributed packages from source.

## Linux and Unix: Installing from Files

If no package manager is available, two alternatives remain:

- **Precompiled binaries** for several distributions can be downloaded from CRAN and installed manually (e.g., with `dpkg` or `rpm`).
- **Building from source**: download the source tarball from CRAN and run the classic sequence:

```{bash,eval=FALSE,echo=TRUE}
tar -xzf R-4.4.2.tar.gz
cd R-4.4.2
./configure
make
sudo make install
```

This is rarely necessary on a desktop, but it is how R reaches unusual platforms — a direct benefit of R being open source.

## Checking Your Installation

Launch R and confirm the console responds:

```{r,echo=TRUE}
1 + 1
sessionInfo()$R.version$version.string
```

If you see sensible output, your installation works and you are ready for Chapter 2, where we tour the user interface.

## Copyright Notice

- **Copyright.** These slides are adapted from *R in a Nutshell: A Desktop Quick Reference* (2nd ed.) by Joseph Adler, O'Reilly Media. All rights reserved by the original author and publisher.

- **Non-commercial use only.** These materials are strictly for educational purposes and may not be used for commercial gain.

- **Attribution.** Any reproduction, distribution, or use of these materials must properly credit the original source.
