---
title: "Programming for Applications"
subtitle: "Chapter 23: Time Series Analysis"
author: "Yu-You Liou (NTU)"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "R in a Nutshell: A Desktop Quick Reference"
logo: "logo.svg"
editor: source
bibliography: bib.bib
suppress-bibliography: true
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch23-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center")
```

# **Data with a Clock**

## Time Series Objects

Observations ordered in time violate the independence assumed everywhere else — they need their own tools [@shumway2011time]. The course data sets include two monthly price series:

```{r,echo=TRUE,cache=TRUE}
temp <- tempfile(fileext = ".rda")
download.file("https://raw.githubusercontent.com/cran/nutshell/master/data/turkey.price.ts.rda", temp, mode = "wb")
load(temp)
download.file("https://raw.githubusercontent.com/cran/nutshell/master/data/ham.price.ts.rda", temp, mode = "wb")
load(temp)
turkey.price.ts
```

```{r,echo=TRUE,out.width="62%"}
plot(turkey.price.ts, ylab = "US cents/lb")
```

## Autocorrelation Functions

How strongly does a series correlate *with its own past*?

```{r,echo=TRUE,out.width="62%"}
acf(turkey.price.ts)
```

- Spikes at lag 12 (and 24): annual seasonality — November discounts echo every year.
- `pacf()` shows *partial* autocorrelations — the direct effect of lag $k$ after removing shorter lags — the standard tool for choosing AR order.

```{r,echo=TRUE,out.width="62%"}
pacf(turkey.price.ts)
```

## Cross-Correlation

Do two series move together, possibly at a delay?

```{r,echo=TRUE,out.width="62%"}
ccf(turkey.price.ts, ham.price.ts, main = "Turkey vs. ham prices")
```

`ccf()` correlates one series with lagged copies of another — peaks away from lag 0 suggest one series leads the other.

## Decomposition and Smoothing

Split a series into trend + seasonal + remainder:

```{r,echo=TRUE,out.width="65%"}
plot(stl(turkey.price.ts, s.window = "periodic"))
```

- `stl()` uses loess smoothing; `decompose()` is the moving-average classic.
- `filter()` computes moving averages directly; `diff()` removes trend by differencing — the "I" in ARIMA.

## ARIMA Models

The ARMA family models a (differenced) series as autoregression plus moving-average noise:

$$x_t = \phi_1 x_{t-1} + \cdots + \phi_p x_{t-p} + \varepsilon_t + \theta_1 \varepsilon_{t-1} + \cdots + \theta_q \varepsilon_{t-q}$$

```{r,echo=TRUE}
fit <- arima(turkey.price.ts, order = c(1, 0, 0),
             seasonal = list(order = c(1, 0, 0), period = 12))
fit
```

- `order = c(p, d, q)`: AR order, differencing, MA order; `seasonal=` adds the seasonal counterpart.
- Choose orders from ACF/PACF shapes, or by AIC comparison across candidates; `tsdiag(fit)` checks the residuals.

## Forecasting

```{r,echo=TRUE,out.width="65%"}
pr <- predict(fit, n.ahead = 24)
ts.plot(turkey.price.ts, pr$pred,
        pr$pred + 2 * pr$se, pr$pred - 2 * pr$se,
        lty = c(1, 2, 3, 3), col = c("black", "red", "grey", "grey"))
```

- `predict()` returns point forecasts and standard errors; the cone widens — honesty about growing uncertainty.
- The `forecast` package automates order selection (`auto.arima()`) and adds exponential smoothing (`ets()`).

## A Time Series Checklist

1. Plot the series; note trend, seasonality, breaks.
2. Stabilize: log-transform if variance grows; difference until stationary.
3. Inspect `acf()`/`pacf()`; propose ARIMA orders.
4. Fit `arima()`; compare candidates by AIC.
5. Validate residuals (`tsdiag`, Ljung–Box) — they should look like white noise.
6. Forecast with intervals, never just the point line.

## Copyright Notice

- **Copyright.** These slides are adapted from *R in a Nutshell: A Desktop Quick Reference* (2nd ed.) by Joseph Adler, O'Reilly Media. All rights reserved by the original author and publisher.

- **Non-commercial use only.** These materials are strictly for educational purposes and may not be used for commercial gain.

- **Attribution.** Any reproduction, distribution, or use of these materials must properly credit the original source.
