---
title: "Programming for Applications"
subtitle: "Chapter 23: Time Series Analysis"
author: "Yu-You Liou (NTU)"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "R in a Nutshell: A Desktop Quick Reference"
logo: "logo.svg"
editor: source
bibliography: bib.bib
suppress-bibliography: true
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch23-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center", fig.height = 4.2)
```

# **Data Over Time**

## Why Time Series Are Different

Time series carry **long-term trends** and **periodic patterns** that ordinary summary statistics miss. Analyzing them needs tools that respect *order* and *lag*. We revisit the turkey-price series from Chapter 7:

```{r,echo=TRUE}
temp <- tempfile(fileext = ".rda")
download.file("https://raw.githubusercontent.com/cran/nutshell/master/data/turkey.price.ts.rda",
              temp, mode = "wb")
load(temp)
```

# Autocorrelation

## acf and pacf

The **autocorrelation function** measures how correlated a series is with its own past, by lag — the signature of seasonality and cycles:

```{r,eval=FALSE,echo=TRUE}
acf(x, lag.max = NULL, type = c("correlation", "covariance", "partial"),
    plot = TRUE, na.action, demean = TRUE, ...)
pacf(x, lag.max, plot, na.action, ...)   # alias: type = "partial"
```

```{r,echo=TRUE,fig.height=3.6}
acf(turkey.price.ts, plot=FALSE)
```

- `acf` plots by default; `plot=FALSE` returns the numbers.
- **`pacf`** (partial autocorrelation) — correlation at a lag *after removing* the influence of shorter lags; it helps identify the order of an AR model.

The turkey series shows strong positive autocorrelation at 12-month lags, negative at 6 — clear annual seasonality.

# Time Series Models

## Autoregressive Models: ar

Time-series models predict future values from past ones. The **autoregressive (AR)** model regresses each value on its predecessors — `ar` even picks the order automatically by AIC:

```{r,eval=FALSE,echo=TRUE}
ar(x, aic = TRUE, order.max = NULL, method = c("yule-walker","burg",
   "ols","mle","yw"), ...)
```

```{r,echo=TRUE}
turkey.price.ts.ar <- ar(turkey.price.ts)
turkey.price.ts.ar$order      # order chosen by AIC
```

- `aic=TRUE` selects the order; `order.max` caps it; `method` chooses the estimation algorithm.
- The fitted object reports the chosen order and the AR coefficients.

## Forecasting with predict

`predict` on an `ar` object projects future values, with standard errors:

```{r,echo=TRUE}
turkey.forecast <- predict(turkey.price.ts.ar, n.ahead=12)
turkey.forecast$pred
```

```{r,echo=TRUE,fig.height=3.4}
ts.plot(turkey.price.ts, turkey.forecast$pred, col=c("black","blue"), lty=c(1,2))
```

## ARIMA and Beyond

For richer dynamics, **`arima`** fits autoregressive integrated moving-average models — the order `(p, d, q)` capturing AR terms, differencing, and MA terms; seasonal terms extend it to **SARIMA**:

```{r,echo=TRUE}
turkey.arima <- arima(turkey.price.ts, order=c(1,0,0),
                      seasonal=list(order=c(1,0,0), period=12))
turkey.arima
predict(turkey.arima, n.ahead=6)$pred
```

Supporting tools: `diff` (differencing to remove trend), `Box.test` (testing residual autocorrelation), `HoltWinters` (exponential smoothing), `StructTS` (structural models), and `decompose`/`stl` (splitting a series into trend, seasonal, and remainder).

::: {.callout-tip}
**The Box-Jenkins rhythm.** Plot the series; check `acf`/`pacf`; difference until stationary; fit an `arima`; check that the *residuals* look like white noise (`acf`, `Box.test`); then forecast. The modern `forecast` and `fable` packages automate much of this (`auto.arima`).
:::

## Additional Topics

The book closes by pointing onward: time series feed into broader analyses — regression with lagged predictors, volatility models (GARCH, package `tseries`), and multivariate series (VAR). The `ts` class (Chapter 7) is the common substrate; specialized classes like `zoo` and `xts` handle irregular spacing and richer date indexing.

## Copyright Notice

- **Copyright.** These slides are adapted from *R in a Nutshell: A Desktop Quick Reference* (2nd ed.) by Joseph Adler, O'Reilly Media. All rights reserved by the original author and publisher.

- **Non-commercial use only.** These materials are strictly for educational purposes and may not be used for commercial gain.

- **Attribution.** Any reproduction, distribution, or use of these materials must properly credit the original source.
