---
title: "Programming for Applications"
subtitle: "Chapter 5: An Overview of the R Language"
author: "Yu-You Liou (NTU)"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "R in a Nutshell: A Desktop Quick Reference"
logo: "logo.svg"
editor: source
bibliography: bib.bib
suppress-bibliography: true
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch5-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center")
```

# **How R Thinks**

## The Big Picture

Chapter 3 showed you R phrase by phrase; this chapter is the grammar lesson. Four ideas organize everything that follows in Part II:

- programs are built from **expressions**;
- expressions operate on **objects**;
- names are **symbols** bound to objects;
- work is done by **functions** — and functions are objects too.

Master these and the detailed syntax of Chapter 6 will feel inevitable rather than arbitrary.

## Expressions

An R program is a sequence of expressions, each evaluated to produce a value:

```{r,echo=TRUE}
x <- 3 * 7          # an assignment expression
if (x > 20) "big" else "small"   # even control flow returns a value
```

- Separate expressions with newlines or semicolons.
- Group several expressions into one with braces `{ }`; the block's value is its last expression.
- Because everything returns a value, expressions compose: you can nest an `if` inside a function call inside an assignment.

## Objects

Every value an expression touches is an **object**: vectors, lists, functions, models, environments.

```{r,echo=TRUE}
v <- c(8, 6, 7)
class(v)
attributes(matrix(1:6, 2, 3))
```

Objects carry **attributes** — metadata such as dimensions, names, or class — that change how generic functions treat them. A matrix is literally a vector wearing a `dim` attribute.

## Symbols

A variable name is itself an object: a **symbol** bound to a value within some environment.

```{r,echo=TRUE}
fortune <- "When in doubt, make a data frame."
fortune
```

- Evaluating the symbol `fortune` looks up what it is bound to.
- The binding lives in an **environment** (Chapter 8); the same name can be bound to different values in different environments.
- This is why R can treat code as data: `quote(x + y)` returns the unevaluated expression, symbols and all.

```{r,echo=TRUE}
quote(x + y)
```

## Functions

A function is an object that accepts **arguments** and returns a **value**:

```{r,echo=TRUE}
f <- function(x, y) {
  x + 2 * y
}
f(1, 10)
f(y = 1, x = 10)        # arguments may be named
```

- Functions can be assigned to variables, stored in lists, passed to other functions, and returned as results — the trait that makes R a *functional* language.
- Default values make arguments optional: `function(x, base = 10)`.

## Objects Are Copied in Assignment

R behaves as though assignment and argument-passing **copy** objects, rather than sharing references:

```{r,echo=TRUE}
w <- c(1, 2, 3)
double_it <- function(z) { z <- z * 2; z }
double_it(w)
w                       # untouched
```

::: {.callout-note}
**Call by value.** A function cannot accidentally modify your data; it works on its own copy. (Internally R delays the physical copy until something is actually modified, so this is cheaper than it sounds.)
:::

## Everything Is an Object

Beyond data, the language's own machinery is made of objects:

```{r,echo=TRUE}
class(mean)             # functions are objects
class(quote(x + y))     # unevaluated code is an object
formals(double_it)      # a function's argument list, as data
```

A function name is just a symbol pointing at a function object — which is why you can do `apply(m, 1, mean)`: you are passing the function itself.

## Special Values

A few special constants appear throughout real data work:

```{r,echo=TRUE}
NA + 1                  # NA: missing value, infects calculations
1 / 0                   # Inf
-1 / 0                  # -Inf
0 / 0                   # NaN: not a number
is.null(NULL)           # NULL: the empty object
```

- `NA` means "unknown"; most summary functions offer `na.rm=TRUE` to skip missing values.
- `NULL` means "nothing at all" — assigning `x$col <- NULL` deletes a component.

## Coercion

When types do not match, R converts — **coerces** — values according to fixed rules, moving toward the more general type:

```{r,echo=TRUE}
c(1, 2, TRUE)           # logical -> numeric
c(1, 2, "three")        # numeric -> character
as.numeric("3.14")      # explicit coercion
```

- Implicit coercion follows the hierarchy logical → integer → numeric → character.
- The `as.*` family (`as.numeric`, `as.character`, `as.factor`, ...) coerces on demand; `I()` (AsIs) suppresses unwanted conversion, e.g., when building data frames.

## The R Interpreter

R is an **interpreted** language: expressions are parsed and evaluated on the fly, with no separate compile step.

- That is what makes the console possible — and what makes R slower than compiled languages for tight loops (Chapter 24 shows the cures, including the built-in byte-code compiler).
- Parsing turns text into expression objects; evaluation walks those objects. You can do both yourself:

```{r,echo=TRUE}
ex <- parse(text = "7 * 6")[[1]]
eval(ex)
```

This "code is data" property powers R's formula notation, `quote()`, and much of its metaprogramming.

## Copyright Notice

- **Copyright.** These slides are adapted from *R in a Nutshell: A Desktop Quick Reference* (2nd ed.) by Joseph Adler, O'Reilly Media. All rights reserved by the original author and publisher.

- **Non-commercial use only.** These materials are strictly for educational purposes and may not be used for commercial gain.

- **Attribution.** Any reproduction, distribution, or use of these materials must properly credit the original source.
