---
title: "Programming for Applications"
subtitle: "Chapter 5: An Overview of the R Language"
author: "Yu-You Liou (NTU)"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "R in a Nutshell: A Desktop Quick Reference"
logo: "logo.svg"
editor: source
bibliography: bib.bib
suppress-bibliography: true
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch5-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center")
```

# **How R Thinks**

## The Grammar Lesson

Learning a computer language is much like learning a spoken one (only simpler). A tourist can survive on memorized phrases; Chapter 3 gave you those phrases. But to *speak* fluently you need grammar — verb conjugation, sentence structure. This chapter is R's grammar lesson, built on four ideas:

- programs are built from **expressions**;
- expressions operate on **objects**;
- names are **symbols** bound to objects;
- work is done by **functions** — and functions are objects too.

Master these and the detailed syntax of Chapter 6 will feel inevitable rather than arbitrary.

## Expressions

R code is a series of **expressions**: assignments, conditionals, arithmetic, and more — each evaluated to produce a value:

```{r,echo=TRUE}
x <- 1
if (1 > 2) "yes" else "no"
127 %% 10
```

Expressions are composed of objects and functions, and may be separated by newlines **or semicolons**:

```{r,echo=TRUE}
"this expression will be printed"; 7 + 13; exp(0+1i*pi)
```

## Objects

All R code manipulates **objects** — "things" represented by the computer. Numeric vectors, character vectors, lists, and functions are all objects:

```{r,echo=TRUE}
c(1, 2, 3, 4, 5)                       # a numeric vector (five elements)
"This is an object too"               # a character vector (one element)
list(c(1,2,3,4,5), "This is an object too", "this whole thing is a list")
function(x, y) { x + y }               # a function is an object as well
```

## Symbols

Formally, variable names in R are **symbols**. Assigning a value to a name binds the symbol to the object *in the current environment* — and, a touch tautologically, an environment is the set of symbols defined in a given context:

```{r,echo=TRUE}
x <- 1     # binds symbol "x" to object 1 in the current environment
```

Chapter 8 treats symbols and environments in full; for now, just remember that a name and the thing it names are different objects.

# Functions

## All Work Is Done by Functions

A **function** takes input objects (its **arguments**) and returns an output object. Every statement — setting variables, doing arithmetic, looping — can be written as a function call. The book's demonstration starts with a humble vector:

```{r,echo=TRUE}
animals <- c("cow", "chicken", "pig", "tuba")
animals[4] <- "duck"     # the normal way to fix element 4
animals
```

That bracket assignment is *parsed into* a call to the function `[<-`. This is the equivalent expression:

```{r,echo=TRUE}
animals <- c("cow", "chicken", "pig", "tuba")
`[<-`(animals, 4, "duck")
```

(One subtle difference: written this way the result isn't auto-printed — but the effect is the same.)

## Why Care That Everything Is a Function Call?

In practice you would never write `` `[<-`(animals, 4, "duck") `` — bracket notation is far easier to read. But knowing the function is there means you can **inspect its code**, **read its help page**, or **write methods with the same name** for your own classes (Chapter 10). More pretty-vs-functional pairs:

```{r,echo=TRUE}
apples <- 3            # pretty assignment
`<-`(oranges, 4)       # functional form of assignment
apples + oranges       # pretty arithmetic
`+`(apples, oranges)   # functional form
if (apples > oranges) "apples are better" else "oranges are better"
`if`(apples > oranges, "apples are better", "oranges are better")
```

```{r,echo=TRUE}
x <- c("apple","orange","banana","pear")
x[2]                   # pretty vector reference
`[`(x, 2)              # functional form
```

## Objects Are Copied in Assignment

In assignment statements, most objects are **immutable**: R copies the object, not merely a reference to it. (Immutability is a virtue — it prevents whole categories of errors, e.g., in multithreaded programs.)

```{r,echo=TRUE}
u <- list(1)
v <- u
u[[1]] <- "hat"
u
v          # v still holds the original
```

This holds for vectors, lists, and most other primitive objects.

## Copying Applies to Function Calls Too

Arguments are copied as well. This function sets the `i`th element of `x` to 4 — and accomplishes nothing visible:

```{r,echo=TRUE}
f <- function(x, i) { x[i] = 4 }
w <- c(10, 11, 12, 13)
f(w, 1)
w          # untouched: w was copied into x
```

Technically, the interpreter copies the object bound to `w` and binds the symbol `x` to the copy; the modification happens only inside the function's context. (Mutable objects and pass-by-reference become possible with environments — Chapter 8.)

::: {.callout-note}
**Semantics vs. performance.** R *behaves as if* every assignment copies, but often modifies objects **in place** — e.g., `v <- 1:100; v[50] <- 27` alters the vector without copying it. Identical meaning, much better speed. Details live in the *R Internals* guide.
:::

## Everything in R Is an Object

Not just data: **functions, symbols, even expressions** are objects. A function name is really a symbol pointing at a function object (a relationship itself stored in an environment object). The same symbol can hold a number one moment and a function the next:

```{r,error=TRUE,echo=TRUE}
x <- 1
x
x(2)                     # x is not (yet) a function
x <- function(i) i^2
x
x(2)
```

You can even write R code that constructs new functions — in principle, a function that rewrites its own definition.

# Special Values

## NA: Missing Values

`NA` ("not available") represents **missing values** — common in loaded text data, and substituted for database `NULL`s. Expanding a structure beyond its defined values fills the new cells with `NA`:

```{r,echo=TRUE}
v <- c(1, 2, 3)
length(v) <- 4
v
```

## Inf, -Inf, and NaN

Numbers too big to represent become **infinity**, signed accordingly — also the result of dividing by zero:

```{r,echo=TRUE}
2 ^ 1024
- 2 ^ 1024
1 / 0
```

Computations that make no sense yield **`NaN`** ("not a number"):

```{r,echo=TRUE}
Inf - Inf
0 / 0
```

## NULL

The symbol `NULL` always points to one special object: the **null object**.

- Often used as a function argument meaning "no value supplied"; some functions also return it.
- `NULL` is **not** the same as `NA`, `Inf`, `-Inf`, or `NaN` — it is not a value at all, but the absence of one.

```{r,echo=TRUE}
is.null(NULL)
```

# Coercion

## Automatic Type Conversion

Call a function with the "wrong" type and R tries to **coerce** the value so the call still works. Two mechanisms operate automatically:

- **formal objects:** generic functions lacking an exact method look for a coercion method to a type that has one (creating such methods: Chapter 10);
- **built-in types:** R converts from more specific to more general types on its own:

```{r,echo=TRUE}
x <- c(1, 2, 3, 4, 5)
typeof(x); class(x)
x[2] <- "hat"          # one character value forces the whole vector over
x
typeof(x); class(x)
```

## The Coercion Rules

- Logical values become numbers: `TRUE` → 1, `FALSE` → 0.
- Values convert to the **simplest type that can represent all the information**.
- The ordering is roughly: logical < integer < numeric < complex < character < list.
- `raw` objects are not converted to other types.
- Object **attributes are dropped** during coercion.
- To inhibit coercion when passing arguments, wrap the value in `I()` (the `AsIs` function) — see `?AsIs`.

::: {.callout-note}
**Why so permissive?** Strongly typed languages (Java) raise exceptions instead of converting. John Chambers, S's designer, recalled that early S "made as many cases work as possible," but the later, formal conclusion was that converting richer types to simpler ones *in all situations* breeds confusing, untrustworthy results. In daily practice — numeric vectors of one type — coercion rarely bites.
:::

# The R Interpreter

## Parse, Substitute, Evaluate

R is an **interpreted** language: the interpreter executes your expressions directly, with no compile step (unlike C, C++, Java; like LISP, Perl, JavaScript). For each expression, the interpreter:

1. **parses** it, translating syntactic sugar into functional form;
2. **substitutes** objects for symbols where appropriate;
3. **evaluates** it, returning an object — recursively for complex expressions.

Example: typing `x <- 1` becomes the call `` `<-`(x, 1) ``, which binds 1 to `x` in the current environment and returns 1.

## Walking Through a Conditional

With `x` bound to 1, consider:

```{r,echo=TRUE}
x <- 1
if (x > 1) "orange" else "apple"
```

The interpreter parses the line as an if-then-else structure, evaluates the **condition** `(x > 1)` first — false — and therefore *never evaluates* `"orange"`; it evaluates the expression after `else` and returns `"apple"`. In special cases like this, **not all arguments are evaluated** — which is why `if` cannot be an ordinary function. And whatever an expression returns at the console gets handed to `print`.

## .Internal and .Primitive

Some functionality lives inside the R system itself:

- Many functions call internal system code via **`.Internal`** — e.g., the graphics workhorse:

```{r,echo=TRUE}
plot.xy
```

- Where even `.Internal`'s calling overhead is too high, functions are implemented entirely internally via **`.Primitive`** — reserved for a few performance-critical basics:

```{r,echo=TRUE}
`<-`
```

# Seeing How R Works

## quote: Parsing Without Evaluating

Because R expressions are R objects, you can parse — or partially evaluate — expressions and inspect the result. The `quote()` function parses its argument but does *not* evaluate it, returning a **language object**:

```{r,echo=TRUE}
typeof(quote(if (x > 1) "orange" else "apple"))
quote(if (x > 1) "orange" else "apple")   # the print method isn't very informative...
```

## The Parse Tree

...but convert the language object **to a list** and the structure appears — this is the expression's parse tree:

```{r,echo=TRUE}
as(quote(if (x > 1) "orange" else "apple"), "list")
```

Apply `typeof` to every element to label the tree (here using `lapply` — and thanks to coercion, the `as` is optional):

```{r,echo=TRUE}
lapply(as(quote(if (x > 1) "orange" else "apple"), "list"), typeof)
```

Observations: the `else` keyword vanishes in parsing; the first element is a **symbol** referring to the `if` function — so even special syntax becomes *function name first, arguments after*, exactly like any call. For constants, the list has a single item: `as.list(quote(1))`.

## Sugar Exposed, and deparse

The bracket lookup `x[2]` and its functional twin parse to the *identical* tree:

```{r,echo=TRUE}
as.list(quote(x[2]))
as.list(quote(`[`(x, 2)))
```

The translation is irreversible — both collapse to one tree — but `deparse` turns a parse tree back into properly formatted R code, choosing the pretty syntax:

```{r,echo=TRUE}
deparse(quote(x[2]))
deparse(quote(`[`(x, 2)))
```

::: {.callout-tip}
**Tools to carry through the book:** `quote`, `substitute`, `typeof`, `class`, and `methods` let you watch the interpreter parse and dispatch. When R surprises you, these are your microscope.
:::

## Copyright Notice

- **Copyright.** These slides are adapted from *R in a Nutshell: A Desktop Quick Reference* (2nd ed.) by Joseph Adler, O'Reilly Media. All rights reserved by the original author and publisher.

- **Non-commercial use only.** These materials are strictly for educational purposes and may not be used for commercial gain.

- **Attribution.** Any reproduction, distribution, or use of these materials must properly credit the original source.
